# Pre-course tasks {#pre-course}

Complete everything in this pre-course tasks section and 
finish with *completing the survey* at the end.

## Installing the latest programs

Given this is an intermediate course, you should already have R and RStudio
installed. However, you may not have the latest versions installed.

1. Have the latest version of [R](https://cloud.r-project.org/index.html)
installed (preferably >4.0.0, but it is fine if you have at least 3.6.0).
1. Have the latest version of [RStudio](https://www.rstudio.com/products/rstudio/download/#download)
installed (preferably the latest 1.3 version, but at least 1.2.5001 is fine).

There are a few other things to install:

1. [Git](https://git-scm.com/downloads). We'll be using Git 
(building off of the Introduction course), so it needs to be installed.
1. For some Windows users, you may need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/Rtools35.exe)
in order for some R packages to be installed (which you'll do shortly).

All these programs are required for the course, even Git.
Git, which is a software program to formally manage file versions, 
is used because of it's popularity and the amount of documentation available for it.
Check out the online book [Happy Git with R](https://happygitwithr.com/),
especially the "Why Git" section, to understand why we are using Git.
Windows users tend to have more trouble with installing Git than macOS or Linux users.
See the section on [Installing Git for Windows](https://happygitwithr.com/install-git.html#install-git-windows) for help.

> *A note to those who have or use work laptops with 
restrictive administrative privileges*: 
You may encounter problems installing software due to administrative reasons
(e.g. you don't have permission to install things).
For issues with updating to the latest version of R or RStudio, 
if you have at least 3.6.0 for R and at least 1.2.5001 for RStudio, 
that should be fine. If you have versions of R and RStudio older than that,
it is necessary for you to request IT update your software if you can't yourself.
Unfortunately, given that Git is not a commonly used software for some organizations,
you may not have it installed and you *will* need to ask IT to install it.
We *require* it for the course, so please make sure to give IT enough time to
be able to install it for you.

Once R, RStudio, and Git have been installed, open RStudio. 
*If at any point during these pre-course tasks you have any troubles*, 
try as best as you can to complete the task and then let me know about the
issues in the pre-course survey (at the end of this section). 
If you continue having problems, indicate on the survey that you need help
and we can try to book a quick video call to fix the problem. 
Otherwise, if you can, come to the course earlier (15-20 min) to get help.

## Installing the R packages

We will be using specific R packages for the course, 
so you will need to install them.
A detailed walkthrough for installing the necessary packages is
available on the [pre-course tasks for installing packages](https://r-cubed-v2.rostools.org/pre-course.html#installing-r-packages)
section of the beginner course,
however, you only need to install the r3 helper package in order to install
all the necessary packages:

1. Install the remotes package: 

    ```r
    install.packages("remotes")
    ```

2. Install the helper package for this course, r3:

    ```r
    remotes::install_gitlab("rostools/r3", upgrade = TRUE)
    ```


*Note*: When you see a command like `something::something()`,
for example with `remotes::install_gitlab()`, you would "read" this as:

> R, can you please use the install_gitlab function from the remotes package.

The normal way of doing this would be to load the package with `library(remotes)` 
and then running the command (`install_gitlab()`).
But by using the `::`, we tell R to directly use a function from a package,
without needing to load the package. We use this trick because we only want to
use the `install_gitlab()` command from the remotes package and not have to load
all the other functions from remotes.

## Setting up Git

Since Git has already been covered in the Introduction course, 
we won't cover it during this course, but we will be using it 
(since version control should be part of any modern data analysis workflow,
we'll be using it as well).
If you have used or currently use Git, you can skip this section.
If you haven't used it, please do these tasks:


1. Follow the [pre-course tasks for Git](https://r-cubed-v2.rostools.org/pre-course.html#setting-up-git-and-github)
(*not* the GitHub tasks) from the beginner's course.
Specifically, type in the RStudio Console:

    ```r
    # A pop-up to type in your name (first and last), 
    # as well as your email
    r3::setup_git_config()
    ```
    
2. Read through the 
[Version Control lesson](https://r-cubed-v2.rostools.org/version-control.html)
of the beginner course.
*You don't need to do any of the exercises or activities*, 
but can if it helps you learn or understand it better.

Regardless of whether you've done the steps above or not, 
*everyone* needs to run

```r
r3::check_setup()
```

and eventually copy and paste the output into one of the survey questions.

## Create an R project for the course

One of the basic steps to reproducibility and modern workflows in data analysis
is to keep everything contained in a single location.
In RStudio, this is done with projects.
Please read all of 
[Section 6.1](https://r-cubed-v2.rostools.org/r-project-management.html#what-is-a-project-and-why-use-it)
from the beginner course to learn about R Projects 
and keeping things self-contained. 
*You don't need to do any of the exercises or activities*.

There are several ways to organise a project folder.
We’ll be using the structure from the package prodigenr. 
The project setup can be done by either:

1. Using RStudio's New Project menu item: "File -> New Project -> New Directory",
scroll down to "Scientific Analysis Project using prodigenr" and name the project
"LearnR3" in the Directory Name, saving it to the "Desktop" with Browse.
2. Or, running the command `prodigenr::setup_project("~/Desktop/LearnR3")` in the Console.

When the RStudio Project opens up again, run these two commands in the Console:

```r
usethis::use_blank_slate()
usethis::use_r("functions", open = FALSE)
```

## R Markdown

We teach and use R Markdown because it is one of the very first steps to being
reproducible and because it is a very powerful tool to doing data analysis.
Please do these two tasks:

1. Please read over the [R Markdown section](https://r-cubed-v2.rostools.org/reproducible-documents.html)
of the beginner course.
If you use R Markdown already, you can skip this step.
2. Open up the `LearnR3` project,
either by clicking the `LearnR3.Rproj` file or
by using the "File -> Open Project" menu.
Run the function below in the Console when RStudio is in the `LearnR3` project,
which will create a new file called `lesson.Rmd` in the `doc/` folder.

    ```r
    r3::create_rmarkdown_doc()
    ```

After the R Markdown document is created, 
please run this function in the Console:

```r
r3::check_project_setup()
```

And paste the output into the survey at the end.

## Course introduction

Most of the course description is found in the [syllabus](syllabus.html).
If you haven't read it, **please read it now**.
Read over what the course will cover, 
what we expect you to learn at the end of it,
and what our basic assumptions are about who you are and what you know.
The final pre-course task is a survey that asks some questions on if you've read
and understood it.

One goal of the course is to teach about open science,
and true to our mission, we practice what we preach.
The course material is publicly accessible (all on this website) and 
[openly licensed](index.html#licensing) so you can use and re-use it for free!
The material and table of contents on the side is listed in the order 
that we will cover in the course.

We have a [Code of Conduct](conduct.html).
If you haven't read it, **read it now**. 
The survey at the end will ask about Conduct.
We want to make sure this course is a supportive 
and safe environment for learning,
so this Code of Conduct is important.

You're almost done.
Please fill out the **[pre-course survey]** to finish this assignment.

[pre-course survey]: https://docs.google.com/forms/d/e/1FAIpQLSe2gpbN9ExCXyeTfJLgu_poemF0iUtkcihnCunXl3YA4e1hHQ/viewform?usp=sf_link

See you at the course!

## The MMASH dataset

1. Learn about filesystems, relative and absolute paths, and how to make use 
of the fs package to navigate files in your project.
1. Learn where to store your raw data so that you can use
scripts as a record of what was done to process the data before analyzing it,
and why that's important.

For this course, we're going to use an openly licensed dataset on monitoring
sleep and activity (MMASH) [@Rossi2020]. We'll switch over to the [MMASH
website][mmash-website] and go over:

1. What is contained in the dataset by looking at the the Data Description.
We'll be making use of the Data Description for the code-along as well as the
exercises. **Take 5 min to quickly look over the Data Description and get more
familiar with it**.
1. The open license and the ability to re-use it. A small note: GDPR makes it
more strict on how to share and use personal data, but it *does not prohibit
sharing it or making it public*! GDPR and Open Data are not in conflict.

[mmash-website]: https://physionet.org/content/mmash/1.0.0/

> *Note*: Sometimes the PhysioNet website is slow. If that's the case,
use **[this alternative link](resources/mmash-page.html)** instead.

After we have looked over the MMASH website, we need to setup where we will store
and prepare the dataset for processing. Here we'll make use of the [usethis]
package to help setup things. usethis is an extremely useful package for managing
R projects and I highly recommend checking out how you can use it more in your
own work. For now, while in your `LearnR3` R Project, go to the Console pane in
RStudio and type out:

[usethis]: https://usethis.r-lib.org/

```{r create-data-raw, eval=FALSE}
usethis::use_data_raw("mmash")
```

What this function does is create a new folder called `data-raw/` and creates
an R script called `mmash.R` in the folder. This is where we will store the 
raw, original MMASH data that we'll get from the website. The R script should
have opened up for you, otherwise, go into the `data-raw/` folder and open up
the new `mmash.R` script.

The first thing we want do is delete all the code that is there by default.
Then we'll create a new line at the top and type out:

```{r dr-script-add-here}
library(here)
```

The [here] package was described in the 
[Management of R Projects](https://r-cubed-v2.rostools.org/r-project-management.html)
of the introductory course. **Take 8 minutes to read the [section about
here][here-r-cubed] and read the next two paragraphs**. here makes it easy to
refer to other files in an R project.

R works based on the current *working directory*, which you can see on the top 
of the RStudio Console pane. When in an RStudio R Project, the working directory
is the folder where the `.Rproj` file is located. When you run scripts in with
`source()` sometimes the working directory will be where the R script is located.
So you can encounter problems with finding files. So instead, by using `here()`,
R knows to start searching for files from the `.Rproj`.

[here]: https://here.r-lib.org/
[here-r-cubed]: https://r-cubed-v2.rostools.org/r-project-management.html#packages-data-and-file-paths

Let's use an example. Below is the folder tree. If we open up RStudio with the
`LearnR3.Rproj` file and run code in the `data-raw/mmash.R`, R runs the commands
assuming everything starts in the `LearnR3/` folder. But! If we run the code in 
the `mmash.R` script by other ways (e.g. not with RStudio, not in an R Project,
or with `source()`), R runs everything assuming it starts in the `data-raw/`.
This can make things tricky. What `here()` does is tell R to first look for the
`.Rproj` file and then start looking for the file we actually want.

```text
LearnR3
├── data
│   └── README.md
├── data-raw
│   └── mmash.R
├── doc
│   ├── lesson.Rmd
│   └── README.md
├── R
│   └── README.md
├── .Rbuildignore
├── .gitignore
├── DESCRIPTION
├── LearnR3.Rproj
└── README.md
```

**Stop reading and we'll go back to coding together**. The first step we want
to take is to download the dataset. From this material, paste this code into the
`data-raw/mmash.R` script:

```{r mmash-link}
mmash_link <- "https://physionet.org/static/published-projects/mmash/multilevel-monitoring-of-activity-and-sleep-in-healthy-people-1.0.0.zip"
```

> *Note*: Sometimes the PhysioNet website is slow. If that's the case,
use `r3::mmash_data_link` instead of the link used above. In this case,
it will look like `mmash_link <- r3::mmash_data_link`.

Then we're going to write out the function `download.file()` to download and
save the zip dataset. We're going to save the zip file to
`data-raw/mmash-data.zip` with the `destfile` argument. This code should be 
written in the `data-raw/mmash.R` file.
Run these lines of code to download the dataset.

```{r download-data, eval=FALSE}
download.file(mmash_link, destfile = here("data-raw/mmash-data.zip"))
```

Because the original dataset is stored elsewhere, we don't need to keep it or 
save it to our Git history. So we'll add the zip file to the Git ignore list.
In the *Console*, type out and run this code. You only need to do this once.

```{r ignore-mmash-zip, eval=FALSE}
usethis::use_git_ignore("data-raw/mmash-data.zip")
```

This is a good time to save the changes we've made to the Git history.
Let's open the Git interface with either the Git icon at the top near the menu 
bar, or with `Ctrl-Alt-M`. When this opens up we'll click the checkbox beside
the `.gitignore` and `data-raw/mmash.R` files. Then we write a commit message
in the text box on the right, something like "Code to download data zip file".
Click the "Commit" button and close the Git interface.

Alright, let's start preparing the dataset. We'll open up the zip file and look
at what is inside (only instructor does this). Inside there is the license file,
another file to check if the download worked correctly (the SHA file),
and another zip of the dataset inside. Because we are starting with the original
raw `mmash-data.zip`, we should record exactly how we process the data set for
use. This also relates to the principal of *"keep your raw data raw"*, as in
don't edit or touch your raw data, let R or other programming language process 
it. This lets you have a history of what was done to the raw data.
During data collection, programs like Excel or Google Sheets are incredibly 
powerful. But after collection is done, don't make edits directly to the data
unless absolutely necessary.

A quick comment about whether you should save your *raw* data in `data-raw/`.
A general guideline is:

- *Do* store it to `data-raw/` if the data will only be used for the one project.
Use the `data-raw/` R script to be the record for how you processed your data
for final analysis work.
- *Don't* save it to `data-raw/` if: 1) there is a central dataset that multiple
people use for multiple projects; or 2) you got the data online. Instead, use
the `data-raw/` R script to be the record for which website you downloaded it
from or from which central location you extracted it from and how you processed
it.
- *Don't save it to a project-specific `data-raw/` folder if you will use the
raw data for multiple projects. Instead, create a central location for the data
*for yourself* so that you can point all other projects to it and use their
individual `data-raw/` R scripts as the record for how you processed the raw data.

Ok, let's start unzipping the zip files. In `data-raw/mmash.R`, continue writing 
below the `download.file()` function. We'll use the `unzip()` function
to unzip the dataset. The main argument for `unzip()` is the zip file,
and the other important one called `exdir` tells `unzip()` the folder we want to
extract the files to. The argument `junkpaths` is used here because we want 
everything extracted to the `data-raw/` folder (don't ask nor know why it's
called "junkpaths").

```{r unzip-main-file}
unzip(here("data-raw/mmash-data.zip"), 
      exdir = here("data-raw"),
      junkpaths = TRUE)
```

Notice the indentations and spacings of the code. Like writing any language,
code should follow a [style guide]. An easy way of following a style is by
selecting your code and using RStudio's builtin style fixer of either
`Ctrl-Shift-A` or "Code -> Reformat Code" menu item.
Ok, next, we want to extract the new `data-raw/MMASH.zip` file. Because we want to
keep the folder structure inside this zip file, we don't use `junkpaths`.

[style guide]: https://style.tidyverse.org

```{r unzip-data-file}
unzip(here("data-raw/MMASH.zip"),
      exdir = here("data-raw"))
```

Almost done! There are several files left over that we don't need, so we'll
also write in the script code to remove these files. We'll use the [fs] package,
which means filesystem, to work with files. First, we delete all the files
we originally extracted (`LICENSE.txt`, `SHA256SUMS.txt`, and `MMASH.zip`)
by using the `file_delete()` function. Then we'll rename the new folder
`data-raw/DataPaper/` to something more explicit like `data-raw/mmash/`
using the `file_move()` function. So the `data-raw/` folder will initially look
like:

[fs]: https://fs.r-lib.org/

```{r, eval=FALSE}
dir_tree("data-raw", recurse = 1)
```

```text
data-raw
├── LICENSE.txt
├── MMASH.zip
├── SHA256SUMS.txt
├── DataPaper
│   ├── user_1
│   ├── user_10
│   ├── ...
│   ├── user_8
│   └── user_9
├── mmash-data.zip
└── mmash.R
```

Then we add these lines of code to the `data-raw/mmash.R` script and run them:

```{r delete-rename-files}
library(fs)
file_delete(here(c("data-raw/MMASH.zip", 
                   "data-raw/SHA256SUMS.txt",
                   "data-raw/LICENSE.txt")))
file_move(here("data-raw/DataPaper"), here("data-raw/mmash"))
```

Afterward, the files and folders in `data-raw/` will look like:

```text
data-raw
├── mmash
│   ├── user_1
│   ├── user_10
│   ├── ...
│   ├── user_8
│   └── user_9
├── mmash-data.zip
└── mmash.R
```

Since we have an R script that downloads the data and processes it for us,
we don't need to have Git track it. So, in the *Console*, type out and run
this command:

```{r git-ignore-mmash, eval=FALSE}
usethis::use_git_ignore("data-raw/mmash/")
```

Now that we have everything prepared, let's add and commit the changes to the 
Git history.
