# Save time, don't repeat yourself {#dry}

![](https://img.shields.io/badge/document%20status-revising-orange?style=flat-square)

```{r setup-03, include=FALSE}
library(here)
library(fs)
library(dplyr)
library(purrr)
```

Here we will cover the second block, "*Workflow*" in Figure \@ref(fig:diagram-overview-2).

```{r diagram-overview-2, fig.cap="Section of the overall workflow we will be covering.", echo=FALSE}
diagram_overview(2)
```

And your folder and file structure should look like:

```
LearnR3
├── data/
│   └── README.md
├── data-raw/
│   ├── mmash-data.zip
│   ├── mmash/
│   │  ├── user_1
│   │  ├── ...
│   │  └── user_22
│   └── mmash.R
├── doc/
│   ├── README.md
│   └── lesson.Rmd
├── R/
│   ├── functions.R
│   └── README.md
├── .Rbuildignore
├── .gitignore
├── DESCRIPTION
├── LearnR3.Rproj
└── README.md
```

## Learning objectives

1. Learn what functions are in R, how to create and use them.
1. Learn about functional programming, vectorization, and functionals,
and how to use them to get more done with less code and less time.
1. Learn a workflow of using R Markdown and `devtools::load_all()` (or
`Ctrl-Shift-L` in RStudio) as a tool and process for developing functions
that can be automatically loaded in for later and easy use.
1. Learn what R package dependency management is and how it can simplify your
data analysis work.
1. Continue practising Git version control to manage changes to your files.

## The basics of a function

**Take 5 min and read this section until it says to stop.**
As mentioned before, all actions in R are functions. For instance, the `+` is a
function, `mean()` is a function, `[]` is a function, and so on. Because R
is open source, that means anyone can see how things work underneath.
So, if we want to see what a function does underneath, we type out the
function name without the `()` into the Console and run it. 
If we do it with the function `sd()` which calculates the standard deviation,
we see:

```{r}
sd
```

How you calculate the standard deviation is the square root of the variance.
In this code, the `var()` is inside the `sqrt()` function, which is exactly what
it should be.

So, if you learn how to create your own functions, it can make doing your work
easier and more efficient because you don't have to repeat yourself later.
Making functions always has a basic structure of:

1. Giving a name to the function (e.g. `mean`).
2. Starting the function call using `function()`, assigning it to the name with
`<-`. This tells R that the name is a function object.
3. Optionally providing arguments to give to the function call, for instance
`function(arg1, arg2, arg3)`. 
4. Filling out the body of the function, with the arguments (if any) contained
inside, that does some action.
5. Optionally, but strongly encouraged, use `return()` to indicate what you want 
the function to output.

While there is no minimum or maximum number of arguments you can provide for a
function (e.g. you could have zero or dozens of arguments), its good practice
for yourself and for others to have as few arguments as possible to get the job
done. So, the structure is:

```{r function-structure, eval=FALSE}
name <- function(arg1, arg2) {
    # body of function
    output <- ... code ....
    return(output)
}
```

**Ok, stop here and we'll go over it together**.
Let's write a simple example. First, create a new Markdown header called 
`## Making a function` and create a code chunk below that with `Ctrl-Alt-I`.
Then, inside the function, let's write this code out:

```{r create-add-function}
add_numbers <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

You can use the new function by running the above code and writing out your new 
function, with arguments to give it.

```{r call-function}
add_numbers(1, 2)
```

The function name is fairly good... `add_numbers` is read as "add numbers".
While we generally want to write code that describes what it does by reading it,
it's also good practice to add some formal documentation to the function. Use
the "Insert Roxygen Skeleton" in the "Code" menu (or by typing
`Ctrl-Shift-Alt-R`) and you can add template documentation right above the
function. It looks like:

```{r roxygen-docs-1}
#' Title
#'
#' @param num1 
#' @param num2 
#'
#' @return
#' @export
#'
#' @examples
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or several words 
that describe the function. Creating a new paragraph below this line allows you 
to add a more detailed description. The other items are:

- `@param num`: These lines describe what each argument is for and what to give it.
- `@return`: This describes what output the function gives. Is it a data.frame?
A plot? What else does the output give?
- `@export`: Tells R that this function should be accessible to the user of your
package. Since we aren't making packages, delete it.
- `@examples`: Any lines below this are used to show examples of how to use the
function. This is very useful when making packages, but not really in this case.
So we'll delete it. Let's write out some documentation for this function:

```{r roxygen-docs-2}
#' Add two numbers together.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#'
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

Once we've created that, let's open up the Git Interface (`Ctrl-Alt-M`) and add
and commit these changes to our history.

## Making a function for vroom

Now that we have a basic understanding of what a function looks like, let's
apply it to something we're doing right now: Importing our data.

Making functions is a series of steps:

1. Write code that works and does what you want.
1. Enclose it as a function with `name <- function() { ... }`, 
with an appropriate and descriptive name.
1. Create arguments in the function call (`function(arg1, arg2)`) with appropriate
and descriptive names, then replace the code with the argument names where 
appropriate. 
1. Rename any objects created to be more generic and include the `return()`
function at the end to indicate what the function will output.
1. Run the function and check that it works.
1. Add the roxygen2 documentation tags (with `Ctrl-Alt-Shift-R` or "Code -> Insert
Roxygen Skeleton" menu item while the cursor is in the function).

So, step one. Let's take the code we wrote for importing the `user_info` data
and convert that as a function:

```{r function-creation-step-1, eval=FALSE}
user_1_info_data <- vroom(
    user_1_info_file,
    col_select = -1,
    col_types = cols(
        Gender = col_character(),
        Weight = col_double(),
        Height = col_double(),
        Age = col_double()
    ),
    .name_repair = snakecase::to_snake_case
)
```

Next we wrap it in the function call and give it an appropriate name. In this case,
`import_user_info` is descriptive and meaningful. Make sure to style it correctly
with `Ctrl-Shift-A`.

```{r function-creation-step-2, eval=FALSE}
import_user_info <- function() {
    user_1_info_data <- vroom(
        user_1_info_file,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
}
```

Then, we add arguments in the function and replace within the code. Here, we
have only one thing that we would change: The file path to the dataset.
So, a good name might be `file_path`.

```{r function-creation-step-3, eval=FALSE}
import_user_info <- function(file_path) {
    user_1_info_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
}
```

Then we clean things up by renaming `user_1_info_data` since we would like to
also import more than just `user_1`. A nice object name would be `info_data`. 
Add the `return()` function at the end with the object you want your function to
output.

```{r function-creation-step-4}
import_user_info <- function(file_path) {
    info_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(info_data)
}
```

Great! Now we need to test it out. Let's try on two datasets, 
two `user_info.csv` files in the `user_1` and `user_2` folders.

```{r function-creation-step-5}
import_user_info(here("data-raw/mmash/user_1/user_info.csv"))
import_user_info(here("data-raw/mmash/user_2/user_info.csv"))
```

Awesome! It works. The final stage is adding the Roxygen documentation.

```{r function-creation-step-6}
#' Import MMASH user info data file.
#'
#' @param file_path Path to user info data file.
#'
#' @return Outputs a data frame.
#'
import_user_info <- function(file_path) {
    info_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(info_data)
}
```

A massive advantage of using functions is that if you want to make a change to all
your code, you can very easily do it by modifying the function and it will
change all your other code too. Now that we have a working function, let's add
and commit it to the Git history with the RStudio Git Interface.

## Exercise: Repeat with the saliva data

Time: 15 min

Take the code you created for importing the saliva data set from 
Exercise \@ref(ex-import-saliva) and make it into a function. A helpful tip:
To move around an R Markdown or R script more easily, open up the 
"Document Outline" on the side by clicking the button in the top right corner of
the R Markdown pane or by using `Ctrl-Shift-O`.

1. Create a new markdown header `## Exercise for importing the saliva data`.
1. Create a code chunk below that (`Ctrl-Alt-I`).
1. Paste the code you used from the exercise into the code chunk and begin
converting it into a function, like we did above.
    - Wrap it with the `function() {...}`
    - Make a meaningful name (e.g. `import_saliva`)
    - Make an argument for the file path and replace it with the code in `vroom()`
    - Rename the output object and include it in the `return()` function
    - Test that it works
    - Create the Roxygen documentation

```{r solution-repeat-saliva, echo=FALSE}
# Excluding the Roxygen docs.
import_saliva <- function(file_path) {
    saliva_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            SAMPLES = col_character(),
            `Cortisol NORM` = col_double(),
            `Melatonin NORM` = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(saliva_data)
}
```


## Continuing the workflow

We've created two functions. Now we need to move those functions from the 
`doc/lesson.Rmd` file and into the `R/`. We do this for a few reasons:

1. To prevent the R Markdown document from becoming too long and having a large
portion of R code over other text.
1. To make it easier to maintain and find things in your project.
1. To make use of the `devtools::load_all()` functionality available with
R Projects that are indicated as Packages.

Because this project is set as a Package, we can use the `load_all()` keyboard
shortcut `Ctrl-Shift-L` in RStudio. But first, we need to make a new file to
store our functions. So, in the *Console*, type out and run:

```{r functions-script, eval=FALSE}
usethis::use_r("functions")
```

This creates a new file called `R/functions.R`. Then we cut and paste the
functions we created into this new file. So, cut *only* the `import_user_info()` 
function we created in `doc/lesson.Rmd`, including the Roxygen documentation,
and paste it into `R/functions.R`.

Once we have it in there, let's test out how the `load_all()` function works.
Let's restart our R session with either `Ctrl-Shift-F10` or from the 
"Session -> Restart R" menu item. Move back into the `doc/lesson.Rmd`
and go to where you wrote:

```{r, eval=FALSE}
import_user_info(here("data-raw/mmash/user_1/user_info.csv"))
```

Then hit `Ctrl-Shift-L`. What happens? You should see a bunch of text pop up
on the side about loading the `LearnR3` project. Ok. Try to run the
`import_user_info()` line. What happens now? You might get an error about
not finding the `vroom()` function. Or if you put `library(vroom)` in the code
chunk called `setup`, you might not get an error. If you did get an error,
that's because R doesn't know what the `vroom()` function is. This is where we
start getting into package dependency management.

What is package dependency management? Whenever you use an R package, you 
depend on it for your code to work. The informal way to show what packages you use
is by using the `library()` function. But if you come back to the project, 
or get a new computer, or someone else is working on your project too,
how will they know, easily, which packages you depend on?
Do they have to search through all your files just to find all `library()`
functions you used and then install those packages individually? A much better
way here is to formally indicate your package dependency so that installing
dependencies is easy. We do this by using the `DESCRIPTION` file.

Open up the `DESCRIPTION` file. You may or may not see something that looks like:

```text
Package: LearnR3
Title: Title of Project
Version: 0.0.0
Authors@R: c(
    person(given = "Luke",
           family = "Johnston",
           role = c("aut", "cre"),
           email = "lwjohnst@gmail.com",
           comment = c(ORCID = "0000-0003-4169-2616"))
    )
Imports: 
    devtools
```

Notice the `Imports:` key. This is where information about packages are added.
To quickly add a package, go to the *Console* and type out:

```r
usethis::use_package("vroom")
```

You will see a bunch of text about adding it to `Imports`. If you look in your
`DESCRIPTION` file now, you'll see something like:

```text
Imports: 
    devtools,
    vroom
```

Now, if you or someone else wants to install all the package dependency,
they can do that going to the *Console* and running:

```r
remotes::install_deps()
```

This function finds the `DESCRIPTION` file and installs all the packages in
`Imports`. Let's add the other dependencies by typing in the *Console*:

```r
usethis::use_package("here")
usethis::use_package("fs")
usethis::use_package("snakecase")
```

Great! Now that we've formally established package dependencies, we also need to
formally declare which package each function comes from inside our functions.
Before getting into the correct way, we need to quickly cover the incorrect way
that you may or may not have seen that others have done on websites or in code.
Sometimes people use `library()` or `require()` inside functions like:

```{r bad-practice-library, eval=FALSE}
add_numbers <- function(num1, num2) {
    library(packagename)
    ...code...
    return(added)
}
```

Or: 

```{r bad-practice-require, eval=FALSE}
add_numbers <- function(num1, num2) {
    require(packagename)
    ...code...
    return(added)
}
```

This is *very* **bad practice** and can have some unintended and serious
consequences that you might not notice or that won't give any warning or error.
The correct way of indicating which package a function comes from is instead by
using `packagename::`, which you've seen and used many times in this course.
We won't get into the reasons why this is incorrect because it can quickly get
quite technical.

Another reason to use `packagename::` for each function from an R package
you use in your own function is that it explicitly tells R (and us the readers)
where the function comes from. Because the same function name can be used by
multiple packages, if you don't explicitly state which package the function is 
from, R will use the function that it can find... which isn't always the function
you *mean*.

We also do this step at the end of making the function because doing it while
we create it can be quite tedious. Alright, let's go into `R/functions.R` and
add `vroom::` to each of the vroom functions we've used:

```{r explicit-function-calls}
import_user_info <- function(file_path) {
    info_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            Gender = vroom::col_character(),
            Weight = vroom::col_double(),
            Height = vroom::col_double(),
            Age = vroom::col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(info_data)
}
```

Test that it works by restart the R session (`Ctrl-Shift-F10` or "Session ->
Restart R"), loading the functions with `Ctrl-Shift-L`, and going to the
*Console* and typing out:

```r
import_user_info(here::here("data-raw/mmash/user_1/user_info.csv"))
```

It should work as expected!
Now that we've done that, let's add and commit the changes made through the
Git interface.

## Exercise: Move and update the rest of the functions

Time: 20 min

Repeat this process and do this for the rest of the code you worked on
previously that imported the `RR.csv`, and `Actigraph.csv` data. 


1. Convert the importing code into functions while in the `doc/lessons.Rmd` file.
Include the Roxygen documentation and use `packagename::` to be explicit about
where the function comes from.
    - Name the new functions `import_rr` and `import_actigraph`.
1. Move the function into `R/functions.R`.
1. Restart R and then use `devtools::load_all()` by using `Ctrl-Shift-L` and
test that the functions work by running them in the *Console*.

Also update the `import_saliva()` function you created by being explicit about
where the functions come from (e.g. with the `packagename::`).

```{r solution-update-other-functions, echo=FALSE}
# Excluding the Roxygen docs.
import_saliva <- function(file_path) {
    saliva_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            SAMPLES = vroom::col_character(),
            `Cortisol NORM` = vroom::col_double(),
            `Melatonin NORM` = vroom::col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(saliva_data)
}
import_rr <- function(file_path) {
    rr_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            ibi_s = vroom::col_double(),
            day = vroom::col_double(),
            # Converts to seconds
            time = vroom::col_time(format = "")
        ),
        .name_repair = snakecase::to_snake_case
    ) 
    return(rr_data)
}
import_actigraph <- function(file_path) {
    actigraph_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            Axis1 = vroom::col_double(),
            Axis2 = vroom::col_double(),
            Axis3 = vroom::col_double(),
            Steps = vroom::col_double(),
            HR = vroom::col_double(),
            `Inclinometer Off` = vroom::col_double(),
            `Inclinometer Standing` = vroom::col_double(),
            `Inclinometer Sitting` = vroom::col_double(),
            `Inclinometer Lying` = vroom::col_double(),
            `Vector Magnitude` = vroom::col_double(),
            day = vroom::col_double(),
            time = vroom::col_time(format = "")
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(actigraph_data)
}
```

## Functional programming

**Please take 15 min to read over this section before we continue.**
Unlike many other programming languages, R's primary strength and approach to
programming is in functional programming. So what is functional programming? It
is a form of programming that:

- Uses functions (like `function()`)
- Applies functions to vectors all at once (called [vectorisation]), rather than
in a loop
- Can use functions as input to other functions (called a [functional])

[vectorisation]: https://bookdown.org/rdpeng/rprogdatascience/vectorized-operations.html
[functional]: https://adv-r.had.co.nz/Functionals.html

We've already covered functions. You've probably already used vectorization
since it is one of R's big strengths. Functions like `mean()`, `sd()`,
`sum()` are vectorized in the sense that they take a vector and do something
to all the values in the vector at once. As a comparison, in other programming
languages, if you wanted to calculate the sum you would need a loop:

```{r using-for-loop}
total_sum <- 0
# a vector
values <- 1:10
for (value in values) {
    total_sum <- value + total_sum
}
total_sum
```

But in R, you can give a function that uses vectorization the entire vector
(e.g. `c(1, 2, 3, 4)`) and R will know what to do with it. 
Figure \@ref(fig:image-vectorization) shows how a function conceptually uses
vectorization.

```{r image-vectorization, fig.cap="A function using vectorization. Modified from the [RStudio purrr cheatsheet][purrr-cheatsheet].", out.width="40%", echo=FALSE}
knitr::include_graphics(here::here("images/vectorization.svg"))
```

Instead, in R, there is the function called `sum()` that takes the entire
vector of `values` and outputs the total sum, without needing a for loop.

```{r}
# Vectorized
sum(values)
```

[purrr-cheatsheet]: https://raw.githubusercontent.com/rstudio/cheatsheets/master/purrr.pdf

A functional on the other hand is a function that can also use a function as
one of its arguments. Figure \@ref(fig:image-functionals) shows how the functional
`map()` from the [purrr] package works by taking a vector (or list), applying a 
function to each of those items, and outputting the results from each function.
The name `map()` doesn't mean a geographic map, it means the mathematical meaning
of map: To use a function on each item in a set of items.

```{r image-functionals, fig.cap="A functional that uses a function to apply it to each item in a vector. Modified from the [RStudio purrr cheatsheet][purrr-cheatsheet].", out.width="75%", echo=FALSE}
knitr::include_graphics(here::here("images/functionals.svg"))
```

[purrr]: https://purrr.tidyverse.org

Here's a simple toy example to show how it works. Here we use `paste()` on each
item of `1:5`.

```{r}
library(purrr)
map(1:5, paste)
```

You'll notice that `map()` outputs a list, with all the `[[1]]` printed. `map()`
will always output a list. Also notice that the `paste()` function is given 
*without* the `()` brackets. Without the brackets, the function can be used
by the `map()` functional and treated like any other object in R. 

Let's say we wanted to paste together the number with the sentence "seconds have
passed". Normally it would look like:

```{r paste-multiple}
paste(1, "seconds have passed")
paste(2, "seconds have passed")
paste(3, "seconds have passed")
paste(4, "seconds have passed")
paste(5, "seconds have passed")
```

Or as a loop:

```{r paste-loop}
for (num in 1:5) {
    sec_passed <- paste(num, "seconds have passed")
    print(sec_passed)
}
```

With `map()`, we'd do this a bit differently. purrr uses a shortcut to allow
you to write functions that do more things to the input vector (e.g. `1:5`).
This shortcut is using `~` to start the function and `.x` as the replacement
for the vector item:

```{r}
map(1:5, ~paste(.x, "seconds have passed"))
```

This is the basics of using functionals. So, with functions, vectorization, and
functionals, you generally should only use for loops as the very last option.
That's because R has better, more expressive, and more powerful tools to
use instead. And while technically, using a loop let's you "not repeat
yourself", it's a lot of typing and code to express a simple idea: *Do an action
on each of these items*. So using functionals let's you repeat yourself 
less `r emo::ji("smile")`

So what does functionals have to do with what we are doing now? Well, our 
`import_user_info()` function only takes in one data file. But we have right now
22 files. We can use functionals to load in all the datasets at once!
**Ok, stop reading and we'll go over this again before continuing with the
coding.**

`r details_for_instructors("
Go over this section briefly by reinforcing what they read. Make sure they
understand the concept of applying something to many things at once. Doing the
code-along should also help reinforce this concept. 

Also highlight that the resources appendix has some links for continued learning
for this and that the RStudio purrr cheatsheet is an amazing resource to use.
")`

The first thing we have to do is add `library(purrr)` to the `setup` code chunk
in the `doc/lesson.Rmd` document. Then we need to add the package dependency
by going to the *Console* and running:

```r
usethis::use_package("purrr")
```

Then, the next step to actually using the `map()` functional is to get a vector
or list of all the dataset files available to us. We turn back to use the fs
package, which has a function called `dir_ls()` that finds files of a certain
pattern. In our case, the pattern is `user_info.csv`. So, let's add `library(fs)`
to the `setup` code chunk. Then, go to the bottom of the `doc/lesson.Rmd`
document, create a new header called `## Using map`, and create a code chunk
below that.

The `dir_ls()` function takes the path that we want to search (`data-raw/mmash/`),
uses the argument `regexp` (short for [regular expression][regex] or also `regex`) to find the pattern,
and `recurse` to look in all subfolders. We'll cover regular expressions a bit
more in next session.

[regex]: https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions

```{r list-user-info-files}
user_info_files <- dir_ls(here("data-raw/mmash/"), 
                          regexp = "user_info.csv", 
                          recurse = TRUE)
```

Then let's see what the output looks like. For the website, we are only showing
the first 3 files. Your output will look slightly different from this.

```{r list-files-head, eval=FALSE}
user_info_files
```

```{r list-files-for-book, echo=FALSE}
head(gsub(".*\\/data-raw", "data-raw", user_info_files), 3)
```

Alright, we now have all the files ready to give to `map()`. So let's try it!

```{r}
user_info_list <- map(user_info_files, import_user_info)
```




Remember, that `map()` always outputs a list, so when we look into this object,
it will give us 22 tibbles (data.frames). Here we'll only show the first one:

```{r}
user_info_list[[1]]
```

This is great because with one line of code we imported all these datasets!
But we're missing an important bit of information: The user ID. 
A powerful feature of the purrr package is that it has other functions to
make working with functionals easier. We know `map()` always outputs a list.
What if you want to output a character vector instead? If we check the help:

```r
?map
```

`r details_for_instructors("Go through this help documentation and talk a bit about it.")`

We see that there are other functions, including a function
called `map_chr()` that seems to output a character vector. 
There are several others that give an output based 
on the ending of `map_`, such as:

- `map_int()` outputs an integer.
- `map_dbl()` outputs a numeric value, called a "double" in programming.
- `map_dfr()` outputs a data frame, combining the list items by row (`r`).
- `map_dfc()` outputs a data frame, combining the list items by column (`c`).

The `map_dfr()` looks like the one we want, since we want all these datasets
together as one. If we look at the help for it,
we see that it has an argument `.id`, which we can use to create a new column
that sets the user ID, or in this case, the file path to the dataset, which
has the user ID information in it. So, let's use it and create a new column
called `file_path_id`.

```{r}
user_info_df <- map_dfr(user_info_files, import_user_info,
                        .id = "file_path_id")
```

Your `file_path_id` variable will look different. Don't worry, we're going to
tidy up the `file_path_id` variable later.

```{r, eval=FALSE}
user_info_df
```

```{r show-data-for-book, echo=FALSE}
user_info_df %>% 
    mutate(file_path_id = gsub(".*\\/data-raw", "data-raw", file_path_id))
```

Now that we have this working, let's add and commit the changes to the Git history.

## Exercise: Make a function for importing other datasets with functionals {#ex-function-import-all-data}

Time: 20 min

We need to do basically the exact same thing for the `saliva.csv`, `RR.csv`, and
`Actigraph.csv` datasets, following this format:

```{r ex-template, eval=FALSE}
user_info_files <- dir_ls(here("data-raw/mmash/"), 
                          regexp = "user_info.csv", 
                          recurse = TRUE)
user_info_df <- map_dfr(user_info_files, import_user_info,
                        .id = "file_path_id")
```

For importing the other datasets, we basically only have to modify two locations
to get this code to import the other datasets: at the `regexp =` argument and
where the `import_user_info` function is. This is the perfect chance to make a
function that you can use for other purposes and that is itself a functional
(since it takes a function as an input). So inside `doc/lesson.Rmd`, convert
this bit of code into a function that works to import the other three datasets.

1. Create a new header `## Exercise: Map on the other datasets` at the bottom of
the document.
1. Create a new code chunk below it.
1. Repeat the steps you've taken previously to create a new function:
    - Wrap the code with `function() { ... }`
    - Name the function `import_multiple_files`
    - Create two new arguments called `file_pattern` and `import_function`
    - Create generic intermediate objects (instead of `user_info_file` and
    `user_info_df`). I'd recommend something like `data_files` and
    `combined_data`
    - Use `return()` at the end of the function for the object you want to output
    - Create and write Roxygen documentation to describe the new function
    - Append `packagename::` to the individual functions (there are three packages
    used: fs, here, and purrr)
    - Run it and check that it works on `saliva.csv`
1 After it works, cut and paste the function into the `R/functions.R` file.
Then restart the R session, run `load_all()`, and test the code out in the
*Console*.
1. Then, write code in your `doc/lesson.Rmd` file where the function used to be
and import in the other datasets. You should now have three new objects for all
the imported datasets, called `saliva_df`, `rr_df`, and `actigraph_df`.
1. Once done, add the changes you've made and commit them to the Git history.

```{r solution-import-multiple-files, echo=FALSE}
# Excluding the Roxygen docs.
import_multiple_files <- function(file_pattern, import_function) {
    data_files <- fs::dir_ls(here::here("data-raw/mmash/"),
                             regexp = file_pattern,
                             recurse = TRUE)
    
    combined_data <- purrr::map_dfr(data_files, import_function,
                                    .id = "file_path_id")
    return(combined_data)
}
```
