# Save time, don't repeat yourself {#dry}

```{r setup-03, include=FALSE, eval=FALSE}
library(here)
library(fs)
library(dplyr)
library(purrr)
```

Here we will cover the second block, "*Workflow*" in Figure \@ref(fig:diagram-overview-2).

```{r diagram-overview-2, fig.cap="Section of the overall workflow we will be covering.", echo=FALSE}
diagram_overview(2)
```

And your folder and file structure should look like:

```
LearnR3
├── data/
│   └── README.md
├── data-raw/
│   ├── mmash-data.zip
│   ├── mmash/
│   │  ├── user_1
│   │  ├── ...
│   │  └── user_22
│   └── mmash.R
├── doc/
│   ├── README.md
│   └── lesson.Rmd
├── R/
│   ├── functions.R
│   └── README.md
├── .gitignore
├── DESCRIPTION
├── LearnR3.Rproj
└── README.md
```

## Learning objectives

1. Learn what functions are in R, how to create and use them.
1. Learn a workflow of using R Markdown, `source()` (or
`Ctrl-Shift-S` in RStudio), and restarting R (`Ctrl-Shift-F10`) 
as a tool and process for developing functions that can be later easily used.
1. Learn what R package dependency management is and how it can simplify your
data analysis work.
1. Continue practising Git version control to manage changes to your files.
1. Debugging...

## The basics of a function

**Take 5 min and read this section until it says to stop.**
As mentioned before, all actions in R are functions. For instance, the `+` is a
function, `mean()` is a function, `[]` is a function, and so on. Because R
is open source, that means anyone can see how things work underneath.
So, if we want to see what a function does underneath, we type out the
function name without the `()` into the Console and run it. 
If we do it with the function `sd()` which calculates the standard deviation,
we see:

```{r}
sd
```

How you calculate the standard deviation is the square root of the variance.
In this code, the `var()` is inside the `sqrt()` function, which is exactly what
it should be.

So, if you learn how to create your own functions, it can make doing your work
easier and more efficient because you don't have to repeat yourself later.
Making functions always has a basic structure of:

1. Giving a name to the function (e.g. `mean`).
2. Starting the function call using `function()`, assigning it to the name with
`<-`. This tells R that the name is a function object.
3. Optionally providing arguments to give to the function call, for instance
`function(arg1, arg2, arg3)`. 
4. Filling out the body of the function, with the arguments (if any) contained
inside, that does some action.
5. Optionally, but strongly encouraged, use `return()` to indicate what you want 
the function to output.

`r details_for_instructors("
Emphasize that we will be using this workflow for creating functions all the
time throughout course and that this workflow is also what you'd use in your
daily work.
")`

While there is no minimum or maximum number of arguments you can provide for a
function (e.g. you could have zero or dozens of arguments), its good practice
for yourself and for others to have as few arguments as possible to get the job
done. So, the structure is:

```{r function-structure, eval=FALSE}
name <- function(arg1, arg2) {
    # body of function
    output <- ... code ....
    return(output)
}
```

**Ok, stop here and we'll go over it together**.
Let's write a simple example. First, create a new Markdown header called 
`## Making a function` and create a code chunk below that with `Ctrl-Alt-I`.
Then, inside the function, let's write this code out:

```{r create-add-function}
add_numbers <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

You can use the new function by running the above code and writing out your new 
function, with arguments to give it.

```{r call-function}
add_numbers(1, 2)
```

The function name is fairly good... `add_numbers` is read as "add numbers".
While we generally want to write code that describes what it does by reading it,
it's also good practice to add some formal documentation to the function. Use
the "Insert Roxygen Skeleton" in the "Code" menu (or by typing
`Ctrl-Shift-Alt-R`) and you can add template documentation right above the
function. It looks like:

```{r roxygen-docs-1}
#' Title
#'
#' @param num1 
#' @param num2 
#'
#' @return
#' @export
#'
#' @examples
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or several words 
that describe the function. Creating a new paragraph below this line allows you 
to add a more detailed description. The other items are:

- `@param num`: These lines describe what each argument is for and what to give it.
- `@return`: This describes what output the function gives. Is it a data.frame?
A plot? What else does the output give?
- `@export`: Tells R that this function should be accessible to the user of your
package. Since we aren't making packages, delete it.
- `@examples`: Any lines below this are used to show examples of how to use the
function. This is very useful when making packages, but not really in this case.
So we'll delete it. Let's write out some documentation for this function:

```{r roxygen-docs-2}
#' Add two numbers together.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#'
add_nums <- function(num1, num2) {
    added <- num1 + num2
    return(added)
}
```

Once we've created that, let's open up the Git Interface (`Ctrl-Alt-M`) and add
and commit these changes to our history.

## Making a function for vroom

Now that we have a basic understanding of what a function looks like, let's
apply it to something we're doing right now: Importing our data.

Making functions is a series of steps:

1. Write code that works and does what you want.
1. Enclose it as a function with `name <- function() { ... }`, 
with an appropriate and descriptive name.
1. Create arguments in the function call (`function(arg1, arg2)`) with appropriate
and descriptive names, then replace the code with the argument names where 
appropriate. 
1. Rename any objects created to be more generic and include the `return()`
function at the end to indicate what the function will output.
1. Run the function and check that it works.
1. Add the roxygen2 documentation tags (with `Ctrl-Alt-Shift-R` or "Code -> Insert
Roxygen Skeleton" menu item while the cursor is in the function).

`r details_for_instructors("
Emphasize that we will be using this workflow for creating functions all the
time throughout course and that this workflow is also what you'd use in your
daily work.
")`

So, step one. Let's take the code we wrote for importing the `user_info` data
and convert that as a function:

```{r function-creation-step-1, eval=FALSE}
user_1_info_data <- vroom(
    user_1_info_file,
    col_select = -1,
    col_types = cols(
        Gender = col_character(),
        Weight = col_double(),
        Height = col_double(),
        Age = col_double()
    ),
    .name_repair = snakecase::to_snake_case
)
```

Next we wrap it in the function call and give it an appropriate name. In this case,
`import_user_info` is descriptive and meaningful. Make sure to style it correctly
with `Ctrl-Shift-A`.

```{r function-creation-step-2, eval=FALSE}
import_user_info <- function() {
    user_1_info_data <- vroom(
        user_1_info_file,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
}
```

Then, we add arguments in the function and replace within the code. Here, we
have only one thing that we would change: The file path to the dataset.
So, a good name might be `file_path`.

```{r function-creation-step-3, eval=FALSE}
import_user_info <- function(file_path) {
    user_1_info_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
}
```

Then we clean things up by renaming `user_1_info_data` since we would like to
also import more than just `user_1`. A nice object name would be `info_data`. 
Add the `return()` function at the end with the object you want your function to
output.

```{r function-creation-step-4}
import_user_info <- function(file_path) {
    info_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(info_data)
}
```

Great! Now we need to test it out. Let's try on two datasets, 
two `user_info.csv` files in the `user_1` and `user_2` folders.

```{r function-creation-step-5}
import_user_info(here("data-raw/mmash/user_1/user_info.csv"))
import_user_info(here("data-raw/mmash/user_2/user_info.csv"))
```

Awesome! It works. The final stage is adding the Roxygen documentation.

```{r function-creation-step-6}
#' Import MMASH user info data file.
#'
#' @param file_path Path to user info data file.
#'
#' @return Outputs a data frame.
#'
import_user_info <- function(file_path) {
    info_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            Gender = col_character(),
            Weight = col_double(),
            Height = col_double(),
            Age = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(info_data)
}
```

A massive advantage of using functions is that if you want to make a change to all
your code, you can very easily do it by modifying the function and it will
change all your other code too. Now that we have a working function, let's add
and commit it to the Git history with the RStudio Git Interface.

## Exercise: Repeat with the saliva data

Time: 15 min

Take the code you created for importing the saliva data set from 
Exercise \@ref(ex-import-saliva) and make it into a function. A helpful tip:
To move around an R Markdown or R script more easily, open up the 
"Document Outline" on the side by clicking the button in the top right corner of
the R Markdown pane or by using `Ctrl-Shift-O`.

1. Create a new markdown header `## Exercise for importing the saliva data`.
1. Create a code chunk below that (`Ctrl-Alt-I`).
1. Paste the code you used from the exercise into the code chunk and begin
converting it into a function, like we did above.
    - Wrap it with the `function() {...}`
    - Make a meaningful name (e.g. `import_saliva`)
    - Make an argument for the file path and replace it with the code in `vroom()`
    - Rename the output object and include it in the `return()` function
    - Test that it works
    - Create the Roxygen documentation

```{r solution-repeat-saliva, echo=FALSE}
# Excluding the Roxygen docs.
import_saliva <- function(file_path) {
    saliva_data <- vroom(
        file_path,
        col_select = -1,
        col_types = cols(
            SAMPLES = col_character(),
            `Cortisol NORM` = col_double(),
            `Melatonin NORM` = col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(saliva_data)
}
```


## Continuing the workflow

We've created two functions. Now we need to move those functions from the 
`doc/lesson.Rmd` file and into the `R/` folder. We do this for a few reasons:

1. To prevent the R Markdown document from becoming too long and having a large
portion of R code over other text.
1. To make it easier to maintain and find things in your project.
1. To make use of the `source()` function.

We need to make a new file to store our functions so its easier to source them.
So, in the *Console*, type out and run:

```{r functions-script, eval=FALSE}
usethis::use_r("functions")
```

This creates a new file called `R/functions.R`. Then we cut and paste the
functions we created into this new file. So, cut *only* the `import_user_info()` 
function we created in `doc/lesson.Rmd`, including the Roxygen documentation,
and paste it into `R/functions.R`.

Once we have it in there, let's test out the workflow.
Restart our R session with either `Ctrl-Shift-F10` or from the 
"Session -> Restart R" menu item. Move back into the `doc/lesson.Rmd`
and add `source(here("R/functions.R"))` to the code chunk called `setup` at the top.
Run that line of code. Then go to where you wrote:

```{r, eval=FALSE}
import_user_info(here("data-raw/mmash/user_1/user_info.csv"))
```

Now run this line. What happens now? You *may* get an error about
not finding the `vroom()` function. If you put `library(vroom)` in the `setup`
code chunk, you might not get an error. If you did get an error,
that's because R doesn't know what the `vroom()` function is. This is where we
start getting into package dependency management.

What is package dependency management? Whenever you use an R package, you 
depend on it for your code to work. The informal way to show what packages you use
is by using the `library()` function. But if you come back to the project, 
or get a new computer, or someone else is working on your project too,
how will they know, easily, which packages you depend on?
Do they have to search through all your files just to find all `library()`
functions you used and then install those packages individually? A much better
way here is to formally indicate your package dependency so that installing
dependencies is easy. We do this by using the `DESCRIPTION` file.

Open up the `DESCRIPTION` file. You may or may not see something that looks like:

```text
Package: LearnR3
Title: Title of Project
Version: 0.0.0
Authors@R: c(
    person(given = "Luke",
           family = "Johnston",
           role = c("aut", "cre"),
           email = "lwjohnst@gmail.com",
           comment = c(ORCID = "0000-0003-4169-2616"))
    )
Imports: 
    devtools
```

Notice the `Imports:` key. This is where information about packages are added.
To quickly add a package, go to the *Console* and type out:

```r
usethis::use_package("vroom")
```

You will see a bunch of text about adding it to `Imports`. If you look in your
`DESCRIPTION` file now, you'll see something like:

```text
Imports: 
    devtools,
    vroom
```

Now, if you or someone else wants to install all the package dependency,
they can do that going to the *Console* and running:

```r
remotes::install_deps()
```

This function finds the `DESCRIPTION` file and installs all the packages in
`Imports`. Let's add the other dependencies by typing in the *Console*:

```r
usethis::use_package("here")
usethis::use_package("fs")
usethis::use_package("snakecase")
```


...First, let's go to the `setup`
code chunk and replace purrr with tidyverse and move `library(tidyverse)` to
the top of the `setup` code chunk. We also have to add tidyverse as a dependency.
Because tidyverse is a large collection of packages, the recommended way to add 
the dependency is with:

```r
usethis::use_package("tidyverse", type = "Depends")
```


Great! Now that we've formally established package dependencies, we also need to
formally declare which package each function comes from inside our functions.
Before getting into the correct way, we need to quickly cover the incorrect way
that you may or may not have seen that others have done on websites or in code.
Sometimes people use `library()` or `require()` inside functions like:

```{r bad-practice-library, eval=FALSE}
add_numbers <- function(num1, num2) {
    library(packagename)
    ...code...
    return(added)
}
```

Or: 

```{r bad-practice-require, eval=FALSE}
add_numbers <- function(num1, num2) {
    require(packagename)
    ...code...
    return(added)
}
```

This is *very* **bad practice** and can have some unintended and serious
consequences that you might not notice or that won't give any warning or error.
The correct way of indicating which package a function comes from is instead by
using `packagename::`, which you've seen and used many times in this course.
We won't get into the reasons why this is incorrect because it can quickly get
quite technical.

Another reason to use `packagename::` for each function from an R package
you use in your own function is that it explicitly tells R (and us the readers)
where the function comes from. Because the same function name can be used by
multiple packages, if you don't explicitly state which package the function is 
from, R will use the function that it can find... which isn't always the function
you *mean*.

We also do this step at the end of making the function because doing it while
we create it can be quite tedious. Alright, let's go into `R/functions.R` and
add `vroom::` to each of the vroom functions we've used:

```{r explicit-function-calls}
import_user_info <- function(file_path) {
    info_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            Gender = vroom::col_character(),
            Weight = vroom::col_double(),
            Height = vroom::col_double(),
            Age = vroom::col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(info_data)
}
```

Test that it works by restart the R session (`Ctrl-Shift-F10` or "Session ->
Restart R"), loading the functions with `Ctrl-Shift-L`, and going to the
*Console* and typing out:

```r
import_user_info(here::here("data-raw/mmash/user_1/user_info.csv"))
```

It should work as expected!
Now that we've done that, let's add and commit the changes made through the
Git interface.

## Exercise: Move and update the rest of the functions

Time: 20 min

Repeat this process and do this for the rest of the code you worked on
previously that imported the `RR.csv`, and `Actigraph.csv` data. 


1. Convert the importing code into functions while in the `doc/lessons.Rmd` file.
Include the Roxygen documentation and use `packagename::` to be explicit about
where the function comes from.
    - Name the new functions `import_rr` and `import_actigraph`.
1. Move the function into `R/functions.R`.
1. Restart R, `source()` the functions file (`Ctrl-Shift-S`), and
test that the functions work by running them in the *Console*.

Also update the `import_saliva()` function you created by being explicit about
where the functions come from (e.g. with the `packagename::`).

```{r solution-update-other-functions, echo=FALSE}
# Excluding the Roxygen docs.
import_saliva <- function(file_path) {
    saliva_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            SAMPLES = vroom::col_character(),
            `Cortisol NORM` = vroom::col_double(),
            `Melatonin NORM` = vroom::col_double()
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(saliva_data)
}
import_rr <- function(file_path) {
    rr_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            ibi_s = vroom::col_double(),
            day = vroom::col_double(),
            # Converts to seconds
            time = vroom::col_time(format = "")
        ),
        .name_repair = snakecase::to_snake_case
    ) 
    return(rr_data)
}
import_actigraph <- function(file_path) {
    actigraph_data <- vroom::vroom(
        file_path,
        col_select = -1,
        col_types = vroom::cols(
            Axis1 = vroom::col_double(),
            Axis2 = vroom::col_double(),
            Axis3 = vroom::col_double(),
            Steps = vroom::col_double(),
            HR = vroom::col_double(),
            `Inclinometer Off` = vroom::col_double(),
            `Inclinometer Standing` = vroom::col_double(),
            `Inclinometer Sitting` = vroom::col_double(),
            `Inclinometer Lying` = vroom::col_double(),
            `Vector Magnitude` = vroom::col_double(),
            day = vroom::col_double(),
            time = vroom::col_time(format = "")
        ),
        .name_repair = snakecase::to_snake_case
    )
    return(actigraph_data)
}
```

## Debugging functions

Debugging is one of those things that seems really scary and difficult, but
once you try it and use it, is actually not as intimidating. To debug, which means
to find and fix problems in your code, there are several ways, the simplest of
which is by adding the `browser()` function to the start of your function and
then re-running the function (either by manually running it or with `Ctrl-Shift-S` to
`source()` the file).

TODO: Update this section
The skills from this are directly usable
when using `browser()`. In the `R/functions.R` file, go to the
... function and click the space to the left of the line
number for the first line *inside* the function. There will be something that
pops up at the top of the pane saying something like:

> "Breakpoints will be activated when the package is built and reloaded."

<!-- TODO: eventually add an image or video.. but for next year. -->

Activate it by running `source()` with `Ctrl-Shift-S` while in the functions
file. The dot should now be solid red, rather than a circle. In the Console,
type out and run the code:

```{r, error=TRUE}
#...
```

You may have also already written that code previously, so you can use the up
arrow in the Console to go back to it. After running the code, a bunch of things
happen: 

- A yellow line will highlight the code in the function, along with a green arrow
on the left of the line number
- The Console will now start with `Browse[1]>` will have text like `debug at ...`
- There will be new buttons on the top of the Console like "Next", "Continue",
and "Stop"
- The Environment pane will be empty and will say "Traceback"

<!-- TODO: Add image or video of this or extend it from previous video? -->
<!-- TODO: Image of debugging changes, with descriptions -->

We are now in debug mode. In this mode you can really investigate what is happening
with your code and how to fix it. The way to figure out what's wrong is by running
the code bit by bit. First let's check that the first works by highlighting
the code (shown below) and running it.

```{r, eval=FALSE}
data
```

After highlighting and running this, we see that it works fine. Alright, next step:

```{r, eval=FALSE}
data %>%
    dplyr::group_by(day)
```

Still is fine. Next:

```{r, eval=FALSE}
data %>%
    dplyr::group_by(day) %>%
    dplyr::summarise(dplyr::across(columns,
                                   list(mean = mean, sd = sd),
                                   na.rm = TRUE))
```

Ah, here is where the problem occurs. The argument `columns` contains the
`age` variable we want to use in `summarise()`... but the problem is that
R is interpreting `age` as an object that exists in the R session when in reality
it only exists in the `mmash` data frame. But it works when we use it outside of 
the function, why not here?

So what's happening? We've encountered a problem due to "[non-standard
evaluation][nse]" (or NSE). NSE is a feature of R and is used quite a lot throughout
R, but is especially used in the tidyverse packages. It's one of the first
things computer scientists complain about when they use R, because it is such a
foreign thing in other programming languages. But NSE is what allows you to use
formulas (e.g. `y ~ x + x2` in modeling) or allows you to type out
`select(Gender, BMI)` or `library(purrr)`. In "standard evaluation", it
would instead be `select("Gender", "BMI")` or `library("purrr")`. So NSE gives
flexibility and ease of use for the user (we don't have to type quotes every
time) when doing data analysis, but can give some headaches when programming in
R, like with making functions. There's more detail about this on the 
[dplyr website][dplyr-programming], which will give a few options, the simplest 
of which is to wrap `columns` inside the `across()` function with `{{}}`. So,
let's "Stop" the debugging, add these curly brackets to `columns`, and remove 
the Breakpoint by clicking it (if it hasn't been removed already).

[nse]: https://adv-r.had.co.nz/Computing-on-the-language.html
[dplyr-programming]: https://dplyr.tidyverse.org/articles/programming.html#warm-up

```{r}
tidy_summarise_by_day <- function(data, columns) {
    data %>%
        dplyr::group_by(day) %>%
        dplyr::summarise(dplyr::across({{columns}},
                                       list(mean = mean, sd = sd),
                                       na.rm = TRUE)) %>% 
        tidyr::pivot_longer(dplyr::ends_with(c("_mean", "_sd")),
                            names_to = c("name", ".value"),
                            names_sep = "_") 
}
```

Now, if we run the function again:

```{r}
load(here("data/mmash.rda"))
tidy_summarise_by_day(mmash, age)
```

Yea! It works! If we want to include other variables:

```{r}
tidy_summarise_by_day(mmash, c(age, weight))
```
