# Extra material {#extra-material}

## Debugging functions

Debugging is one of activities that seems really scary and difficult, but
once you try it and use it, is not nearly as intimidating as it seemed. 
To debug, which means to find and fix problems in your code, there are several
ways, the simplest of which is by inserting the `browser()` function into the
the start of your function, re-running the function (either by manually
running it or with `Ctrl-Shift-S` to `source()` the file), and using it again.

For instance, we have a function like this:

```{r}
test_debugging <- function(number) {
    number + number
}
```

To start the debugger, insert the `browser()` function into your function:

```{r, eval=FALSE}
test_debugging <- function(number) {
    browser()
    number + number
}
```

And re-run and use the function again, which will pop up a new debugging
panel in RStudio. Sadly, we can't show this on the website since it only
works in RStudio (we may add a video as some point). When you are in
the debugger will open up and it will show a few things:

<!-- TODO: eventually add an image or video.. but for next year. -->

- A yellow line will highlight the code in the function, along with a green arrow
on the left of the line number.
- The Console will now start with `Browse[1]>` and will have text like 
`debug at ...`.
- There will be new buttons on the top of the Console like "Next", "Continue",
and "Stop".
- The Environment pane will be empty and will say "Traceback".

<!-- TODO: Add image or video of this or extend it from previous video? -->
<!-- TODO: Image of debugging changes, with descriptions -->

In this mode you can really investigate what is happening
with your code and how to fix it. The way to figure out what's wrong is by running
the code bit by bit. This debug environment is empty except for the actions that
occur within it, so it really can help figure things out.

## Non-standard evaluation (NSE)

Writing your own functions that use tidyverse functions, you may eventually
encounter an error that might not be very easy to figure out. Here's an example
where you want to use your own arguments in `filter()`.

```{r test-nse, error=TRUE}
test_nse <- function(data, filter_condition) {
    data %>% 
        dplyr::filter(filter_condition)
}

CO2 %>% 
    test_nse(conc > 100)
```

The error occurs of something called "[non-standard evaluation][nse]" (or NSE).
NSE is a feature of R and is used quite a lot throughout R (e.g. `library()`),
but is especially used in the tidyverse packages. It's one of the first
things computer scientists complain about when they use R, because it is such a
foreign thing in other programming languages. But NSE is what allows you to use
formulas (e.g. `y ~ x + x2` in modeling) or allows you to type out
`select(Gender, BMI)` or `library(purrr)`. In "standard evaluation", these
would instead be `select("Gender", "BMI")` or `library("purrr")`. So NSE gives
flexibility and ease of use for the user (we don't have to type quotes every
time) when doing data analysis, but can give some headaches when programming in
R, like when making functions. There's more detail about this on the 
[dplyr website][dplyr-programming], which will give a few options to deal with
NSE while programming with tidyverse packages, the simplest of which is to wrap
the argument with `{{}}`. 

[nse]: https://adv-r.had.co.nz/Computing-on-the-language.html
[dplyr-programming]: https://dplyr.tidyverse.org/articles/programming.html#warm-up

```{r test-nse-fixed, error=TRUE}
test_nse <- function(data, filter_condition) {
    data %>% 
        dplyr::filter({{ filter_condition }})
}

CO2 %>% 
    test_nse(conc > 100)

CO2 %>% 
    test_nse(uptake < 30)
```
