# Resources 

## For this course

### Potential exercise solutions

These are *potential* exercise solutions of R code only.

```{r all-exercise-solutions, results="asis", echo=FALSE, eval=TRUE, message=FALSE}
library(magrittr)
source(here::here("R/extract-solutions.R"))
solutions <- purrr::map(
    c("02-importing.Rmd", "03-dry.Rmd", "04-dplyr.Rmd"),
    extract_solution_chunks
)
unlist(solutions) %>% 
    cat(sep = "\n")
```

## For continued learning

**Free online books**:

- [Efficient R Programming](https://csgillespie.github.io/efficientR/): Excellent book on being efficient when writing R code.
- [Advanced R](https://adv-r.hadley.nz/): Detailed book on advanced features of R.
- [R Packages](https://r-pkgs.org/): Learn how to create R Packages from the
basics.
- [R for Data Science](https://r4ds.had.co.nz/): 
Excellent open and online resource for using R for data analysis and data science.
- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/):
Great overview of using R for Data Science, with more of a focus on the programming
side of things
- [ModernDive: Statistical Inference via Data Science](https://moderndive.com/): 
Great book on using statistics and data science methods in R
- [Happy Git and GitHub for the useR](https://happygitwithr.com/) (**highly recommended**):
Specifically useful is the chapter on [Daily Workflows](https://happygitwithr.com/repeated-amend.html) using Git
- [Online book for R Markdown](https://bookdown.org/yihui/rmarkdown/):
The go-to reference for learning and using R Markdown
- [Course material for a statistics class](https://stat545.com/):
Excellent course material for teaching statistics and R.

**Quick references**:

- [RStudio cheatsheets](https://www.rstudio.com/resources/cheatsheets/):
Multiple, high-quality cheatsheets you can print off to use as a handy reference
- [Tidyverse style guide](https://style.tidyverse.org/):
To learn about how to write well styled code in R
- [Tidyverse design philosophy of writing code](https://design.tidyverse.org/)

**Articles**:

- [Good enough practicies in scientific computing](https://doi.org/10.1371/journal.pcbi.1005510):
An article listing and describing some practices to use when writing code
- [Best practices in scientific computing](https://doi.org/10.1371/journal.pbio.1001745)

**General sites**

- [Case studies](https://www.practicereproducibleresearch.org/) and lessons for doing reproducibility
- [Organizing R Source Code](https://www.r-bloggers.com/r-best-practices-r-you-writing-the-r-way/)
- [RStudio tutorial on using R Markdown](https://rmarkdown.rstudio.com/lesson-1.html)
- [Markdown syntax guide](https://rmarkdown.rstudio.com/authoring_basics.html)
- [Pandoc Markdown Manual](https://pandoc.org/MANUAL.html#pandocs-markdown) (R
Markdown uses [pandoc])

**Interactive sites or resources for hands-on learning**:

- [Interactive tutorials for using R, within R](https://swirlstats.com/)
- [RStudio's Learning Primers](https://rstudio.cloud/learn/primers)

**Getting help**:
    
- [StackOverflow for tidyr](https://stackoverflow.com/questions/tagged/tidyr)
- [StackOverflow for dplyr](https://stackoverflow.com/questions/tagged/dplyr)
- Tip: Combine auto-completion with `::` to find new functions 
and documentation on the functions 
(e.g. try typing `base::` and then hitting Tab to show a list of all functions found in base R)

## Useful R packages

```{r table-common-packages, echo=FALSE, eval=TRUE}
knitr::kable(r3::useful_packages_list, caption = "Useful and common packages to use in data analysis.")
```
