---
search: false
format: rostools-theme-revealjs
bibliography: "../includes/references.bib"
---

# Welcome to the Intermediate R^3^ workshop!

- :heavy_check_mark: Accept the GitHub Organization invite
- :heavy_check_mark: Pick a group from the basket and go to that table
- :heavy_check_mark: Introduce yourself to your table mates
- :heavy_check_mark: Open your `LearnR3.Rproj` RStudio project

::: notes
Introduce teachers and helpers.
:::

# Motivation for this workshop {.center}

## How do you *exactly* do data processing and analysis? What's the actual workflow? :thinking: {.center}

::: notes
When I started out doing research during my Masters, I always wondered *how* do
researchers go about doing data analysis... what was their workflow *exactly*
like. No one ever really taught that. Even on online tutorials, mostly pieces of
code or how to use code are taught... but never teaching the bigger picture...
How do researchers in their daily work write R code and do their data analysis?

This question is the reason why the overall workflow is the primary focus in
this workshop and partly with the beginner one. I try to focus on the bigger
picture and the overall workflow you would do for doing data analysis.

So, why is it that there isn't much information on how researchers do data
analysis?
:::

# :shrug::confused:...Because code sharing is almost non-existent in science {.center}

Very few papers provide code [@Leek2017a, @Considine2017a]

::: notes
It's because code sharing basically doesn't exist in the vast majority of
scientific fields. You've probably read a methods section in a paper and
wondered how exactly they did it.
:::

## Code sharing: From scientific principle of "reproducibility" {.center}

...often confused with "replicability" [@Plesser2018a] (see also American
Statistical Association
[statement](https://www.amstat.org/asa/files/pdfs/POL-ReproducibleResearchRecommendations.pdf))

. . .

::: columns
::::: {.column width="50%"}
### Replicability

- Repeating a study by *independently* performing another identical study
- Linked to the "irreproducibility crisis" (aka "irreplicability crisis")
:::::

::::: {.column width="50%"}
### Reproducibility

- Generating the exact same results when using the same data and code
- *Question*: If we can't even *reproduce* a study's results, how can we expect
  to replicate it?
:::::
:::

## How can we check reproducibility if no code is given? {.center}

This is a little bit of a rhetorical question :stuck_out_tongue_closed_eyes:

## Very low reproducibility in most of science [@Samuel2024, @Trisovic2022] {.smaller}

::: {layout-ncol="2"}
```{r}
#| echo: false
#| fig-height: 6

library(tidyverse)
theme_set(
  theme_minimal() +
    theme(
      plot.background = element_rect(fill = "#E9E9E9"),
      axis.title = element_blank(),
      axis.text.y = element_text(size = 18),
      axis.text.x = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank()
    )
)

search <- tribble(
  ~item, ~description, ~value, ~order,
  "Articles", "Found among all articles", 3467, 1,
  "GitHub\nrepos", "Those with a GitHub link and at least one notebook", 2660, 2
) |>
  mutate(
    item = fct_rev(fct_reorder(item, order))
  )

ggplot(search, aes(y = item, x = value, label = value)) +
  geom_col(fill = "#203C6E") +
  geom_text(nudge_x = 200, size = 5.5)
```

```{r}
#| echo: false
#| fig-height: 6

repro_results <- tribble(
  ~item, ~description, ~value, ~order,
  "Total notebooks", "Total notebook files found in all the repositories", 27271, 1,
  "Notebook repo\nwas valid", "File names were correct, dependency information was available, no local modules were needed.", 15817, 2,
  "Could install\ndependencies", "Could install dependencies without errors.", 10388, 3,
  "Finished executing\nwithout errors", "Could run without any errors.", 1203, 4,
  "Same results\nas paper", "When the results of the paper matched what the execution results output.", 879, 5
) |>
  mutate(
    percent = str_c(value, "\n(", round(value / max(value) * 100), "%)"),
    item = fct_rev(fct_reorder(item, order))
  )

ggplot(repro_results, aes(y = item, x = value, label = percent)) +
  geom_col(fill = "#203C6E") +
  geom_text(nudge_x = 1700, size = 5.5)
```
:::

## Scientific culture is not well-prepared for analytic and computation era {.center}

Especially with the rise of "AI" tools

## With "AI" tools, generating (or "vibing") lots of code is easy... but... {.center}

## Writing code has never been the hard part {.center}

Designing, knowing what to do, planning, making decisions with trade-offs, etc have always been the hard parts.

::: notes
I want to stress this part... As beginner just learning how to write code, it
feels like writing code is the hard part. But as you get more experienced, you
realise that writing code is actually the easy part. The hard part is all the
other things...  designing, planning, communication, making decisions that come
with trade-offs, and even just knowing what you want.
:::

## LLM tools are just tools, that predict an output based on an average of their data {.center}

::: notes
Do we want to be aiming for and happy with average?
:::

## If you aren't experienced in code, how can you assess it? {.center}

::: notes
Problem with AI tools, it predicts the average next items. Key word,
average. Do you want to go for average? Is average code even good?
There's a lot of shit code in the world, that, the AI has been trained
on.
:::

## These issues can be fixed by creating and nurturing a culture of openness {.center}

::: notes
All of this is because of a problem with our culture in research. We aren't
open, we don't really share, and don't often follow basic principles of science.
To fix this, we need to start creating and nurturing a better and healthier
culture. We all can be involved in that, we all have that power to do something,
even if its small thing.
:::


## Goal of this course: Start changing the culture by providing the training {.center}

# Workshop overview

## Where does this course fit in the research lifecycle? {.center}

```{mermaid}
flowchart LR
  Collection --> Processing:::focus --> Analysis --> Writing --> Publishing
  classDef focus stroke:#f96,stroke-width:4px;
```

## Setup and layout {.smaller}

::::: columns
::: {.column width="50%"}
- Workshop is mix of:
    - "Code-alongs" (we type and explain, you type along)
    - Hands-on coding, discussing, and reading exercises
    - Extra hands-on exercises (we'll move on once everyone finished the main ones)
    - A team project at the end
-   Less about coding, more about:
    - Connecting and collaborating with others
    - Thinking about designing first, coding second
:::

::: {.column width="50%"}
- Networking and connecting:
    - Draw table name from bowl each morning
    - Introduce yourself to those at your table
    - During lunch, sit beside someone you don't know
    - Discussion activities
- Feedback survey for every session
:::
:::::

::: notes
Explain a bit more about the reading, why doing it, and that this workshop in
particular has a lot more of it than more advanced workshops.

Since this is also a space to socialise and network, we strongly encourage you
to sit besides someone you don't know during lunch. This is also why we have you
pick a piece of paper that has a table name on it every day, so that you can sit
with different people each day.

We collect feedback at the end of each session through the survey. It is
extremely helpful and has greatly improved the workshop material over the years.
So please, give us feedback!! We're all in this together, learning and improving
:grin:
:::

## Getting or asking for help :raising_hand_woman:

::::: columns
::: {.column width="50%"}
- Put the origami hat :tophat: on your laptop to get help
- There are lots of helpers
- Table mates, help out too
:::

::: {.column width="50%"}
- We're all learning here!
- This is a supportive and safe environment
- Remember our **Code of Conduct**
:::
:::::

## Practice using origami hats: Have you opened RStudio to `LearnR3.Rproj`? {.center}

# Activities

## :walking_woman::walking_man: How do you perceive your skill in R? {.center}

::: notes
Along the wall, arrange to one side is "novice/basic" and other side is
"advanced".
:::

## :walking_woman::walking_man: What percent of time do you spend doing data processing (if you do any)? {.center}

::: notes
Along the wall, arrange to one side is "0%" and other side is "100%".

If you don't do any go to the 0% side.
:::

## :walking_woman::walking_man: Do you wish you did more or less data processing? {.center}

## Alright, return to your seats! :chair: {.center}

# References {.smaller}
