<!--
NOTE: This is separate here in order to have a solutions section.
Include `solutions: true` in header to show solutions.
-->

## :technologist: Exercise: Finish setting the dependencies

**Time: \~10 minutes.**

1.  While we added `{readr}` to the function, we haven't added it to the
    `DESCRIPTION` file yet. In the **Console**, use
    `usethis::use_package()` to add the `{readr}` package to the
    `DESCRIPTION` file.

2.  There is one other function we use in the `import_cgm()` function.
    Find it, figure out what package it comes from, use `::` to
    explicitly state the package, and add it to `DESCRIPTION` file using
    `usethis::use_package()` in the **Console**.

::: {.content-hidden unless-meta="solutions"}

This is what the function should look like after using `::`.

```{r solution-includes-set-function-dependencies}
#| echo: false
import_cgm <- function(file_path) {
  cgm <- file_path |>
    readr::read_csv(
      show_col_types = FALSE,
      name_repair = snakecase::to_snake_case,
      n_max = 100
    )
  return(cgm)
}
```
:::

3.  There is one other package we've used that we haven't added to the
    `DESCRIPTION` file. We used the package in the `data-raw/dime.R`
    file. Open that file, which you can do with
    {{< var keybind.open-file >}} and typing "dime.R" and selecting the
    file from the menu. In that file, find the package we used and add
    it to the `DESCRIPTION` file using `usethis::use_package()` in the
    **Console**.

::: {.content-hidden unless-meta="solutions"}
Run these in the **Console** to add the packages to the `DESCRIPTION` file.

```{r solution-includes-set-project-dependencies}
#| filename: Console
#| eval: false
usethis::use_package("readr")
usethis::use_package("snakecase")
usethis::use_package("fs")
`
``
:::

```{r admin-includes-print-desc-add-others}
#| echo: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_package("readr")
    usethis::use_package("snakecase")
    usethis::use_package("fs")
  },
  quiet = TRUE
)
```

4.  Finally, style the code using {{< var keybind.styler >}}, render the
    `docs/learning.qmd` file with {{< var keybind.render >}} to test
    that everything still works, and then add and commit the changes to
    Git with {{< var keybind.git >}}.

{{< text_snippet sticky_up >}}
