# Basic setup and workflow {#intro}

## Introduction to course

**[Introduction slides](slides/introduction.html)**

```{r slides-page-frame, echo=FALSE}
knitr::include_url("slides/introduction.html", height = "375px")
```

`r details_for_instructors("The slides contain speaking notes that you can view by pressing 'p' on the keyboard.")`

## Overview of workflow

**Take 10 min to read over this section before we go over it together**.

This section provides an bigger picture view of what we will be doing, 
why we want to do it, how we will be going about doing it, and what it will look
like in the end. Firstly, what we ultimately want to do is process the MMASH
data so that we have a single dataset to work with for later (hypothetical)
analyses. We also want to make sure that whatever processing we do to the data
is reproducible. So everything we'll do throughout the course is done in order
to achieve our aim of making a single data and to make things reproducible.

Both the folder and file structures below as well as the 
Figure \@ref(fig:diagram-overview-0) show exactly what we will be doing and how
it will look like at the file level (not at the R code level).
Hopefully with this overview, you can better understand where we are and where
we want to get to. A comment on naming: whenever folder names are given,
they always end in `/`, for instance `data/` or `doc/`.

Right now, everyone's initial project structure should look like:

```
LearnR3
├── data/
│   └── README.md
├── data-raw/
│   ├── mmash-data.zip
│   ├── mmash/
│   │  ├── user_1
│   │  ├── ...
│   │  └── user_22
│   └── mmash.R
├── doc/
│   ├── README.md
│   └── lesson.Rmd
├── R/
│   ├── functions.R
│   └── README.md
├── .gitignore
├── DESCRIPTION
├── LearnR3.Rproj
├── README.md
└── TODO.md
```

At the end of this workshop, it should look something like:

```
LearnR3
├── data/
│   ├── README.md
│   └── mmash.rda
├── data-raw/
│   ├── mmash-data.zip
│   ├── mmash/
│   │  ├── user_1
│   │  ├── ...
│   │  └── user_22
│   └── mmash.R
├── doc/
│   ├── README.md
│   ├── lesson.html
│   └── lesson.Rmd
├── R/
│   ├── README.md
│   └── functions.R
├── .gitignore
├── DESCRIPTION
├── LearnR3.Rproj
└── README.md
```

Why do we structure it this way? Because we want to follow some standard conventions
in R, like having a `DESCRIPTION` file, keeping raw data in the `data-raw/` folder,
and keeping R scripts in the `R/`. We also want to keep things structured to make
it easier for others and ourselves to reproducible the work.

Our workflow will generally look like Figure \@ref(fig:diagram-overview-0),
with each block representing one or two sessions. We've already done a bit
of the first block, *"Download raw data"*.

```{r diagram-overview-0, fig.cap="Overview of the workflow we will be using and covering.", echo=FALSE}
knitr::include_graphics("images/diagram-overview-0.svg")
```

Our workflow and process will be something like:

- Make an R script (`data-raw/mmash.R`) to download the dataset (`data-raw/mmash/`),
which you already did in the pre-course tasks. 
    - We do this to keep the code for downloading and processing the data
    together with the raw data.
- Use R Markdown (`doc/lesson.Rmd`) to write and test out code, convert it into
a function, test it, and then move it into `R/functions.R`.
    - We use R Markdown to prototype code because 1) it's easier from a teaching
    perspective to interweave text and comments with code as we do the
    code-alongs, 2) to make it easier to translate your work into a format for easy
    dissemination (e.g. webpage, Word Doc, or PDF), and 3) because we want to
    you to practice and reinforce skills in using and writing R Markdown since
    it is a key component to doing reproducible research in R. Another alternative
    way of working is to prototype ("write and test") code in a temporary R
    script in the `R/` folder.
    - We keep the functions in a separate file because we will frequently `source()`
    from it as we prototype and test out code in the R Markdown file. It also 
    creates a clear separation between "finalized" code and prototyping code.
- Use a combination of restarting R (`Ctrl-Shift-F10` or "Session -> Restart
R") and using `source()` (`Ctrl-Shift-S` while in `R/functions.R`) to run the
functions inside of `R/functions.R`.
    - We restart R because it is the only sure way there is to clearing up the
    R workspace and starting for a "clean plate". And for reproducibility,
    we should always aim to working from a clean plate.
- Replace the old code in the R Markdown document (`doc/lesson.Rmd`) with the
new function.
- Finish building functions that prepare the dataset, use them in the cleaning
script (`data-raw/mmash.R`), run the script and save the data as an `.rda`
file (`data/mmash.rda`).
    - Like with the `R/functions.R` script, we move the finalized data
    processing code over from the R Markdown into the `data-raw/mmash.R`
    script to have a clear separation of completed code and prototyping code.
- Remove the old code from the R Markdown document (`doc/lessons.Rmd`).
    - Once we've finished prototyping code and moved it over into a more "final"
    location, we remove left over code because it isn't necessary and we don't
    want our files to be littered with left over code. Again, like with restarting
    R often, we want to work from a clean plate.
- Whenever we complete a task, we *add and commit* those file changes and save
them into the Git history.
    - We use Git because it is the best way of keeping track of what was done to
    your files, when, and why. It keeps your work transparent and makes it easier
    to share your code by uploading to GitHub. Using version control should be
    a standard practice to doing better science since it fits with the philosophy
    of doing science.

**Stop reading here**.

`r details_for_instructors("
Go over the text again, to reinforce the workflow, and the reasons why we do certain
things.
")`

## Exercise: Discuss the types of workflows you use

Time: 15 min

As a group, discuss what type of workflow you use for your own work. How does
it compare with the others in the group? How does it compare with what has been
described so far? Groups will briefly share what they've discussed.

## Setting up our project

Now that we've gone over the overview, let's get our project ready for the next
steps. But first, we need to do a few things. Since we did all that work in the
pre-course tasks downloading the data and unzipping it, we need to save these
changes to the Git history. Open the Git interface with either the Git icon at
the top near the menu bar, or with `Ctrl-Alt-M`. When the Git inferface opens up
we'll click the checkbox beside the `.gitignore` and `data-raw/mmash.R` files.
Then we write a commit message in the text box on the right, something like
"Code to download data zip file". Click the "Commit" button and close the Git
interface.

Next, delete one of the files we don't need by using the [fs] package.

[fs]: https://fs.r-lib.org/

In the **Console**, type out:

```{r remove-extra-files, eval=FALSE}
fs::file_delete("TODO.md")
```

Then, open up the `README.md` and fix some of the TODO items.
After cleaning everything up, now we need to use Git to add and commit all the 
current files into the history. Open up the Git interface in RStudio with
`Ctrl-Alt-M` or through the Git button. Write a message in the commit textbox
saying "Added initial files". Now we're ready for the next session!
