---
execute:
  eval: false
---

# Using nested conditionals to help clean columns {#sec-case-when}

::: {.callout-warning appearance="default"}
🚧 We are doing **major** changes to this workshop, so much of the content will be changed. 🚧
:::

```{r setup}
#| include: false
library(here)
library(fs)
library(tidyverse)
load(here("_temp/functionals.RData"))
source(here("R/functions.R"))
```

## Learning objectives

{{< include /includes/objectives/_case-when.qmd >}}

## Cleaning with nested conditionals

::: callout-note
## Reading task: \~6 minutes

There are many ways to clean up this particular problem, but probably
the easiest, most explicit, and programmatically accurate way of doing
it would be with the function `case_when()`. This function works by
providing it with a series of logical conditions and an associated
output if the condition is true. Each condition is processed
*sequentially*, meaning that if a condition is TRUE, the output won't be
overridden for later conditions. The general form of `case_when()` looks
like:

```{r}
#| eval: false
case_when(
  variable1 == condition1 ~ output,
  variable2 == condition2 ~ output,
  # (Optional) Otherwise
  TRUE ~ final_output
)
```

The optional ending is only necessary if you want a certain output if
none of your conditions are met. Because conditions are processed
*sequentially* and because it is the *last* condition, by setting it as
`TRUE` the final output will used. If this last `TRUE` condition is not
used then by default, the final output would be a missing value. A
(silly) example using age might be:

```{r}
#| eval: false
case_when(
  age > 20 ~ "old",
  age <= 20 ~ "young",
  # For final condition
  TRUE ~ "unborn!"
)
```

If instead you want one of the conditions to be `NA`, you need to set
the appropriate `NA` value:

```{r}
#| eval: false
case_when(
  age > 20 ~ "old",
  age <= 20 ~ NA_character_,
  # For final condition
  TRUE ~ "unborn!"
)
```

Alternatively, if we want missing age values to output `NA` values at
the end (instead of `"unborn!"`), we would exclude the final condition:

```{r}
#| eval: false
case_when(
  age > 20 ~ "old",
  age <= 20 ~ "young"
)
```

With `{dplyr}` functions like `case_when()`, it requires you be explicit
about the type of output each condition has since all the outputs must
match (e.g. all character or all numeric). This prevents you from
accidentally mixing e.g. numeric output with character output. Missing
values also have data types:

-   `NA_character_` (character)
-   `NA_real_` (numeric)
-   `NA_integer_` (integer)

Assuming the final output is `NA`, in a pipeline this would look like
how you normally would use `mutate()`:

```{r}
user_info_df |>
  mutate(age_category = case_when(
    age > 20 ~ "old",
    age <= 20 ~ "young"
  ))
```
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Briefly review the content again, to reinforce what they read.
:::

While still in the `docs/learning.qmd` file, we can use the `case_when()`
function to set `"before sleep"` as day 1 and `"wake up"` as day 2 by
creating a new column called `day`. (We will use `NA_real_` because the
other `day` columns are numeric, not integer.)

```{r}
#| filename: "docs/learning.qmd"
saliva_with_day_df <- saliva_df |>
  mutate(day = case_when(
    samples == "before sleep" ~ 1,
    samples == "wake up" ~ 2
  ))
saliva_with_day_df
```

...Now, let's use the `reduce()` with `full_join()` again:

```{r}
#| filename: "docs/learning.qmd"
list(
  user_info_df,
  saliva_with_day_df,
  summarised_rr_df,
  summarised_actigraph_df
) |>
  reduce(full_join)
```

We now have two rows per participant! Let's add and commit the changes
to the Git history with {{< var keybind.commit >}}.

## Summary

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Quickly cover this before finishing the session and when starting the
next session.
:::

```{r store-session-code}
#| include: false
save.image(here::here("_temp/dplyr-joins.RData"))
```
