# Cleaning characters and dates {#sec-characters-dates}

```{r setup}
#| include: false
library(here)
library(fs)
library(tidyverse)
load(here("_temp/functionals.RData"))
source(here("R/functions.R"))
```

## Learning objectives

{{< include /includes/objectives/_characters-dates.qmd >}}

## :speech_balloon: Discussion activity: Why can't we join datasets together yet?

::: {.callout-note collapse="true"}
## :teacher: Teacher note

The issues are related to the:

-   IDs, where the CGM and sleep data have `file_path_id` while
    `participant_details.csv` has `id`.
-   Dates, where the CGM data has `device_timestamp`, the sleep data has
    `date`, and the `participant_details.csv` has `start_date` and
    `end_date`.
-   Both the CGM and sleep data is measured at different times, but the
    same date, so a direct join wouldn't work.
-   The participant details dates are in two columns rather than one.
:::

**Time: \~8 minutes.**

We now have a function that imports all our CGM and sleep data, as well
as `import_dime()` that could import the `participant_details.csv` file.
The next direction we want to go to is to join them in some way so that
we can be able to analyse them together. But right now, we can't really
do that. There are two main reasons. Look at the data below and try to
identify them---don't peek ahead! :grin: Then discuss with a neighbour
to see if you can come to a consensus.

```{r admin-display-all-data}
#| echo: false
cat("CGM data")
here("data-raw/dime/cgm") |>
  import_csv_files() |>
  trim_filepath_for_book() |>
  print(n = 6)
cat("Sleep data")
here("data-raw/dime/sleep") |>
  import_csv_files() |>
  trim_filepath_for_book() |>
  print(n = 6)
cat("Participant details")
here("data-raw/dime/participant_details.csv") |>
  import_dime() |>
  print(n = 6)
```

1.  For 1 minute, look at the above data and try to identify the reasons
    why we can't (easily) join the datasets together right now.
2.  For \~4 minutes, discuss what you've thought about with your
    neighbour
3.  Then for about 3 minutes we will all share and discuss together.

## :book: Reading task: Cleaning character data with regular expressions

**Time: \~5 minutes.**

Before we go into joining datasets together, we have to do a bit of
processing first. Specifically, we want to get the participant ID from
the `file_path_id` character data since all the datasets have an ID (in
some form). Whenever you are processing and cleaning data, you will very
likely encounter and deal with character data. A wonderful package to
use for working with character data is called `{stringr}`, which we'll
use to get the participant ID from the `file_path_id` column. The
package is called `{stringr}` because in programming, a "string" is
defined as any collection of character objects. So we will be using
"string" from now on.

The main "engine" behind the functions in `{stringr}` is called [regular
expressions](https://en.wikipedia.org/wiki/Regular_expression) (or regex
for short). These expressions are powerful, very concise ways of finding
patterns in text. But they are also *very very difficult* to learn,
write, and read, even for experienced users. That's because certain
strings like `[` or `?` have special meanings. For instance, `[aeiou]`
is regex for "find one string in the object that is either a, e, i, o,
or u". The `[]` in this case means "find the string in between the two
brackets". We won't cover regex too much in this workshop, but some
great resources for learning them are the [R for Data Science regex
section](https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions),
the [stringr regex
page](https://stringr.tidyverse.org/articles/regular-expressions.html),
as well as in the help doc `?regex`. But we will cover some of the more
commonly used regex patterns, like:

| regex | Meaning |
|----|----|
| `\\d` | find one digit (0-9) |
| `\\D` | find one non-digit |
| `[0-9]` | find one digit between 0 and 9 (the `-` is a range) |
| `[a-z]` | find one letter between a and z |
| `[A-Z]` | find one letter between A and Z |
| `[a-zA-Z]` | find one letter between a and z or A and Z |
| `\\.` | find one dot string |
| `$` | find the end of the string, e.g. `\\.$` means find a dot that is at the end of the string |
| `^` | find the start of the string, e.g. `^\\d` means find a digit at the start of the string |
| `?` | maybe find the string before the `?`, e.g. `\\d?` means maybe find one digit |
| `*` | find the string before it, but maybe many times, e.g. `\\d*` means find zero or many digits one after the other |
| `+` | find the string one or more times before `{}`, e.g. `\\d+` means find one or more digits one after the other |
| `{n}` | find the same string `n` times before `{}`, e.g. `\\d{2}` means find two digits one after the other |
| `[:alpha:]` | find one letter (a-z or A-Z) |
| `[:digit:]` | find one digit (0-9) |
| `[:alnum:]` | find one alphanumeric (a-z, A-Z, 0-9) |

::: callout-note
The `\` string is a special string in regex, but in R it means "escape
or use this as the string". For instance, `\"` in R means "use the `"`
string", like if you wanted to write a message like:

```{r}
print("Analysis has \"nearly\" finished.")
```

Without the `\`, R would see this instead:

``` r
print("Analysis has "nearly" finished.")
```

And think `"Analysis has "` is the first string, that `nearly` is an
object, and the `" finished"` is the second string object.

But in regex, the `\` string is a special command that can mean
different things. For instance, `\d` means "find a digit" (0-9) and `\D`
means "find a non-digit". So when using it in R, it would be `\\d` and
`\\D` so that regex "sees", e.g., `\d`
:::

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: Brainstorm a regex that will get the participant ID

**Time: \~8 minutes.**

https://github.com/rstudio/cheatsheets/blob/main/strings.pdf

Using what you just read from above, try to think of a regex that would
get the participant ID from the file path. Try not to look ahead nor at
the potential solutions! :grin: :wink:

When the time is up, we'll share some ideas and go over what the regex
*could* be (there is actually many different ways to do this). With your
neighbour, do the following tasks:

1.  Looking at the `file_path_id` column, list what is similar in the
    user ID between rows and what is different.
2.  Discuss and verbally describe (in English, not regex) what text
    pattern you might use to extract the user ID.
3.  Use the list below to think about how you might convert the English
    description of the text pattern to a regex. This will probably be
    very hard, but try anyway.
    -   When strings are written as is, regex will find those strings,
        e.g. `user` will find only `user`.
    -   Use `[]` to find one possible string of the several between the
        brackets. E.g. `[12]` means 1 *or* 2 or `[ab]` means "a" or "b".
        To find a range of numbers or letters, use `-` in between the
        start and end ranges, e.g. `[1-3]` means 1 to 3 or `[a-c]` means
        "a" to "c".
    -   Use `?` if the string might be there or not. E.g. `ab?` means
        "a" and maybe "b" follows it or `1[1-2]?` means 1 and maybe 1 or
        2 will follow it.

Once you've done these tasks, we'll discuss all together and go over
what the regex would be to extract the user ID.

::: {.callout-note appearance="minimal" collapse="true"}
## Potential regex solutions

-   `[0-9][0-9][0-9]\\.csv$`
-   `[0-9]+\\.csv$`
-   `[0-9]{3}\\.csv$`
-   `\\d+\\.csv$`
-   `\\d{3}\\.csv$`
-   `[:digit:]+\\.csv$`
-   `[:digit:]{3}\\.csv$`

For the rest of the workshop, we will use the regex `[:digit:]+\\.csv$`
to get the participant ID from the file path. We use this one because it
is more readable than the others and it allows us to get participant IDs
that are one to many digits long.
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Make sure to reinforce that while regex is incredibly complicated, there
are some basic things you can do with it that are quite powerful.

More or less, this section and exercise are to introduce the idea and
concept of regex, but not to really teach it since that is well beyond
the scope of this workshop and this time frame.

Go over the solution. The explanation is that the pattern will find
anything that has `\\.csv$` at the end that is preceded by some numbers.
:::

## Using regular expressions to extract text

Now that we've identified a possible regex to use to extract the
participant ID, let's try it out on the `cgm_data`. Before we do though,
let's make it easier on us to use the `cgm_data` in our Quarto document
by importing it in the `setup` chunk. So, in the `docs/learning.qmd`
file, go to the top of the file, find the `setup` chunk, and in the
bottom of it below the `library()` functions, add the following code:

```{r add-cgm-data-to-setup}
#| filename: "docs/learning.qmd"
cgm_data <- here("data-raw/dime/cgm") |>
  import_csv_files()
```

Then we can run this code chunk by using {{< var keybind.run-code >}} so
that we have the `cgm_data` object available to us to work with. Then we
can continue!

We want to create a new column called `id` for the participant ID rather
than modify the current `file_path_id` column, so we will use the
`mutate()` function from the `{dplyr}` package to create this new
column. We'll use the `str_extract()` function from the `{stringr}`
package to "extract a string" by using the regex `[:digit:]+\\.csv$`
that we discussed from the activity above.

We're also using an argument to `mutate()` you might not have seen
previously, called `.before`. This will insert the new `id` column
before the column we use and we do this entirely for visual reasons,
since it is easier to see the newly created column when we run the code.
In your `docs/learning.qmd` file, create a new header called
`## Using regex for ID` at the bottom of the document, and create a new
code chunk below that with {{< var keybind.chunk >}}.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Walk through writing this code, briefly explain/remind how to use
mutate, and about the `{stringr}` function.
:::

First we'll pipe `cgm_data` into `mutate()` to create a new column
called `id` that will contain the participant ID. We'll use the
`str_extract()` function to extract the participant ID from the
`file_path_id` column using the regex `[:digit:]+\\.csv$`.

```{r get-id}
#| filename: "docs/learning.qmd"
#| eval: false
# Note: your file paths and data may look slightly different.
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ),
    .before = file_path_id
  )
```

```{r admin-get-id}
#| echo: false
cgm_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ),
    .before = file_path_id
  )
```

Cool! But we have `.csv` at the end that we don't want. We can use the
`str_remove()` function to remove the `.csv` from the string. So we can
pipe our output into that function:

```{r get-id-remove-csv}
#| filename: "docs/learning.qmd"
#| eval: false
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$"),
    .before = file_path_id
  )
```

```{r admin-get-id-remove-csv}
#| echo: false
cgm_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$"),
    .before = file_path_id
  )
```

Nice! Almost there though. The `id` column is actually set as a
character type, but we will need it as an integer to join with
`participant_details.csv`. So we can use the `as.integer()` function to
convert it to an integer:

```{r get-id-as-integer}
#| filename: "docs/learning.qmd"
#| eval: false
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$") |>
      as.integer(),
    .before = file_path_id
  )
```

```{r admin-get-id-as-integer}
#| echo: false
cgm_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$") |>
      as.integer(),
    .before = file_path_id
  )
```

Finally, we don't actually want to keep the `file_path_id` column, so we
can use the `select()` function to drop it. We can do this by using the
`-` operator to drop the column. So we can pipe our output into that
function:

```{r drop-file-path-id}
#| filename: "docs/learning.qmd"
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$") |>
      as.integer(),
    .before = file_path_id
  ) |>
  select(-file_path_id)
```

We've done it! We now have a new column called `id` that contains the
participant ID from the `file_path_id` column. Now, time to make it into
a function!

## :technologist: Exercise: Convert 'get ID' code into a function

**Time: \~15 minutes.**

You now have code that takes the data that has the `file_path_id` column
and gets the participant ID from it. You want to be able to easily use
that on the sleep data too. So, as you've done, convert this code in the
`docs/learning.qmd` file into a function.

{{< include /includes/solutions/_characters-dates-get-id-function.qmd >}}

By the end of this exercise, the code to import both the CGM and sleep
data should look like this in the `setup` chunk of the
`docs/learning.qmd` file:

```{r get-participant-id-interface}
# CGM data
cgm_data <- here("data-raw/dime/cgm") |>
  import_csv_files() |>
  get_participant_id()
# Sleep data
sleep_data <- here("data-raw/dime/sleep") |>
  import_csv_files() |>
  get_participant_id()
```

And when you run the `cgm_data` and `sleep_data` in the **Console**, it
should look something like:

```{r show-cgm-sleep-with-id}
#| filename: "Console"
cgm_data
sleep_data
```

So, convert the code we just wrote into a function by following these
steps:

1.  Assign `function()` to a new named function called
    `get_participant_id` and include one argument in `function()` with
    the name `data`.
2.  Put the code we just wrote into the body of the function. Make sure
    to `return()` the output at the end of the function and to create
    Roxygen documentation. Replace all the relevant variables with the
    `data` argument.
3.  Add `stringr` and `dplyr` as a package dependency by using
    `usethis::use_package()` in the **Console**.
4.  Explicitly link the functions you are using in this new function to
    their package by using `::` (e.g. `dplyr::` and `stringr::`).
    Remember that you can find which package a function belongs by using
    the `?` help documentation and looking at the very top to see the
    package name.
5.  Test that the function works by using `get_participant_id(cgm_data)`
    or `get_participant_id(sleep_data)`.
6.  After creating the function and testing that it works, move (**cut**
    and paste) the function into `R/functions.R`.
7.  Run `{styler}` in both the `R/functions.R` file and
    `docs/learning.qmd` with {{< var keybind.style >}}.
8.  In the `setup` chunk in the `docs/learning.qmd` file, pipe `|>` the
    output of `import_csv_files()` into the `get_participant_id()`
    function for both the CGM and sleep data (as shown above).
9.  Render the `docs/learning.qmd` file to make sure things remain
    reproducible with {{< var keybind.render >}}.
10. Restart your R session with {{< var keybind.restart-r >}} to make
    sure later work is in a clean state.
11. Add and commit the changes to the Git history with
    {{< var keybind.commit >}}. Then push to GitHub.

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: How to join by dates?

**Time: \~10 minutes.**

Now that we have the participant ID, we are nearly able to join the CGM
data with the sleep data. But before we do that, we need to make sure
that the date columns are in the same format. The CGM data has a date
column called `device_timestamp`, which is a date-time column, while the
sleep data has a date column called `date`, which is a date column.

But they also both have different times for when the data was recorded
by the devices. We can't join by exact time, since the CGM data is not
recorded at exactly the same time as the sleep data is. So how do we go
about joining these two datasets together? There is no right or correct
answer as it heavily depends on how you might want to analyse the data
later on.

1.  For 1 minute, consider theoretically how you would change the date
    columns so that they can be joined together.
2.  For 3 minutes, discuss and brainstorm with your neighbour how you
    might do this and try to come to a consensus.
3.  For 2 minutes, we will all share our ideas all together as a group.

::: {.callout-note collapse="true"}
## :teacher: Teacher note

There is no right answer here, but you can discuss some ideas like:

-   Create a date column and an hour column, later group by ID, date,
    and hour, summarise the values from each dataset first, and then
    join the two datasets together.
-   Create a date column, group by ID and date, summarise the values,
    and then join the two datasets together.

We would like to have more data points to work with, so we will go with
the first option.
:::

## Finding the right date

Working with dates can be surprisingly incredibly difficult and tricky.
There's timezones, daylight savings time, leap years, and different date
formats that all can cause a dozen different problems. In R, there is a
wonderful package called `{lubridate}` that helps with working with
dates and times. It can't solve all date and time problems, but it can
help with a many of them.

Thankfully, `{lubridate}` is part of the `{tidyverse}` ecosystem, so we
don't need to explicitly load it.

In the `docs/learning.qmd` file, at the bottom of the document create a
new code chunk called `## Working with dates` and below that create a
new code chunk with {{< var keybind.chunk >}}. We'll use `mutate()`
again to create some new columns, as well as `.before` to put it before
the `device_timestamp`.

As we discussed in the activity above, we will create a new `date` and
`hour` column. `{lubridate}` has the function `as_date()` to convert a
column into just a date (dropping time if it has it). It also has
`as_time()`, but since we want to get *just* the hour, it also has a
function to do that too: `hour()`. There are many other helper functions
like `minute()`, `year()`, `month()`, and so on. For now, we will keep
the `device_timestamp` column, so we have it as reference, but we will
drop it later on.

```{r}
cgm_data |>
  mutate(
    date = as_date(device_timestamp),
    hour = hour(device_timestamp),
    .before = device_timestamp
  )
```

Like we normally do, we can convert this into a function because we need
the same things from the sleep data too. So, let's use our workflow
we've done several times before and try it out.

Since we are using `{lubridate}`, let's add it as a package dependency
first:

```{r}
#| filename: Console
#| eval: false
usethis::use_package("lubridate")
```

Now, let's make the function!

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Use the workflow we've done rather than directly writing the function
below as is. Meaning, start with making the function structure, cut and
paste the code into it, add the `package::` to the functions, add the
Roxygen documentation, and then test it out as shown in the code below.
:::

```{r convert-function-prepare-dates}
#| filename: "docs/learning.qmd"
#| error: true
#' Prepare the date columns in DIME CGM and sleep data for joining.
#'
#' @param data The data that has the datetime column.
#' @param column The datetime column to convert to date and hour.
#'
#' @returns A tibble/data.frame
#'
prepare_dates <- function(data, column) {
  prepared_dates <- data |>
    dplyr::mutate(
      date = lubridate::as_date(column),
      hour = lubridate::hour(column),
      .before = column
    )
  return(prepared_dates)
}
cgm_data |>
  prepare_dates(device_timestamp)
```

Hmmm, we encountered an error! This error is something that you would
very quickly encounter as you make functions using `{tidyverse}`
functions. Read the next section to learn about what is happening.

## :book: Reading task: Non-standard evaluation (NSE)

::: {.callout-note collapse="true"}
## :teacher: Teacher note

After they've read it, briefly review it again and try to explain what
NSE is in a slightly different way. Try to emphasise that NSE is
built-in to so many functions in R and that using `{{ }}` with
`{tidyverse}` functions can help with using it in their own functions.
:::

**Time: \~5 minutes.**

As you have noticed, we encountered an error when trying to write, what
we think, is correct code in our function. This error,
`"object not found"` is not easy to figure out. The error occurs because
of something called "[non-standard
evaluation](https://adv-r.had.co.nz/Computing-on-the-language.html)" (or
NSE). NSE is a feature of R and is used quite a lot throughout R (e.g.
`library()`). The `{tidyverse}` packages in particular use it quite a
lot. It is a concept that isn't commonly found in other programming
language.

In R, NSE is what allows you to use formulas (e.g. `y ~ x + x2`) in
statistical models or allows you to type out `select(Gender, BMI)` or
`library(purrr)`. In "standard evaluation", these would instead be
`select("Gender", "BMI")` or `library("purrr")`. So NSE gives
flexibility and ease of use for the user (we don't have to type quotes
every time) when doing data analysis, but can be quite a headache when
trying to program in R, like when making functions. There's more detail
about this on the [dplyr
website](https://dplyr.tidyverse.org/articles/programming.html#warm-up),
which will give a few options to deal with NSE while programming with
`{tidyverse}` packages. The simplest way to use NSE in your own
functions when using `{tidyverse}` packages is to wrap the variable with
`{{ }}` (called "curly-curly"). For example, using a simplified version
of the code we wrote above, instead of this:

```{r}
#| eval: false
prepare_dates <- function(data, column) {
  data |>
    mutate(date = as_date(column))
}
```

You would use this:

```{r}
#| eval: false
prepare_dates <- function(data, column) {
  data |>
    mutate(date = as_date({{ column }}))
}
```

By wrapping `column` with `{{ }}`, we are telling R that we are giving
it an unquoted variable and to give that variable to the `as_date()`
function as the name of the column. It gets very technical very quickly,
but the main thing to remember is that when you make functions that use
`{tidyverse}` functions and you get this type of error, it is likely
because of NSE. So you can use `{{ }}` to wrap the variable name in your
function.

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Using NSE in your function

**Time: \~10 minutes.**

Using what you just read about NSE, fix the `prepare_dates()` to
correctly use the `column` argument.

Before fixing the function though, you'll need to fix the `sleep_data`
date column. We made `prepare_dates()` to create a new column called
`date` and `hour`, but the sleep data already has a column called
`date`. But, this name is incorrect, since it actually contains datetime
data. So you need to rename it to `datetime`. To do this, you can use
the `rename()` function from the `{dplyr}` package. So, in the `setup`
code chunk in `docs/learning.qmd`, add a pipe after
`get_participant_id()` into `rename(datetime = date)`, like:

```{r rename-sleep-date-setup}
#| filename: "docs/learning.qmd"
sleep_data <- here("data-raw/dime/sleep") |>
  import_csv_files() |>
  get_participant_id() |>
  rename(datetime = date)
```

At the end of this exercise, the `setup` code chunk should look like
this:

{{< include /includes/solutions/_characters-dates-using-nse.qmd >}}

```{r setup-cgm-sleep-with-nse}
#| filename: "docs/learning.qmd"
cgm_data <- here("data-raw/dime/cgm") |>
  import_csv_files() |>
  get_participant_id() |>
  prepare_dates(device_timestamp)

sleep_data <- here("data-raw/dime/sleep") |>
  import_csv_files() |>
  get_participant_id() |>
  rename(datetime = date) |>
  prepare_dates(datetime)
```

And if you were to run the `cgm_data` and `sleep_data` in the
**Console**, it should look something like:

```{r show-cgm-sleep-with-dates}
#| filename: "Console"
cgm_data
sleep_data
```

To get to that point, complete these tasks:

1.  Wrap `column` with `{{ }}` the three times it is used *inside* the
    function *body*.
2.  Test that the function works by using
    `prepare_dates(cgm_data, device_timestamp)` and
    `prepare_dates(sleep_data, datetime)`.
3.  Cut and paste the function into `R/functions.R`.
4.  Run `{styler}` on both `R/functions.R` and `docs/learning.qmd` with
    {{< var keybind.style >}}.
5.  Render the `docs/learning.qmd` file to make sure things remain
    reproducible with {{< var keybind.render >}}.
6.  Restart your R session with {{< var keybind.restart-r >}} so that
    your environment is cleaned up and empty.
7.  Add and commit the changes to the Git history with
    {{< var keybind.commit >}} and then push to GitHub.

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

-   While very difficult to learn and use, regular expressions (regex or
    regexp) are incredibly powerful at processing character (also called
    strings) data. Some strings have special meanings like `[]`, `?`,
    `*`, and `+` that can be used to find patterns in string data.
-   Use the `{stringr}` package to work with string data as it has many
    functions to help use regex to process string data, like
    `str_extract()` or `str_remove()`.
-   Use the `{lubridate}` package when working with dates and times, as
    there are many functions in it to help simplify date processing,
    like `as_date()` or `hour()`.
-   Non-standard evaluation (NSE) is a great feature of R but also can
    be frustrating. When using `{tidyverse}` functions in your own
    function, use `{{}}` to wrap the variable that gives the error to
    fix the problem and use NSE correctly.

## :speech_balloon: Discussion activity: How have you worked with dates and characters?

**Time: \~10 minutes.**

As we prepare for the next thing (either lunch or the next session), get
up (and if it's nice, go outside), and discuss with your neighbour or
group the following questions:

1.  Have you worked with characters and/or dates before in your work?
    How was that experience? Easy, hard?
2.  Can you see some useful ways you could use regex in your work?

```{r admin-store-session-code}
#| include: false
save.image(here::here("_temp/characters-dates.RData"))
```

{{< include /_extensions/rostools/r3-theme/includes/code-appendix-r.qmd >}}
