---
execute:
  eval: false
---

# Cleaning characters and dates {#sec-characters-dates}

```{r setup}
#| include: false
library(here)
library(fs)
library(tidyverse)
load(here("_temp/functionals.RData"))
source(here("R/functions.R"))
```

::: callout-caution
:construction_worker: This session is being actively developed and
edited. It will be added here once it is done and ready.
:::

## Learning objectives

{{< include /includes/objectives/_characters-dates.qmd >}}

::: content-hidden

## :speech_balloon: Discussion activity: Why can't we join datasets together yet?

::: {.callout-note collapse="true"}
## :teacher: Teacher note

The issues are related to the:

- IDs, where the CGM and sleep data have `file_path_id` while
  `participant_details.csv` has `id`.
- Dates, where the CGM data has `device_timestamp`, the sleep data has
  `date`, and the `participant_details.csv` has `start_date` and
  `end_date`.
- Both the CGM and sleep data is measured at different times, but the
  same date, so a direct join wouldn't work.
- The participant details dates are in two columns rather than one.

:::

**Time: \~8 minutes.**

We now have a function that imports all our CGM and sleep data, as well
as `read()` that could read in the `data-raw/nurses-stress/survey-results.csv.gz` file.
The next direction we want to go to is to join them in some way so that
we can be able to analyse them together. But right now, we can't really
do that. There are two main reasons. Look at the data below and try to
identify them---don't peek ahead! :grin: Then discuss with a neighbour
to see if you can come to a consensus.

```{r admin-display-all-data}
#| echo: false
cat("CGM data")
read_all("HR.csv.gz") |>
  trim_filepath_for_book() |>
  print(n = 6)
cat("Sleep data")
here("data-raw/dime/sleep") |>
  import_csv_files() |>
  trim_filepath_for_book() |>
  print(n = 6)
cat("Participant details")
here("data-raw/nurses-stress/survey-results.csv.gz") |>
  read() |>
  print(n = 6)
```

1. For 1 minute, look at the above data and try to identify the reasons
   why we can't (easily) join the datasets together right now.
2. For \~4 minutes, discuss what you've thought about with your
   neighbour
3. Then for about 3 minutes we will all share and discuss together.

## :book: Reading task: Cleaning character data with regular expressions

**Time: \~5 minutes.**

Before we go into joining datasets together, we have to do a bit of
processing first. Specifically, we want to get the participant ID from
the `file_path_id` character data since all the datasets have an ID (in
some form). Whenever you are processing and cleaning data, you will very
likely encounter and deal with character data. A wonderful package to
use for working with character data is called `{stringr}`, which we'll
use to get the participant ID from the `file_path_id` column. The
package is called `{stringr}` because in programming, a "string" is
defined as any collection of character objects. So we will be using
"string" from now on.

The main "engine" behind the functions in `{stringr}` is called
[regular expressions](https://en.wikipedia.org/wiki/Regular_expression)
(or regex for short). These expressions are powerful, very concise ways
of finding patterns in text. But they are also *very very difficult* to
learn, write, and read, even for experienced users. That's because
certain strings like `[` or `?` have special meanings. For instance,
`[aeiou]` is regex for "find one string in the object that is either a,
e, i, o, or u". The `[]` in this case means "find the string in between
the two brackets". We won't cover regex too much in this workshop, but
some great resources for learning them are the
[R for Data Science regex section](https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions),
the
[stringr regex page](https://stringr.tidyverse.org/articles/regular-expressions.html),
as well as in the help doc `?regex`. There's also a nice
[cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf)
on it. But we will cover some of the more commonly used regex patterns,
like:

| regex       | Meaning                                                                                                         |
|-------------|-----------------------------------------------------------------------------------------------------------------|
| `\\d`       | find one digit (0-9)                                                                                            |
| `\\D`       | find one non-digit                                                                                              |
| `[0-9]`     | find one digit between 0 and 9 (the `-` is a range)                                                             |
| `[a-z]`     | find one letter between a and z                                                                                 |
| `[A-Z]`     | find one letter between A and Z                                                                                 |
| `[a-zA-Z]`  | find one letter between a and z or A and Z                                                                      |
| `\\.`       | find one dot string                                                                                             |
| `$`         | find the end of the string, e.g. `\\.$` means find a dot that is at the end of the string                       |
| `^`         | find the start of the string, e.g. `^\\d` means find a digit at the start of the string                         |
| `?`         | maybe find the string before the `?`, e.g. `\\d?` means maybe find one digit                                    |
| `*`         | find the string before it, but maybe many times, e.g. `\\d*` means find zero or many digits one after the other |
| `+`         | find the string one or more times before `{}`, e.g. `\\d+` means find one or more digits one after the other    |
| `{n}`       | find the same string `n` times before `{}`, e.g. `\\d{2}` means find two digits one after the other             |
| `[:alpha:]` | find one letter (a-z or A-Z)                                                                                    |
| `[:digit:]` | find one digit (0-9)                                                                                            |
| `[:alnum:]` | find one alphanumeric (a-z, A-Z, 0-9)                                                                           |

::: callout-note
The `\` string is a special string in regex, but in R it means "escape
or use this literally as the string". For instance, `\"` in R means "use
the `"` string", like if you wanted to write a message like:

```{r}
print("Analysis has \"nearly\" finished.")
```

Without the `\`, R would see this instead:

``` r
print("Analysis has "nearly" finished.")
```

And think `"Analysis has "` is the first string, that `nearly` is an
object, and the `" finished"` is the second string object.

But in regex, the `\` string is a special command that can mean
different things. For instance, `\d` means "find a digit" (0-9) and `\D`
means "find a non-digit". So when using it in R, it would be `\\d` and
`\\D` so that regex "sees", e.g., `\d`
:::

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: Brainstorm a regex that will get the participant ID

**Time: \~8 minutes.**

Using what you just read from above, try to think of a regex that would
get the participant ID from the file path. Try not to look ahead nor at
the potential solutions! :grin: :wink:

When the time is up, we'll share some ideas and go over what the regex
*could* be (there is actually many different ways to do this). With your
neighbour, do the following tasks:

1. Looking at the `file_path_id` column, list what is similar in the
   user ID between rows and what is different.
2. Discuss and verbally describe (in English, not regex) what text
   pattern you might use to extract the user ID.
3. Use the list below to think about how you might convert the English
   description of the text pattern to a regex. This will probably be
   very hard, but try anyway.
   - When strings are written as is, regex will find those strings, e.g.
     `user` will find only `user`.
   - Use `[]` to find one possible string of the several between the
     brackets. E.g. `[12]` means 1 *or* 2 or `[ab]` means "a" or "b". To
     find a range of numbers or letters, use `-` in between the start
     and end ranges, e.g. `[1-3]` means 1 to 3 or `[a-c]` means "a" to
     "c".
   - Use `?` if the string might be there or not. E.g. `ab?` means "a"
     and maybe "b" follows it or `1[1-2]?` means 1 and maybe 1 or 2 will
     follow it.

Once you've done these tasks, we'll discuss all together and go over
what the regex would be to extract the user ID.

::: {.callout-note appearance="minimal" collapse="true"}
## Potential regex solutions

- `[0-9][0-9][0-9]\\.csv$`
- `[0-9]+\\.csv$`
- `[0-9]{3}\\.csv$`
- `\\d+\\.csv$`
- `\\d{3}\\.csv$`
- `[:digit:]+\\.csv$`
- `[:digit:]{3}\\.csv$`

For the rest of the workshop, we will use the regex `[:digit:]+\\.csv$`
to get the participant ID from the file path. We use this one because it
is more readable than the others and it allows us to get participant IDs
that are one to many digits long.
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Make sure to reinforce that while regex is incredibly complicated, there
are some basic things you can do with it that are quite powerful.

More or less, this section and exercise are to introduce the idea and
concept of regex, but not to really teach it since that is well beyond
the scope of this workshop and this time frame.

Go over the solution. The explanation is that the pattern will find
anything that has `\\.csv$` at the end that is preceded by some numbers.
:::

## Using regular expressions to extract text

Now that we've identified a possible regex to use to extract the
participant ID, let's try it out on the `cgm_data`. Before we do though,
let's make it easier on us to use the `cgm_data` in our Quarto document
by importing it in the `setup` chunk. So, in the `docs/learning.qmd`
file, go to the top of the file, find the `setup` chunk, and in the
bottom of it below the `library()` functions, add the following code:

```{r add-cgm-data-to-setup}
#| filename: "docs/learning.qmd"
cgm_data <- here("data-raw/dime/cgm") |>
  import_csv_files()
```

Then we can run this code chunk by using {{< var keybind.run-code >}} so
that we have the `cgm_data` object available to us to work with. Then we
can continue!

We want to create a new column called `id` for the participant ID rather
than modify the current `file_path_id` column, so we will use the
`mutate()` function from the `{dplyr}` package to create this new
column. We'll use the `str_extract()` function from the `{stringr}`
package to "extract a string" by using the regex `[:digit:]+\\.csv$`
that we discussed from the activity above.

We're also using an argument to `mutate()` you might not have seen
previously, called `.before`. This will insert the new `id` column
before the column we use and we do this entirely for visual reasons,
since it is easier to see the newly created column when we run the code.
In your `docs/learning.qmd` file, create a new header called
`## Using regex for ID` at the bottom of the document, and create a new
code chunk below that with {{< var keybind.chunk >}}.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Walk through writing this code, briefly explain/remind how to use
mutate, and about the `{stringr}` function.
:::

First we'll pipe `cgm_data` into `mutate()` to create a new column
called `id` that will contain the participant ID. We'll use the
`str_extract()` function to extract the participant ID from the
`file_path_id` column using the regex `[:digit:]+\\.csv$`.

```{r get-id}
#| filename: "docs/learning.qmd"
#| eval: false
# Note: your file paths and data may look slightly different.
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ),
    .before = file_path_id
  )
```

```{r admin-get-id}
#| echo: false
cgm_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ),
    .before = file_path_id
  )
```

Cool! But we have `.csv` at the end that we don't want. We can use the
`str_remove()` function to remove the `.csv` from the string. So we can
pipe our output into that function:

```{r get-id-remove-csv}
#| filename: "docs/learning.qmd"
#| eval: false
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$"),
    .before = file_path_id
  )
```

```{r admin-get-id-remove-csv}
#| echo: false
cgm_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$"),
    .before = file_path_id
  )
```

Nice! Almost there though. The `id` column is actually set as a
character type, but we will need it as an integer to join with
`participant_details.csv`. So we can use the `as.integer()` function to
convert it to an integer:

```{r get-id-as-integer}
#| filename: "docs/learning.qmd"
#| eval: false
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$") |>
      as.integer(),
    .before = file_path_id
  )
```

```{r admin-get-id-as-integer}
#| echo: false
cgm_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$") |>
      as.integer(),
    .before = file_path_id
  )
```

Finally, we don't actually want to keep the `file_path_id` column, so we
can use the `select()` function to drop it. We can do this by using the
`-` operator to drop the column. So we can pipe our output into that
function:

```{r drop-file-path-id}
#| filename: "docs/learning.qmd"
cgm_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "[:digit:]+\\.csv$"
    ) |>
      str_remove("\\.csv$") |>
      as.integer(),
    .before = file_path_id
  ) |>
  select(-file_path_id)
```

We've done it! We now have a new column called `id` that contains the
participant ID from the `file_path_id` column. Now, time to make it into
a function!

## :technologist: Exercise: Convert 'get ID' code into a function

**Time: \~15 minutes.**

You now have code that takes the data that has the `file_path_id` column
and gets the participant ID from it. You want to be able to easily use
that on the sleep data too. So, as you've done, convert this code in the
`docs/learning.qmd` file into a function.

{{< include /includes/solutions/_characters-dates-get-id-function.qmd >}}

By the end of this exercise, the code to import both the CGM and sleep
data should look like this in the `setup` chunk of the
`docs/learning.qmd` file:

```{r get-participant-id-interface}
# CGM data
cgm_data <- here("data-raw/dime/cgm") |>
  import_csv_files() |>
  get_participant_id()
# Sleep data
sleep_data <- here("data-raw/dime/sleep") |>
  import_csv_files() |>
  get_participant_id()
```

And when you run the `cgm_data` and `sleep_data` in the **Console**, it
should look something like:

```{r show-cgm-sleep-with-id}
#| filename: "Console"
cgm_data
sleep_data
```

So, convert the code we just wrote into a function by following these
steps:

1. Assign `function()` to a new named function called
   `get_participant_id` and include one argument in `function()` with
   the name `data`.
2. Put the code we just wrote into the body of the function. Make sure
   to `return()` the output at the end of the function and to create
   Roxygen documentation. Replace all the relevant variables with the
   `data` argument.
3. Add `stringr` and `dplyr` as a package dependency by using
   `usethis::use_package()` in the **Console**.
4. Explicitly link the functions you are using in this new function to
   their package by using `::` (e.g. `dplyr::` and `stringr::`).
   Remember that you can find which package a function belongs by using
   the `?` help documentation and looking at the very top to see the
   package name.
5. Test that the function works by using `get_participant_id(cgm_data)`
   or `get_participant_id(sleep_data)`.
6. After creating the function and testing that it works, move (**cut**
   and paste) the function into `R/functions.R`.
7. Run `{styler}` in both the `R/functions.R` file and
   `docs/learning.qmd` with {{< var keybind.style >}}.
8. In the `setup` chunk in the `docs/learning.qmd` file, pipe `|>` the
   output of `import_csv_files()` into the `get_participant_id()`
   function for both the CGM and sleep data (as shown above).
9. Render the `docs/learning.qmd` file to make sure things remain
   reproducible with {{< var keybind.render >}}.
10. Restart your R session with {{< var keybind.restart-r >}} to make
    sure later work is in a clean state.
11. Add and commit the changes to the Git history with
    {{< var keybind.commit >}}. Then push to GitHub.

{{< text_snippet sticky_up >}}



## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- While very difficult to learn and use, regular expressions (regex or
  regexp) are incredibly powerful at processing character (also called
  strings) data. Some strings have special meanings like `[]`, `?`, `*`,
  and `+` that can be used to find patterns in string data.
- Use the `{stringr}` package to work with string data as it has many
  functions to help use regex to process string data, like
  `str_extract()` or `str_remove()`.


```{r admin-store-session-code}
#| include: false
save.image(here::here("_temp/characters-dates.RData"))
```

{{< include /includes/_code-appendix.qmd >}}

:::
