# Processing and joining datasets for cleaning {#sec-dplyr-joins}

```{r setup}
#| include: false
library(here)
library(fs)
library(tidyverse)
load(here("_temp/functionals.RData"))
source(here("R/functions.R"))
options(dplyr.summarise.inform = FALSE)
```

Here we will continue using the "*Workflow*" block and start moving over
to the third block, "*Create project data*", in
@fig-overview-create-project-data.

![Section of the overall workflow we will be
covering.](/images/overview-create-project-data.svg){#fig-overview-create-project-data}

## Learning objectives

The learning objectives of this session are:

1.  Learn what regular expressions are and how to use them on character
    data.
2.  Learn about and apply the various ways data can be joined.
3.  Apply functionals when repeatedly joining multiple datasets.
4.  Apply the function `case_when()` when you need nested conditionals
    for cleaning.
5.  Use the `usethis::use_data()` function to save the final, fully
    joined dataset as an `.Rda` file in `data/`.

## Processing character data

::: callout-note
## Reading task: \~5 minutes

Before we go into joining datasets together, we have to do a bit of
processing first. Specifically, we want to get the user ID from the
`file_path_id` character data. Whenever you are processing and cleaning
data, you will very likely encounter and deal with character data. A
wonderful package to use for working with character data is called
`{stringr}`, which we'll use to extract the user ID from the
`file_path_id` column.

The main driver behind the functions in `{stringr}` are [regular
expressions](https://en.wikipedia.org/wiki/Regular_expression) (or regex
for short). These expressions are powerful, very concise ways of finding
patterns in text. Because they are so concise, though, they are also
*very very difficult* to learn, write, and read, even for experienced
users. That's because certain characters like `[` or `?` have special
meanings. For instance, `[aeiou]` is regex for "find one character in a
string that is either a, e, i, o, or u". The `[]` in this case mean
"find the character in between the two brackets". We won't cover regex
too much in this course, some great resources for learning them are the
[R for Data Science regex
section](https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions),
the [stringr regex
page](https://stringr.tidyverse.org/articles/regular-expressions.html),
as well as in the help doc `?regex`.

We've already used them a bit in the `dir_ls()` function with the
`regexp` argument to find our data files. In the case of the regex in
our use of `dir_ls()`, we had wanted to find, for instance, the pattern
`"user_info.csv"` in all the folders and files. But in this case, we
want to extract the user ID pattern, from `user_1` to `user_22`. So how
would we go about extracting this pattern?
:::

## Exercise: Brainstorm a regex that will match for the user ID

> Time: 10 minutes.

In your groups do these tasks. Try not to look ahead, nor in the
solution section :wink:! When the time is up, we'll share some ideas and
go over what the regex will be.

1.  Looking at the `file_path_id` column, list what is similar in the
    user ID between rows and what is different.
2.  Discuss and verbally describe (in English, not regex) what text
    pattern you might use to extract the user ID.
3.  Use the list below to think about how you might convert the English
    description of the text pattern to a regex. This will probably be
    very hard, but try anyway.
    -   When characters are written as is, regex will find those
        characters, e.g. `user` will find only `user`.
    -   Use `[]` to find one possible character of the several between
        the brackets. E.g. `[12]` means 1 *or* 2 or `[ab]` means "a" or
        "b". To find a range of numbers or letters, use `-` in between
        the start and end ranges, e.g. `[1-3]` means 1 to 3 or `[a-c]`
        means "a" to "c".
    -   Use `?` if the character might be there or not. E.g. `ab?` means
        "a" and maybe "b" follows it or `1[1-2]?` means 1 and maybe 1 or
        2 will follow it.

Once you've done these tasks, we'll discuss all together and go over
what the regex would be to extract the user ID.

```{r solution-regex}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
"user_[1-9][0-9]?"
```

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Make sure to reinforce that while regex is incredibly complicated, there
are some basic things you can do with it that are quite powerful.

More or less, this section and exercise are to introduce the idea and
concept of regex, but not to really teach it since that is well beyond
the scope of this course and this time frame.

Go over the solution. Explanation is that the pattern will find anything
that has `user_` followed by a number from 1 to 9 and maybe followed by
another number from 0 to 9.
:::

## Using regular expressions to extract text

Now that we've identified a possible regex to use to extract the user
ID, let's test it out on the `user_info_df` data. Once it works, we will
convert it into a function and move (**cut and paste**) it into the
`R/functions.R` file.

Since we will create a new column for the user ID, we will use the
`mutate()` function from the `{dplyr}` package. We'll use the
`str_extract()` function from the `{stringr}` package to "extract a
string" by using the regex `user_[1-9][0-9]?` that we discussed from the
exercise. Since we're going to use `{stringr}`, so let's add it as a
package dependency by typing this in the **Console**:

``` {.r filename="Console"}
usethis::use_package("stringr")
```

We're also using an argument to `mutate()` you might not have seen
previously, called `.before`. This will insert the new `user_id` column
before the column we use and we do this entirely for visual reasons,
since it is easier to see the newly created column when we run the code.
In your `doc/learning.qmd` file, create a new header called
`## Using regex for user ID` at the bottom of the document, and create a
new code chunk below that.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Walk through writing this code, briefly explain/remind how to use
mutate, and about the `{stringr}` function.
:::

```{r extract-user-id}
#| filename: "doc/learning.qmd"
#| eval: false
user_info_df <- import_multiple_files("user_info.csv", import_user_info)
# Note: your file paths and data may look slightly different.
user_info_df |>
  mutate(
    user_id = str_extract(
      file_path_id,
      "user_[1-9][0-9]?"
    ),
    .before = file_path_id
  )
```

```{r extract-user-id-eval}
#| echo: false
user_info_df |>
  trim_filepath_for_book() |>
  mutate(
    user_id = str_extract(
      file_path_id,
      "user_[1-9][0-9]?"
    ),
    .before = file_path_id
  )
```

Since we don't need the `file_path_id` column anymore, let's drop it
using `select()` and `-`.

```{r drop-file-path-id}
#| filename: "doc/learning.qmd"
user_info_df |>
  mutate(
    user_id = str_extract(
      file_path_id,
      "user_[1-9][0-9]?"
    ),
    .before = file_path_id
  ) |>
  select(-file_path_id)
```

## Exercise: Convert ID extractor code into a function

> Time: 15 minutes.

We now have code that takes the data that has the `file_path_id` column
and extracts the user ID from it. **First step**: While in the
`doc/learning.qmd` file, convert this code into a function, using the
same process you've done previously.

1.  Call the new function `extract_user_id` and add one argument called
    `imported_data`.
    -   Remember to output the code into an object and `return()` it at
        the end of the function.
    -   Include Roxygen documentation.
2.  After writing it and testing that the function works, move (**cut**
    and paste) the function into `R/functions.R`.
3.  Run `{styler}` while in the `R/functions.R` file with
    {{< var keybind.styler >}}.
4.  Replace the code in the `doc/learning.qmd` file with the function
    name so it looks like `extract_user_id(user_info_df)`, restart the R
    session, source everything with `source()` with
    {{< var keybind.source >}}, and run the new function in the code
    chunk inside the `doc/learning.qmd` to test that it works. This
    should automatically run the `setup` code chunk, otherwise, run that
    chunk if it doesn't.
5.  Knit / render the `doc/learning.qmd` file to make sure things remain
    reproducible with {{< var keybind.render >}}.
6.  Add and commit the changes to the Git history with
    {{< var keybind.git >}}.

::: callout-tip
If you don't know what package a function comes from when you need to
append the package when using `::`, you can find out what the package is
by using the help documentation `?functionname` (can also be done by
pressing F1 when the cursor is over the function). The package name is
at the very top left corner, surrounded by `{ }`.
:::

Use this code as a guide to help complete this exercise:

``` {.r filename="doc/learning.qmd"}
extract_user_id <- ___(___) {
    ___ <- ___ |>
        ___mutate(
            user_id = ___str_extract(file_path_id,
                                     "user_[0-9][0-9]?"),
            .before = file_path_id
        ) |>
        ___select(-file_path_id)
    return(___)
}

# This tests that it works:
# extract_user_id(user_info_df)
```

```{r solution-user-id-extract}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
#' Extract user ID from data with file path column.
#'
#' @param imported_data Data with `file_path_id` column.
#'
#' @return A data.frame/tibble.
#'
extract_user_id <- function(imported_data) {
  extracted_id <- imported_data |>
    dplyr::mutate(
      user_id = stringr::str_extract(
        file_path_id,
        "user_[0-9][0-9]?"
      ),
      .before = file_path_id
    ) |>
    dplyr::select(-file_path_id)
  return(extracted_id)
}

# This tests that it works:
extract_user_id(user_info_df)
```

## Modifying existing functions as part of the processing workflow

Now that we've created a new function to extract the user ID from the
file path variable, we need to actually use it within our processing
pipeline. Since we want this function to work on all the datasets that
we will import, we need to add it to the `import_multiple_files()`
function. We'll go to the `import_multiple_files()` function in
`R/functions.R` and use the `|>` to add it after using the
`list_rbind()` function. The code should look something like:

```{r add-extract-user-to-import}
#| filename: "R/functions.R"
import_multiple_files <- function(file_pattern, import_function) {
  data_files <- fs::dir_ls(here::here("data-raw/mmash/"),
    regexp = file_pattern,
    recurse = TRUE
  )

  combined_data <- purrr::map(data_files, import_function) |>
    purrr::list_rbind(names_to = "file_path_id") |>
    extract_user_id() # Add the function here.
  return(combined_data)
}
```

We'll re-source the functions with `source()` using
{{< var keybind.source >}}. Then re-run these pieces of code you wrote
during the exercise in @sec-ex-function-import-all-data to update them
based on the new code in the `import_multiple_files()` function. Add
this to your `doc/learning.qmd` file for now.

```{r}
#| filename: "doc/learning.qmd"
user_info_df <- import_multiple_files("user_info.csv", import_user_info)
saliva_df <- import_multiple_files("saliva.csv", import_saliva)
rr_df <- import_multiple_files("RR.csv", import_rr)
actigraph_df <- import_multiple_files("Actigraph.csv", import_actigraph)
```

As well as adding the `summarised_rr_df` and `summarised_actigraph_df`
to use `user_id` instead of `file_path_id`:

```{r}
#| filename: "doc/learning.qmd"
summarised_rr_df <- rr_df |>
  group_by(user_id, day) |>
  summarise(across(ibi_s, list(
    mean = \(x) mean(x, na.rm = TRUE),
    sd = \(x) sd(x, na.rm = TRUE)
  ))) |>
  ungroup()

summarised_actigraph_df <- actigraph_df |>
  group_by(user_id, day) |>
  # These statistics will probably be different for you
  summarise(across(hr, list(
    mean = \(x) mean(x, na.rm = TRUE),
    sd = \(x) sd(x, na.rm = TRUE)
  ))) |>
  ungroup()
```

Let's render the `doc/learning.qmd` document using
{{< var keybind.render >}}to make sure everything still runs fine. Then,
**add and commit** all the changed files into the Git history with
{{< var keybind.git >}}.

## Join datasets together

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Walk through and describe these images and the different type of joins
after they've read it.
:::

::: callout-note
## Reading task: \~10 minutes

The ability to join datasets together is a fundamental component of data
processing and transformation. In our case, we want to add the datasets
together so we eventually have preferably one main dataset to work with.

There are many ways to join datasets in `{dplyr}` that are described in
`?dplyr::join`. The more common ones that are implemented in the
`{dplyr}` package are:

-   `left_join(x, y)`: Join all rows and columns in `y` that match rows
    and columns in `x`. *Columns* that exist in `y` but not `x` are
    joined to `x`.

![Left joining in `{dplyr}`. Notice how the row with `D` in column `A`
in the blue data is not included in the outputted data on the right.
Modified from the RStudio `{dplyr}`
[cheatsheet](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf).](/images/left-join.png){#fig-left-join
width="90%"}

-   `right_join(x, y)`: The opposite of `left_join()`. Join all rows and
    columns in `x` that match rows and columns in `y`. *Columns* that
    exist in `x` but not `y` are joined to `y`.

![Right joining in `{dplyr}`. Notice how the row with `C` in column `A`
in the green data is not included in the outputted data on the right.
Modified from the RStudio `{dplyr}`
[cheatsheet](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf).](/images/right-join.png){#fig-right-join
width="90%"}

-   `full_join(x, y)`: Join all rows and columns in `y` that match rows
    and columns in `x`. Columns *and* **rows** that exist in `y` but not
    `x` are joined to `x`.

![Full joining in `{dplyr}`. Notice how all rows and columns are
included in the outputted data on the right, and that some missingness
is introduced because those values don't exist when the data are
combined in this way. Modified from the RStudio `{dplyr}`
[cheatsheet](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf).](/images/full-join.png){#fig-full-join
width="90%"}

In our case, we want to use `full_join()`, since we want all the data
from both datasets. This function takes two datasets and lets you
indicate which column to join by using the `by` argument. Here, both
datasets have the column `user_id` so we will join by them.

```{r}
full_join(user_info_df, saliva_df, by = "user_id")
```

`full_join()` is useful if we want to include all values from both
datasets, as long as each participant ("user") had data collected from
that dataset. When the two datasets have rows that don't match, we will
get missingness in that row, but that's ok in this case.

We also eventually have other datasets to join together later on. Since
`full_join()` can only take two datasets at a time, do we then just keep
using `full_join()` until all the other datasets are combined? What if
we get more data later on? Well, that's where more functional
programming comes in. Again, we have a simple goal: For a set of data
frames, join them all together. Here we use another functional
programming concept called `reduce()`. Like `map()`, which "maps" a
function onto a set of items, `reduce()` applies a function to each item
of a vector or list, each time reducing the set of items down until only
one remains: the output. Let's use an example with our simple function
`add_numbers()` we had created before (but later deleted) and add up 1
to 5. Since `add_numbers()` only takes two numbers, we have to give it
two numbers at a time and repeat until we reach 5.

```{r}
# Add from 1 to 5
first <- add_numbers(1, 2)
second <- add_numbers(first, 3)
third <- add_numbers(second, 4)
add_numbers(third, 5)
```

Instead, we can use reduce to do the same thing:

```{r}
reduce(1:5, add_numbers)
```

@fig-reduce visually shows what is happening within `reduce()`.

![A functional that iteratively uses a function on a set of items until
only one output remains. Notice how the output of the first iteration of
the `func()` function is placed in the first position of `func()` in the
next iteration, and so on. Modified from the RStudio purrr
[cheatsheet](https://posit.co/wp-content/uploads/2022/10/purrr.pdf).](/images/reduce.png){#fig-reduce
width="90%"}

If we look at `?reduce`, we see that `reduce()`, like `map()`, takes
either a vector or a list as an input. Since data frames can only be put
together as a list and not as a vector (a data frame has vectors for
columns and so can't be a vector itself), we need to combine the
datasets together in a `list()` and reduce them with `full_join()`.
:::

Let's code this together, using `reduce()`, `full_join()`, and `list()`
while in the `doc/learning.qmd` file.

```{r}
#| filename: "doc/learning.qmd"
list(
  user_info_df,
  saliva_df
) |>
  reduce(full_join)
```

We now have the data in a form that would make sense to join it with the
other datasets. So lets try it:

```{r}
#| filename: "doc/learning.qmd"
list(
  user_info_df,
  saliva_df,
  summarised_rr_df
) |>
  reduce(full_join)
```

Hmm, but wait, we now have four rows of each user, when we should have
only two, one for each day. By looking at each dataset we joined, we can
find that the `saliva_df` doesn't have a `day` column and instead has a
`samples` column. We'll need to add a day column in order to join
properly with the RR dataset. For this, we'll learn about using nested
conditionals.

## Cleaning with nested conditionals

::: callout-note
## Reading task: \~6 minutes

There are many ways to clean up this particular problem, but probably
the easiest, most explicit, and programmatically accurate way of doing
it would be with the function `case_when()`. This function works by
providing it with a series of logical conditions and an associated
output if the condition is true. Each condition is processed
*sequentially*, meaning that if a condition is TRUE, the output won't be
overridden for later conditions. The general form of `case_when()` looks
like:

```{r}
#| eval: false
case_when(
  variable1 == condition1 ~ output,
  variable2 == condition2 ~ output,
  # (Optional) Otherwise
  TRUE ~ final_output
)
```

The optional ending is only necessary if you want a certain output if
none of your conditions are met. Because conditions are processed
*sequentially* and because it is the *last* condition, by setting it as
`TRUE` the final output will used. If this last `TRUE` condition is not
used then by default, the final output would be a missing value. A
(silly) example using age might be:

```{r}
#| eval: false
case_when(
  age > 20 ~ "old",
  age <= 20 ~ "young",
  # For final condition
  TRUE ~ "unborn!"
)
```

If instead you want one of the conditions to be `NA`, you need to set
the appropriate `NA` value:

```{r}
#| eval: false
case_when(
  age > 20 ~ "old",
  age <= 20 ~ NA_character_,
  # For final condition
  TRUE ~ "unborn!"
)
```

Alternatively, if we want missing age values to output `NA` values at
the end (instead of `"unborn!"`), we would exclude the final condition:

```{r}
#| eval: false
case_when(
  age > 20 ~ "old",
  age <= 20 ~ "young"
)
```

With `{dplyr}` functions like `case_when()`, it requires you be explicit
about the type of output each condition has since all the outputs must
match (e.g. all character or all numeric). This prevents you from
accidentally mixing e.g. numeric output with character output. Missing
values also have data types:

-   `NA_character_` (character)
-   `NA_real_` (numeric)
-   `NA_integer_` (integer)

Assuming the final output is `NA`, in a pipeline this would look like
how you normally would use `mutate()`:

```{r}
user_info_df |>
  mutate(age_category = case_when(
    age > 20 ~ "old",
    age <= 20 ~ "young"
  ))
```
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Briefly review the content again, to reinforce what they read.
:::

While still in the `doc/learning.qmd` file, we can use the `case_when()`
function to set `"before sleep"` as day 1 and `"wake up"` as day 2 by
creating a new column called `day`. (We will use `NA_real_` because the
other `day` columns are numeric, not integer.)

```{r}
#| filename: "doc/learning.qmd"
saliva_with_day_df <- saliva_df |>
  mutate(day = case_when(
    samples == "before sleep" ~ 1,
    samples == "wake up" ~ 2
  ))
saliva_with_day_df
```

...Now, let's use the `reduce()` with `full_join()` again:

```{r}
#| filename: "doc/learning.qmd"
list(
  user_info_df,
  saliva_df,
  summarised_rr_df,
  summarised_actigraph_df
) |>
  reduce(full_join)
```

We now have two rows per participant! Let's add and commit the changes
to the Git history with {{< var keybind.git >}}.

## Wrangling data into final form

Now that we've got several datasets processed and joined, it's time to
bring it all together and put it into the `data-raw/mmash.R` script so
we can create a final working dataset.

Open up the `data-raw/mmash.R` file and the top of the file move the
code `library(fs)` to go with the other packages as well. It should look
something like this now:

```{r}
#| filename: "data-raw/mmash.R"
library(here)
library(tidyverse)
library(fs)
source(here("R/functions.R"))
```

Making sure to have the `download.file()` commented out, we'll include
some code to delete the created `data-raw/mmash/` folder so this script
can run cleanly each time. Place the code right below where the
`download.file()` code is.

<!-- TODO: Remove this after course ends, since the same info is in pre-course now -->

```{r data-raw-mmash-comment-out}
#| filename: "data-raw/mmash.R"
#| eval: false
# Download
mmash_link <- "https://physionet.org/static/published-projects/mmash/multilevel-monitoring-of-activity-and-sleep-in-healthy-people-1.0.0.zip"
# download.file(mmash_link, destfile = here("data-raw/mmash-data.zip"))

# Remove previous `mmash/` folder to have clean update
dir_delete(here("data-raw/mmash/"))
```

Go into the `doc/learning.qmd` and cut the code used to create the
`saliva_with_day_df` as well as the code to `full_join()` all the
datasets together with `reduce()` and paste it at the bottom of the
`data-raw/mmash.R` script. Assign the output into a new variable called
`mmash`, like this:

```{r data-raw-mmash-to-final}
#| filename: "data-raw/mmash.R"
saliva_with_day_df <- saliva_df |>
  mutate(day = case_when(
    samples == "before sleep" ~ 1,
    samples == "wake up" ~ 2,
    TRUE ~ NA_real_
  ))

mmash <- list(
  user_info_df,
  saliva_df,
  summarised_rr_df,
  summarised_actigraph_df
) |>
  reduce(full_join)
```

Lastly, we have to save this final dataset into the `data/` folder.
We'll use the function `usethis::use_data()` to create the folder and
save the data as an `.rda` file. We'll add this code to the very bottom
of the script:

```{r save-data}
#| filename: "data-raw/mmash.R"
usethis::use_data(mmash, overwrite = TRUE)
```

We're adding `overwrite = TRUE` so every time we re-run this script, the
dataset will be saved. Alternatively, we could (but won't in this
course) save it as a `.csv` file with:

```{r}
#| eval: false
write_csv(mmash, here("data/mmash.csv"))
```

And later load it in with `read_csv()` (since it is so fast). Alright,
we're finished creating this dataset! Let's generate it by:

-   First running `{styler}` with {{< var keybind.styler >}}.
-   Restarting the R session with {{< var keybind.restart-r >}}.
-   Sourcing the `data-raw/mmash.R` script with
    {{< var keybind.source >}}.

We now have a final dataset to start working on! The main way to load
data is with `load(here::here("data/mmash.rda"))`. Go into the
`doc/learning.qmd` file and delete **everything** again, except for the
YAML header and `setup` code chunk, so that we are ready for the next
session. Lastly, **add and commit** all the changes, including adding
the final `mmash.rda` data file, to the Git history by using
{{< var keybind.git >}}.

## Summary

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Quickly cover this before finishing the session and when starting the
next session.
:::

-   While very difficult to learn and use, regular expressions (regex or
    regexp) are incredibly powerfully at processing character data.
-   Use `left_join()`, `right_join()`, and `full_join()` to join two
    datasets together.
-   Use the functional `reduce()` to iteratively apply a function to a
    set of items in order to end up with one item (e.g. join more than
    two datasets into one final dataset).
-   Use `case_when()` instead of nesting multiple "if else" conditions
    whenever you need to do slightly more complicated conditionals.

```{r store-session-code}
#| include: false
rm(actigraph_df, rr_df)
save.image(here::here("_temp/dplyr-joins.RData"))
```
