# Doing many things at once with functionals {#sec-functionals}

```{r setup}
#| include: false
library(here)
library(tidyverse)
library(fs)
source(here("R/functions.R"))
load(here("_temp/robustness.RData"))
```

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Briefly go over the bigger picture (found in the introduction section)
and remind everyone the 'what' and 'why' of what we are doing.
:::

## Learning objectives

{{< include /includes/objectives/_functionals.qmd >}}

## :speech_balloon: Discussion activity: Recall and share what you've learned so far

**Time: \~6 minutes.**

A very effective way to learn is to recall and describe to someone else
what you've learned. So before we continue this session, take some time
to think about what you've learned from yesterday.

1. Take 1-2 minute and try to recall as much as you can about what you'd
   done over the last two days. Without looking at your notes or the
   other sections of the website, try to remember things about importing
   and robust functions.
2. Then, for 4-5 minutes, share with your neighbour what you remember
   and try to describe it to each other. Maybe you will each remember
   different things.

## :book: Reading task: Functional programming

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Go over this section briefly by reinforcing what they read, especially
reinforcing the concepts shown in the image. Make sure they understand
the concept of applying something to many things at once and that
functionals are better coding patterns to use compared to loops. Doing
the code-along should also help reinforce this concept.

Also highlight that the resources appendix has some links for continued
learning for this and that the Posit `{purrr}` cheat sheet is an amazing
resource to use.
:::

**Time: \~10 minutes.**

Unlike many other programming languages, R's primary strength and
approach to programming is in functional programming. So what is
"functional programming"? At its simplest, functional programming is a
way of thinking about and doing programming that is declarative rather
than imperative. A functional ("declarative") way of thinking is
actually how we humans intuitively think about the world. Unfortunately,
many programming languages are imperative, and even the computer at its
core needs imperative instructions. This is part of the reason why
programming is so hard for humans.

A useful analogy is that declarative/functional programming is similar
to how you talk to an adult about what you want done and letting the
adult do it, e.g. "move these chairs to this room". "Imperative"
programming is how you might talk to a very young child to make sure it
is done exactly as you say, without deviations from those instructions.
Using the chair example, it would be like saying "take this chair, walk
over to that room, drop off the chair in the corner, come back, take
that next chair, walk back to the room and put it beside the other
chair, ..." and so on. There's a lot of room for error because, e.g.
"take" or "walk" or "drop" can mean many different things.

Functional programming, in R, is programming that:

- Uses functions (like `function()`).
- Applies actions as a function to vectors all at once (called
  [vectorisation](https://bookdown.org/rdpeng/rprogdatascience/vectorized-operations.html)),
  rather than one at a time.
  - Vectors are multiple items, like a sequence of numbers from 1 to 5,
    that are bundled together, for instance a column for body weight in
    a dataset is a vector of numbers.
- Can use functions as input to other functions to then output a vector
  (called a [functional](https://adv-r.hadley.nz/functionals.html) or
  "higher-order function").

We've already covered functions. You've definitely already used
vectorization since it is one of R's big strengths. For instance,
functions like `mean()`, `sd()`, or `sum()` are vectorized in that you
give them a vector of numbers and they do something to all the values in
the vector at once. In vectorized functions, you can give the function
an entire vector (e.g. `c(1, 2, 3, 4)`) and R will know what to do with
it. @fig-vectorization shows how a function conceptually uses
vectorization.

![A function using vectorization. Notice how a set of items is included *all at once* in the `func()` function and outputs a single item on the right. Modified from the [Posit purrr cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/purrr.pdf).
Image license is CC-BY-SA
4.0.](/images/vectorization.png){#fig-vectorization width="45%"}

For example, in R, there is a vectorized function called `sum()` that
takes the entire vector of `values` and outputs the total sum.

```{r}
values <- 1:10
# Vectorized
sum(values)
```

In many other programming languages, you would need to use a loop to
calculate the sum because the language doesn't support vectorization. In
R, a loop to calculate the sum would look like this:

```{r}
total_sum <- 0
# a vector
values <- 1:10
for (value in values) {
  total_sum <- value + total_sum
}
total_sum
```

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Emphasize this next paragraph.
:::

Writing effective and correct loops is surprisingly difficult and
tricky. Because of this and because there are better and easier ways of
writing R code to replace loops, we **strongly** recommend not using
loops. If you think you need to, you probably don't. This is why we will
**not** be covering loops in this workshop.

A **functional** is R's native way of doing loops. A functional is a
function where you *give* it a function as one of its arguments.
@fig-functionals shows how the functional `map()` from the `{purrr}`
package works by taking a vector (or list), applying a function to each
of those items, and outputting the results from each function. The name
`map()` doesn't mean a geographic map, it is the mathematical meaning of
map: To use a function on each item in a set of items.

![A functional, in this case `map()`, applies a function to each item in a vector. Notice how each of the green coloured boxes are placed into the `func()` function and outputs the same number of blue boxes as there are green boxes. Modified from the [Posit purrr cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/purrr.pdf).
Image license is CC-BY-SA
4.0.](/images/functionals.png){#fig-functionals width="90%"}

Here's a simple example to show how it works. We'll use `paste()` on
each item of `1:5` to simple output the same number as the input. We
will load in `{purrr}` to be clear which package it comes from, but
`{purrr}` is loaded with the `{tidyverse}` package, so you don't need to
load it if you are using the `{tidyverse}` package.

```{r}
library(purrr)
map(1:5, paste)
```

You'll notice that `map()` outputs a list, with all the `[[1]]` printed.
A list is a specific type of object that can contain different types of
objects. For example, a data frame is actually a special type of list
where each column is a list item that contains the values of that
column. `map()` will always output a list, so it is very predictable.
Notice also that the `paste()` function is given *without* the `()`
brackets. Without the brackets, the function can be used by the `map()`
functional and treated like any other object in R. Remember how we said
that a function is an action when it has `()` at the end. In this case,
we do *not* want R to do the action *yet*, we want R to use the action
on each item in the `map()`. In order for `map()` to do that, you need
to give the function as an object so it can do the action later.

::: callout-tip
A useful analogy here is imagine one person writing instructions on
several pieces of paper along with box of things related to the
instructions. This person then gives each other person in a group one
piece of paper and one thing inside the box. Both the box, *as well as*
the piece of paper are objects, since no action has happened yet. But
once each person gets the box and the piece of paper and reads the
instructions, they will then do an action based on the instructions to
the thing from the box.

That piece of paper is like a function without the `()`, and when the
person reads it and follows the instructions, the function becomes an
action with `()`.
:::

Let's say we wanted to paste together the number with the sentence
"seconds have passed". Normally it would look like:

```{r}
paste(1, "seconds have passed")
paste(2, "seconds have passed")
paste(3, "seconds have passed")
paste(4, "seconds have passed")
paste(5, "seconds have passed")
```

Or as a loop:

```{r}
for (num in 1:5) {
  sec_passed <- paste(num, "seconds have passed")
  print(sec_passed)
}
```

Using `map()` we can give it something called "anonymous" functions.
"Anonymous" because it isn't named with `name <- function()` like we've
done before. Anonymous functions are functions that you use once and
don't remember in your environment (like named functions do). As you
might guess, you can make an anonymous function by simply *not* naming
it! For instance:

```{r}
#| output: false
function(number) {
  paste(number, "seconds have passed")
}
```

There is a shortened version of this using `\(x)`:

```{r}
#| output: false
\(number) paste(number, "seconds have passed")
```

Both forms of anonymous functions are equivalent and can be used in
`map()`:

```{r}
map(1:5, function(number) paste(number, "seconds have passed"))
# Or with the short version
map(1:5, \(number) paste(number, "seconds have passed"))
```

So `map()` will take each number in `1:5`, put it into the anonymous
function in the `number` argument we made, and then run the `paste()`
function on it. The output will be a list of the results of the
`paste()` function.

::: callout-note
`{purrr}` also supports the use of a syntax shortcut to write anonymous
functions. This shortcut is using `~` (tilde) to start the function and
`.x` as the replacement for the vector item. `.x` is used instead of `x`
in order to avoid potential name collisions in functions where `x` is an
function argument (for example in `ggplot2::aes()`, where `x` can be
used to define the x-axis mapping for a graph). Here is the same example
as above, now using the `~` shortcut:

```{r}
map(1:5, ~ paste(.x, "seconds have passed"))
```

The `~` was made before `\()` was made in R. After this new, native R
version was made, the `{purrr}` authors now strongly recommend using it
rather then their `~` for anonymous functions. The `\()` is also a bit
clearer because you give it the name of the argument, e.g. `number`,
rather than be forced to use `.x`.
:::

`map()` will always output a list, but sometimes we might want to output
a different data type. If we look into the help documentation with
`?map`, it shows several other types of map that all start with `map_`:

- `map_chr()` outputs a character vector.
- `map_int()` outputs an integer vector.
- `map_dbl()` outputs a numeric value, called a "double" in programming.

This is the basics of functionals like `map()`. Functions,
vectorization, and functionals provide expressive and powerful
approaches to a simple task:
*Doing an action on each item in a set of items*. And while technically
using a for loop lets you "not repeat yourself", they tend to be more
error prone and harder to write and read compared to these other tools.
For some alternative explanations of this, see
@sec-alternative-loop-explanation.

This approach of doing an action on each item in a set of items also
allows us to use another powerful "design pattern": If you make a
function to do something to one item, you can do it for all items. So
your focus on writing code should try to aim on building a solid, robust
function for one thing, and then use functionals to very easily apply
that to many things at once with nearly no extra effort. For example, do
you have a dozen different models to run on the same data? Focus on
building a solid, robust function that runs one model, then arrange the
input to be a list of the different model formulas, and then use
functionals to run all your models at once. There are many other ways of
applying this thinking to designing and writing your code.

{{< text_snippet sticky_up >}}

## Using functionals to import multiple datasets

But what does functionals have to do with what we are doing now? Well,
we have `read()` that only reads in one CSV file. But we have many more!
Well, we can use functionals to read them all at once :grin:

Before we can use `map()` though, we need to have a vector or list of
the file paths for all the CSV files we want to read in. For that, we
can use the `{fs}` package, which has a function called `dir_ls()` that
finds files in a folder (directory).

So, let's add `library(fs)` to the `setup` code chunk. Then, go to the
bottom of the `docs/learning.qmd` document, create a new header called
`## Using map`, and create a code chunk below that with
{{< var keybind.chunk >}}

The `dir_ls()` function, which stands for "directory list", takes the
path that we want to search, in this case `data-raw/nurses-stress/`, an
argument to search all sub-folders with `recurse = TRUE`, and has an
argument called `regexp` to tell it what type of files to search for.
This `regexp` argument is short for "regular expression", which we will
cover a bit more in a later session. Regular expressions are very
powerful pattern finding tools, but they can also get very complicated.
Thankfully in our case, we only want to find all CSV files in the
`data-raw/nurses-stress/` folder that are called `HR.csv.gz`. So we can
use that in `regexp`. Let's pipe the output of the `here()` path to the
stress folder into `dir_ls()` by assigning it to a new variable
`hr_file` so we can use it later.

```{r list-hr-files}
#| filename: "docs/learning.qmd"
hr_files <- here("data-raw/nurses-stress/") |>
  dir_ls(regexp = "HR.csv.gz", recurse = TRUE)
```

Let's see what the output looks like. The paths you see on your computer
will be different from what are shown here (it is a different computer).
Plus to keep it simpler, on the website we will only show the first 3
files.

```{r list-hr-files-show}
#| filename: "docs/learning.qmd"
#| eval: false
hr_files
```

```{r admin-list-files-for-book}
#| echo: false
head(gsub(".*\\/data-raw", "data-raw", hr_files), 3)
```

So we have a list of all the CSV files we want to read in. Alright, we
now have all the files ready to give to `map()`, so let's read in
**all** HR files into R!

```{r map-read-hr}
#| filename: "docs/learning.qmd"
hr_data <- hr_files |>
  map(read)
```

Notice that we give `map()` the function `read` *without* the `()`.
That's because we need to give the functional `map()` the function as
*an object*, not as an *action*. After you run it, `map()` will take the
`read` function as an object and apply it to each HR file as an action.

Remember, that `map()` always outputs a list, so when we look into this
object, it will give us `r length(hr_data)` tibbles (data frames). For
the website we'll only show the first two tibbles:

```{r map-import-hr-show}
#| filename: "docs/learning.qmd"
hr_data[1:2]
```

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Remind everyone that we still only import the first 100 rows of each
data file. So if some of the data itself seems looks odd or so little
data, that is the reason why. Remind them that we do this to more
quickly prototype and test code out.
:::

This is great because with one line of code we imported all these
datasets and made them into one data frame! But we're missing an
important bit of information: The participant ID. The `{purrr}` package
has many other powerful functions to make it easier to work with
functionals. We know `map()` always outputs a list. But what we want is
a single tibble at the end that also contains the participant ID.

There are two functions that take a list of data frames and convert them
into a single data frame. They are called `list_rbind()` to bind
("stack") the data frames by rows or `list_cbind()` to bind ("stack")
the data frames by columns. In our case, we want to bind (stack) by
rows, so we will use `list_rbind()` by piping the output of the `map()`
code we wrote into `list_rbind()`.

```{r use-list-rbind}
#| filename: "docs/learning.qmd"
hr_data <- hr_files |>
  map(read) |>
  list_rbind()
hr_data
```

But, hmm, we don't have the participant ID in the data frame. This is
because `list_rbind()` doesn't know how to get that information, since
it isn't included in the data frames. If we look at the help for
`list_rbind()`, we will see that it has an argument called `names_to`.
This argument lets us create a new column that is based on the name of
the list item, which in our case is the file path. This file path also
has the participant ID information in it, but it also has the full file
path in it too, which isn't exactly what we want. So we'll have to fix
it later. But first, let's start with adding this information to the
data frame as a new column called `file_path_id`. So we will add the
`names_to` argument to the code we've already written.

```{r use-list-rbind-with-paths}
#| filename: "docs/learning.qmd"
hr_data <- hr_files |>
  map(read) |>
  list_rbind(names_to = "file_path_id")
```

And then look at it by running `hr_data` below the code:

```{r use-list-rbind-with-paths-show}
#| filename: "docs/learning.qmd"
#| eval: false
hr_data
```

```{r admin-show-rbind-with-path-for-book}
#| echo: false
hr_data |>
  trim_filepath_for_book()
```

The `file_path_id` variable will look different on everyone's computer.
Don't worry, we're going to tidy up the `file_path_id` variable later in
another session. Looking at the code and thinking about our "design
pattern" of "do something to one thing, then do it for all things", we
can see that if we wanted to read in the `IBI` or `BVP` or other files
found inside `stress/`, we would only need to make one change to the
code above, which is to change the `regexp` argument in `dir_ls()` to
find the files we want. So, it's your turn to try to convert it to a
function and read in the other datasets.

Before moving on to the exercise, let's style with
{{< var keybind.style >}} the document, render with
{{< var keybind.render >}}, and finally **add and commit** the changes
to the Git history using {{< var keybind.commit >}} before pushing to
GitHub.

## :technologist: Exercise: Convert the map with reading code into a function

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Because later sessions depend on this code, after they've finished the
exercise, walk through with them the solution. So that we are all on the
same page.
:::

**Time: \~15 minutes.**

::: callout-tip
As you add more code and text to your `docs/learning.qmd` file, it will
start to get longer. A helpful tip to move around a Quarto or R script
more easily is to open up the "Document Outline" on the side by clicking
the button in the top right corner of the Quarto pane or by using
{{< var keybind.outline >}}.
:::

We've made some code that does what we want, now the next step is to
convert it into a function so we can use it on the other datasets within
`stress/`. The code we've written only reads the `HR` data and it so far
looks like this:

```{r}
#| eval: false
hr_files <- here("data-raw/nurses-stress/") |>
  dir_ls(regexp = "HR.csv.gz", recurse = TRUE)

hr_data <- hr_files |>
  map(read) |>
  list_rbind(names_to = "file_path_id")

hr_data
```

We want to be able to do the same thing for the other datasets like
`BVP` or `IBI`, but we don't want to repeat ourselves. What we want is a
new function that looks like the code chunk below that we can use to
import the other datasets.

```{r}
#| eval: false
hr_data <- read_all("HR.csv.gz")
ibi_data <- read_all("IBI.csv.gz")
bvp_data <- read_all("BVP.csv.gz")
```

Recall that the structure of making functions below and use this template to help you get started:

``` r
___ <- function(___) {
  # Code that does something
  ___ <- ___

  ___ <- ___

  return(___)
}
```

Complete the tasks below to convert the code into a stable, robust
function that you can re-use.

1. We've used two new packages in this session, `{fs}` and `{purrr}`.
   First add these packages to the `DESCRIPTION` file by using
   `usethis::use_package()` in the **Console**.
2. In the bottom of `docs/learning.qmd`, create a new header
   `## Exercise: Convert map to function and use on sleep`.
3. Below the header, create a new code chunk using
   {{< var keybind.chunk >}}.
4. As you've already done in the previous sessions, convert this code
   into a function. You can always refer to the @sec-functions and
   @sec-robustness sessions.
   - Name the function `read_all`.
   - Use the `function() { ... }` code to create a new function
   - Inside `function()`, add an argument called `filename`.
   - Paste the code into the function body.
   - Rename `hr_files` to simply `files`.
   - Rename `hr_data` to simply `data`.
   - Include a `return()` at the end.
   - Add `package_name::` to the individual functions used. To find
     which package a function comes from, use `?functionname` to see the
     help documentation (there is four functions to add `::` to).
   - Create and write Roxygen documentation to describe the new function
     by using {{< var keybind.roxygen >}}.
   - Run it and check that it works using the code shown above e.g.
     `read_all("HR.csv.gz")`.
5. Render the document with {{< var keybind.render >}} to check that the
   function works.
6. **Cut** and paste the function into the `R/functions.R` file. Then go
   back to the `docs/learning.qmd` file and (making sure the old
   function is deleted or the code chunk with it has `eval: false`),
   render again with {{< var keybind.render >}} to check that the
   function works after moving it.
7. Run `{styler}` while in the `R/functions.R` file with
   {{< var keybind.style >}}.
8. Once done, **add** the changes you've made and **commit** them to the
   Git history, using {{< var keybind.commit >}} and then push to
   GitHub.

```{r}
#| label: solution-read-multiple-files
#| echo: false
#' Read all `.csv.gz` files in the `stress/` folder into one data frame.
#'
#' @param filename The name of files in the sub-folders that we
#'    want to read in.
#'
#' @returns A single data frame/tibble.
#'
read_all <- function(filename) {
  files <- here::here("data-raw/stress/") |>
    fs::dir_ls(regexp = filename, recurse = TRUE)

  data <- files |>
    purrr::map(read) |>
    purrr::list_rbind(names_to = "file_path_id")

  return(data)
}

read_all("HR.csv.gz")
read_all("IBI.csv.gz")
```

{{< text_snippet sticky_up >}}

## :technologist: Extra exercise: Include a new `max_rows` argument to `read_all()`

**Time: ~10 minutes.**

If you finished the previous exercise early, you can try this exercise,
which is much harder. To make it a bit harder, we also won't provide a
scaffold.

Right now, the `read_all()` can only read in the first 100 rows of each
file, because we've put the default value of `max_rows` in `read()` to
100. If we wanted to read in all the data, we would have to change that
default value, which isn't ideal nor robust of a feature for
`read_all()`. Instead, it would be nice if we could set the `max_rows`
argument in `read_all()`, like:

```r
read_all("HR.csv.gz", max_rows = 1000)
read_all("HR.csv.gz", max_rows = 10)
read_all("HR.csv.gz", max_rows = Inf)
```

In the `R/functions.R` file, complete these tasks of updating the
`read_all()` function while using the workflow of editing,
`source()`'ing the `R/functions.R` file, and testing the function in the
**Console**. Complete these tasks:

1. Add a new argument to the `read_all()` function in the
   `function(...)` called `max_rows` with a default value of 100.
2. Review the section you read at the start of this session on
   functionals and "anonymous functions".
3. Use an anonymous function in the `map()` code to give the `max_rows`
   argument to the `read()` function's `max_rows` argument. You will
   need to use the `\(x)` syntax for this. Instead of `x`, call it
   `file` to be more descriptive. In the anonymous function, you will
   need to use both `file` and `max_rows` in `read()`.
4. Test that `read_all("HR.csv.gz", max_rows = 10)` works (in the
   **Console**) and only reads in 10 rows of each file, by
   `source()`'ing the `R/functions.R` file with
   {{< var keybind.source >}} whenever you make a change to the
   `read_all()` function.
5. Once it works, update the Roxygen docs by adding `@params max_rows`
   to the `read_all()` documentation.
6. Render the `docs/learning.qmd` file to check that it works in the
   document with {{< var keybind.render >}}.
7. Style the `R/functions.R` file with {{< var keybind.style >}}.
8. Lastly, commit your changes to Git with {{< var keybind.commit >}}
   and push to GitHub.

```{r}
#| label: solution-read-all-max-rows
#| echo: false
#' Read all `.csv.gz` files in the `stress/` folder into one data frame.
#'
#' @param filename The name of files in the sub-folders that we
#'    want to read in.
#' @param max_rows The maximum number of rows to read in for each file.
#'
#' @returns A single data frame/tibble.
#'
read_all <- function(filename, max_rows = 100) {
  files <- here::here("data-raw/stress/") |>
    fs::dir_ls(regexp = filename, recurse = TRUE)

  data <- files |>
    purrr::map(\(file) read(file, max_rows = max_rows)) |>
    purrr::list_rbind(names_to = "file_path_id")

  return(data)
}

read_all("HR.csv.gz", max_rows = 10)
read_all("HR.csv.gz", max_rows = 1000)
```

{{< text_snippet sticky_up >}}

## :technologist: Extra exercise: Move the `here::here()` path out of the function

**Time: ~5 minutes.**

If you finished the previous exercise early, you can try this exercise,
which is much harder. Like with the extra exercise above, to make it a
bit harder, we won't provide a scaffold.

In general, it isn't good practice to hard code things in functions, as
it makes them less robust to changes. For instance, if we ever renamed
the `data-raw/` folder or the `nurses-stress/` folder, the `read_all()`
function (and any other function that might use `here::here()` to those
folders) would break and stop working. So instead, it is better to move
those types of things out of the function and into a common location in
the "global" environment (meaning outside of a function). That way, if
you ever need to change the path, you only need to change the value of
that one variable. So in this exercise, you will move that
`here::here()` path out of the function and into a variable in the
global environment, and then use that variable in the function. All
while the function continues to work as usual.

While this approach is quite powerful when there are some variables that
many functions use, it can also cause some problems as you might
accidentally create a variable with the same name as the one you used in
the function. We can avoid this by using a specific naming style for
"global" variables by capitalising them and to hide them by using a `.`
at the start of the variable name. So try it out by following these
steps:

1. Open the `R/functions.R` file and at the top of the file, create a
   new "section" with a header called `# Global variables -----`. The
   `-----` is a common style to make it easier to visually see the
   section in the code and to make it appear in the "Document Outline"
   in RStudio.
2. Below that header, create a new variable called `.DATASET_DIR` and
   assign it the value of `here::here("data-raw/stress/")`. The `.` at
   the start of the variable name makes it hidden in the environment and
   harder to accidentally change or use again. The capitalisation makes
   it clear that it is a "global" variable that is used inside functions
   and isn't just a variable for one function.
3. Go into the `read_all()` function body and replace the
   `here::here("data-raw/stress/")` with the new variable
   `.DATASET_DIR`.
4. Test that `read_all()` works (in the **Console**) by `source()`'ing
   the `R/functions.R` file with {{< var keybind.source >}} and running
   `read_all("HR.csv.gz")` whenever you make a change to the
   `read_all()` function.
5. When it works, style the `R/functions.R` file with
   {{< var keybind.style >}}.
6. Lastly, commit your changes to Git with {{< var keybind.commit >}}
   and push to GitHub.

::: {.callout-note collapse="true"}
```{r}
#| label: solution-read-multiple-files
#| echo: false
.DATASET_DIR <- here::here("data-raw/stress/") # <1>

# In the function body of `read_all()`.
read_all <- function(filename, max_rows = 100) {
  files <- .DATASET_DIR |> # <2>
    fs::dir_ls(regexp = filename, recurse = TRUE)

  data <- files |>
    purrr::map(\(file) read(file, max_rows = max_rows)) |>
    purrr::list_rbind(names_to = "file_path_id")

  return(data)
}

read_all("HR.csv.gz", max_rows = 10)
read_all("HR.csv.gz", max_rows = 1000)
```

1. Place it at the top of `R/functions.R` so that it can be used by
   other functions below it. Use a `.` at the start to make it hidden so
   it isn't visible in the environment and so is harder to accidentally
   change or use again. Capitalise it to make it clear that it is a
   "global" variable that is used inside functions and isn't just a
   variable for one function.
2. Replace the `here::here()` in the function body with the new variable
   `.DATASET_DIR`. This way, if we ever need to change the path to the
   data, we only need to change it in one place, which is more robust
   and easier to maintain.

:::

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- R is a functional programming language:
  - It has functions that take an input, do an action, and give an
    output.
  - It uses vectorisation that apply a function to multiple items (in a
    vector) all at once rather than using loops.
  - It uses functionals that allow functions to use other functions as
    input.
- Thinking in terms of declaratively telling a computer to do actions,
  rather than telling a computer how to do things step-by-step, is a
  powerful way to think about programming, especially in R.
- Design how you write your code by making functions to do an action to
  one type of object (like a data frame), putting similar types of
  objects (like a data frame or vector) into a list, and then using
  functionals like `map()` to do the action to all of them at once, to
  be effective and efficient with doing lots of things with very little
  code.
- Use the `{purrr}` package and its function `map()` when you want to
  repeat a function on multiple items at once.
- Use `list_rbind()` to combine multiple data frames into one data frame
  by stacking them on top of each other.

## :speech_balloon: Discussion activity: Incorporating functionals in your own work

**Time: \~6 minutes.**

As we prepare for the next session and taking the break, get up and
discuss with your neighbour or group the following questions:

- What are some things you do repetitively that you think might benefit
  from using functionals on?
- Consider your code you've written for other projects or tasks you have
  in other projects. How might you use functional programming to help
  your work? How might you change existing code to use them? Might it be
  a lot of work?

```{r admin-store-session-code}
#| include: false
save.image(here::here("_temp/functionals.RData"))
```

{{< include /includes/_code-appendix.qmd >}}
