---
execute:
  eval: true
---

# Bundling code into functions {#sec-functions}

```{r setup}
#| include: false
library(fs)
library(tidyverse)
library(snakecase)
library(here)
```

## Learning objectives

{{< include /includes/objectives/_functions.qmd >}}

## :book: Reading task: The basics of a function

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Repeat and reinforce the part of what functions are made of, their
structure, that all actions are functions, and all functions are objects
(but not that all objects are functions).
:::

**Time: \~5 minutes.**

The first thing to know about R is that everything is an "object" and
that some objects can do an action. These objects that do an action are
called functions. You've heard or read about functions before during the
workshop, but what is a function?

A function is a bundled sequence of steps that achieve a specific action
and you can usually tell if an object is an action if it has a `()` at
the end of it's name. For example, `mean()` is function to calculate the
mean or `sd()` is a function to calculate the standard deviation. It
isn't always true that functions end in `()` though, which you'll read
about shortly. For instance, the `+` is a function that adds two numbers
together, the `[]` is a function that is used to subset or extract an
item from a list of items like getting a column from a data frame, or
`<-` is a function that creates a new object from some value.

All created functions have the same structure: they are assigned a
function name with `<-`, it uses `function()` to give it its parameters
or arguments, and it has the internal sequence of steps called the
function body that are wrapped in `{}`:

``` {.r filename="Console"}
function_name <- function(argument1, argument2) {
  # body of function with R code
}
```

Notice that this uses two functions to create this `function_name`
function object:

-   `<-` is the action (function) that will create the new object
    `function_name`.
-   `function()` is the action (function) to tell R that this object is
    an action (function) whenever it is used with a `()` at the end,
    e.g. `function_name()`.

Because R is open source, anyone can see how things work underneath. So,
if you want to see what a function does underneath, you would type out
the function name without the `()` into the **Console** and run it. If
we do it with the function `sd()` which calculates the standard
deviation, we see:

```{r show-sd-internals}
#| filename: "Console"
sd
```

Here you see `sd()` as the arguments `x` and `na.rm`. In the function
body it shows how it calculates the standard deviation, which is the
square root of the variance. In this code, the `var()` is inside the
`sqrt()` function, which is exactly what it should be if you know the
math (you don't need to).

Normally you can tell if something is a function if it has `()` at the
end of the name. But there are special functions, like `+`, `[]`, or
even `<-`, that do an action but that don't have `()` at the end. These
are called *operator* functions. Which is why the `<-` is called the
assignment operator, because it assigns something to a new object. To
see how they work internally, you would wrap \` around the operator. So
for `+` or `<-` it would be:

```{r show-operators-internals}
#| filename: "Console"
`+`
`<-`
```

You'll see something called `.Primitive`. Often operators are written in
very low level computer code to work, which are called "primitives".
This is *way* beyond the scope of this workshop to explain primitives
and so we won't go into what that means and why.

To show that they are a function, you can even use them with their `()`
version like this:

```{r show-how-operators-work}
1 + 2
`+`(1, 2)

`<-`(x, 1)
x
x <- 1
x
```

But hopefully you can see that using it with the `()` isn't very nice to
read or use!

If you can learn to make your own functions, it can help make your life
and work much easier and more efficient! That's because you can make a
sequence of actions that you can then reuse again and again. And
luckily, you will be making many functions throughout this workshop.
Making a function always follows a basic structure:

1.  Give a name to the function (e.g. `mean`).
2.  Use `function()` to tell R the new object will be a function and
    assigning it to the name with `<-`.
3.  Optionally provide arguments to the function object, for example
    `function(argument1, argument2, argument3)`.
4.  Fill out the body of the function, with the arguments (if any)
    contained inside, that does some sequence of actions.
5.  Optionally, use `return()` to indicate what final output you want
    the function to have. For learning purposes, we'll always use
    `return()` to help show us what is the final function output but it
    isn't necessary.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Emphasize that we will be using this workflow for creating functions all
the time throughout workshop and that this workflow is also what you'd
use in your daily work.
:::

While there is no minimum or maximum number of arguments you can provide
for a function (e.g. you could have zero or dozens of arguments), its
generally good practice and design to have as few arguments as necessary
to get the job done. Part of making functions is to reduce your own and
others cognitive load when working with or reading code. The fewer
arguments you use, the lower the cognitive load. So, the structure is:

``` r
name <- function(argument1, argument2) {
    # body of function
    output <- ... code ....
    return(output)
}
```

Writing your own functions can be absolutely amazing and fun and
powerful, but you also often want to pull your hair out with frustration
at errors that are difficult to understand and fix. One of the best ways
to deal with this is by making functions that are small and simple, and
testing them as you use them. The smaller they are, the less chance you
will have that there will be an error or issue that you can't figure
out. There's also some formal debugging steps you can do but due to time
and to the challenge of making meaningful debugging exercises since
solutions to problems are very dependent on the project and context,
there is some extra material in @sec-extra-material that you can look
over in your own time. It contains some instructions on debugging and
dealing with some common problems you might encounter with R.

{{< text_snippet sticky_up >}}

## Creating our first function

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Take your time slowly going over this, especially taking about the
Roxygen documentation template.
:::

Let's create a really basic function to show the process. First, create
a new Markdown header called `## Making a function to add` and create a
code chunk below that with {{< var keybind.chunk >}} . Then, inside the
code chunk, we'll write this code:

```{r create-add-function}
#| filename: "docs/learning.qmd"
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

You can use the new function by running the above code and writing out
your new function, with arguments to give it.

```{r call-function}
#| filename: "docs/learning.qmd"
add_numbers(1, 2)
```

The function name is fairly good; `add_numbers` is read as "add
numbers". While we generally want to write code that describes what it
does by reading it, it's also good practice to add some formal
documentation to the function. Use the "Insert Roxygen Skeleton" in the
"Code" menu, by typing {{< var keybind.roxygen >}}, and you can add
template documentation right above the function. Make sure your cursor
is within the function in order for the Roxygen template to be added to
your function. It looks like:

```{r roxygen-docs-1}
#| filename: "docs/learning.qmd"
#' Title
#'
#' @param num1
#' @param num2
#'
#' @return
#' @export
#'
#' @examples
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or
several words that describe the function. Creating a new paragraph below
this line allows you to add a more detailed description. The other items
are:

-   `@param num`: These lines describe what each argument (also called
    parameter) is for and what to give it.
-   `@return`: This describes what output the function gives. Is it a
    data.frame? A plot? What else does the output give?
-   `@export`: Tells R that this function should be accessible to the
    user of your package. Since we aren't making packages, delete it.
-   `@examples`: Any lines below this are used to show examples of how
    to use the function. This is very useful when making packages, but
    not really in this case. So we'll delete it. Let's write out some
    documentation for this function:

```{r roxygen-docs-2}
#| filename: "docs/learning.qmd"
#' Add two numbers together.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#'
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

Once we've created that and before moving on, let's style our code with
{{< var keybind.styler >}}, render the Quarto document with
{{< var keybind.render >}}, and then open up the Git Interface with
{{< var keybind.git >}} to **add and commit** these changes to the Git
history.

## Making a function for importing our data

Now that we have a basic understanding of what a function looks like,
let's apply that what we're doing right now: Importing our data.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Emphasize that we will be using this workflow for creating functions all
the time throughout workshop and is also a common workflow when making
functions in R.
:::

Making a function is a series of steps:

1.  Write code that works and does what you want.
2.  Enclose it as a function with `name <- function() { ... }`, with an
    appropriate and descriptive name.
3.  Create arguments in the function call
    (`function(argument1, argument2)`) with appropriate and descriptive
    names, then replace the code in the function body with the argument
    names where appropriate.
4.  Rename any objects created to be more generic and include the
    `return()` function at the end to indicate what the function will
    output.
5.  Run the function and check that it works.
6.  Add the Roxygen documentation tags with {{< var keybind.roxygen >}}
    while the cursor is in the function.

In `docs/learning.qmd`, create a new Markdown header called
`## Import 101's cgm data with a function` and create a code chunk below
that with {{< var keybind.chunk >}} .

So, step one. Let's copy and paste the code we previously for importing
the `cgm_101` data and convert that as a function:

```{r function-creation-step-1}
#| filename: "docs/learning.qmd"
#| output: false
cgm_101 <- here("data-raw/dime/cgm/101.csv") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100,
  )
```

Next we wrap it in the function call and give it an appropriate name. In
this case, `import_cgm` is descriptive and meaningful.

```{r function-creation-step-2}
#| filename: "docs/learning.qmd"
#| eval: false
import_cgm <- function() {
  cgm_101 <- here("data-raw/dime/cgm/101.csv") |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100,
    )
}
```

Then, we add arguments in the function and replace within the code.
Here, we have only one thing that we would change: The file path to the
dataset. So, a good name might be `file_path`. It's also good practice
to not hard code the use of `here()` within a function. Instead, it's
good design to give functions a full file path that it can use
internally. Then when we use the function, we would use `here()` with
the correct path in the function argument. So it would be:

```{r function-creation-step-3}
#| filename: "docs/learning.qmd"
#| eval: false
import_cgm <- function(file_path) {
  cgm_101 <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100,
    )
}
```

Then we simplify things internally by renaming `cgm_101` to simply
`cgm`, since we would like to be able to import other participant CGM
data later. Finally, we will add the `return()` function at the end with
the object we want the function to output.

```{r function-creation-step-4}
#| filename: "docs/learning.qmd"
import_cgm <- function(file_path) {
  cgm <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100,
    )
  return(cgm)
}
```

Great! Now we need to test it out. Let's try on two `cgm` datasets for
101's and 102's files:

```{r function-creation-step-5}
#| filename: "docs/learning.qmd"
import_cgm(here("data-raw/dime/cgm/101.csv"))
import_cgm(here("data-raw/dime/cgm/102.csv"))
```

Awesome! It works :tada: The final stage is to add the Roxygen
documentation.

```{r function-creation-step-6}
#| filename: "docs/learning.qmd"
#' Import one participants CGM data from the DIME dataset.
#'
#' @param file_path Path to the CGM data file.
#'
#' @return Outputs a data frame/tibble.
#'
import_cgm <- function(file_path) {
  cgm <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100,
    )
  return(cgm)
}
```

A massive advantage of using functions is that if you want to make a
change to what your code does, like if you fix a mistake, you can very
easily do it by modifying the function and it will change all your other
code too.

Now that we have a working function, run `{styler}` with
{{< var keybind.styler >}}, render with {{< var keybind.render >}}, and
then add and commit the changes to the Git history with
{{< var keybind.git >}}.

## :technologist: Exercise: Convert the sleep code into a function

**Time: \~10 minutes.**

We've converted the code to import the CGM data into a function. Now,
let's do the same for the sleep data. Use the code you made from the
exercise in @sec-exercise-import-sleep and convert that into a function
using the same steps we did from above.

::: {.callout-tip appearance="default"}
A helpful tip: To move around a Quarto or R script more easily, open up
the "Document Outline" on the side by clicking the button in the top
right corner of the Quarto pane or by using {{< var keybind.outline >}}.
:::

1.  Create a new Markdown header at the bottom of `docs/learning.qmd`
    called `## Exercise to make function to import sleep data`.
2.  Below the Markdown header, create a new code chunk with
    {{< var keybind.chunk >}}.
3.  Paste the code you used from the exercise in
    @sec-exercise-import-sleep into the code chunk and begin converting
    it into a function, like we did above.
    -   Wrap it with the `function() {...}`.
    -   Make a meaningful name (use `import_sleep`).
    -   Make an argument for the file path (`file_path`) and replace the
        `here("")` code with `file_path`.
    -   Rename the output object from `sleep_101` to simply `sleep`.
    -   End the function with using `return()` to output the `sleep`
        object.
    -   Use the function and run it in the code chunk to test that it
        works on both `101` and `102`'s sleep data.
    -   Add the Roxygen documentation with {{< var keybind.roxygen >}}
        and then fill in the details, like we did above.
4.  Run `{styler}` while in the `docs/learning.qmd` file with
    {{< var keybind.styler >}}.
5.  Render the Quarto document with {{< var keybind.render >}}.
6.  Finally, **add and commit** the changes to the Git history, using
    {{< var keybind.git >}}.

The code that you write to test that the function works should look and
what it should output should be similar to the below code chunk:

```{r solution-repeat-saliva}
#| echo: false
#' Import a participant's sleep data from DIME.
#'
#' @param file_path Path to the participant's sleep file.
#'
#' @return Outputs a data frame/tibble.
#'
import_sleep <- function(file_path) {
  sleep <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100,
    )
  return(sleep)
}
```

```{r solution-sleep-function-output}
import_sleep(here("data-raw/dime/sleep/101.csv"))
import_sleep(here("data-raw/dime/sleep/102.csv"))
```

<!-- TODO: The function might be the same as cgm... might revise once all sessions are done -->

{{< text_snippet sticky_up >}}

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Mention that the `import_sleep()` function is identical to the
`import_cgm()`. Briefly say that in the next session we will go over
making more general functions.
:::

## Key takeaways

-   Everything in R is an object.
-   Every action in R is a function and every function is an object.
-   Functions contain a sequence of steps that do actions to an object.
-   Functions have five components:
    -   The three required ones are using `function() { }`, the code in
        the function body between the `{ }`, and an output (usually set
        with `return()`).
    -   The two optional ones are assigning the function as a new object
        with `<-` and the function arguments put within `function()`.
-   Document functions by using Roxygen.
-   Keep functions small and simple, so it is easier to test and fix
    them.
-   Use few arguments in functions to reduce cognitive load.

```{r store-session-code}
#| include: false
save.image(here::here("_temp/functions.RData"))
```

## :speech_balloon: Discussion activity: What are some tasks that could be functions?

**Time: \~6 minutes.**

As we prepare for the next session, get up, walk around, and discuss
with your neighbour some of the following questions:

-   What are some tasks you do that are repetitive or that you do
    multiple times with very small changes each time you do the task?
-   How might you use functions in your work? Can you think of specific
    tasks or situations where you could use one?
