# Doing many things at once with functionals {#sec-functionals}

```{r setup}
#| include: false
library(here)
library(tidyverse)
library(fs)
source(here("R/functions.R"))
load(here("_temp/robustness.RData"))
```

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Briefly go over the bigger picture (found in the introduction section)
and remind everyone the 'what' and 'why' of what we are doing.
:::

## Learning objectives

{{< include /includes/objectives/_functionals.qmd >}}

## :speech_balloon: Discussion activity: Recall and share what you've learned so far

**Time: \~6 minutes.**

A very effective way to learn is to recall and describe to someone else
what you've learned. So before we continue this session, take some time
to think about what you've learned from yesterday.

1.  Take 1-2 minute and try to recall as much as you can about what
    you'd done over the last two days. Without looking at your notes or
    the other sections of the website, try to remember things about
    importing and robust functions.
2.  Then, for 4-5 minutes, share with your neighbour what you remember
    and try to describe it to each other. Maybe you will each remember
    different things.

## :book: Reading task: Functional programming

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Go over this section briefly by reinforcing what they read, especially
reinforcing the concepts shown in the image. Make sure they understand
the concept of applying something to many things at once and that
functionals are better coding patterns to use compared to loops. Doing
the code-along should also help reinforce this concept.

Also highlight that the resources appendix has some links for continued
learning for this and that the Posit `{purrr}` cheat sheet is an amazing
resource to use.
:::

**Time: \~10 minutes.**

Unlike many other programming languages, R's primary strength and
approach to programming is in functional programming. So what is
"functional programming"? At its simplest, functional programming is a
way of thinking about and doing programming that is declarative rather
than imperative. A functional ("declarative") way of thinking is
actually how we humans intuitively think about the world. Unfortunately,
many programming languages are imperative, and even the computer at its
core needs imperative instructions. This is part of the reason why
programming is so hard for humans.

A useful analogy is that functional programming is like how you talk to
an adult about what you want done and letting the adult do it, e.g.
"move these chairs to this room". "Imperative" is how you'd talk to a
very young child to make sure it is done exactly as you say, without
deviations from those instructions. Using the chair example, it would
like saying "take this chair, walk over to that room, drop of the chair
in the corner, come back, take that next chair, walk back to the room
and put it beside the other chair, ..." and so on. There's a lot of room
for error because, e.g. "take" or "walk" or "drop" can mean many
different things.

Functional programming, in R, is programming that:

-   Uses functions (like `function()`).
-   Applies actions as a function to vectors all at once (called
    [vectorisation](https://bookdown.org/rdpeng/rprogdatascience/vectorized-operations.html)),
    rather than one at a time.
    -   Vectors are multiple items, like a sequence of numbers from 1 to
        5, that are bundled together, for instance a column for body
        weight in a dataset is a vector of numbers.
-   Can use functions as input to other functions to then output a
    vector (called a
    [functional](https://adv-r.hadley.nz/functionals.html) or
    "higher-order function").

We've already covered functions. You've definitely already used
vectorization since it is one of R's big strengths. For instance,
functions like `mean()`, `sd()`, or `sum()` are vectorized in that you
give them a vector of numbers and they do something to all the values in
the vector at once. In vectorized functions, you can give the function
an entire vector (e.g. `c(1, 2, 3, 4)`) and R will know what to do with
it. @fig-vectorization shows how a function conceptually uses
vectorization.

![A function using vectorization. Notice how a set of items is included
*all at once* in the `func()` function and outputs a single item on the
right. Modified from the [Posit purrr cheat
sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/purrr.pdf).
Image license is CC-BY-SA
4.0.](/images/vectorization.png){#fig-vectorization
width="45%"}

For example, in R, there is a vectorized function called `sum()` that
takes the entire vector of `values` and outputs the total sum.

```{r example-vec-sum}
values <- 1:10
# Vectorized
sum(values)
```

In many other programming languages, you would need to use a loop to
calculate the sum because the language doesn't support vectorization. In
R, a loop to calculate the sum would look like this:

```{r using-for-loop}
total_sum <- 0
# a vector
values <- 1:10
for (value in values) {
  total_sum <- value + total_sum
}
total_sum
```

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Emphasize this next paragraph.
:::

Writing effective and correct loops is surprisingly difficult and
tricky. Because of this and because there are better and easier ways of
writing R code to replace loops, we **strongly** recommend not using
loops. If you think you need to, you probably don't. This is why we will
**not** be covering loops in this workshop.

A **functional** is R's native way of doing loops. A functional is a
function where you *give* it a function as one of its arguments.
@fig-functionals shows how the functional `map()` from the `{purrr}`
package works by taking a vector (or list), applying a function to each
of those items, and outputting the results from each function. The name
`map()` doesn't mean a geographic map, it is the mathematical meaning of
map: To use a function on each item in a set of items.

![A functional, in this case `map()`, applies a function to each item in
a vector. Notice how each of the green coloured boxes are placed into
the `func()` function and outputs the same number of blue boxes as there
are green boxes. Modified from the [Posit purrr cheat
sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/purrr.pdf).
Image license is CC-BY-SA
4.0.](/images/functionals.png){#fig-functionals width="90%"}

Here's a simple example to show how it works. We'll use `paste()` on
each item of `1:5` to simple output the same number as the input. We
will load in `{purrr}` to be clear which package it comes from, but
`{purrr}` is loaded with the `{tidyverse}` package, so you don't need to
load it if you are using the `{tidyverse}` package.

```{r}
library(purrr)
map(1:5, paste)
```

You'll notice that `map()` outputs a list, with all the `[[1]]` printed.
A list is a specific type of object that can contain different types of
objects. For example, a data frame is actually a special type of list
where which column is a list item that contains the values of that
column. `map()` will always output a list, so it is very predictable.
Notice also that the `paste()` function is given *without* the `()`
brackets. Without the brackets, the function can be used by the `map()`
functional and treated like any other object in R. Remember how we said
that a function is an action when it has `()` at the end. In this case,
we do *not* want R to do the action *yet*, we want R to use the action
on each item in the `map()`. In order `map()` to do that, you need to
give the function as an object so it can do the action later.

Let's say we wanted to paste together the number with the sentence
"seconds have passed". Normally it would look like:

```{r paste-multiple}
paste(1, "seconds have passed")
paste(2, "seconds have passed")
paste(3, "seconds have passed")
paste(4, "seconds have passed")
paste(5, "seconds have passed")
```

Or as a loop:

```{r paste-loop}
for (num in 1:5) {
  sec_passed <- paste(num, "seconds have passed")
  print(sec_passed)
}
```

Using `map()` we can give it something called "anonymous" functions.
"Anonymous" because it isn't named with `name <- function()` like we've
done before. Anonymous functions are functions that you use once and
don't remember in the your environment (like named functions do). As you
might guess, you can make an anonymous function by simply *not* naming
it! For instance:

```{r anonymous-func}
#| output: false
function(number) {
  paste(number, "seconds have passed")
}
```

There is a shortened version of this using `\(x)`:

```{r anonymous-func-short}
#| output: false
\(number) paste(number, "seconds have passed")
```

Both forms of anonymous functions are equivalent and can be used in
`map()`:

```{r}
map(1:5, function(number) paste(number, "seconds have passed"))
# Or with the short version
map(1:5, \(number) paste(number, "seconds have passed"))
```

So `map()` will take each number in `1:5`, put it into the anonymous
function in the `number` argument we made, and then run the `paste()`
function on it. The output will be a list of the results of the
`paste()` function.

::: callout-note
`{purrr}` also supports the use of a syntax shortcut to write anonymous
functions. This shortcut is using `~` (tilde) to start the function and
`.x` as the replacement for the vector item. `.x` is used instead of `x`
in order to avoid potential name collisions in functions where `x` is an
function argument (for example in `ggplot2::aes()`, where `x` can be
used to define the x-axis mapping for a graph). Here is the same example
as above, now using the `~` shortcut:

```{r}
map(1:5, ~ paste(.x, "seconds have passed"))
```

The `~` was made before `\()` was made in R. After this new, native R
version was made, the `{purrr}` authors now strongly recommend using it
rather then their `~` for anonymous functions. The `\()` is also a bit
clearer because you give it the name of the argument, e.g. `number`,
rather than be forced to use `.x`.
:::

`map()` will always output a list, but sometimes we might want to output
a different data type. If we look into the help documentation with
`?map`, it shows several other types of map that all start with `map_`:

-   `map_chr()` outputs a character vector.
-   `map_int()` outputs an integer vector.
-   `map_dbl()` outputs a numeric value, called a "double" in
    programming.

This is the basics of functionals like `map()`. Functions,
vectorization, and functionals provide expressive and powerful
approaches to a simple task: *Doing an action on each item in a set of
items*. And while technically using a for loop lets you "not repeat
yourself", they tend to be more error prone and harder to write and read
compared to these other tools. For some alternative explanations of
this, see @sec-alternative-loop-explanation.

{{< text_snippet sticky_up >}}

## Using functionals to import multiple datasets

But what does functionals have to do with what we are doing now? Well,
we have the `import_dime()` that only imports one CSV file. But we have
many more! Well, we can use functionals to import them all at once
:grin:

Before we can use `map()` though, we need to have a vector or list of
the file paths for all the CSV files we want to import. For that, we can
use the `{fs}` package, which has a function called `dir_ls()` that
finds files in a folder (directory).

So, let's add `library(fs)` to the `setup` code chunk. Then, go to the
bottom of the `docs/learning.qmd` document, create a new header called
`## Using map`, and create a code chunk below that with
{{< var keybind.chunk >}}

The `dir_ls()` function takes the path that we want to search, in this
case `data-raw/dime/cgm` and has an argument called `glob` to tell it
what type of files to search for. In our case, we want to search for all
CSV files in the `data-raw/dime/cgm/` folder. So we will pipe the output
of the path to the CGM folder with `here()` into `dir_ls()`:

```{r list-cgm-files}
#| filename: "docs/learning.qmd"
cgm_files <- here("data-raw/dime/cgm/") |>
  dir_ls(glob = "*.csv")
```

Then we can see what the output is. The paths you see on your computer
will be different from what are shown here (it is a different computer).
Plus to keep it simpler, on the website we will only show the first 6
files.

```{r list-cgm-files-show}
#| filename: "docs/learning.qmd"
#| eval: false
cgm_files
```

```{r admin-list-files-for-book}
#| echo: false
head(gsub(".*\\/data-raw", "data-raw", cgm_files), 3)
```

Alright, we now have all the files ready to give to `map()`.

```{r map-import-cgm}
#| filename: "docs/learning.qmd"
cgm_data <- cgm_files |>
  map(import_dime)
```

Remember, that `map()` always outputs a list, so when we look into this
object, it will give us `r length(cgm_data)` tibbles (data frames). For
the website we'll only show the first two tibbles:

```{r map-import-cgm-show}
#| filename: "docs/learning.qmd"
cgm_data[1:2]
```

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Remind everyone that we still only import the first 100 rows of each
data file. So if some of the data itself seems looks odd or so little
data, that is the reason why. Remind them that we do this to more
quickly prototype and test code out.
:::

This is great because with one line of code we imported all these
datasets and made them into one data frame! But we're missing an
important bit of information: The participant ID. The `{purrr}` package
has many other powerful functions to make it easier to work with
functionals. We know `map()` always outputs a list. But what we want is
a single tibble at the end that also contains the participant ID.

There are two functions that take a list of data frames and convert them
into a single data frame. They are called `list_rbind()` to bind
("stack") the data frames by rows or `list_cbind()` to bind ("stack")
the data frames by columns. In our case, we want to bind (stack) by
rows, so we will use `list_rbind()` by piping the output of the `map()`
code we wrote into `list_rbind()`.

```{r use-list-rbind}
#| filename: "docs/learning.qmd"
cgm_data <- cgm_files |>
  map(import_dime) |>
  list_rbind()
cgm_data
```

But, hmm, we don't have the participant ID in the data frame. This is
because `list_rbind()` doesn't know how to get that information, since
it isn't included in the data frames. If we look at the help for
`list_rbind()`, we will see that it has an argument called `names_to`.
This argument lets us create a new column that is based on the name of
the list item, which in our case is the file path. This file path also
has the participant ID information in it, but it also has the full file
path in it too, which isn't exactly what we want. So we'll have to fix
it later. But first, let's start with adding this information to the
data frame as a new column called `file_path_id`. So we will add the
`names_to` argument to the code we've already written.

```{r use-list-rbind-with-paths}
#| filename: "docs/learning.qmd"
cgm_data <- cgm_files |>
  map(import_dime) |>
  list_rbind(names_to = "file_path_id")
```

And then look at it by running `cgm_data` below the code:

```{r use-list-rbind-with-paths-show}
#| filename: "docs/learning.qmd"
#| eval: false
cgm_data
```

```{r admin-show-rbind-with-path-for-book}
#| echo: false
cgm_data |>
  trim_filepath_for_book()
```

The `file_path_id` variable will look different on everyone's computer.
Don't worry, we're going to tidy up the `file_path_id` variable later in
another session.

Before moving on to the exercise, let's style with
{{< var keybind.style >}} the document, render with
{{< var keybind.render >}}, and finally **add and commit** the changes
to the Git history using {{< var keybind.commit >}}.

## :technologist: Exercise: Convert into a function to use it on the sleep data

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Because later sessions depend on this code, after they've finished the
exercise, walk through with them the solution. So that we are all on the
same page.
:::

**Time: \~20 minutes.**

We've made some code that does what we want, now the next step is to
convert it into a function so we can use it on both the `cgm` and
`sleep` datasets. The code we've written to import only the `cgm` data
looks like this:

```{r}
#| eval: false
cgm_files <- here("data-raw/dime/cgm/") |>
  dir_ls(glob = "*.csv")
cgm_data <- cgm_files |>
  map(import_dime) |>
  list_rbind(names_to = "file_path_id")
cgm_data
```

We want to be able to do the same thing for the `sleep` data, but we
don't want to repeat ourselves. What we want is a new function that
looks like the below code chunk and that outputs a similar data frame,
though the `file_path_id` values will be different:

{{< include /includes/solutions/_functionals-import-multiple-files.qmd >}}

```{r}
#| eval: false
here("data-raw/dime/cgm/") |>
  import_csv_files()
here("data-raw/dime/sleep/") |>
  import_csv_files()
```

```{r}
#| echo: false

here("data-raw/dime/cgm/") |>
  import_csv_files() |>
  trim_filepath_for_book()
here("data-raw/dime/sleep/") |>
  import_csv_files() |>
  trim_filepath_for_book()
```

Complete the tasks below to convert this code into a stable, robust
function that you can re-use.

1.  We've used two packages here, `{fs}` and `{purrr}`. You will first
    need to add these packages to the `DESCRIPTION` file by using
    `usethis::use_package()` in the **Console**.
2.  In the bottom of `docs/learning.qmd`, create a new header
    `## Exercise: Convert map to function and use on sleep`.
3.  Below the header, create a new code chunk using
    {{< var keybind.chunk >}}.
4.  As you've already done in the previous sessions, convert this code
    into a function. You can always refer to the @sec-functions and
    @sec-robustness sessions.
    -   Name the function `import_csv_files`.
    -   Use the `function() { ... }` code to create a new function
    -   Inside `function()`, add an argument called `folder_path`.
    -   Paste the code into the function body.
    -   Replace `here("data-raw/dime/cgm")` with `folder_path`.
    -   Rename `cgm_files` to `files`.
    -   Rename `cgm_data` to `data`.
    -   Include a `return()` at the end.
    -   Add `package_name::` to the individual functions used. To find
        which package a function comes from, use `?functionname` to see
        the help documentation.
    -   Create and write Roxygen documentation to describe the new
        function by using {{< var keybind.roxygen >}}.
    -   Run it and check that it works using the code shown above.
5.  Render the document with {{< var keybind.render >}} to check that
    the function works.
6.  **Cut** and paste the function into the `R/functions.R` file. Then
    go back to the `docs/learning.qmd` file and (making sure the old
    function is deleted), render again with {{< var keybind.render >}}
    to check that the function works after moving it.
7.  Run `{styler}` while in the `R/functions.R` file with
    {{< var keybind.style >}}.
8.  Once done, **add** the changes you've made and **commit** them to
    the Git history, using {{< var keybind.commit >}}.

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

-   R is a functional programming language:
    -   It has functions that take an input, do an action, and give an
        output.
    -   It uses vectorisation that apply a function to multiple items
        (in a vector) all at once rather than using loops.
    -   It uses functionals that allow functions to use other functions
        as input.
-   Thinking in terms of declaratively telling a computer to do actions,
    rather than telling a computer how to do things step-by-step, is a
    powerful way to think about programming, especially in R.
-   in terms taking an object like a data frame or list, doing an
    action, outputting a data frame or list, and doing more actions on
    that output is
-   Use the `{purrr}` package and its function `map()` when you want to
    repeat a function on multiple items at once.
-   Use `list_rbind()` to combine multiple data frames into one data
    frame by stacking them on top of each other.

## :speech_balloon: Discussion activity: Incorporating functionals in your own work

**Time: \~6 minutes.**

As we prepare for the next session and taking the break, get up and
discuss with your neighbour or group the following questions:

-   What are some things you do repetitively that you think might
    benefit from using functionals on?
-   Consider your code you've written for other projects or tasks you
    have in other projects. How might you use functional programming to
    help your work? How might you change existing code to use them?
    Might it be a lot of work?

```{r admin-store-session-code}
#| include: false
save.image(here::here("_temp/functionals.RData"))
```

{{< include /_extensions/rostools/r3-theme/includes/code-appendix-r.qmd >}}
