# Save time, don't repeat yourself: Making functions {#sec-dry-functions}

```{r setup}
#| include: false
library(fs)
library(tidyverse)
library(readr)
library(here)
load(here("_temp/importing.RData"))
temp_path <- fs::path_temp()
project_path <- fs::path(temp_path, "LearnR3")

untar(
  tarfile = here::here("_temp", "pre-course.tar.gz"),
  exdir = project_path
)
```

Here we will cover the second block, "*Workflow*" in
@fig-overview-workflow.

![Section of the overall workflow we will be
covering.](/images/overview-workflow.svg){#fig-overview-workflow}

## Learning objectives

The overall learning outcome for this session is to:

1.  Describe a basic workflow for creating functions and then apply this
    workflow to import data.

Specific objectives are to:

1.  Describe and identify the individual components of a function as
    well as the workflow to creating them, and then use this workflow to
    make one to import some data.
2.  Describe a workflow for developing code that combines using Quarto,
    `source()`'ing with {{< var keybind.source >}}, and restarting R
    with {{< var keybind.restart-r >}}, explain its usefulness in
    assessing correctness and re-usability of code, and then apply this
    workflow when developing functions.
3.  Explain what R package dependency management is, why it is necessary
    when writing code and ensuring reproducibility, and apply tools like
    `usethis::use_package()` to manage dependencies in an R project.
4.  Describe and apply a workflow of prototyping code in a Quarto
    document and then moving it into a script (called, e.g.,
    `functions.R`) once finished and tested, then using `source()` to
    load the functions into the R session.
5.  Continue practicing Git version control to manage changes to your
    files.

## The basics of a function

::: callout-note
## Reading task: \~5 minutes

We've mentioned functions multiple times, but what is a function? At its
core, a function in R is anything that does an action. A function is a
bundled sequence of steps that achieve a specific action. For instance,
the `+` (to add) is a function, `mean()` is a function, `[]` (to subset
or extract) is a function, and so on. In simple terms, functions are
made of a function call, its arguments, and the function body:

``` {.r filename="Console"}
function(argument1, argument2){ 
  # body of function with R code
}
```

Because R is open source, anyone can see how things work underneath. So,
if we want to see what a function does underneath, we type out the
function name without the `()` into the **Console** and run it. If we do
it with the function `sd()` which calculates the standard deviation, we
see:

```{r show-sd-internals}
#| filename: "Console"
sd
```

Here you see `sd()` as the arguments `x` and `na.rm`. Within the
function body is how it calculates the standard deviation, which is the
square root of the variance. In this code, the `var()` is inside the
`sqrt()` function, which is exactly what it should be.

So, if you learn how to create your own functions, it can make doing
your work easier and more efficient because you don't have to repeat
yourself later. Making functions always has a basic structure of:

1.  Giving a name to the function (e.g. `mean`).
2.  Starting the function call using `function()`, assigning it to the
    name with `<-`. This tells R that the name is a function object.
3.  Optionally providing arguments to give to the function call, for
    instance `function(argument1, argument2, argument3)`.
4.  Filling out the body of the function, with the arguments (if any)
    contained inside, that does some action.
5.  Optionally, use `return()` to indicate what you want the function to
    output. For learning purposes, we'll always use `return()` to help
    show us what is the final function output.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Emphasize that we will be using this workflow for creating functions all
the time throughout course and that this workflow is also what you'd use
in your daily work.
:::

While there is no minimum or maximum number of arguments you can provide
for a function (e.g. you could have zero or dozens of arguments), its
good practice for yourself and for others to have as few arguments as
necessary to get the job done. So, the structure is:

``` {.r filename="Console"}
name <- function(argument1, argument2) {
    # body of function
    output <- ... code ....
    return(output)
}
```

Writing your own functions can be absolutely amazing and fun and
powerful, but you also often want to pull your hair out with frustration
at errors that are difficult to understand and fix. The best way to deal
with this is by debugging. Due to time and to the challenge of making
meaningful debugging exercises (solutions to problems are very dependent
on the project), read @sec-extra-material in your own time for some
instructions on debugging and dealing with another common problem you
might encounter with R.
:::

Let's write a simple example. First, create a new Markdown header called
`## Making a function` and create a code chunk below that with
{{< var keybind.chunk >}} . Then, inside the code chunk, we'll write
this code out:

```{r create-add-function}
#| filename: "doc/learning.qmd"
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

You can use the new function by running the above code and writing out
your new function, with arguments to give it.

```{r call-function}
#| filename: "doc/learning.qmd"
add_numbers(1, 2)
```

The function name is fairly good; `add_numbers` is read as "add
numbers". While we generally want to write code that describes what it
does by reading it, it's also good practice to add some formal
documentation to the function. Use the "Insert Roxygen Skeleton" in the
"Code" menu, by typing {{< var keybind.roxygen >}}, and you can add
template documentation right above the function. It looks like:

```{r roxygen-docs-1}
#| filename: "doc/learning.qmd"
#' Title
#'
#' @param num1
#' @param num2
#'
#' @return
#' @export
#'
#' @examples
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or
several words that describe the function. Creating a new paragraph below
this line allows you to add a more detailed description. The other items
are:

-   `@param num`: These lines describe what each argument (also called
    parameter) is for and what to give it.
-   `@return`: This describes what output the function gives. Is it a
    data.frame? A plot? What else does the output give?
-   `@export`: Tells R that this function should be accessible to the
    user of your package. Since we aren't making packages, delete it.
-   `@examples`: Any lines below this are used to show examples of how
    to use the function. This is very useful when making packages, but
    not really in this case. So we'll delete it. Let's write out some
    documentation for this function:

```{r roxygen-docs-2}
#| filename: "doc/learning.qmd"
#' Add two numbers together.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @return Returns the sum of the two numbers.
#'
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

Once we've created that, let's open up the Git Interface with
{{< var keybind.git >}} and **add and commit** these changes to our
history.

## Making a function for importing our data

Now that we have a basic understanding of what a function looks like,
let's apply it to something we're doing right now: Importing our data.

Making functions is a series of steps:

1.  Write code that works and does what you want.
2.  Enclose it as a function with `name <- function() { ... }`, with an
    appropriate and descriptive name.
3.  Create arguments in the function call
    (`function(argument1, argument2)`) with appropriate and descriptive
    names, then replace the code with the argument names where
    appropriate.
4.  Rename any objects created to be more generic and include the
    `return()` function at the end to indicate what the function will
    output.
5.  Run the function and check that it works.
6.  Add the Roxygen documentation tags with {{< var keybind.roxygen >}}
    while the cursor is in the function.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Emphasize that we will be using this workflow for creating functions all
the time throughout course and that this workflow is also what you'd use
in your daily work.
:::

In `doc/learning.qmd`, create a new Markdown header called
`## Import the user data with a function` and create a code chunk below
that with {{< var keybind.chunk >}} .

So, step one. Let's take the code we wrote for importing the `user_info`
data and convert that as a function:

```{r function-creation-step-1}
#| filename: "doc/learning.qmd"
#| eval: false
user_1_info_data <- read_csv(
  user_1_info_file,
  col_select = -1,
  col_types = cols(
    gender = col_character(),
    weight = col_double(),
    height = col_double(),
    age = col_double(),
    .delim = ","
  ),
  name_repair = snakecase::to_snake_case
)
```

Next we wrap it in the function call and give it an appropriate name. In
this case, `import_user_info` is descriptive and meaningful.

```{r function-creation-step-2}
#| filename: "doc/learning.qmd"
#| eval: false
import_user_info <- function() {
  user_1_info_data <- read_csv(
    user_1_info_file,
    col_select = -1,
    col_types = cols(
      gender = col_character(),
      weight = col_double(),
      height = col_double(),
      age = col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
}
```

Then, we add arguments in the function and replace within the code.
Here, we have only one thing that we would change: The file path to the
dataset. So, a good name might be `file_path`.

```{r function-creation-step-3}
#| filename: "doc/learning.qmd"
#| eval: false
import_user_info <- function(file_path) {
  user_1_info_data <- read_csv(
    file_path,
    col_select = -1,
    col_types = cols(
      gender = col_character(),
      weight = col_double(),
      height = col_double(),
      age = col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
}
```

Then we clean things up by renaming `user_1_info_data` since we would
like to also import more than just `user_1`. A nice object name would be
`info_data`. Add the `return()` function at the end with the object you
want your function to output.

```{r function-creation-step-4}
#| filename: "doc/learning.qmd"
import_user_info <- function(file_path) {
  info_data <- read_csv(
    file_path,
    col_select = -1,
    col_types = cols(
      gender = col_character(),
      weight = col_double(),
      height = col_double(),
      age = col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(info_data)
}
```

Great! Now we need to test it out. Let's try on two datasets, two
`user_info.csv` files in the `user_1` and `user_2` folders.

```{r function-creation-step-5}
#| filename: "doc/learning.qmd"
import_user_info(here("data-raw/mmash/user_1/user_info.csv"))
import_user_info(here("data-raw/mmash/user_2/user_info.csv"))
```

Awesome! It works. The final stage is adding the Roxygen documentation.

```{r function-creation-step-6}
#| filename: "doc/learning.qmd"
#' Import MMASH user info data file.
#'
#' @param file_path Path to user info data file.
#'
#' @return Outputs a data frame/tibble.
#'
import_user_info <- function(file_path) {
  info_data <- read_csv(
    file_path,
    col_select = -1,
    col_types = cols(
      gender = col_character(),
      weight = col_double(),
      height = col_double(),
      age = col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(info_data)
}
```

A massive advantage of using functions is that if you want to make a
change to all your code, you can very easily do it by modifying the
function and it will change all your other code too. Now that we have a
working function, run `{styler}` with {{< var keybind.styler >}} and
then let's **add and commit** it to the Git history with the RStudio Git
Interface by using {{< var keybind.git >}}.

## Exercise: Repeat with the saliva data

> Time: 15 minutes.

Take the code you created for importing the saliva data set from
@sec-ex-import-saliva (*not* the code related to using `spec()`) and
make it into a function. It looks like the code below.

``` {.r filename="doc/learning.qmd"}
user_1_saliva_data <- read_csv(
    user_1_saliva_file,
    col_select = -1,
    col_types = cols(
        samples = col_character(),
        cortisol_norm = col_double(),
        melatonin_norm = col_double()  
    ), 
    name_repair = snakecase::to_snake_case
)
```

::: {.callout-tip appearance="default"}
A helpful tip: To move around a Quarto or R script more easily, open up
the "Document Outline" on the side by clicking the button in the top
right corner of the Quarto pane or by using {{< var keybind.outline >}}.
:::

1.  Create a new markdown header
    `## Exercise for importing the saliva data as a function`.
2.  Create a new code chunk below the header with
    {{< var keybind.chunk >}}.
3.  Paste the code you used from the exercise (shown above) into the
    code chunk and begin converting it into a function, like we did
    above.
    -   Wrap it with the `function() {...}`
    -   Make a meaningful name (use `import_saliva`)
    -   Make an argument for the file path (`file_path`) and replace
        `user_1_saliva_file` with `file_path` in the `read_csv()` code
    -   Rename the output object to `saliva_data` and put it in the
        `return()` function
    -   Run the function and then test that it works
    -   Create the Roxygen documentation by adding it with
        {{< var keybind.roxygen >}}
4.  Run `{styler}` while in the `doc/learning.qmd` file with
    {{< var keybind.styler >}}.
5.  Then, **add and commit** the changes to the Git history, using
    {{< var keybind.git >}}.

Use the below code as a guide:

``` {.r filename="doc/learning.qmd"}
# Need to also add the Roxygen documentation 
import_saliva <- function(file_path) {
    # Paste the code to import saliva data you created 
    # from previous exercise
    saliva_data <- ___(
        ___
    )
    return(saliva_data)
}

# Test that the function works
___(here("data-raw/mmash/user_1/saliva.csv"))
```

```{r solution-repeat-saliva}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
#' Import the MMASH saliva file.
#'
#' @param file_path Path to the user saliva data file.
#'
#' @return Outputs a data frame/tibble.
#'
import_saliva <- function(file_path) {
  saliva_data <- read_csv(
    file_path,
    col_select = -1,
    col_types = cols(
      samples = col_character(),
      cortisol_norm = col_double(),
      melatonin_norm = col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(saliva_data)
}

# Test that this works
# import_saliva(here("data-raw/mmash/user_1/saliva.csv"))
```

## Continuing the workflow

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Really emphasize to **cut** and paste, so that the function in the
`doc/learning.qmd` file is **deleted** and no longer kept in the Quarto
document.
:::

We've created two functions. Now we need to move those functions from
the `doc/learning.qmd` file and into the `R/` folder by **cutting and
pasting** (not just copying). We do this for a few reasons:

1.  To prevent the Quarto document from becoming too long and having a
    large portion of R code over other text.
2.  To make it easier to maintain and find things in your project.
3.  To make use of the `source()` function.

We want to store our functions in the file `R/functions.R` script so its
easier to source them. Cut and paste *only* the `import_user_info()`
function we created in `doc/learning.qmd`, including the Roxygen
documentation, and paste it into the newly created `R/functions.R`.

Once we have it in there, let's test out the workflow. Restart our R
session with by either going to the "Session -\> Restart R" menu item or
by using the keybindings {{< var keybind.restart-r >}}. Move back into
the `doc/learning.qmd` and add `source(here("R/functions.R"))` to the
code chunk called `setup` at the top. Run all the code inside the
`setup` code chunk. Then go to where you wrote:

``` {.r filename="doc/learning.qmd"}
import_user_info(here::here("data-raw/mmash/user_1/user_info.csv"))
```

Now run this line. What happens now? You *may* get an error about not
finding the `read_csv()` function. If you put `library(tidyverse)` in
the `setup` code chunk, you might not get an error. If you did get an
error, that's because R doesn't know what the `read_csv()` function is.
This is where we start getting into package dependency management.

What is package dependency management? Whenever you use an R package,
you depend on it for your code to work. The informal way to show what
packages you use is by using the `library()` function. But if you come
back to the project, or get a new computer, or someone else is working
on your project too, how will they know which packages your project
depends on? Do they have to search through all your files just to find
all `library()` functions you used and then install those packages
individually? A much better way here is to formally indicate your
package dependency so that installing dependencies is easy. We do this
by making use of the `DESCRIPTION` file.

Open up the `DESCRIPTION` file. You may or may not see something that
looks like:

```{r admin-print-desc}
#| echo: false
#| comment: ""
usethis::with_project(
  path = project_path,
  {
    desc::desc_print()
  },
  quiet = TRUE
)
```

If it doesn't look like this, replace all of your current text with the
text above. Notice the `Imports:` key. This is where information about
packages are added. To quickly add a package, go to the **Console** and
type out:

``` {.r filename="Console"}
usethis::use_package("readr")
```

```{r admin-use-package-readr}
#| echo: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_package("readr")
  },
  quiet = TRUE
)
```

You will see a bunch of text about adding it to `Imports`. If you look
in your `DESCRIPTION` file now, you'll see something like:

```{r admin-print-desc-add-readr}
#| echo: false
#| comment: ""
usethis::with_project(
  path = project_path,
  {
    desc::desc_print()
  },
  quiet = TRUE
)
```

Now, if you or someone else wants to install all the packages your
project depends on, they can do that by going to the **Console** and
running:

``` {.r filename="Console"}
pak::pak()
```

This function finds the `DESCRIPTION` file and installs all the packages
in `Imports`. Let's add the other dependencies by typing in the
**Console**:

``` {.r filename="Console"}
usethis::use_package("here")
usethis::use_package("fs")
usethis::use_package("snakecase")
```

```{r admin-use-packages-several}
#| echo: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_package("here")
    usethis::use_package("fs")
    usethis::use_package("snakecase")
  },
  quiet = TRUE
)
```

Since we will also make use of the tidyverse set of packages later in
the course, we'll also add tidyverse as a dependency. Since the
tidyverse is a large collection of packages, the recommended way to add
this particular dependency is with:

``` {.r filename="Console"}
usethis::use_package("tidyverse", type = "Depends")
```

```{r admin-use-packages-tidyverse}
#| echo: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_package("tidyverse", type = "Depends")
  },
  quiet = TRUE
)
```

If you look in the `DESCRIPTION` file now, you see that the new
`Depends` field has been added with `tidyverse` right below it. There
are fairly technical reasons why we need to put tidyverse in the
`Depends` field that you don't need to know about for this course, aside
from the fact that it is a common practice in R projects. At least in
this context, we use the `Depends` field for tidyverse because of one
big reason: the `usethis::use_package()` function will complain if we
try to put tidyverse in the `Imports` and it recommends putting it in
the `Depends` field.

```{r admin-print-desc-add-tidyverse}
#| echo: false
#| comment: ""
usethis::with_project(
  path = project_path,
  {
    desc::desc_print()
  },
  quiet = TRUE
)
```

Great! Now that we've formally established package dependencies in our
project, we also need to formally declare which package each function
comes from inside our own functions.

::: callout-note
## Reading task: \~5 minutes

Before getting into the correct way, we need to quickly cover the
incorrect way that you may or may not have seen how others have done it
on websites or in script files. Sometimes people use `library()` or
`require()` inside functions like:

``` r
add_numbers <- function(num1, num2) {
    library(packagename)
    ...code...
    return(added)
}
```

Or:

``` r
add_numbers <- function(num1, num2) {
    require(packagename)
    ...code...
    return(added)
}
```

This is *very* **bad practice** and can have some unintended and serious
consequences that you might not notice or that won't give any warning or
error. The correct way of indicating which package a function comes from
is instead by using `packagename::functionname`, which you've seen and
used many times in this course. We won't get into the reasons why this
is incorrect because it can quickly get quite technical.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

You can also talk about why `require()` shouldn't be used compared to
`library()`. The problem with `require()` is that if the package can't
be loaded, it won't throw an error, it only *checks* if the package is
available and will otherwise continue running the code. On the other
hand, `library()` will throw an error if it can't find the package,
which is what you expect if your code depends on a package.
:::

Another reason to use `packagename::functionname` for each function from
an R package you use in your own function is that it explicitly tells R
(and us the readers) where the function comes from. Because the same
function name can be used by multiple packages, if you don't explicitly
state which package the function is from, R will use the function that
it finds first - which isn't always the function you meant to use. We
also do this step at the end of making the function because doing it
while we create it can be quite tedious.
:::

Alright, let's go into `R/functions.R` and add `readr::` to each of the
`{readr}` functions we've used:

```{r explicit-function-calls}
#| filename: "R/functions.R"
import_user_info <- function(file_path) {
  info_data <- readr::read_csv(
    file_path,
    col_select = -1,
    col_types = readr::cols(
      gender = readr::col_character(),
      weight = readr::col_double(),
      height = readr::col_double(),
      age = readr::col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(info_data)
}
```

Test that it works by restart the R session with
{{< var keybind.restart-r >}} and source the file with
{{< var keybind.source >}}, then go to the **Console** and type out:

``` {.r filename="Console"}
import_user_info(here::here("data-raw/mmash/user_1/user_info.csv"))
```

It should work as expected! Now that we've done that, let's **add and
commit** the changes made through the Git interface, using
{{< var keybind.git >}}.

## Exercise: Move and update the rest of the functions

> Time: \~30 minutes.

Repeat this process of making functions by doing this to the rest of the
code you worked on previously that imported the `RR.csv` and
`Actigraph.csv` data.

1.  Convert the importing code into functions while in the
    `doc/learning.qmd` file. Include the Roxygen documentation and use
    `packagename::` to be explicit about where the function comes from.

    -   Name the new functions `import_rr` and `import_actigraph`.

2.  Move (by **cutting** and pasting) **only** the function into
    `R/functions.R`.

3.  Restart R, `source()` the functions file, using
    {{< var keybind.source >}}, and test that the functions work by
    running them in the **Console**. The below code should run without a
    problem if you did it right:

    ``` {.r filename="Console"}
    import_rr(here::here("data-raw/mmash/user_1/RR.csv"))
    import_actigraph(here::here("data-raw/mmash/user_1/Actigraph.csv"))
    ```

4.  Run `{styler}` while in the `R/functions.R` file with
    {{< var keybind.styler >}}.

Also update the `import_saliva()` function you created by being explicit
about where the functions come from (e.g. with the `packagename::`).
Then **cut and paste** this function along with its Roxygen
documentation over into the `R/functions.R` file. Run `{styler}` with
{{< var keybind.styler >}} and afterwards, **add and commit** the
changes to the Git history, using {{< var keybind.git >}}.

Use this code template as a guide for making the functions.

``` r
# Insert Roxygen documentation too
___ <- function(___) {
    ___ <- ___::___(
        ___,
        col_select = ___,
        col_types = ___::cols(
            ___,
            .delim = ","
        ),
        name_repair = snakecase::to_snake_case
    )
    return(___)
}
```

```{r solution-update-other-functions}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
#' Import the MMASH saliva file
#'
#' @param file_path Path to the user saliva data file.
#'
#' @return Outputs a data frame/tibble.
#'
import_saliva <- function(file_path) {
  saliva_data <- readr::read_csv(
    file_path,
    col_select = -1,
    col_types = readr::cols(
      samples = readr::col_character(),
      cortisol_norm = readr::col_double(),
      melatonin_norm = readr::col_double(),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(saliva_data)
}

#' Import the MMASH RR file (heart beat-to-beat interval).
#'
#' @param file_path Path to the user RR data file.
#'
#' @return Outputs a data frame/tibble.
#'
import_rr <- function(file_path) {
  rr_data <- readr::read_csv(
    file_path,
    col_select = -1,
    col_types = readr::cols(
      ibi_s = readr::col_double(),
      day = readr::col_double(),
      # Converts to seconds
      time = readr::col_time(format = ""),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(rr_data)
}

#' Import the MMASH Actigraph file (accelerometer).
#'
#' @param file_path Path to the user Actigraph data file.
#'
#' @return Outputs a data frame/tibble.
#'
import_actigraph <- function(file_path) {
  actigraph_data <- readr::read_csv(
    file_path,
    col_select = -1,
    col_types = readr::cols(
      axis_1 = readr::col_double(),
      axis_2 = readr::col_double(),
      axis_3 = readr::col_double(),
      steps = readr::col_double(),
      hr = readr::col_double(),
      inclinometer_off = readr::col_double(),
      inclinometer_standing = readr::col_double(),
      inclinometer_sitting = readr::col_double(),
      inclinometer_lying = readr::col_double(),
      vector_magnitude = readr::col_double(),
      day = readr::col_double(),
      time = readr::col_time(format = ""),
      .delim = ","
    ),
    name_repair = snakecase::to_snake_case
  )
  return(actigraph_data)
}
```

## Exercise: Brainstorm and discuss why and what you could make as a function

> Time: 15 minutes.

You've learned the basics of making your own, custom function. Now, as a
group, brainstorm and discuss some ways that you might make functions in
your own work to help reduce repetition. What type of code might you
make as a function for your own project? Do you think others, maybe in
your research group, might use this function too? Afterwards, all the
groups will briefly share what they thought of.

## Summary

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Quickly cover this before finishing the session and when starting the
next session.
:::

-   Functions in R are anything that does an action.
-   Functions have five components:
    -   The three required ones are the function call with
        `function() { }`, the function body between the `{ }`, and an
        output (usually set with `return()`).
    -   The two optional ones are assigning the function to a named
        object with `<-` and the function arguments put within
        `function()`.
-   Write function documentation by using Roxygen.
-   Use `use_package()` for the `DESCRIPTION` file as well as
    `packagename::functionname()` to explicit state the packages your
    function depends on.
    -   Don't use `library()` or `require()` *within* your function.

```{r store-session-code}
#| include: false
save.image(here::here("_temp/functions.RData"))
```
