# Bundling code into functions {#sec-functions}

```{r}
#| label: setup
#| include: false

library(fs)
library(tidyverse)
library(snakecase)
library(here)
```

We've written simple code that reads in one of our data files and does
some simple cleaning to it. But we have about 2000 CSV files to read in.
So even with our simple code, isn't going to be very efficient if we
just copy and paste that code to read in all those files. This session
is about bundling our code into functions that will make it easier to
read more files with less code.

## Learning objectives

{{< include /includes/objectives/_functions.qmd >}}

## :book: Reading task: The basics of a function

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Repeat and reinforce the part of what functions are made of, their
structure, that all actions are functions, and all functions are objects
(but not that all objects are functions).
:::

**Time: \~5 minutes.**

The first thing to know about R is that everything is an "object" and
that some objects can do an action. These objects that do an action are
called functions. You've heard or read about functions before during the
workshop, but what is a function?

A function is a bundled sequence of steps that achieve a specific action
and you can usually tell if an object is an action if it has a `()` at
the end of it's name. For example, `mean()` is function to calculate the
mean or `sd()` is a function to calculate the standard deviation. It
isn't always true that functions end in `()` though, which you'll read
about shortly. For instance, the `+` is a function that adds two numbers
together, the `[]` is a function that is used to subset or extract an
item from a list of items like getting a column from a data frame, or
`<-` is a function that creates a new object from some value.

All created functions have the same structure: they are assigned a
function name with `<-`, it uses `function()` to give it its parameters
or arguments, and it has the internal sequence of steps called the
function body that are wrapped in `{}`:

```r
function_name <- function(argument1, argument2) {
  # body of function with R code
}
```

Notice that this uses two functions to create this `function_name`
function object:

- `<-` is the action (function) that will create the new object
  `function_name`.
- `function()` is the action (function) to tell R that this object is an
  action (function) whenever it is used with a `()` at the end, e.g.
  `function_name()`.

Because R is open source, anyone can see how things work underneath. So,
if you want to see what a function does underneath, you would type out
the function name without the `()` into the **Console** and run it. If
we do it with the function `sd()` which calculates the standard
deviation, we see:

```{r}
#| label: show-sd-internals
#| filename: "Console"

sd
```

Here you see `sd()` as the arguments `x` and `na.rm`. In the function
body it shows how it calculates the standard deviation, which is the
square root of the variance. In this code, the `var()` is inside the
`sqrt()` function, which is exactly what it should be if you know the
math (you don't need to).

Normally you can tell if something is a function if it has `()` at the
end of the name. But there are special functions, like `+`, `[]`, or
even `<-`, that do an action but that don't have `()` at the end. These
are called *operator* functions. Which is why the `<-` is called the
assignment operator, because it assigns something to a new object. To
see how they work internally, you would wrap \` around the operator. So
for `+` or `<-` it would be:

```{r}
#| label: show-operators-internals
#| filename: "Console"

`+`
`<-`
```

You'll see something called `.Primitive`. Often operators are written in
very low level computer code to work, which are called "primitives".
This is *way* beyond the scope of this workshop to explain primitives
and so we won't go into what that means and why.

To show that they are a function, you can even use them with their `()`
version like this:

```{r}
#| label: show-how-operators-work

1 + 2
`+`(1, 2)

`<-`(x, 1)
x
x <- 1
x
```

But hopefully you can see that using it with the `()` isn't very nice to
read or use!

If you can learn to make your own functions, it can help make your life
and work much easier and more efficient! That's because you can make a
sequence of actions that you can then reuse again and again. And
luckily, you will be making many functions throughout this workshop.
Making a function always follows a basic structure:

1. Give a name to the function (e.g. `mean`).
2. Use `function()` to tell R the new object will be a function and
   assigning it to the name with `<-`.
3. Optionally provide arguments to the function object, for example
   `function(argument1, argument2, argument3)`.
4. Fill out the body of the function, with the arguments (if any)
   contained inside, that does some sequence of actions.
5. Optionally, use `return()` to indicate what final output you want the
   function to have. For learning purposes, we'll always use `return()`
   to help show us what is the final function output but it isn't
   necessary.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Emphasize that we will be using this workflow for creating functions all
the time throughout workshop and that this workflow is also what you'd
use in your daily work.
:::

While there is no minimum or maximum number of arguments you can provide
for a function (e.g. you could have zero or dozens of arguments), its
generally good practice and design to have as few arguments as necessary
to get the job done. Part of making functions is to reduce your own and
others cognitive load when working with or reading code. The fewer
arguments you use, the lower the cognitive load. So, the structure is:

```r
name <- function(argument1, argument2) {
    # body of function
    output <- ... code ....
    return(output)
}
```

Writing your own functions can be absolutely amazing and fun and
powerful, but you also often want to pull your hair out with frustration
at errors that are difficult to understand and fix. One of the best ways
to deal with this is by making functions that are small and simple, and
testing them as you use them. The smaller they are, the less chance you
will have that there will be an error or issue that you can't figure
out. There's also some formal debugging steps you can do but due to time
and to the challenge of making meaningful debugging exercises, as
solutions to problems are very dependent on the project and context, we
won't be going over debugging. There is some extra material in
@sec-extra-material that you can look over in your own time that has
some instructions on debugging and dealing with some common problems you
might encounter with R.

{{< text_snippet sticky_up >}}

## Creating our first function

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Take your time slowly going over this, especially taking about the
Roxygen documentation template.
:::

Let's create a really basic function to show the process. First, create
a new Markdown header called `## Making a function to add numbers` and
create a code chunk below that with {{< var keybind.chunk >}}. Then,
inside the code chunk, we'll write this code:

```{r}
#| label: create-add-function
#| filename: "docs/learning.qmd"

add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

You can use the new function by running the above code and writing out
your new function, with arguments to give it.

```{r}
#| label: call-function
#| filename: "docs/learning.qmd"

add_numbers(1, 2)
```

The function name is fairly good; `add_numbers` is read as "add
numbers". While we generally want to write code that describes what it
does by reading it, it's also good practice to add some formal
documentation to the function. Use the "Insert Roxygen Skeleton" in the
"Code" menu, by typing {{< var keybind.roxygen >}}, and you can add
template documentation right above the function. Make sure your cursor
is within the function in order for the Roxygen template to be added to
your function. It looks like:

```{r}
#| label: roxygen-docs-1
#| filename: "docs/learning.qmd"

#' Title
#'
#' @param num1
#' @param num2
#'
#' @returns
#' @export
#'
#' @examples
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

In the `Title` area, this is where you type out a brief sentence or
several words that describe the function. Creating a new paragraph below
this line allows you to add a more detailed description. The other items
are:

- `@param num`: These lines describe what each argument (also called
  parameter) is for and what to give it.
- `@return`: This describes what output the function gives. Is it a
  data.frame? A plot? What else does the output give?
- `@export`: Tells R that this function should be accessible to the user
  of your package. Since we aren't making packages, delete it.
- `@examples`: Any lines below this are used to show examples of how to
  use the function. This is very useful when making packages, but not
  really in this case. So we'll delete it. Let's write out some
  documentation for this function:

```{r}
#| label: roxygen-docs-2
#| filename: "docs/learning.qmd"

#' Add two numbers together.
#'
#' @param num1 A number here.
#' @param num2 A number here.
#'
#' @returns Returns the sum of the two numbers.
#'
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

Once we've created that and before moving on, let's style our code with
{{< var keybind.style >}}, render the Quarto document with
{{< var keybind.render >}}, and then open up the Git Interface with
{{< var keybind.commit >}} to **add and commit** these changes to the
Git history before then pushing to GitHub.

## :book: Reading task: Workflow for prototyping and creating functions

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Highlight the workflow and diagram. Reinforce this workflow and that we
will be using it all throughout this workshop.
:::

**Time: \~6 minutes.**

At the level of the code, the way you prototype code is to:

1. Write the code out in Quarto so that it does what you want.
2. Convert that code into a function.
3. Test that the function works either in the Quarto document or in the
   R Console.
4. Fix the function if it doesn't work.
5. Restart the R console with {{< var keybind.restart-r >}} or render
   with {{< var keybind.render >}} to test that the function works in a
   clean environment.
6. Whenever the function works, *add and commit* the changes to the Git
   history with {{< var keybind.commit >}} (or commit after you move the
   function to the `R/functions.R` script, which we will talk about in
   the next session).

```{mermaid}
%%| echo: false
%%| label: "fig-workflow-prototyping"
%%| fig-cap: "Workflow for prototyping code in Quarto, converting to a function, testing it, rendering or restarting, and committing to Git."
%%| fig-alt: "A diagram representing the workflow for prototyping code in Quarto, converting it to a function, testing it, rendering or restarting, and committing to Git."
flowchart TD
    code("Write code<br>in Quarto") --> as_function("Convert to<br>function")
    as_function --> test("Test function in<br>Quarto or Console")
    test --> fix_function("Fix function")
    test -- "Commit" --> git("Git history")
    fix_function --> test
    fix_function --> either([Either:])
    either -. As needed .-> restart("Restart R<br>session")
    either -. As needed .-> render("Render<br>Quarto")
    restart -.-> test
    render -.-> test
```

Either restarting R or rendering the Quarto document is the only way
there is to be certain the R workspace is in a clean state. When code
runs after a clean state, it improves the chances that your code and
project will be reproducible.

We use Git because it is the best way of keeping track of what was done
to your files, when, and why. It helps to keep your work transparent and
makes it easier for you to share your code by uploading to GitHub. Using
version control should be a standard practice to doing better science
since it fits with the philosophy of doing science (e.g., transparency,
reproducibility, and documentation).

{{< text_snippet sticky_up >}}

## Making a function for importing our data

Now that we have a basic understanding of what a function looks like,
let's apply that to what we're doing right now: Importing our data.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Emphasize that we will be using this workflow for creating functions all
the time throughout workshop and is also a common workflow when making
functions in R.
:::

While you read about the general workflow above, the more detailed steps
for making a function is to:

1. Write code that works and does what you want.
2. Enclose it as a function with `name <- function() { ... }`, with an
   appropriate and descriptive name.
3. Create arguments in the function call
   (`function(argument1, argument2)`) with appropriate and descriptive
   names, then replace the code in the function body with the argument
   names where appropriate.
4. Rename any objects created to be more generic and include the
   `return()` function at the end to indicate what the function will
   output.
5. Run the function and check that it works.
6. Add the Roxygen documentation tags with {{< var keybind.roxygen >}}
   while the cursor is in the function.

In `docs/learning.qmd`, create a new Markdown header called
`## Import 5C's HR data with a function` and create a code chunk below
that with {{< var keybind.chunk >}}.

So, step one. Let's copy and paste the code we previously for importing
the `HR.csv.gz` data from participant `5C` and convert that as a
function:

```{r}
#| label: function-creation-step-1
#| filename: "docs/learning.qmd"
#| output: false

here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100
  )
```

Next we wrap it in the function call and give it an appropriate name.
Since we've used this code to read in different types of data, the name
should reflect that it is fairly generic, at least for our data. So we
could simply call it `read`.

::: callout-caution
We want to be careful using very generic names for functions, as these
are often already used by R or in other packages. We can always check if
a name is used by R by typing `?function_name` in the Console. If it
shows you documentation, then it is used and it's best to use another
name for our function.
:::

```{r}
#| label: function-creation-step-2
#| filename: "docs/learning.qmd"
#| eval: false

read <- function() {
  here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100
    )
}
```

Then, we add arguments in the function and replace within the code. We
want to use the `read()` function to read a file path that has data in
it. So we can add an argument to our function for the file path to read
from. So, a good name might be `file_path`.

It's rarely good to hard code things into functions, like the use of
`here()`, otherwise the function will always only read from that file.
Instead, it's good design to give functions a full file path that it can
use internally. Then when we use the function, we would use `here()`
with the correct path in the function argument.

So we would replace `here("...")` with `file_path`:

```{r}
#| label: function-creation-step-3
#| filename: "docs/learning.qmd"
#| eval: false

read <- function(file_path) {
  file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100
    )
}
```

While it isn't required, for learning purposes we will use the
`return()` function to explicitly indicate what the function outputs at
the end. In our case, we want it to output the data frame that we just
read in. Since we haven't assigned the output of the `read_csv()` to an
object, let's do that first and than use `return()` to output that
object at the end. Let's keep it generic again and call it `data`:

```{r}
#| label: function-creation-step-4
#| filename: "docs/learning.qmd"

read <- function(file_path) {
  data <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100
    )
  return(data)
}
```

Great! Now we need to test it out. Let's try it on two `HR.csv.gz`
datasets from `5C`:

```{r}
#| label: function-creation-step-5
#| filename: "docs/learning.qmd"

here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
  read()
here("data-raw/nurses-stress/stress/5C/5C_1586886712/HR.csv.gz") |>
  read()
```

Awesome! It works :tada: The final stage is to add the Roxygen
documentation.

```{r}
#| label: function-creation-step-6
#| filename: "docs/learning.qmd"

#' Read in one nurses' stress data file.
#'
#' @param file_path Path to the data file.
#'
#' @returns Outputs a data frame/tibble.
#'
read <- function(file_path) {
  data <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = 100
    )
  return(data)
}
```

A massive advantage of using functions is that if you want to make a
change to what your code does, like if you fix a mistake, you can very
easily do it by modifying the function and it will change all your other
code too.

Now that we have a working function, run `{styler}` with
{{< var keybind.style >}}, render with {{< var keybind.render >}}, and
then add and commit the changes to the Git history with
{{< var keybind.commit >}} before then pushing to GitHub.

## :book: Reading task: Principles to make general-purpose and reusable functions

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Briefly reinforce what they read by slowly going through these points
below about making generalised functions. Emphasise the principles
below, especially the "do one thing" and "keep it small".
:::

**Time: \~3 minutes.**

There are a few ways to make a function more general-purpose and
reusable, while still being useful for your specific purposes. These
principles are:

1. Have the function's input (the arguments) be generic enough to take
   different types of objects as long as they are the same "type". In
   our case, we have a function that takes a path "type". A path is a
   very general input, so we can keep it as is.
2. Have the function's output be a common "type", like a vector or data
   frame. When working in R, it's a very good practice to have functions
   output a data frame, since many functions, especially in the
   `{tidyverse}`, take a data frame as input.
3. Make the first argument of the function be something that can be
   "pipe-able". That way you can chain together your functions with the
   `|>` operator. In general, if the function does some action to a data
   frame, then the data frame should be the first argument called `data`
   so that it works well with piping from `{tidyverse}` functions.
4. Make your function do one conceptual thing well. For example, read
   data from a file and make it cleaner. Or convert all columns that are
   characters into numbers.
5. Keep the function small. It is easier to be reused, easier to test,
   and easier to debug when it has fewer lines of code.

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Add another function argument to read different number of rows

**Time: \~10 minutes.**

We've converted the code to read in different data files from the
nurses' stress dataset into a function. Now, let's modify it to be a bit
more general-purpose so that we can also test out reading in different
number of rows so we don't have to keep changing the `n_max` value
inside of the function.

Use the `read()` function code we wrote above and revise it so it can be
used like the below:

```{r}
#| label: exercise-read-max-rows
#| eval: false

# Outputs only 100 rows
here("data-raw/nurses-stress/stress/5C/5C_1586886712/HR.csv.gz") |>
  read()

# Outputs only 1000 rows
here("data-raw/nurses-stress/stress/5C/5C_1586886712/HR.csv.gz") |>
  read(max_rows = 1000)

# Outputs all rows (`Inf` means infinity)
here("data-raw/nurses-stress/stress/5C/5C_1586886712/HR.csv.gz") |>
  read(max_rows = Inf)
```

1. In the `read()` function, add a new argument called `max_rows` after
   the `file_path` argument inside `function()`. We'll called it
   `max_rows` because that's a more descriptive name for it that
   `n_max`.
2. Inside the function body, replace the `n_max = 100` with
   `n_max = max_rows` so that the function will use the value given in
   the `max_rows` argument when it is used.
3. Since we'd like this argument to be optional, meaning that we don't
   have to use it (like the first code example above), we have to give
   the argument a *default value* in `function()`. Since we are using
   `100` already, let's use that value by putting `max_rows = 100` in
   `function()`.
4. Copy and paste the example code above to below the function code,
   then run the function code again along with the pasted example code.
   You should see that the first code example outputs 100 rows, the
   second outputs 1000 rows, and the third outputs all rows.
5. Update the Roxygen documentation to include the new `max_rows`
   argument and its description. You can copy and paste the existing
   line with the `@param` for `file_path` and then modify it for
   `max_rows`.
6. Run `{styler}` while in the `docs/learning.qmd` file with
   {{< var keybind.style >}}.
7. Render the Quarto document with {{< var keybind.render >}}.
8. Finally, **add and commit** the changes to the Git history, using
   {{< var keybind.commit >}}. Then push to GitHub.

```{r}
#| label: solution-read-max-rows
#| echo: false

#' Read in one nurses' stress data file.
#'
#' @param file_path Path to the data file.
#' @param max_rows Maximum number of rows to read.
#'
#' @returns Outputs a data frame/tibble.
#'
read <- function(file_path, max_rows = 100) {
  data <- file_path |>
    read_csv(
      show_col_types = FALSE,
      name_repair = to_snake_case,
      n_max = max_rows
    )
  return(data)
}
```

{{< text_snippet sticky_up >}}

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Go over this solution with everyone as we will use it in the next
session.
:::

## :speech_balloon: Discussion activity: What are some tasks that could be functions?

**Time: \~6 minutes.**

As we prepare for the next session and the break, get up, walk around,
and discuss with your neighbour some of the following questions:

- What are some tasks you do that are repetitive or that you do multiple
  times with very small changes each time you do the task?
- How might you use functions in your work? Can you think of specific
  tasks or situations where you could use one?

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- Everything in R is an object.
- Every action in R is a function and every function is an object.
- Functions contain a sequence of steps that do actions to an object.
- Functions have five components:
  - The three required ones are using `function() { }`, the code in the
    function body between the `{ }`, and an output (usually set with
    `return()`).
  - The two optional ones are assigning the function as a new object
    with `<-` and the function arguments put within `function()`.
- Document functions by using Roxygen.
- Keep functions small and simple, so it is easier to test and fix them.
- Use few arguments in functions to reduce cognitive load.

```{r}
#| label: admin-store-session-code
#| include: false

save.image(here::here("_temp/functions.RData"))
```

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
