# Hands-on exercises {#hands-on-work}

::: callout-warning
We're still testing to see what best works for this section.
:::

This session is all about reinforcing the skills you've learned in this
course by continuing to use them. You can either continue working on the
MMASH dataset and complete the exercises below or you could try to apply
these skills to another dataset (such as your own). If you continue with
the MMASH, we strongly encourage you to work together with your group to
ask questions and to help each other out in understanding what to do and
what to code (we are also here for help).

## Quick recap of workflow {#sec-general-workflow}

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

You can go over this point verbally, reiterating what they've learned so
far.
:::

You now have some skills and tools to allow you to reproducibly import,
process, clean, join, and eventually analyze your datasets. Listed below
are the general workflows we've covered and that you can use as a
guideline to complete the following (optional) exercises and group work.

-   Import with the `read_csv()`, and optionally use `spec()` to
    `read_csv()` again to select specific variables.
-   Convert importing into a function in a Quarto document, move to the
    `R/function.R` script, restarting R, and `source()`.
-   Test that joining datasets into a final form works properly while in
    a Quarto document, then cut and paste the code into a data
    processing R script in the `data-raw/` folder (alternatively, you
    can directly write and test code while in the `data-raw/` R script).
-   Restart R and generate the `.rda` dataset in the `data/` folder by
    sourcing the `data-raw/` R script.
-   Restart R, load the new dataset with `load()` and put the loading
    code into a Quarto document.
-   Add any additional cleaning code to the data processing R script in
    `data-raw/` and update the `.rda` dataset in `data/` whenever you
    encounter problems in the dataset.
-   Write R in code chunks in a Quarto document to further analyze your
    data and check reproducibility by often rendering to HTML.
    -   Part of this workflow is to also write R code to output in a way
        that looks nice in the HTML (or Word) formats by mostly creating
        tables or figures of the R output.
-   Use Git often by adding and committing into the history so you never
    lose stuff and can keep track of changes to your files.

## Import and process the activity data

We have a few other datasets that we could join together, but would
likely require more processing in order to appropriately join with the
other datasets. Complete these tasks in the `doc/learning.qmd` file:

1.  Create a new header called `## Exercise: Importing activity data`
2.  Create a new code chunk below this new header
    {{< var keybind.chunk >}}.
3.  Starting the workflow from the beginning (e.g. with the `spec()`
    process), write code that imports the `Activity.csv` data into R.
4.  Convert this code into a new function using the workflow you've used
    from this course:
    -   Call the new function `import_activity`.
    -   Include one argument called `file_path`.
    -   Test that it works.
    -   Add Roxygen documentation with {{< var keybind.roxygen >}} and
        explicit package links (`::`) with the functions.
    -   Move the newly created function into the `R/functions.R`.
    -   Use the new function in `doc/learning.qmd` and use `source()`
        {{< var keybind.source >}} to run it.
5.  Import all the `user_` datasets with `import_multiple_files()` and
    the `import_activity()` function.
6.  Pipe the results into `mutate()` and create a new column called
    `activity_seconds` that is based on subtracting `end` and `start`.
    -   Use `?mutate` and check the examples in the help document that
        pops up if you don't recall how to use this function.
7.  You'll notice that the `activity` column is numeric. Look into the
    [Data Description](https://physionet.org/content/mmash/1.0.0/) and
    find out what each column represents and what the numbers mean in
    the column `activity`. Then think about or complete these tasks:
    -   What is the advantage and disadvantage of using numbers instead
        of text to describe categorical data like in the `activity`
        column?
    -   Using the `case_when()` function (within `mutate()`) we learned
        about, convert the `activity` numbers into more meaningful
        character data.
8.  Render the Quarto document with {{< var keybind.render >}} to see if
    it is reproducible.
9.  Run styler with {{< var keybind.styler >}}.
10. **Add and commit** your changes to the Git history with
    {{< var keybind.git >}}.

```{r solution-import-activity}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
import_activity <- function(file_path) {
  activity_data <- readr::read_csv(
    file_path,
    col_select = -1,
    col_types = readr::cols(
      activity = readr::col_double(),
      start = readr::col_time(format = ""),
      end = readr::col_time(format = ""),
      day = readr::col_double()
    ),
    name_repair = snakecase::to_snake_case
  )
  return(activity_data)
}

activity_df <- import_multiple_files("Activity.csv", import_activity)

activity_df |>
  mutate(activity_duration = end - start)
```

## Fix up some remaining data issues

There are some issues with the data still. Try to work through these
tasks to clean the data up more.

1.  Use the code below to examine the cleaned data for user 21. Notice
    some problems? Go into the raw data as well as the data
    documentation on the [MMASH
    website](https://physionet.org/content/mmash/1.0.0/) and try to
    figure how what happened. How you might fix the problem? Try your
    fix within the `data-raw/mmash.R` cleaning script.

    ```{r}
    #| filename: "data-raw/mmash.R"
    #| eval: false
    mmash |>
      filter(user_id == "user_21")
    ```

    -   **Hint**: Look into user 21's folder at their `.csv` files they
        have. Compare to the other users. Are the same number of files
        there?
    -   **Hint**: In the `reduce()` and `full_join()` code, try
        re-arranging the order that the datasets are listed. Does order
        matter?

2.  If you do `count(mmash, day)` in the Console while the `mmash`
    dataset is loaded, you'll see that the `day` column has a weird
    value of `-29`. Use `filter()` to see which rows and users are
    affected. Look into the data documentation on the [MMASH
    website](https://physionet.org/content/mmash/1.0.0/) and see if you
    can find an explanation for this. How would you go about solving
    this issue? Write up code to fix this issue in the
    `data-raw/mmash.R` script.

    -   **Hint**: Find the users that have this value and look into
        their data `.csv` files that have `day` (or `Day`) variables.
        Which files have this `-29` in the `day` column? Where does the
        `-29` value start in that data file? Can you guess what the
        issue was and what `-29` is or should be?

## Process and join sleep and questionnaire data

There are still a few datasets that you can join in with the current
datasets like `sleep.csv` and `questionnaire.csv`. Using the workflows
in @sec-general-workflow as a guide, start from the beginning and
import, process, clean, make functions, and join these two datasets in
with the others so that they get included in the `data/mmash.rda` final
dataset. Afterwards, do some descriptive analysis using the function
`tidy_summarise_by_day()`.

::: {.callout-tip appearance="default"}
User 11 has no sleep data, so you will eventually have to drop user 11
from the dataset before summarizing the data.
:::

## Create a second dataset of only the Actigraph and RR data

The Actigraph and RR datasets contain a lot of interesting and useful
data that gets destroyed when we first summarise and then join them with
the other datasets. While we can't meaningfully join all this data with
the other datasets, we can join them on their own as separate datasets.

Using the workflows in @sec-general-workflow as a guide, start from the
beginning and import, process, clean, make functions, and create a final
dataset of only the `Actigraph.csv` and `RR.csv` datasets.

-   Join only these two datasets by `user_id`, `day`, and `time`.
-   Name the new dataset `actigraph_rr` and save it to `data/` by using
    another `usethis::use_data()` line in the `data-raw/mmash.R` script.

Than think of how some of these questions might be answered, using the
[Data Description](https://physionet.org/content/mmash/1.0.0/) as a
guide:

1.  When are people most likely to be lying down (`inclinometer_lying`)?
    How long do they lie down for?
2.  When are people most likely to be sitting down
    (`inclinometer_sitting`)? Does this also correspond to a lower heart
    rate or a lower interbeat interval (from the `RR` data) compared to
    when standing?
3.  Are people who have more activity throughout the day also reporting
    better sleep quality (like `number_of_awakenings`)?
4.  Does the self-reported activity match the accelerometry data?

## Other datasets to try out

If you have completed the exercises above and still wanted to practice
the skills, here are some other datasets you can use (aside from your
own):

-   [MIMIC-IV demo data in the OMOP Common Data
    Model](https://physionet.org/content/mimic-iv-demo-omop/0.9/)
-   [RR interval time series from healthy
    subjects](https://physionet.org/content/rr-interval-healthy-subjects/1.0.0/)
-   [Data on the 2021 Olympics in
    Tokyo](https://www.kaggle.com/arjunprasadsarkhel/2021-olympics-in-tokyo)
-   Find your own dataset from
    [PhysioNet](https://physionet.org/about/database/#open) or
    [Zenodo](https://zenodo.org/search?page=1&size=20&q=diabetes&access_right=open&file_type=xlsx&file_type=csv&file_type=zip&type=dataset#).
-   Work on the game ["Murder
    Mystery"](https://r-murder-mystery.askeland.me/) (which
    Anders Askeland converted into a game using R).
