# Pre-process data as you import it {#sec-importing}

```{r}
#| label: setup
#| include: false

library(readr)
library(here)
library(snakecase)
```

Before you can do anything to data in R, you need to actually get it
into R in the first place! This session is all about reading in one data
file and doing some very simple cleaning to it as we read it in. Keeping
in mind, there are nearly 3000 CSV files to read in, so aiming to do
things efficiently is very important here.

## Learning objectives

{{< include /includes/objectives/_importing.qmd >}}

## :book: Reading task: Use Quarto as a sandbox to prototype code

::: {.callout-note collapse="true"}
## :teacher: Teacher note

After they've read this, reinforce this workflow and highlight it's
strengths. Draw upon your own personal experience if you are able to.
And emphasize it also simplifies the teaching-learning role as there is
only two files we all have to keep track of: the R functions file that
we will cover later and the Quarto file.
:::

**Time: \~5 minutes.**

Quarto is an incredibly powerful tool when working to ensure your code
is reproducible. For this workshop, the Quarto document we use is at
`docs/learning.qmd`.

With Quarto, it's easy to quickly test code out and make sure it works.
Think of using the Quarto file as a sandbox to test out and play with
code, without fear of messing things up. And with Quarto, when you
render the document to, for example, HTML it runs all the code from
beginning to end in a new environment (so completely clean). Which means
you can test a part of the reproducibility of your code. Because of this
feature, it allows you to detect errors and issues with the code, making
Quarto an excellent prototyping tool. The workflow is represented below
in @fig-quarto-prototyping.

```{mermaid}
%%| echo: false
%%| label: "fig-quarto-prototyping"
%%| fig-cap: "Workflow for using Quarto to prototype code, assess reproducibility, and fix issues."
%%| fig-alt: "A diagram representing the workflow for using Quarto to prototype code, with boxes for the Quarto file, rendering, and HTML output."
flowchart TD
  quarto[/"Quarto:<br>docs/learning.qmd"/] --> prototype(Prototyping<br>R code)
  prototype -- "Check<br>reproducibility" --> render("Render")
  render --> html[/"HTML file:<br>docs/learning.html"/]
  render -- "Continue<br>or fix issues" --> quarto
```

For this workshop, we also use Quarto because it's easier from a
teaching perspective. You, as the learner, can easily mix together text
and notes to yourself as you learn, alongside the code you write out
during the code-alongs and the exercises. We also use it to force you to
practice working within Quarto documents, which is a key component to
reproducibility, in a learning and supportive environment.

::: callout-tip
Since you can mix text and code together, you can also write your notes
for the workshop in the Quarto document! :grin: So if you want to come
back and look at your notes later, this would be the place to put them.
:::

{{< text_snippet sticky_up >}}

## :book: Reading task: Packages for importing specific data file formats

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Mainly emphasize here that there are so many different file formats and
also different packages to import them into R. It's hard to teach all
the different ways to import data as it will be likely different for
different fields of research, research groups, or institutions. But we
will be focusing on CSV since it is one of the most common file formats.
:::

**Time: \~5 minutes.**

Before you can effectively analyze your project's data, you will very
likely need to first clean and process it into a format and structure
that fits your project's specific needs. So this first phase of an
analysis project is to import the data into R, check it out, and clean
and process it. Eventually, you want to have a version of the raw (or
original) data saved in `data/` ready for your analysis.

There are many different ways to import data into R, and it all depends
on what file format the data is stored in. For most file formats, there
is usually a specific R package designed to import it. For instance, the
below is a list of packages to load common file formats:

- `{haven}`: For reading (also known as importing or loading) in SAS,
  SPSS, and Stata files.
- `{readxl}`: For reading in Excel spreadsheets with `.xls` or `.xlsx`
  file endings.
- `{googlesheets4}`: For reading in Google Sheets from their cloud
  service.
- `utils::read.delim()`: This function comes from the core R package
  `{utils}` and includes other functions like `utils::read.csv()`.
- `data.table::fread()`: From the `{data.table}` package, this function
  is used to load in CSV files.
- `arrow::read_parquet()`: This function is used to read in Parquet
  files, which are an increasingly common file format for storing large
  datasets. This is also becoming more common in data analysis work
  because it can easily connect the imported Parquet data into the
  powerful [DuckDB](https://duckdb.org/) engine through
  [duckplyr](https://duckplyr.tidyverse.org/index.html). We won't cover
  this during the workshop, but like Parquet, it is also becoming
  increasingly common to use when working with large datasets. The
  advantage of using
  [duckplyr](https://duckplyr.tidyverse.org/index.html) is that it uses
  the same syntax as `{dplyr}` so it is very easy to switch between
  using it and using `{dplyr}`.
- `{readr}`: A package that is part of the `{tidyverse}` that is
  designed to load in data stored in text-based files such as CSV with
  `readr::read_csv()` and is incredibly fast at importing data. Notice
  the `_` in the function name, compared to the `.` in the
  `utils::read.csv()` function name above. This small difference can
  confuse people, and it is often the cause of the first errors we have
  when teaching this workshop.

The most common type of file format for data that you will likely
encounter is a CSV file (or Excel), as spreadsheet-style structured data
is often the most common way to record and store data. CSVs are also
open and transparent since they are simply text files that are both
human-readable and machine-readable. You don't need special software to
open them, unlike, for example, needing Excel to open Excel
spreadsheets. Because CSV is so common, we will be using
`readr::read_csv()` and won't cover the other packages listed above. We
list these other packages for your reference and information only.

We're going to use the `{readr}` package because it is well integrated
within the `{tidyverse}` universe of packages. It's also very fast at
loading in data, which for large datasets can be one of the more
time-consuming parts of starting an analysis.

The packages `{readr}`, `{vroom}`, `{haven}`, `{readxl}`, and
`{googlesheets4}` all are very similar in how you use them and their
documentation are almost identical. So the skills you learn in this
workshop with `{readr}` can mostly be applied to these other packages.
And because `{readr}` (which the other packages are based on) has been
around for awhile, there is a large amount of support and help for using
it.

If your data is in CSV format, `{readr}` is perfect.

{{< text_snippet sticky_up >}}

## Importing in the raw data

Alright, now is the time to importing some data! Open up the
`docs/learning.qmd`. A quick way of opening a file in RStudio is to type
{{< var keybind.open-file >}} (dot or period) and then typing `learning`
which should bring up the file we want to open. Hitting enter should now
open the file.

Now with `docs/learning.qmd` open, we should see a `setup` code chunk
already in the file. This code chunk is where we will put the
`library()` code for loading the `{tidyverse}` package, which has
`{readr}` bundled with it, as well as `library(here)`. It should look
like this:

````markdown
```{{r setup}}
library(tidyverse)
library(here)
```
````

This `setup` code chunk is a special, named code chunk. Whenever you
first open a Quarto file or after you restart your R session and then
start running R code, RStudio will first look through the Quarto file
for any code chunk named `setup`. When it finds one, it will run the R
code in that chunk first.

So it's good practice to put `library()` functions as well as functions
that load your data in the `setup` code chunk. For this session, we will
only put the `library()` functions there.

After adding the code to this chunk, run it with
{{< var keybind.run-code >}} so that the packages get loaded. Then,
create a new header below the `setup` code chunk by typing out
`## Importing raw data`, followed by creating a new code chunk right
below it using {{< var keybind.chunk >}}.

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Very briefly explain what `here()` is doing and why it's important to
use it when working with projects. Specifically, that it ensures
reproducibility and that the project is self-contained.
:::

The first argument to the `read_csv()` function is the path to the file
you want to import. We use the `here()` function to help us with this.
The `here()` function is a function that helps us find the path to the
file we want to import, relative to the R Project file. Any time we want
a file path, we use `here()` to ensure our project is reproducible and
self-contained.

::: callout-tip
We covered the `here()` function in more detail in the
[introductory](https://r-cubed-intro.rostools.org/sessions/import-data.html)
workshop, so won't go over it in much detail here.
:::

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Emphasize and take your time explaining the `|>` pipe operator,
describing the output of the function goes into the first argument of
the next function.
:::

We'll use the `|>` pipe operator to pipe the output of `here()` into
`read_csv()`. The pipe operator is a way to give the output of one
function as the input to the next function. We will use this a lot
throughout the workshop.

::: callout-note
You may have seen the `{magrittr}` pipe `%>%` before. We are using the
base `|>` R pipe operator instead because more documentation and
packages are using or relying on it. In terms of functionality, they are
nearly the same, with some small differences. It ultimately doesn't
matter which one you use, but we're using the base R `|>` pipe to be
consistent with other documentation and with the general trend to
recommend it over the `{magrittr}` pipe.
:::

There's a lot of data we could import, so let's start with importing one
data file that contains the heart rate data in `HR.csv.gz` for
participant `5C` in the folder `5C_1586886626/`. As we mentioned in the
introduction session, the data is stored in the gzipped format to save
space and the folder naming scheme is `<id>_<Epoch timestamp>`. [Epoch
timestamp](https://en.wikipedia.org/wiki/Unix_time) is a way to
represent time as a number of seconds since January 1, 1970 and is a
common, though imperfect, way to store timestamps in many systems.

```{r}
#| label: first-load-cgm-101
#| filename: "docs/learning.qmd"

here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
  read_csv()
```

Run this code with {{< var keybind.run-code >}}. Before looking at the
data, let's look through the message it creates. The columns data types
(whether it is a character, a date, or a number) inside of a CSV file is
not clear since CSV is plain text without metadata attached. So when you
load data from a CSV file into R with `read_csv()`, the function will
try to guess what the data types are for each column. Usually it guesses
accurately, but it isn't perfect. So this message is telling us what it
guessed the data types of the columns are.

To remove this message, there are two things we can do:

1. Explicitly tell `read_csv()` what the column names and data types are
   using the argument `col_types`. This is a very tedious, but often
   very powerful and useful step depending on the issues your data may
   have.
2. Tell `read_csv()` to just guess and to hide the message using
   `show_col_types = FALSE`. Usually, `read_csv()` does a good job of
   guessing, so for this workshop we will use this option.

So in the code, we will add `show_col_types = FALSE` to `read_csv()` and
then run the code:

```{r}
#| label: quiet-read-message-cgm-101
#| filename: "docs/learning.qmd"

here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
  read_csv(show_col_types = FALSE)
```

When we run it, we no longer see the message. Great! Now let's focus on
looking at the data itself.

One thing to notice is that the column names are in a format that, while
usable, are not ideal. For example, the first column is called
`CollectionDatetime`, which is in the
[`PascalCase`](https://en.wikipedia.org/wiki/Camel_case) style (which
itself is a version of `camelCase`). For human readability, some type of
space between words makes it easier to read and `PascalCase` has no
spaces between words. So, we will need to fix the column names up and
use a common style that is easier to read. That style is called
`snake_case` and it uses `_` for spaces and is all lower case. Snake
case is recommended in the [tidyverse style
guide](https://style.tidyverse.org/) and there's even a package called
`{snakecase}` that can help us with this!

But how do we fix this as we import the data? Let's look at the help
docs of `{readr}`: `?read_csv`. Scroll down in the help document and we
will see an argument called `name_repair` that handles naming of
columns. There are several options here, none of which will fix to
`snake_case`. But there is one that says "function: apply custom name
repair". Since there is already a `{snakecase}` package with some
functions to help with converting to snake case, we will use this option
to give the `name_repair` the function from the `{snakecase}`.

Since we will make use of the `{snakecase}` package, let's load it with
`library()`. We'll put it into our `setup` code chunk, so go to the top
of your `docs/learning.qmd` file and add it there:

````markdown
```{{r setup}}
library(tidyverse)
library(here)
library(snakecase)
```
````

Now we can run this code with {{< var keybind.run-code >}}. This loads
the `{snakecase}` package and makes it available for us to use in our
code. Let's go back to the bottom of our `docs/learning.qmd` file to
continue working on our importing code.

The function from `{snakecase}` that we want to use in the `name_repair`
argument is the `to_snake_case()` function. To convert the variable
names to snake case, we would add `name_repair = to_snake_case` to the
code. Notice the lack of `()` when using the `to_snake_case()` function.
We'll explain more about this in later sessions, so ignore it for now.

```{r}
#| label: use-name-repair
#| filename: "docs/learning.qmd"

here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case
  )
```

Looking at the data, we can see that the column names have been renamed
to `snake_case`. Awesome, that was an easy fix so far!

## Import less data while prototyping code

As you write code, you'll probably be continually reading in the data as
you test the code out. If the data is quite large, reading it into R can
sometimes take a few seconds. So if it is slow to read in, it will make
it harder for you to smoothly and quickly write code, try things out,
and fix things. Not just harder from a technical point of view, but also
from a motivation and cognitive load perspective. If running code takes
a few seconds, you will be less likely to *want* to test things out, fix
things, and get things right. There will be that cognitive inertia and
resistance.

Not only is it slow to read in a large dataset, but it is also slower to
process and wrangle it. This is especially true when:

- you first read the data into R, as larger data will be slower to load
  into R than smaller data.
- your work computer is not very powerful or has other programs
  (installed or setup by your IT) that slow your computer down.

In both those cases, if you read in less data, it will mean that running
the code will be faster. You'll also have less stress and frustration
because you will be able to get faster feedback on whether your code
works or not. So a good practice as you write and test code is to only
import some of the data. Then when you are sure it works as you want it
to and when you are ready to do a full analysis, you can import all the
data and run the code with the full data.

So, until the end of the workshop, we will only import fewer rows of
data than the whole dataset. `read_csv()` has an argument called `n_max`
that allows you to specify how many rows to import, which we can read
about by looking at the `?read_csv` help documentation.

For the data we are importing right now, let's add this `n_max` argument
to only load in the first 100 rows. It can be less or more, but it's
good to get enough data that `read_csv()` can effectively guess the
columns data types but not so large as to slow things down. In the code
we've already written, add `n_max = 100` to the `read_csv()` function:

```{r}
#| label: read-fewer-rows-prototyping
#| filename: "docs/learning.qmd"

here("data-raw/nurses-stress/stress/5C/5C_1586886626/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100
  )
```

In this particular case, this data file doesn't have a lot of data, so
using `n_max` doesn't help at all here. But, most of the other data
files in the nurses' stress dataset are much larger. In the exercise
below, you'll try reading in larger data files with and without `n_max`.
But before moving on to the exercise, let's:

- Tidy up the code using `{styler}` with {{< var keybind.style >}}
- Make sure everything is reproducible in the `docs/learning.qmd` file
  by rendering the Quarto document to an an HTML file. Click the
  "Render" button at the top of the Source pane or by typing
  {{< var keybind.render >}}. If it generates an HTML document without
  problems, we know our code is at least starting to be reproducible.
- **Add and commit** the changes to the Git history through the RStudio
  Git interface with {{< var keybind.commit >}}. Then click the push
  button while in the interface to push the changes to GitHub.

## :technologist: Exercise: Read in other data files {#sec-exercise-read-other-data}

**Time: \~15 minutes.**

An important part of learning is trying what we've just covered on your
own and in a practical way. So, using the same steps we just followed,
read in a few other data files. Start with reading in without using
`n_max`. Then, read in the same data file but with `n_max` to see the
difference. Use the code template below to copy and paste into your
`docs/learning.qmd` file to help you get started:

```r
here("___") |>
  ___(
    ___ = ___,
    ___ = ___
  )
```

1. Create a new header at the bottom of the `docs/learning.qmd` file and
   call it `## Exercise: Reading other files`.
2. Below the header, create a new code chunk with
   {{< var keybind.chunk >}}.
3. Copy and paste the code template above into the new code chunk. Begin
   replacing the `___` with the correct R functions or other
   information.
4. Start by reading in another `HR.csv.gz` file from the same `5C`
   participant, but instead use the folder `5C_1586886712/`.
   - Put the path to the `HR.csv.gz` file in the `here()` function.
   - Use `read_csv()` (not `read.csv()`).
   - Use `show_col_types = FALSE` to hide the message.
   - Convert the columns `to_snake_case` using the `name_repair`
     argument in `read_csv()`. Remember to not use the `()` at the end
     of the function.
   - Run the code and check out the data.
5. Do the same thing by copying the template again and reading in the
   same `HR.csv.gz` file from `5C`, but this time use `n_max = 100`.
6. Now try a different participant. Read in the `HR.csv.gz` file for
   participant `6B` in the folder `6B_1587569373/` with and without
   `n_max = 100`.
7. This time, read in the `EDA.csv.gz` file for participant `6B` in the
   folder `6B_1587569373/` with and without `n_max = 100`.
8. Once you have the code working, run `{styler}` with
   {{< var keybind.style >}}, render the Quarto document to HTML to
   check reproducibility using {{< var keybind.render >}}, and then use
   the RStudio Git interface to **add and commit** the changes to the
   Git history using {{< var keybind.commit >}}. Then, push to GitHub.

```{r}
#| label: solution-read-other-data
#| echo: false
#| output: false

here("data-raw/nurses-stress/stress/5C/5C_1586886712/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case
  )

here("data-raw/nurses-stress/stress/5C/5C_1586886712/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100
  )

here("data-raw/nurses-stress/stress/6B/6B_1587569373/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case
  )

here("data-raw/nurses-stress/stress/6B/6B_1587569373/HR.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100
  )

here("data-raw/nurses-stress/stress/6B/6B_1587569373/EDA.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case
  )

here("data-raw/nurses-stress/stress/6B/6B_1587569373/EDA.csv.gz") |>
  read_csv(
    show_col_types = FALSE,
    name_repair = to_snake_case,
    n_max = 100
  )
```

::: callout-note
You might notice something annoying already... this code is quite
repetitive. If you wanted to read in more data files, you can imagine
that this becomes quite tedious, not to mention error-prone and
difficult to change (if you decided to change how you've read the data).
Next session we will learn how to "not repeat ourselves".
:::

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- Use the appropriate package when importing a specific data file (e.g.
  for `.xlsx` use `{readxl}`).
- Use `read_csv()` to read plain text type files (e.g. csv) fast.
- Fix column names to match a style using `name_repair =` argument in
  `read_csv()`, such as as using `to_snake_case()` to convert to
  `snake_case`.
- For very large datasets, only read the first 100-1000 rows using
  `n_max =` to quickly iterate and test code before reading the full
  dataset.

{{< include /includes/_code-appendix.qmd >}}
