# Pre-process data as you import it {#sec-importing}

```{r setup}
#| include: false
library(readr)
library(here)
temp_path <- fs::path_temp()
project_path <- fs::path(temp_path, "LearnR3")

untar(
  tarfile = here::here("_temp", "pre-workshop.tar.gz"),
  exdir = project_path
)

usethis::with_project(
  project_path,
  {
    fs::file_delete("TODO.md")
    gert::git_add(".")
    gert::git_commit("Added initial files")
  }
)
```

## Learning objectives

{{< include /includes/_learning-objectives-importing.qmd >}}

## Importing in the raw data

The ultimate goal for the beginning phases of a data analysis project is
to eventually save a version of the raw data that is specific to your
research questions and enables you to conduct your analyses. The first
step to processing data is to import it into R so we can work on it. So
for now, we'll open up the `doc/learning.qmd` file so we can start
building and testing out the code. There should be a `setup` code chunk
already be in the file, where we will put the `library()` code for
loading the `{tidyverse}` package, which has `{readr}` bundled with it,
as well as `library(here)`. It should look like this:

```{{r setup}}
library(tidyverse)
library(here)
```

This `setup` code chunk is a special, named code chunk that tells R to
run this code chunk first whenever you open this Quarto file and run
code inside of the file. It's in this `setup` code chunk that we will
add `library()` functions when we want to load other packages. After
adding this code chunk, create a new header below the `setup` code chunk
by typing out `## Importing raw data`, followed by creating a new code
chunk right below it using {{< var keybind.chunk >}}.

::: callout-note
## Reading task: \~5 minutes

What is `{readr}`? It is a package designed to load in data,
specifically text-based data files such as CSV. In R there are several
packages that you can use to load in data and of different types of file
formats. We won't cover these other packages, so you can use this as a
reference for when or if you ever have to load other file types:

-   `{haven}`: For reading (also known as importing or loading) in SAS,
    SPSS, and Stata files.
-   `{readxl}`: For reading in Excel spreadsheets with `.xls` or `.xlsx`
    file endings.
-   `{googlesheets4}`: For reading in Google Sheets from their cloud
    service.
-   `{vroom}`: We originally taught this package, but since then it has
    been merged in with the `{readr}` package, which is why we now teach
    that instead. This package was designed to read in text-based data
    files really fast.
-   `utils::read.delim()`: This function comes from the core R package
    utils and includes other functions like `utils::read.csv()` (note
    the `.` in the name, rather than `_` in the `readr::read_csv()`
    version).
-   `data.table::fread()`: From the `{data.table}` package, this
    function is used to load in CSV files.

We're using the `{readr}` package because it is well integrated within
the `{tidyverse}` universe of packages. It also runs `{vroom}`
internally. How fast is `{vroom}`? The website has a [benchmark
page](https://vroom.r-lib.org/articles/benchmarks.html) showing how fast
it is. For many people, loading in the data can be one of the most
time-consuming parts of starting an analysis. Hopefully by using this
package, that time can be reduced.

The packages `{readr}`, `{vroom}`, `{haven}`, `{readxl}`, and
`{googlesheets4}` all are very similar in how you use them and their
documentation are almost identical. So the skills you learn in this
session with `{readr}` can mostly be applied to these other packages.
And because `{readr}` (which the other packages are based on) has been
around for awhile, there is a large amount of support and help for using
it.

If your data is in CSV format, `{readr}` is perfect. The CSV file format
is a commonly used format for data because it is open and transparent,
readable by any computer, and doesn't depend on any special software to
open (unlike for e.g. Excel spreadsheets).
:::

Let's first start by creating an object that has the file path to the
dataset, then we'll use `read_csv()` to import that dataset.

```{r first-load-in-user-info}
#| filename: "doc/learning.qmd"
user_1_info_file <- here("data-raw/mmash/user_1/user_info.csv")
user_1_info_data <- read_csv(user_1_info_file)
```

You'll see the output mention using `spec()` to use in the argument
`col_types`. And that it has 5 columns, one called `...1`. If we look at
the CSV file though, we see that there are only four columns with names,
but that technically there is a first empty column without a column
header. If you recall, the `user_info.csv` file has an empty column
name. Looking at the [data
dictionary](https://physionet.org/content/mmash/1.0.0/) it doesn't seem
there is any reference to this column, so it likely isn't important.
More than likely, `{readr}` is complaining about this empty column name
and the use of `...1` to represent it. We don't need it, so we'll get
rid of it when we load in the dataset. But how? Look at the help
documentation again. Go to the **Console** and type out:

``` {.r filename="Console"}
?readr::read_csv
```

Looking at the list of arguments, there is an argument called
`col_select` that sounds like we could use that to keep or drop columns.
It says that it is used similar to `dplyr::select()`, which normally is
used with actual column names. Our column doesn't have a name, that's
the problem. Next check the Example section of the help. Scrolling down,
you'll eventually see:

> `read_csv(input_file, col_select = c(1, 3, 11))`

So, it takes numbers! With `dplyr::select()`, using the `-` before the
column name (or number) means to drop the column, so in this case, we
could drop the first column with `col_select = -1`!

```{r second-load-in-info-no-error}
#| filename: "doc/learning.qmd"
user_1_info_data <- read_csv(
  user_1_info_file,
  col_select = -1
)
```

Hmm, we still have the message. You may or may not see this message
depending on the version of your packages (the most updated packages
will show this), but you can see it in the code chunk above.

This is `read_csv()` letting you know that a column was renamed because
it found an empty column with no column name and is letting you know it
had to rename it. Even though we excluded it with `col_select = -1`,
that only excludes it after reading the column names, before importing
the data.

To remove this new message, we need to tell `read_csv()` exactly how to
handle renaming cases. Let's look at the help docs of `{readr}`:
`?read_csv`. Scroll down and we see an argument called `name_repair`
that handles naming of columns. It says that it treats problematic
column names, of which a missing column name is definitely a problem.
There are several options here, but the one I want to focus on is the
comment about "function: apply custom name repair". This is an important
one because we eventually want to rename the columns to match the [style
guide](https://style.tidyverse.org/) by using `snake_case`, since there
is a package to do that called `{snakecase}`.

In the **Console**, type out `snakecase::` and hit Tab. You'll see a
list of possible functions to use. We want to use the snakecase one, so
scroll down and find the `to_snake_case()`. That's the one we want to
use. So to remove the messages and convert the variable names to snake
case, we would add `name_repair = snakecase::to_snake_case` to the code.
Notice the lack of `()` when using the function. We'll explain more
about this in later sessions.

```{r use-name-repair}
#| filename: "doc/learning.qmd"
user_1_info_data <- read_csv(
  user_1_info_file,
  col_select = -1,
  name_repair = snakecase::to_snake_case
)
```

We can now look at the data:

```{r print-user-info-data}
#| filename: "Console"
user_1_info_data
```

Before moving on to the exercise, run `{styler}` with
{{< var keybind.styler >}}, then **add and commit** the changes to the
Git history through the RStudio Git interface with
{{< var keybind.git >}}.

## Exercise: Import the saliva data {#sec-ex-import-saliva}

> Time: 15 minutes.

Practice importing data files by doing the same process with the saliva
data.

1.  Create a new header at the bottom of the `doc/learning.qmd` file and
    call it `## Exercise: Import the saliva data`.
2.  Below the header, create a new code chunk with
    {{< var keybind.chunk >}}.
3.  Copy and paste the code template below into the new code chunk.
    Begin replacing the `___` with the correct R functions or other
    information.
4.  Once you have the code working, run `{styler}` with
    {{< var keybind.styler >}} and then use the RStudio Git interface to
    **add and commit** the changes to the Git history using
    {{< var keybind.git >}}.

``` {.r filename="doc/learning.qmd"}
user_1_saliva_file <- here("data-raw/mmash/user_1/___")

user_1_saliva_data <- read_csv(
    user_1_saliva_file,
    col_select = ___,
    name_repair = ___
)
```

```{r solution-import-the-saliva-data}
#| output: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
user_1_saliva_file <- here("data-raw/mmash/user_1/saliva.csv")

user_1_saliva_data <- read_csv(
  user_1_saliva_file,
  col_select = -1,
  name_repair = snakecase::to_snake_case
)
```

## Importing larger datasets

Sometimes you may have a dataset that's just a bit too large. Sometimes
`{readr}` may not have enough information to guess the data type of the
column. Or maybe there are hundreds or thousands of columns in your data
and you only want to import specific columns. In these cases, we can do
a trick: read in the first few lines of the dataset to prototype your code
to make sure it works. Reading fewer data into R is much faster than
reading in the entire dataset.

Let's do this on the `RR.csv` file. We can see from the file size that
it is bigger than most of the other files (\~2Mb). So, we'll use this
technique to decide what we want to keep. First, create a new header
`## Import larger datasets` and a new code chunk below it using
{{< var keybind.chunk >}}.

Do the same thing that we've been doing, but this time we are going to
use the argument `n_max`, which tells `{readr}` how many rows to read
into R. In this case, let's read in 100, since that is the amount
`{readr}` will guess until. This dataset, like the others, has an empty
column that we will drop.

```{r import-some-of-data-trick}
#| filename: "doc/learning.qmd"
user_1_rr_file <- here("data-raw/mmash/user_1/RR.csv")
user_1_rr_data <- read_csv(
  user_1_rr_file,
  n_max = 100,
  col_select = -1,
  name_repair = snakecase::to_snake_case
)
```

We'll use `n_max` to help us prototype our code, since it is faster to
import only a few rows. We can look at the data:

```{r}
#| filename: "Console"
user_1_rr_data
```

To make sure everything is so far reproducible within the
`doc/learning.qmd` file, we will render the Quarto document to output an
HTML file. Click the "Render" button at the top of the Source pane or by
typing {{< var keybind.render >}}. If it generates an HTML document
without problems, we know our code is at least starting to be
reproducible.

## Exercise: Import the Actigraph data

> Time: 15 minutes.

Practice some more. Do the same thing with the `Actigraph.csv` dataset
as we did for the `RR.csv`. But first:

1.  Create a new header at the bottom of the `doc/learning.qmd` file and
    call it `## Exercise: Import the Actigraph data`.
2.  Below the header, create a new code chunk with
    {{< var keybind.chunk >}}.

Use the *same technique* as we used for the `RR.csv` data and read in
the `Actigraph.csv` file from `user_1/`.

1.  Set the file path to the dataset with `here()`.
2.  Read in a max of 100 rows for `n_max` and exclude the first column
    with `col_select = -1`.
3.  Render the Quarto document with {{< var keybind.render >}} to
    regenerate the HTML document.
4.  If everything works, run `{styler}` with {{< var keybind.styler >}}
    and then **add and commit** the new changes to the Git history with
    {{< var keybind.git >}}

Use this template as a guide for completing this exercise.

``` {.r filename="doc/learning.qmd"}
user_1_actigraph_file <- here("data-raw/mmash/user_1/___")
user_1_actigraph_data <- ___(
    ___,
    n_max = ___,
    col_select = ___,
    name_repair = ___
)
```

```{r solution-import-the-actigraph-data}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
# Use first 100 or so lines to get spec
user_1_actigraph_file <- here("data-raw/mmash/user_1/Actigraph.csv")
user_1_actigraph_data <- read_csv(
  user_1_actigraph_file,
  n_max = 100,
  col_select = -1,
  name_repair = snakecase::to_snake_case
)
```

## :speech_balloon: Discussion activity: Experiences with cleaning data

**Time: \~10 minutes.**

As we prepare for the next session, get up and discuss with you
neighbour some of the following questions:

-   Share a positive or exciting experience you had with cleaning,
    processing, and preparing data for later analysis. It can be recent
    or in the past, but as long as it was a memorable and positive. What
    made it positive? Do you try to repeat that experience, or at least
    the feeling of it?
-   Share a challenging experience or barriers that you encounter while
    cleaning, processing, and preparing data. Why was it challenging or
    had barriers and how could you limit those for for next time?

## Summary

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Quickly cover this before finishing the session and when starting the
next session.
:::

-   Use the appropriate package when importing a specific data file
    (e.g. for `.xlsx` use `{readxl}`).
-   Use `read_csv()` for fast importing of plain text type files (e.g.
    csv).
-   Use `col_select =` arguments in `read_csv()` to import only the data
    you need.
-   Fix column names to match a style using `name_repair =` argument in
    `read_csv()`.
-   For very large datasets, importing in the first 100-1000 rows using
    `n_max =` to test that importing works fine before importing the
    full dataset.

```{r store-session-code}
#| include: false
save.image(here::here("_temp/importing.RData"))
```
