# Joining data together, all at one {#sec-joins}

```{r setup}
#| include: false
library(here)
library(tidyverse)
load(here("_temp/pivots.RData"))
```

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Briefly go over the bigger picture (found in the introduction section)
and remind everyone the 'what' and 'why' of what we are doing, and where
we are at.
:::

## Learning objectives

{{< include /includes/objectives/_joins.qmd >}}

## :speech_balloon: Discussion activity: Recall and share what you've learned so far

**Time: \~6 minutes.**

A very effective way to learn is to recall and describe to someone else
what you've learned. So before we continue this session, take some time
to think about what you've learned in the last two days.

1.  Take 1-2 minute and try to recall as much as you can about what
    you'd done over the last two days. Without looking at your notes or
    the other sections of the website, try to remember things about
    importing, robust functions, functionals, regex,
    split-apply-combine, and pivots.
2.  Then, for 4-5 minutes, share with your neighbour what you remember
    and try to describe it to each other. Maybe you will each remember
    different things.

## :book: Reading task: The many ways to join data

**Time: \~10 minutes.**

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Walk through and describe these images and the different type of joins
after they've read it.
:::

The ability to join datasets together is a fundamental component of data
processing and transformation. In many cases, it is required to do
before you can meaningfully do any work. Like in our case, we have
essentially three datasets that aren't particularly useful on their own
and only really benefit by being joined together. We join them to be
able to make comparisons, analyse the variables together, and to work on
one dataset rather than three.

But joining datasets can be tricky. There are many ways to join datasets
and you often have to do several checks of the data and prepare it in
the right way for your needs, like we've done, to get a useful and
correct output in the end. It can be really easy to not realise that
some data was dropped or not included or that extra data was added. It
can be extremely frustrating to check and confirm things, but ultimately
necessary to continue with doing your analyses and work correctly.

There are many ways to join datasets in `{dplyr}`, all of which are
described in the `?dplyr::join` and `?dplyr::semi_join` help
documentation. While there are several, only a few are most commonly
used. These are `left_join()`, `right_join()`, and `full_join()`.

`left_join(x, y)`

:   Join all rows and columns in `y` that match rows and columns in `x`.
    *Columns* that exist in `y` but not `x` are joined to `x`. The
    number of rows in the joined data is equal to the number of rows in
    `x`. If there are some rows in `y` that don't match with `x`, those
    rows are not included, and so are assigned `NA` values in the output
    data.

![Left joining in `{dplyr}`. Notice how the last row in the blue data
(the row with `d` in column `A`) is not included in the outputted data
on the right. Modified from the Posit `{dplyr}` [cheat
sheet](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf).
Image license is CC-BY-SA.](/images/left-join.png){#fig-left-join
width="90%"
fig-alt="A diagram showing two data frames, one green and one blue, with three columns each. Each data frame has one different column and one different row. The output data frame shows that all columns are output, but only the rows that match with the left data frame (the green one) are included in the output."}

`right_join(x, y)`

:   The opposite of `left_join()`. Join all rows and columns in `x` that
    match rows and columns in `y`. *Columns* that exist in `x` but not
    `y` are joined to `y`. Like `left_join()`, but the opposite, the
    number of rows in the joined data is equal to the number of rows in
    `y`. If there are some rows in `x` that don't match with `y`, those
    rows are not included and are given an `NA` value. This function can
    be useful when used while piping with `|>` when you are processing
    data in a sequence of steps and want to continue piping but join a
    different way then `left_join()`.

![Right joining in `{dplyr}`. Notice how the last row in the green data
(the row with `c` in column `A`) is not included in the outputted data
on the right. Modified from the Posit `{dplyr}` [cheat
sheet](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf).
Image license is
CC-BY-SA.](/images/right-join.png){#fig-right-join width="90%"
fig-alt="A diagram showing two data frames, one green and one blue, with three columns each. Each data frame has one different column and one different row. The output data frame shows that all columns are output, but only the rows that match with the right data frame (the blue one) are included in the output."}

`full_join(x, y)`

:   Join all rows and columns in `y` that match rows and columns in `x`.
    Columns *and* **rows** that exist in `y` but not `x` are joined to
    `x`. A full join keeps all the data from both `x` and `y`. This can
    be useful if you want to make sure you have all the data from both
    datasets, even if it results in more missing values in the output
    data.

![Full joining in `{dplyr}`. Notice how all rows and columns are
included in the outputted data on the right, and that some missingness
is introduced because those values don't exist when the data are
combined in this way. Modified from the Posit `{dplyr}` [cheat
sheet](https://posit.co/wp-content/uploads/2022/10/data-transformation-1.pdf).
Image license is CC-BY-SA.](/images/full-join.png){#fig-full-join
width="90%"
fig-alt="A diagram showing two data frames, one green and one blue, with three columns each. Each data frame has one different column and one different row. The output data frame shows that all columns are output, and all rows are included in the output, even if they don't match with the other data frame."}

{{< text_snippet sticky_up >}}

## Joining with dplyr functions

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Remind everyone that there will be some things that look odd when
joining, but it's likely due to us only importing the first 100 rows
from all the data files.

Walk through the code, the output, and highlight what remains after the
join.
:::

Let's use some of `{dplyr}`'s join functions! We'll join the
`participant_details` and `cgm_data` datasets together. We'll probably
want to use `full_join()`, but let's test out the other two joins first.
In your `docs/learning.qmd` file, go to the bottom of the file and make
a new header called `## Joining datasets`. Below that make a new code
chunk with {{< var keybind.chunk >}}. We'll start with using `left_join()`
to join the two datasets together.

First, let's look at the two datasets before joining them.

```{r}
#| filename: "docs/learning.qmd"
participant_details
cgm_data
```

Notice the number of rows in each dataset. The `participant_details`
dataset has `r nrow(participant_details)` rows and the `cgm_data`
dataset has `r nrow(cgm_data)` rows. When we `left_join()` them, there
should no more than `r nrow(participant_details) + nrow(cgm_data)` rows
(`cgm_data` plus `participant_details`) in the output. There will likely
be less because, remember, we have only imported the first 100 rows of
each dataset. Let's join them with `left_join()`.

```{r left-join}
#| filename: "docs/learning.qmd"
participant_details |>
  left_join(cgm_data)
```

Here, because `cgm_data` has the `hour` column, that means it expands
all the data from `participant_details` for the dates that it joins by.
The number of rows now is
`r nrow(left_join(participant_details, cgm_data))`, which is slightly
less than we expected but it also isn't surprising. Some rows in
`cgm_data` got dropped because we haven't yet imported all the data.

Because theoretically, the same `id` and `date` values should be present
in both `cgm_data` and `participant_details`, using either `left_join()`
or `full_join()` should be similar results. It should *not* give more
than `r nrow(participant_details) + nrow(cgm_data)` rows though.

But let's try `right_join()` first to see what happens. This is simpler,
since it will only join what is in the right side, e.g. `cgm_data`. That
means there should be exactly `r nrow(cgm_data)` rows in the output.

```{r right-join}
#| filename: "docs/learning.qmd"
participant_details |>
  right_join(cgm_data)
```

We see that, yes, it does have the same number of rows. We also see that
there are no missing values in the glucose columns. That's because the
`participant_details` dataset doesn't have glucose data and we know
there is no missing values in `cgm_data` for it. So there will be no
missing values in the output.

Lastly, we'll fully join. Like `left_join()`, the number of rows can not
be more than the sum of the number of rows in both datasets. But it can
be less, since, by joining there will be overlap between the two
datasets. At a minimum, there can not be less rows than the largest of
the two datasets, which is in this case `participant_details` at
`r nrow(participant_details)` rows. Let's try `full_join()` to see what
happens.

```{r full-join}
#| filename: "docs/learning.qmd"
participant_details |>
  full_join(cgm_data)
```

We can see that the number of rows is
`r nrow(full_join(participant_details, cgm_data))`, which is still less
than the sum of the two datasets. But again, it isn't surprising to us
here, since we know there is some overlap between the two datasets.

The `full_join()` function is most often used because we often want to
see all the data we have, even if there are some missing values.
Especially considering that some statistical models can use missing
values as extra information.

We also eventually want to join by the sleep data. Since `full_join()`
can only take two datasets at a time, we could write multiple
`full_join()` statements in a row. But what if we get other data later
on? Well, that's where more functional programming comes in. Again, we
have a simple goal: For a set of data frames, join them all together.

## :book: Reading task: Using the reduce functional

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Once they've read it, walk through the diagram again and try to explain
it slightly differently/in your own words.
:::

**Time: \~5 minutes.**

Here we use another functional programming concept from the `{purrr}`
package called `reduce()`. Like `map()`, which "maps" a function onto a
set of items, `reduce()` applies a function to each item of a vector or
list, each time reducing the set of items down until only one remains:
the output. In fact, this is how functions like `mean()` or `sum()`
work, as they take a vector and give you a single number. Likewise,
`summarise()` does the same thing, it takes a long vector (within groups
if you are using `group_by()` and gives you back a single number.

Let's use an example with our simple function `add_numbers()` we had
created earlier on, which we include below as a reminder. Let's say we
want to add up all the numbers from 1 to 5. Since `add_numbers()` only
takes two numbers, we have to give it two numbers at a time and repeat
until we reach 5.

```{r}
add_numbers <- function(num1, num2) {
  added <- num1 + num2
  return(added)
}
```

```{r}
# Add from 1 to 5
first <- add_numbers(1, 2)
second <- add_numbers(first, 3)
third <- add_numbers(second, 4)
add_numbers(third, 5)
```

We could also use the pipe `|>` operator to do the same thing:

```{r}
# Add from 1 to 5
add_numbers(1, 2) |>
  add_numbers(3) |>
  add_numbers(4) |>
  add_numbers(5)
```

This is quite a bit of code to read and write. Instead, we can use the
`reduce()` functional to do the same thing:

```{r}
reduce(1:5, add_numbers)
```

@fig-reduce below visually shows what is happening within `reduce()`.

![A functional that iteratively uses a function on a set of items until
only one output remains. Notice how the output of the first iteration of
`func()` is placed in the first position of `func()` in the next
iteration, and so on. Modified from the Posit purrr [cheat
sheet](https://posit.co/wp-content/uploads/2022/10/purrr.pdf). Image
license is CC-BY-SA](/images/reduce.png){#fig-reduce width="90%"
fig-alt="A diagram showing a block of green sub-blocks with the use of the reduce functional. Each sub-block is used in the first two positions of the function, then repeats with the output going into the first position of the next function. The final output is a single block of blue to represent everything being combined into one thing at the end."}

If we look at the documentation for `reduce()` by writing `?reduce` in
the Console, we see that`reduce()`, like `map()`, takes either a vector
or a list as an input. Since data frames can only be put together as a
list and not as a vector (a data frame has vectors for columns and so it
can't be a vector itself), we need to combine the datasets together in a
`list()` to be able to reduce them with `full_join()`.

{{< text_snippet sticky_up >}}

## Using `full_join()` with `reduce()`

Let's code this together, using `reduce()`, `full_join()`, and `list()`
while in the `docs/learning.qmd` file. In the same code chunk, we will
join the datasets together using `reduce()` and `full_join()`. This
time, we will assign the output to a new object called `dime_data`.

```{r reduce-full-join}
#| filename: "docs/learning.qmd"
dime_data <- list(participant_details, cgm_data) |>
  reduce(full_join)
dime_data
```

We now have the data in a form that would make sense to join it with the
other datasets. In the exercise, you will add the sleep data to join
everything together.

What's the advantage of using `reduce()` here? For us right now, not too
much. The amount of code is nearly identical if we just used several
`full_join()` in a row. But the advantage is that it is much easier to
include more datasets in the future, since it is only one extra data
frame object to include in the `list()`.

## :technologist: Exercise: Join the sleep data and check the output

**Time: \~12 minutes.**

Now that you've joined two datasets together, time to add the sleep data
too! It will be very easy to do, because we've already set it up to add
extra data to the `list()`. But, there still needs to be some checks.
After joining, the final `dime_data` should look something like:

{{< include /includes/solutions/_joins-sleep-data.qmd >}}

```{r}
#| filename: "Console"
dime_data
```

Complete these tasks to get to this point and to check the data:

1.  In the bottom of `docs/learning.qmd`, add a new header called
    `## Joining sleep data` and add a new code chunk below with
    {{< var keybind.chunk >}}.
2.  Start with checking the `sleep_data` by running it in the Console or
    in the code chunk. How many rows and columns does it have right now?
    Any missingness? (There is, lots of it, but where and how are things
    missing?)
3.  Calculate (in your head, on a piece of paper, with R using `nrow()`
    and `+`) what the maximum number of rows the fully joined datasets
    should have. You will need to add up all three datasets. That number
    is how many rows the fully joined datasets should, *at most*, have.
    Then, find what the smallest dataset. That number is the minimum
    number of rows the fully joined should have.
4.  In the code chunk, copy and paste (or manually type out) the code we
    used above to join the datasets with `reduce()`. Then, include the
    `sleep_data` at the end of the `list()`. Run the code with
    {{< var keybind.run-code >}}.
5.  Look through the newly joined data. Do some basic checks. Use
    `View(dime_data)` to get an overview and use the filtering/sorting
    buttons to explore the data a bit. What are some things you notice?
    Is there a pattern to the missingness? What about all the columns,
    anything to notice there? Do the glucose values and sleep values
    overlap at all?
    -   Keep in mind, since we only imported the first 100 rows, there
        will be some things that don't match up. But even with this,
        there are odd things that we will need to address.
    -   If you'd like, discuss with your neighbour about some things
        that could be fixed or improved.

{{< text_snippet sticky_up >}}

::: {.callout-note collapse="true"}
## :teacher: Teacher note

After they finish the exercise, highlight some of the issues with the
data, that we will then cover in the next session.
:::

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

-   Knowing how to join datasets together is a fundamental part of data
    processing and transformation. But it can be also tricky and a
    source of unexpected issues later on if not done well or not
    checked.
-   Use `left_join()`, `right_join()`, and `full_join()` to join two
    datasets together.
-   Use the functional `reduce()` to iteratively apply a function to a
    set of items in order to end up with one item (e.g. join more than
    two datasets into one final dataset).

```{r store-session-code}
#| include: false
save.image(here::here("_temp/dplyr-joins.RData"))
```

{{< include /_extensions/rostools/r3-theme/includes/code-appendix-r.qmd >}}
