# Joining data together {#sec-joins}

```{r}
#| label: setup
#| include: false

library(here)
library(tidyverse)
# To trigger downlit
library(dplyr)
load(here("_temp/pivots.RData"))
```

We now have several datasets that have a common set of columns, like
`id` and `datetime_id`, along with the a common datetime unit (minutes).
Now we can actually join all the datasets together into a single
dataset, which is what we will do during this session.

## Learning objectives

{{< include /includes/objectives/_joins.qmd >}}

## :speech_balloon: Discussion activity: Recall and share what you've learned so far

**Time: \~6 minutes.**

A very effective way to learn is to recall and describe to someone else
what you've learned. So before we continue this session, take some time
to think about what you've learned from yesterday.

1. Take 1-2 minute and try to recall as much as you can about what you'd
   done over the last two days. Without looking at your notes or the
   other sections of the website, try to remember things about importing
   and robust functions.
2. Then, for 4-5 minutes, share with your neighbour what you remember
   and try to describe it to each other. Maybe you will each remember
   different things.

## :book: Reading task: The many ways to join data

**Time: \~10 minutes.**

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Walk through and describe these images and the different type of joins
after they've read it.
:::

The ability to join datasets together is a fundamental component of data
processing and transformation. In many cases, it is required to do
before you can meaningfully do any work. Like in our case, we have
essentially three datasets that aren't particularly useful on their own
and only really benefit by being joined together. We join them to be
able to make comparisons, analyse the variables together, and to work on
one dataset rather than three.

But joining datasets can be tricky. There are many ways to join datasets
and you often have to do several checks of the data and prepare it in
the right way for your needs, like we've done, to get a useful and
correct output in the end. It can be really easy to not realise that
some data was dropped or not included or that extra data was added. It
can be extremely frustrating to check and confirm things, but ultimately
necessary to continue with doing your analyses and work correctly.

There are many ways to join datasets in `{dplyr}`, all of which are
described in the `?dplyr::join` and `?dplyr::semi_join` help
documentation. While there are several, only a few are most commonly
used. These are `left_join()`, `right_join()`, and `full_join()`.

`left_join(x, y)`

:   Join all rows and columns in `y` that match rows and columns in `x`.
    *Columns* that exist in `y` but not `x` are joined to `x`. The
    number of rows in the joined data is equal to the number of rows in
    `x`. If there are some rows in `y` that don't match with `x`, those
    rows are not included, and so are assigned `NA` values in the output
    data.

![Left joining in `{dplyr}`. Notice how the last row in the blue data
(the row with `d` in column `A`) is not included in the outputted data
on the right. Modified from the Posit `{dplyr}` [cheat
sheet](https://opensource.posit.co/resources/cheatsheets/data-transformation/data-transformation.pdf).
Image license is
CC-BY-SA.](/images/left-join.svg){#fig-left-join
width="90%" fig-alt="A diagram showing two data frames, one green and
one blue, with three columns each. Each data frame has one different
column and one different row. The output data frame shows that all
columns are output, but only the rows that match with the left data
frame (the green one) are included in the output."}

`right_join(x, y)`

:   The opposite of `left_join()`. Join all rows and columns in `x` that
    match rows and columns in `y`. *Columns* that exist in `x` but not
    `y` are joined to `y`. Like `left_join()`, but the opposite, the
    number of rows in the joined data is equal to the number of rows in
    `y`. If there are some rows in `x` that don't match with `y`, those
    rows are not included and are given an `NA` value. This function can
    be useful when used while piping with `|>` when you are processing
    data in a sequence of steps and want to continue piping but join a
    different way then `left_join()`.

![Right joining in `{dplyr}`. Notice how the last row in the green data
(the row with `c` in column `A`) is not included in the outputted data
on the right. Modified from the Posit `{dplyr}` [cheat
sheet](https://opensource.posit.co/resources/cheatsheets/data-transformation/data-transformation.pdf).
Image license is
CC-BY-SA.](/images/right-join.svg){#fig-right-join
width="90%" fig-alt="A diagram showing two data frames, one green and
one blue, with three columns each. Each data frame has one different
column and one different row. The output data frame shows that all
columns are output, but only the rows that match with the right data
frame (the blue one) are included in the output."}

`full_join(x, y)`

:   Join all rows and columns in `y` that match rows and columns in `x`.
    Columns *and* **rows** that exist in `y` but not `x` are joined to
    `x`. A full join keeps all the data from both `x` and `y`. This can
    be useful if you want to make sure you have all the data from both
    datasets, even if it results in more missing values in the output
    data.

![Full joining in `{dplyr}`. Notice how all rows and columns are
included in the outputted data on the right, and that some missingness
is introduced because those values don't exist when the data are
combined in this way. Modified from the Posit `{dplyr}` [cheat
sheet](https://opensource.posit.co/resources/cheatsheets/data-transformation/data-transformation.pdf).
Image license is
CC-BY-SA.](/images/full-join.svg){#fig-full-join
width="90%" fig-alt="A diagram showing two data frames, one green and
one blue, with three columns each. Each data frame has one different
column and one different row. The output data frame shows that all
columns are output, and all rows are included in the output, even if
they don't match with the other data frame."}

{{< text_snippet sticky_up >}}

## Joining with dplyr functions

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Remind everyone that there will be some things that look odd when
joining, but it's likely due to us only importing the first 100 rows
from all the data files.

Walk through the code, the output, and highlight what remains after the
join.
:::

Let's use some of `{dplyr}`'s join functions! We'll start by joining the
`survey_long` with the `survey_tidy` datasets together. We'll probably
want to use `full_join()`, but let's test out the other two joins first.
In your `docs/learning.qmd` file, go to the bottom of the file and make
a new header called `## Joining datasets`. Below that make a new code
chunk with {{< var keybind.chunk >}}. We'll start with using
`left_join()` to join the two datasets together.

First, let's look at the two datasets before joining them.

```{r}
#| filename: "docs/learning.qmd"

survey_long
survey_tidy
```

Notice the number of rows in each dataset. The `survey_long` dataset has
`r nrow(survey_long)` rows and the `survey_tidy` dataset has
`r nrow(survey_tidy)` rows. When we `left_join()` them, there should no
more than `r nrow(survey_long) + nrow(survey_tidy)` rows (`survey_tidy`
plus `survey_long`) in the output. There will likely be less because,
remember, we have only imported the first 100 rows of each dataset.
Let's join them with `left_join()`.

```{r}
#| label: left-join
#| filename: "docs/learning.qmd"

survey_long |>
  left_join(survey_tidy)
```

Ok, we got a warning about "many-to-many" joins. When you get this
warning, this means there are duplicate ID values. This type of thing
usually requires manual investigation and correction. For now, let's
just identify which IDs are the duplicates. A very easy way to check
that is by combining `count()` and `filter()` from `{dplyr}`. Let's
check for duplicates in the `id` and `datetime_id` columns, since those
are our unique observational identifiers. Let's start with the
`count()`.

```{r}
#| label: count-ids

survey_tidy |>
  count(id, datetime_id)
```

See how all the `n` values that we can see are `1`? That means for that
combination of `id` and `datetime_id`, there is only one. Which is what
we want. To find which are duplicates we use `filter()` to find any rows
where `n > 1`.

```{r}
#| label: check-ids

survey_tidy |>
  count(id, datetime_id) |>
  filter(n > 1)
```

Alright, it seems we have some duplicate identifiers. For this workshop,
we won't do anything about it, as it requires reviewing the
documentation, understanding the context that the data was created in,
and then deciding how to handle those duplicates. But this information
tells us *why* we are receiving the warning. We'll just ignore the
warnings for now.

Let's continue and try with the `right_join()`. This should give us
about the same number of rows as `survey_tidy`, which is
`r nrow(survey_tidy)`, plus those duplicates.

```{r}
#| label: right-join
#| filename: "docs/learning.qmd"

survey_long |>
  right_join(survey_tidy)
```

We see that it has the same number of rows as when using `left_join()`.
This isn't always the case, but in our case it is.

Lastly, we'll use `full_join()`. Like `left_join()`, the number of rows
can not be more than the sum of the number of rows in both datasets. But
it can be less, since, by joining there will be overlap between the two
datasets. At a minimum, there can not be fewer rows in the joined data
than in the largest of the two datasets, which is in this case
`survey_long` at `r nrow(survey_long)` rows. Let's try `full_join()` to
see what happens.

```{r}
#| label: full-join
#| filename: "docs/learning.qmd"

survey_long |>
  full_join(survey_tidy)
```

We can see that the number of rows is
`nrow(full_join(survey_long, survey_tidy))`, which is still less than
the sum of the two datasets. We know from the warnings that there are
some duplicates, but we'll keep ignoring those for now.

The `full_join()` function is most often used because we often want to
see all the data we have, even if there are some missing values.
Especially considering that some statistical models can use missing
values as extra information. Let's continue the pipe and join the
`hr_data` dataset as well.

```{r}
#| label: assign-full-join
#| filename: "docs/learning.qmd"

survey_long |>
  full_join(survey_tidy) |>
  full_join(hr_data)
```

We now have the start of a single dataset that we can eventually use for
analyses. Let's do some checks first to make sure everything joined
together as we expect. Let's assign the output of the joins to
`nurses_stress`.

```{r}
#| label: check-joined-data
#| filename: "docs/learning.qmd"

nurses_stress <- survey_long |>
  full_join(survey_tidy) |>
  full_join(hr_data)
```

Then use `arrange()` of the `id` and `collection_datetime` columns
before piping to `View()` to make it easier to check the data.

```{r}
#| label: view-joined-data
#| eval: false
#| filename: "docs/learning.qmd"

nurses_stress |>
  arrange(id, collection_datetime) |>
  View()
```

Ok, it looks like we have a lot of missing values. Do we expect that?
Maybe. So far we've only read in the first 100 rows the HR-type data, so
this could be the reason. Until we read in everything, we won't
completely know. But there might also be missing values given that the
HR-type data was recorded for many days in a row, while the survey data
seems to only contain data on dates that the survey was filled out. But
remember, at this stage, we're only prototyping and testing out the
code, it doesn't have to make sense just yet, as long as the code runs.
For now it's time for you to create the final dataset and then you can
check it out in more detail.

## :technologist: Exercise: Join the rest of the data together

**Time: \~10 minutes.**

Time to put it all together in the `docs/cleaning.qmd` file.

1. Restart your R session with {{< var keybind.restart-r >}} to clear
   your environment and make sure you have a clean slate.
2. Open up the `docs/cleaning.qmd` file and scroll down to the bottom of
   the file. Make a new header called `## Joining datasets`.
3. Below that, make a new code chunk with {{< var keybind.chunk >}}.
4. In that code chunk, copy and paste the `full_join()` code you just
   wrote to join the `survey_long`, `survey_tidy`, and `hr_data`
   datasets together.
5. Continue the pipe and using `full_join()` to join the rest of the
   datasets, which are `ibi_data`, `temp_data`, `bvp_data`, and
   `eda_data`.
6. Assign the output of the joins to `nurses_stress`.
7. Run the code in `docs/cleaning.qmd` from beginning to end, then run
   `View(nurses_stress)` in the Console to check the data.
8. Style the with `styler::style_file()`.
9. Render the document with {{< var keybind.render >}} to make sure
   everything runs.
10. Commit the changes with {{< var keybind.commit >}} and push to
    GitHub.

::: {.callout-note collapse="true" appearance="minimal" title="**Click for the solution**. Only click if you are struggling or are out of time."}
```{r}
#| label: admin-read-hr-type-data
#| include: false

hr_data <- read_all("HR.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()

ibi_data <- read_all("IBI.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()

temp_data <- read_all("TEMP.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()

bvp_data <- read_all("BVP.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()

eda_data <- read_all("EDA.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()
```

This should be at the bottom of the `docs/cleaning.qmd` file.

```{r}
#| label: solution-join-all
#| output: false
#| filename: "docs/cleaning.qmd"

nurses_stress <- survey_long |>
  full_join(survey_tidy) |>
  full_join(hr_data) |>
  full_join(ibi_data) |>
  full_join(temp_data) |>
  full_join(bvp_data) |>
  full_join(eda_data)
```

And then run the code to check the data in the Console.

```{r}
#| label: solution-check-joined-data
#| filename: "Console"
#| eval: false

View(nurses_stress)
```
:::

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Save the final dataset to the `data/` folder

**Time: \~10 minutes.**

The final step is to save the dataset to the `data/` folder. But, we
also have to make sure to now read in all the data, not just the first
100 rows.

1. At the bottom of the last code chunk in the `docs/cleaning.qmd` file,
   add the code to save the `nurses_stress` dataset to the `data/`
   folder with `write_csv()`. Use the `here()` function to specify the
   path to save it to. Save it to `data/nurses-stress.csv`. Why save it
   there? So that you know that that is the "ready data" and not the raw
   data. Run the code to make sure that it saves correctly.
2. Since we don't want to track this file in Git, run the code in the
   Console to ignore it with `usethis::use_git_ignore()`.
3. Open up the `R/functions.R` file and find the `read()` function. Find
   the line that says `max_rows = 100` and replace it with
   `max_rows = Inf` to read in all the data. (If you did one of the
   previous exercises to add `max_rows` argument to `read_all()`, then
   you will need to change that instead.)
4. Go back into `docs/cleaning.qmd` and render the `docs/cleaning.qmd`
   file with {{< var keybind.render >}} to make sure everything runs and
   the dataset is saved correctly. **WARNING**: This will take a long
   time to run, so make sure this is the last thing you do here. If your
   computer isn't fast enough or powerful enough, you can change the
   `max_rows` to `10000` or something similar so that you can at least
   running all the code and output a large dataset.
5. Commit the changes with {{< var keybind.commit >}} and push to
   GitHub.

::: {.callout-note collapse="true" appearance="minimal" title="**Click for the solution**. Only click if you are struggling or are out of time."}
```{r}
#| label: solution-save-nurses-stress
#| eval: false
#| filename: "docs/cleaning.qmd"

write_csv(nurses_stress, here("data/nurses-stress.csv"))
```

And run the code in the Console to ignore the file.

```{r}
#| label: solution-ignore-nurses-stress
#| eval: false
#| filename: "Console"

usethis::use_git_ignore("data/nurses-stress.csv")
```
:::

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- Knowing how to join datasets together is a fundamental part of data
  processing and transformation. But it can be also tricky and a source
  of unexpected issues later on if not done well or not checked.
- Use `left_join()`, `right_join()`, and `full_join()` to join two
  datasets together.

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
