# Quickly re-arranging data with pivots {#sec-pivots}

```{r setup}
#| include: false
# Used this only during writing to run code dependent on previous chapters.
library(here)
library(fs)
library(tidyverse)
source(here("R/functions.R"))
load(here("_temp/dplyr-joins.RData"))
load(here("data/mmash.rda"))
options(dplyr.summarise.inform = FALSE)
```

Here we will continue using the *Workflow* block as we cover the fourth
block, "*Work with project data*" in @fig-overview-analyze-project-data.

![Section of the overall workflow we will be
covering.](/images/overview-analyze-project-data.svg){#fig-overview-analyze-project-data}

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Briefly go over the bigger picture (found in the introduction section)
and remind everyone the 'what' and 'why' of what we are doing.
:::

## Learning objectives

The overall learning outcome for this session is to:

1.  Describe what long and wide data are, what pivoting means in this
    context, and demonstrate how to switch between the two.

Specific objectives are to:

1.  Describe the concept of "pivoting" data to convert data in a "long"
    format to a "wide" format and vice versa.
2.  Identify situations when data in the long or wide format are more
    appropriate to use.
3.  Apply the `pivot_longer()` and `pivot_wider()` functions from the
    `{purrr}` package to pivot data.
4.  Continue extending functions to make use of pivots for processing
    and cleaning data.

## Setup for the analysis in Quarto

We now have a working dataset to start doing some simple analyses on in
the Quarto document. A recommended workflow with Quarto is to often
render / "knit" it and make sure your analysis is reproducible (while on
your computer). We already cleaned it up from the previous session.

We will now add the `load()` code right below the `source()` function in
the `setup` code chunk:

```{r}
#| filename: "doc/learning.qmd"
#| eval: false
source(here("R/functions.R"))
load(here("data/mmash.rda"))
```

As we write more R code and do some simple analyses of the data, we are
going to be knitting fairly often (depending on how long the analysis
takes of course). The main reason for this is to ensure that whatever
you are writing and coding will at least be reproducible on your
computer, since Quarto is designed to ensure the document is
reproducible.

## Re-arranging data for easier summarizing

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Let them read through this section and then walk through it again and
explain it a bit more, making use of the tables and graphs. Doing both
reading and listening again will help reinforce the concept of pivoting,
which is usually quite difficult to grasp for those new to it.
:::

::: callout-note
## Reading task: \~6 minutes

Now that we have the final dataset to work with, we want to explore it a
bit with some simple descriptive statistics. One extremely useful and
powerful tool to summarizing data is by "pivoting" your data. Pivoting
is when you convert data between longer forms (more rows) and wider
forms (more columns). The `{tidyr}` package within `{tidyverse}`
contains two wonderful functions for pivoting: `pivot_longer()` and
`pivot_wider()`. There is a well written documentation on pivoting in
the [tidyr website](https://tidyr.tidyverse.org/articles/pivot.html)
that can explain more about it. The first thing we'll use, and probably
the more commonly used in general, is `pivot_longer()`. This function is
commonly used because entering data in the wide form is easier and more
time efficient than entering data in long form. For instance, if you
were measuring glucose values over time in participants, you might enter
data in like this:

```{r table-example-wide}
#| echo: false
example_wide <- tribble(
  ~person_id, ~glucose_0, ~glucose_30, ~glucose_60,
  1, 5.6, 7.8, 4.5,
  2, 4.7, 9.5, 5.3,
  3, 5.1, 10.2, 4.2
)

example_wide |>
  knitr::kable(
    caption = "Example of a **wide** dataset that is useful for data entry.",
    align = "c"
  )
```

However, when it comes time to analyze the data, this wide form is very
inefficient and difficult to computationally and statistically work
with. So, we do data entry in wide and use functions like
`pivot_longer()` to get the data ready for analysis. @fig-pivot-longer
visually shows what happens when you pivot from wide to long.

![Pivot longer in tidyr. New columns are called 'name' and 'value'.
Notice how the values in `A` and `B` columns are stacked on top of each
other in the newly created `V`
column.](/images/pivot-longer.png){#fig-pivot-longer width="90%"}

If you had, for instance, an ID column for each participant, the
pivoting would look like what is shown in @fig-pivot-longer-id.

![Pivot longer in tidyr, excluding an 'id' column. New columns are
called 'name' and 'value', as well as the old 'id' column. Notice how,
unlike the previous image, the `id` column is excluded when pivotting
into the data on the
right.](/images/pivot-longer-id.png){#fig-pivot-longer-id
width="90%"}

Pivoting is a conceptually challenging thing to grasp, so don't be
disheartened if you can't understand how it works yet. As you practice
using it, you will understand it. With `pivot_longer()`, the first
argument is the data itself. The other arguments are:

1.  `cols`: The columns to use to convert to long form. The input is a
    vector made using `c()` that contains the column names, like you
    would use in `select()` (e.g. you can use the `select_helpers` like
    `starts_with()`, or `-` minus to exclude).
2.  `names_to`: Optional, the default is `name`. If provided, it will be
    the name of the newly created column (as a quoted character) that
    contains the original column names.
3.  `values_to`: Optional, the default is `value`. Like `names_to`, sets
    the name of the new columns.

Both `pivot_longer()` and its opposite `pivot_wider()`, which we will
cover later in the session, are incredibly powerful functions. We can't
show close to everything it can do in this course, but if you want to
learn more, read up on the
[documentation](https://tidyr.tidyverse.org/articles/pivot.html) for it.
:::

Since we'll be using `{tidyr}`, we will need to add it as a dependency
by running this in the Console:

```{r}
#| filename: "Console"
#| eval: false
usethis::use_package("tidyr")
```

Let's try this out with `mmash`. In your `doc/learning.qmd` file, create
a new header called `## Pivot longer` and create a new code chunk below
that with {{< var keybind.chunk >}}. Now we can start typing in our
code:

```{r pivot-everything-error}
#| filename: "doc/learning.qmd"
#| error: true
mmash |>
  # pivot every column
  pivot_longer(everything())
```

Why the error? We get an error because we are trying to mix data types.
We can't have character data and number data in the same column. Let's
pivot only numbers.

```{r pivot-numeric}
#| filename: "doc/learning.qmd"
mmash |>
  pivot_longer(where(is.numeric))
```

Nice! But not super useful. We can exclude specific columns from
pivoting with `-` before the column name, for instance with `user_id`
and `day`. Let's drop the `samples` column before pivoting since `day`
gives us the same information:

```{r pivot-longer-example}
#| filename: "doc/learning.qmd"
mmash |>
  select(-samples) |>
  pivot_longer(c(-user_id, -day, -gender))
```

## Exercise: Summarise your data after pivoting

> Time: 15 minutes.

Using the `group_by()` and `summarise()` functions we learned in
@sec-summarise-with-functionals, complete these tasks starting from this
code.

``` {.r filename="doc/learning.qmd"}
mmash |> 
  select(-samples) |> 
  pivot_longer(c(-user_id, -day, -gender)) |> 
  ___
```

1.  Continuing the `|>` from `pivot_longer()`, use `group_by()` to group
    the data by `gender`, `day`, and `name` (the long form column
    produced from `pivot_longer()`).
2.  After grouping with `group_by()`, use `summarise()` and `across()`
    on the `value` column and find the mean and standard deviation (put
    them into a named list like we did previously). Don't forget to use
    `na.rm = TRUE` to exclude missing values.
3.  Stop the grouping effect with `ungroup()`.
4.  Run `{styler}` while in the `doc/learning.qmd` file with
    {{< var keybind.styler >}}.
5.  Render the Quarto file into HTML with {{< var keybind.render >}}.
6.  Open up the Git interface and **add and commit** the changes to
    `doc/learning.qmd` with {{< var keybind.git >}}

```{r solution-pivot-summarize}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
mmash |>
  select(-samples) |>
  pivot_longer(c(-user_id, -day, -gender)) |>
  group_by(gender, day, name) |>
  summarise(across(
    value,
    list(
      mean = \(x) mean(x, na.rm = TRUE),
      sd = \(x) sd(x, na.rm = TRUE)
    )
  )) |>
  ungroup()
```

## Pivot data to wider form

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Like with the pivoting to long section, let them read through this
section first and then go over it again to verbally explain it more,
making use of the graphs to help illustrate what is happening. Doing
both reading and listening will help reinforce the concepts.
:::

::: callout-note
## Reading task: \~6 minutes

After using `pivot_longer()` on the summarised data, it looks nice, but
it could be better. Right now it is in a pretty long form, but for
showing as a table, having columns for either `gender` or `day` would
make it easier to compare the mean and SD values we obtain. This is
where we can use `pivot_wider()` to get the data wider rather than long.
The arguments for `pivot_wider()` are very similar to those in
`pivot_longer()`, except instead of `names_to` and `values_to`, they are
called `names_from` and `values_from`. Like with many R functions, the
first argument is the data and the other arguments are:

1.  `id_cols`: This is optional as it will default to all column names.
    This argument tells `pivot_wider()` to use the given columns as the
    identifiers for when converting. Unlike `pivot_longer()` which
    doesn't require some type of "key" or "id" column to convert to long
    form, the conversion to wide form requires some type of "key" or
    "id" column because `pivot_wider()` needs to know which rows belong
    with each other.
2.  `names_from`: Similar to the `pivot_longer()`, this is the name of
    the column that has the values that will make up the new columns.
    Unlike with the `names_to` argument in `pivot_longer()` which takes
    a character string as input, the column name for `names_from` must
    be *unquoted* because you are selecting a column that already exists
    in the dataset.
3.  `values_from`: Same as `names_from`, this is the column name (that
    exists and must be given *unquoted*) for the values that will be in
    the new columns.

@fig-pivot-wider visually shows what's happening when using
`pivot_wider()`.

![Pivot wider in tidyr, where a set of stacked "groups" in the data on
the left are placed side-by-side as new columns in the output data on
the right. Notice how the values in the column `N`, which is used in the
`names_from` argument, are used as the names for the new columns `A` and
`B` in the new data.](/images/pivot-wider.png){#fig-pivot-wider
width="90%"}
:::

In our case, we want either `gender` or `day` as columns with the mean
and SD values. Let's use `pivot_wider()` on `day` to see differences
between days.

```{r pivot-wider-from-summarised}
#| filename: "doc/learning.qmd"
#| error: true
mmash |>
  select(-samples) |>
  pivot_longer(c(-user_id, -day, -gender)) |>
  group_by(gender, day, name) |>
  summarise(across(
    value,
    list(
      mean = \(x) mean(x, na.rm = TRUE),
      sd = \(x) sd(x, na.rm = TRUE)
    )
  )) |>
  ungroup() |>
  pivot_wider(names_from = day)
```

Hmm, didn't work. Nothing has been pivoted to wider. That's because we
are missing the `value_from` argument. Since we actually have the two
`value_mean` and `value_sd` columns that have "values" in them, we need
to tell `pivot_wider()` to use those two columns. Since `values_from`
works similar to `select()`, we can use `starts_with()` to select the
columns starting with `"values"`.

```{r pivot-wider-from-summarised-starts-with}
#| filename: "doc/learning.qmd"
mmash |>
  select(-samples) |>
  pivot_longer(c(-user_id, -day, -gender)) |>
  group_by(gender, day, name) |>
  summarise(across(
    value,
    list(
      mean = \(x) mean(x, na.rm = TRUE),
      sd = \(x) sd(x, na.rm = TRUE)
    )
  )) |>
  ungroup() |>
  pivot_wider(names_from = day, values_from = starts_with("value"))
```

Now we have a different problem. There are missing values in both the
`day` and `gender` columns that, at least in this case, we don't want
pivoted. Shouldn't they be removed when we include `na.rm = TRUE` in our
code? The function of `na.rm = TRUE` is not to remove `NA` values, but
to instead tell R to not include variables in `mmash` that are `NA` when
calculating the mean and standard deviation. In this particular case,
the columns `value_mean_NA` or `value_mean_-29` have `NA` or `NaN`
values because there are no other values in the data *other than* `NA`.
Since we don't actually care about missing days (or the random `-29`
day), we can remove missing values with the function called `drop_na()`.
We also don't care about missing `gender` values, so we'll drop them as
well. Add it in the pipe right before `group_by()`.

```{r pivot-wider-from-summarised-drop-na}
#| filename: "doc/learning.qmd"
mmash |>
  select(-samples) |>
  pivot_longer(c(-user_id, -day, -gender)) |>
  drop_na(day, gender) |>
  group_by(gender, day, name) |>
  summarise(across(
    value,
    list(
      mean = \(x) mean(x, na.rm = TRUE),
      sd = \(x) sd(x, na.rm = TRUE)
    )
  )) |>
  ungroup() |>
  pivot_wider(names_from = day, values_from = starts_with("value"))
```

Now that that works, let's render the document to HTML with
{{< var keybind.render >}}.

## Exercise: Convert this code into a function

> Time: 15 minutes.

Using the same workflow we've been doing throughout this course, convert
the code we just wrote above into a function.

1.  Name the function `tidy_summarise_by_day`.
2.  Create one argument called `data`. Create a new variable inside the
    function called `daily_summary` and put it in `return()` so the
    function outputs it.
3.  Test that the function works.
4.  Add Roxygen documentation with {{< var keybind.roxygen >}} and use
    explicit function calls with `packagename::`.
    -   Don't forget, you can use `?functionname` to find out which
        package the function comes from.
5.  Move (**cut** and paste) the newly created function over into the
    `R/functions.R` file.
6.  Run `{styler}` while in the `R/functions.R` file with
    {{< var keybind.styler >}}.
7.  Restart R with {{< var keybind.restart-r >}}, go into the
    `doc/learning.qmd` file and run the `setup` code chunk in the Quarto
    document with the `source()` and `load()` commands. Then test that
    the new function works in a code chunk at the bottom of the
    document.
8.  Render the Quarto document to HTML with {{< var keybind.render >}}.
9.  Add and commit the changes to the Git history with
    {{< var keybind.git >}}.

Use this code to refresh your memory on how functions are structured and
to use as a starting point:

``` r
___ <- function(___) {
    
}
```

```{r solution-summarise-function}
#| eval: true
#| output: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
#' Calculate tidy summary statistics by day.
#'
#' @param data The MMASH dataset.
#'
#' @return A data.frame/tibble.
#'
tidy_summarise_by_day <- function(data) {
  daily_summary <- data |>
    dplyr::select(-samples) |>
    tidyr::pivot_longer(c(-user_id, -day, -gender)) |>
    tidyr::drop_na(day, gender) |>
    dplyr::group_by(gender, day, name) |>
    dplyr::summarise(dplyr::across(
      value,
      list(
        mean = \(x) mean(x, na.rm = TRUE),
        sd = \(x) sd(x, na.rm = TRUE)
      )
    )) |>
    dplyr::ungroup() |>
    tidyr::pivot_wider(
      names_from = day,
      values_from = dplyr::starts_with("value")
    )
  return(daily_summary)
}

# Testing that the function works.
mmash |>
  tidy_summarise_by_day()
```

## Extending the function to use other statistics and to be tidier

Now that we've made the tidy summary code into a function, let's make it
more generic so we can use other summary statistics and to have the
output be a bit tidier. For instance, it would be nice to be able to do
something like this (we'd still have to use `na.rm` though):

``` r
mmash |> 
    tidy_summarise_by_day(\(x) median(x, na.rm = TRUE))
mmash |> 
    tidy_summarise_by_day(\(x) max(x, na.rm = TRUE))
mmash |> 
    tidy_summarise_by_day(
      list(
        median = \(x) median(x, na.rm = TRUE),
        max = \(x) max(x, na.rm = TRUE)
      )
    )
```

Before we get to adding this functionality, let's first make it so the
function has a tidier output. Specifically, we want to round the values
so they are easier to read. Go into the `R/functions.R` script to the
`tidy_summarize_by_day()` function. We'll create a new line right after
the `dplyr::summarise()` function, after the `|>` pipe. Since we want to
round values of existing columns, we need to use `mutate()`. And like we
used `across()` in `summarise()`, we can also use `across()` within
`mutate()` on specific columns. In our case, we want to round columns
that `start_with()` the word `"value"` to 2 digits.

```{r rounding-summary-function-output}
#| filename: "R/functions.R"
tidy_summarise_by_day <- function(data) {
   daily_summary <- data |>
    dplyr::select(-samples) |>
    tidyr::pivot_longer(c(-user_id, -day, -gender)) |>
    tidyr::drop_na(day, gender) |>
    dplyr::group_by(gender, day, name) |>
    dplyr::summarise(dplyr::across(
      value,
      list(
        mean = \(x) mean(x, na.rm = TRUE),
        sd = \(x) sd(x, na.rm = TRUE)
      )
    )) |>
    dplyr::ungroup() |>
    dplyr::mutate(dplyr::across(
      dplyr::starts_with("value"),
      \(x) round(x, digits = 2)
    )) |>
    tidyr::pivot_wider(
      names_from = day,
      values_from = dplyr::starts_with("value")
    )
return(daily_summary)
}
```

Source the `R/functions.R` file with {{< var keybind.source >}}, then
test out the function in the Console:

```{r}
#| filename: "Console"
tidy_summarise_by_day(mmash)
```

That's much easier to read with the values rounded. Now let's add the
ability to change the summary statistics function to something else.
This is a surprisingly easy thing so before we do that, let's take a few
minutes to brainstorm how we can achieve this.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Get the groups to chat together for about 5 minutes to think about how
they'd do that. Ask that they don't look ahead in the text. After that,
discuss some ways to add the functionality.
:::

Now that we've discussed this and come to a conclusion, let's update the
function.

```{r generalize-summary-function}
#| filename: "R/functions.R"
tidy_summarise_by_day <- function(data, summary_fn) {
  daily_summary <- data |>
    dplyr::select(-samples) |>
    tidyr::pivot_longer(c(-user_id, -day, -gender)) |>
    tidyr::drop_na(day, gender) |>
    dplyr::group_by(gender, day, name) |>
    dplyr::summarise(dplyr::across(
      value,
      summary_fn
    )) |>
    dplyr::ungroup() |>
    dplyr::mutate(dplyr::across(
      dplyr::starts_with("value"),
      \(x) round(x, digits = 2)
    )) |>
    tidyr::pivot_wider(
      names_from = day,
      values_from = dplyr::starts_with("value")
    )
return(daily_summary)
}
```

Source the `R/functions.R` file with {{< var keybind.source >}}, then
test out the function in the Console:

```{r}
#| filename: "Console"
tidy_summarise_by_day(mmash, \(x) max(x, na.rm = TRUE))
```

Now that it works, let's add some summary statistics to the
`doc/learning.qmd` file.

```{r test-tidier-summary}
#| filename: "doc/learning.qmd"
mmash |>
  tidy_summarise_by_day(\(x) median(x, na.rm = TRUE))
mmash |>
  tidy_summarise_by_day(list(
    median = \(x) median(x, na.rm = TRUE),
    max = \(x) max(x, na.rm = TRUE)
  ))
```

Before continuing, let's render the Quarto document with
{{< var keybind.render >}} to check reproducibility.

## Making prettier output in Quarto

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Can go over this quite quickly after they've (optionally) finished the
previous exercise.
:::

What we created is nice and all, but since we are working in a Quarto
document and generating to HTML, let's make it easier for others
(including yourself) to read the document. Let's make the output as an
actual table. We can do that with `knitr::kable()` (meaning "knitr
table"). We can also add a table caption with the `caption` argument.

```{r knitr-table}
#| filename: "doc/learning.qmd"
mmash |>
  tidy_summarise_by_day(list(
    mean = \(x) mean(x, na.rm = TRUE),
    min = \(x) min(x, na.rm = TRUE),
    max = \(x) max(x, na.rm = TRUE)
  )) |>
  knitr::kable(caption = "Descriptive statistics of some variables.")
```

Then render the document with {{< var keybind.render >}} and check out
the HTML file. So pretty! :grin: Well, there's lots of things to fix up,
but its a good starting place. Let's finish by running `{styler}` with
{{< var keybind.styler >}} and then committing the changes to the Git
history with {{< var keybind.git >}}.

## Summary

-   Data is usually structured to varying degrees as wide or long
    format.
    -   Use `pivot_longer()` to convert from wide to long.
    -   Use `pivot_wider()` to convert from long to wide.

