# Pivoting your data between long and wide {#sec-pivots}

```{r}
#| label: setup
#| include: false

# Used this only during writing to run code dependent on previous chapters.
library(here)
library(tidyverse)
load(here("_temp/split-apply-combine.RData"))
# hr_data <- read_all("HR.csv.gz") |>
#   get_participant_id() |>
#   summarise_by_datetime()
```

In the HR-type datasets, we now have the participant ID that matches the
participant ID in the survey results data. We also have minutes as the
unit of time for the datetime column that matches the unit of time in
the time columns of the survey results data. But, the HR data is in a
longer format, where there are multiple rows for each participant, while
the survey result data is in wider format but also has multiple rows per
participant. Right now, we are not able to (meaningfully) join these
datasets together because of this incompatibility in the format of the
data. So in this session we will go over how to pivot data between
long(er) and wide(r) formats.

## Learning objectives

{{< include /includes/objectives/_pivots.qmd >}}

## :book: Reading task: Pivoting data for easier wrangling

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Briefly walk through it again after they've read it and explain it a bit
more, making use of the tables and graphs. Pivoting is often a difficult
concept to grasp, to it's good to go over it a few times. Emphasise the
difficulty of it, to remind them it is ok to not get it.
:::

**Time: \~6 minutes.**

Rarely is our data in the exact format we need or want it to be in to
effectively process, analyse, or visualize it. Often we'll need to do
some heavy reshaping. One very powerful way to reshape data is with
"pivoting". Pivoting is when you take a dataset's rows and convert them
to columns or convert columns to rows, and in many different
combinations of this.

The `{tidyr}` package within the `{tidyverse}` contains two wonderful
functions for pivoting: `pivot_longer()` and `pivot_wider()`. There is a
well written documentation on pivoting in the [tidyr
website](https://tidyr.tidyverse.org/articles/pivot.html) that can
explain more about it that you can read at another time.

Most use cases of pivoting are to convert to the longer form, which you
can do with `pivot_longer()` and is what we will cover first in this
session. There are many reasons pivoting longer is the most common pivot
direction.

An example of why you often need to pivot longer is because data entry
is easier to do in the wider format. Entering data, especially through a
spreadsheet tool like Excel, is so much easier in the wider format and
more time efficient than entering data in long form. Consider the
scenario where you are entering data from a participant for glucose
values that were measured every 30 minutes. So you might enter data like
this:

  | person_id | glucose_0 | glucose_30 | glucose_60 |
  | :-------- | :-------- | :--------- | :--------- |
  | 1         | 5.6       | 7.8        | 4.5        |
  | 2         | 4.7       | 9.5        | 5.3        |
  | 3         | 5.1       | 10.2       | 4.2        |

  : This is a **wide** dataset that is useful for data entry.

If you are working with a spreadsheet, you might enter the glucose value
for one participant at one time point, then hit the tab key to move to
the next cell and add the next glucose value. Even outside of a
spreadsheet, the way data is structured from survey software is often
also in the wide format. However, when it comes time to analyze the
data, this wide format is very inefficient and difficult to
computationally and statistically work with. So, we do data entry in
wide format and use functions like `pivot_longer()` to get the data
ready for analysis. See @fig-pivot-longer below that visually shows what
happens when you pivot from wide to long.

![Pivot longer in `{tidyr}`. New columns are called `name` and `value`.
Notice how the values in `A` and `B` columns are stacked on top of each
other in the newly created `V`
column.](/images/pivot-longer.svg){#fig-pivot-longer
width="90%" fig-alt="A diagram showing the use of the pivot longer
function, where a table with two columns, A and B, is converted to a
table with two columns, name and value. The original columns values are
stacked in the value column and the original columns names are stacked
in the name column."}

If you had, for instance, an ID column for each participant, the
pivoting would look like what is shown in @fig-pivot-longer-id.

![Pivot longer in `{tidyr}`, excluding an `id` column. New columns are
called `name` and `value`, as well as the old `id` column. Notice how,
unlike the previous image, the `id` column is excluded when pivoting
into the data on the
right.](/images/pivot-longer-id.svg){#fig-pivot-longer-id
width="90%" fig-alt="A diagram showing the use of the pivot longer
function, where a table with three columns, id, A and B, is converted to
a table with three columns, id, name and value. The original columns
values are stacked in the value column and the original columns names
are stacked in the name column, but the original id column remains, but
with the ids stacked on top of each other."}

Pivoting is a conceptually challenging thing to grasp, so don't be
discouraged if you don't understand how it works yet or of the ways you
could use it. As you practice using it, you will slowly begin to
understand it. With `pivot_longer()`, the first argument is the data
itself. The other arguments are:

1. `cols`: The columns to convert to long form. The input is a vector
   made using `c()` that contains the column names, like you would use
   in `select()` (e.g. you can use the `{tidyselect}` helpers like
   `starts_with()`, or `-` minus to exclude).
2. `names_to`: Optional, the default is `name`. If given, it will be the
   name of the newly created column (as a quoted character) that
   contains the original column names.
3. `values_to`: Optional, the default is `value`. Like `names_to`, sets
   the name of the new columns.

Both `pivot_longer()` and its opposite `pivot_wider()`, which we will
cover later in the session, are incredibly powerful functions. We can't
show close to everything it can do in this workshop, but if you want to
learn more, read up on the
[documentation](https://tidyr.tidyverse.org/articles/pivot.html) for it.

{{< text_snippet sticky_up >}}

## Pivot to long form

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Emphasise that there are a lot of really powerful ways you can use
`pivot_longer()` and to point out looking to the documentation to see
some examples of that.

There are several code chunks below. You should **not** create a new
code chunk for each one. Instead, write the code within a single chunk
while teaching, unless the text explicitly says otherwise.
:::

Over the last few sessions, we have prepared our HR-type data. But it is
in long format, which in this case is in a tidy format (where each value
in a row and column represents a distinct observation at a given time).
And while our survey results data is also in a tidy format, it is in a
wider format and doesn't allow us to join with the HR-type data.

While everyone read in the `survey-results.csv.gz` data in a previous
session's exercise, let's make sure we're all together and read it in
again. So, go to the bottom of the `docs/learning.qmd` file and create a
new header called `## Pivot longer`. Below that, create a new code chunk
with {{< var keybind.chunk >}}. Then, in that code chunk, read in the
survey results data again with and run the code with
{{< var keybind.run-code >}}:

```{r}
#| label: read-survey-results
#| filename: "docs/learning.qmd"

survey_data <- here("data-raw/nurses-stress/survey-results.csv.gz") |>
  read()
survey_data
```

Let's examine the data and make some observations about it. We see that
each participant has potentially multiple rows of data, one row for
times when they were surveyed on their stress levels. There are also
multiple date and time columns: `start_time`, `end_time`, `duration`,
and `date`. After that are columns for the responses to the survey
questions for that survey time period.

What we want is to have a single column for the ID and a single column
for a combined datetime in a way that matches the format of the HR-type
data. That means we have to pivot the date and time columns into a long
format and then expand them so that we have a row for each minute within
the start and end time. Before we can pivot though, we need to process
the data a bit. First, the `date` column is actually a character column.
So we need to convert that to an actual date. Second, because we want a
datetime column, we will need to create two new columns for
`start_datetime` and `end_datetime`. Let's fix the date first.

```{r}
#| label: convert-date-column
#| filename: "docs/learning.qmd"

survey_data |>
  mutate(date = as_date(date))
```

Hmm, that didn't work as expected, now the dates are all `NA`! Dates are
really tricky to work with and this is an example of that. The reason it
didn't work is because the `date` column is in the format of
`month/day/year`, but the `as_date()` function is expecting it to be in
the format of `year-month-day`. Thankfully, the `{lubridate}` package
has the `mdy()` function (for month-day-year) that can handle this
format. So we can use that instead.

```{r}
#| label: convert-date-column-fixed
#| filename: "docs/learning.qmd"

survey_data |>
  mutate(date = mdy(date))
```

Nice! Ok, let's create those datetime columns now. We can use the
`as_datetime()` function to convert the date and time columns into
datetimes. Within `as_datetime()`, we can use `paste()` to combine the
date and time columns together into a single character string, which can
then be converted to a datetime. Let's do that for both the start and
end datetime columns. We'll use the `.before` argument to place the new
columns before the `start_time` column, mostly to make it easier to see
them.

```{r}
#| label: create-datetime-columns
#| filename: "docs/learning.qmd"

survey_data |>
  mutate(
    date = mdy(date),
    start_datetime = as_datetime(paste(date, start_time)),
    end_datetime = as_datetime(paste(date, end_time)),
    .before = start_time
  )
```

Great, we've combined them together. Now we don't need the `date`,
`start_time`, and `end_time` columns anymore since we combined them
together, so we can drop those. And since the start and end datetimes
provide the same information as `duration`, we can drop that. But we
still have some unique way of identify each observation of participant
at a given time. So let's add an explicit copy of `start_datetime` and
call it `datetime_id`, to ensure we always have an explicit unique set
of identifiers for later processing. Let's also assign the data to a new
variable called `survey_tidy` so we can work on this data frame for the
rest of the session.

```{r}
#| label: keep-only-datetime-columns
#| filename: "docs/learning.qmd"

survey_tidy <- survey_data |>
  mutate(
    date = mdy(date),
    start_datetime = as_datetime(paste(date, start_time)),
    end_datetime = as_datetime(paste(date, end_time)),
    datetime_id = start_datetime,
    .before = start_time
  ) |>
  select(-c(date, start_time, end_time, duration))
```

Great, we have the datetimes in the format we want and no columns that
we don't need. But now, we need to get the data into the format to match
the HR-type data. We'll need at least a column for the participant ID
and a column for the datetime, which includes both start and end time.
Since we want to have a common dataset to join all data together, let's
select only the columns we need for that, which are the `id`,
`datetime_id`, `start_datetime`, and `end_datetime` columns. It also
makes it easier to see what we're pivoting.

```{r}
#| label: select-for-pivoting
#| filename: "docs/learning.qmd"

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime)
```

And then we can trying to pivot the data longer:

```{r}
#| label: try-pivot-longer
#| filename: "docs/learning.qmd"
#| error: true

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime) |>
  pivot_longer()
```

This give an error because we haven't told it which columns to use yet.
We want to pivot only the date columns. We can do this by using `c()`
for each of the columns we want to pivot.

```{r}
#| label: pivot-only-date-columns
#| filename: "docs/learning.qmd"

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime) |>
  pivot_longer(c(start_datetime, end_datetime))
```

It seems to have worked. We see that there are two new columns created
called `name` and `value`. These aren't very descriptive names though.
If we look at the help documentation for `pivot_longer()`, we can see
that it has an argument called `names_to` and `values_to` that allow use
to rename these new columns to something else. If we look at the `name`
column, it contains the original column names `start_datetime` and
`end_datetime`. With the `id`, the `value` (which is the actual datetime
value), and the `duration` column, we have a unique "key" for each
observation of a participant at a given date and time. So we don't
actually need the `name` column. With `names_to` we can set it to `NULL`
so it doesn't create it. Let's do that. We'll have to remove the `name`
value in the `relocate()` function too.

```{r}
#| label: pivot-only-date-columns-no-name
#| filename: "docs/learning.qmd"

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime) |>
  pivot_longer(c(start_datetime, end_datetime), names_to = NULL)
```

Better, but let's rename the `value` column to `collection_datetime`, to
match the other HR-type datasets. We can do that with the `values_to`
argument in `pivot_longer()`.

```{r}
#| label: pivot-only-date-columns-rename-value
#| filename: "docs/learning.qmd"

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime) |>
  pivot_longer(c(start_datetime, end_datetime), names_to = NULL, values_to = "collection_datetime")
```

We're getting there. But our HR-type data has rows in minutes, but here
we only have the start and end times. So now we need to convert the
`collection_datetime` column so that it has rows per minute, from the
start time to the end time. The `{tidyr}` package has many wonderful
functions that help tidy up data, including a function for this
particular situation. The `complete()` function will "complete" (fill
in) the columns you give it, converting implicit missing values to
explicit missing values. This is exactly what we're "missing" here: the
rows for each minute for each participant. Like `mutate()` and
`summarise()`, the `complete()` function works by assigning values to
columns. But unlike `mutate()` and `summarise()`, you can't create new
columns with `complete()`. You have to apply it to an existing column,
which is the `collection_datetime` column.

We also need to give this function the actual values that should be
present in the specific column, which is the sequence of datetimes from
the start to the end time by 1 minute. In this case, we can use the
`seq()` function from the start time (`min()`) to the end time
(`max()`), with a step of 60 seconds (since `seq()` interprets a
datetime as seconds). Let's try it out:

```{r}
#| label: complete-dates-seconds
#| filename: "docs/learning.qmd"

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime) |>
  pivot_longer(
c(start_datetime, end_datetime),
    names_to = NULL,
    values_to = "collection_datetime"
  ) |>
  complete(
    collection_datetime = seq(
      min(collection_datetime),
      max(collection_datetime),
      by = 60
    )
  )
```

Ok, so it made a row for each minute, but now we have a lot of other
missing data. And it seems that it took the `min()` and `max()` of
**all** the datetime values, rather than the min and max for each
participant. We want to fill in the missing datetimes for each
participant for each period of time. So we can use `group_by()` to group
the data for everything except the `collection_datetime` column, which
is the column we want to complete. Rather than writing out all the
columns to group by, we can use the `pick()` function from `{dplyr}` to
select all the columns except for the `collection_datetime` column.
We'll also `ungroup()` afterwards:

```{r}
#| label: group-and-complete
#| filename: "docs/learning.qmd"

survey_tidy |>
  select(id, datetime_id, start_datetime, end_datetime) |>
  pivot_longer(
    c(start_datetime, end_datetime),
    names_to = NULL,
    values_to = "collection_datetime"
  ) |>
  group_by(pick(-collection_datetime)) |>
  complete(
    collection_datetime = seq(
      min(collection_datetime),
      max(collection_datetime),
      by = 60
    )
  ) |>
  ungroup()
```

Nice! We now have our dataset in a format that allows us to join it with
the other datasets, including both the original survey results data and
the HR-type data. Time to make it into a function and include within the
`docs/cleaning.qmd` file.

## :technologist: Exercise: Convert into two new functions

**Time: \~20 minutes.**

Time to convert the code into functions. There were two distinct "steps"
we did to the survey data. The first was tidying up the dates and times
to prepare it for pivoting and the second was pivoting into a longer
format. So convert those into two separate functions. They should be
used as shown below:

```r
survey_tidy <- tidy_survey_dates(survey_data)
survey_long <- survey_to_long(survey_tidy)
```

You'll eventually take those functions, put them into `R/functions.R`
and use them in the `docs/cleaning.qmd` file to process the survey data.
Start with the date-tidying code:

1. As you've done throughout this workshop, convert the code into a
   function called `tidy_survey_dates`, with one argument `data` inside
   `function()`.
2. Copy and paste the code we just wrote into the body of the function.
   Replace the dataset name `survey_data` with the `data` argument and
   assign the output to `tidied`.
3. Explicitly link each function you use inside the function to its
   packages with `::`. There should be three packages used within the
   function.
4. There's one package you haven't added to the dependencies. Add it
   with `usethis::use_package()`.
5. Add Roxygen documentation with {{< var keybind.roxygen >}}.
6. Run `{styler}` in your `docs/learning.qmd` with
   {{< var keybind.style >}}.
7. Run the function with {{< var keybind.run-code >}} and test it out in
   the Console with `tidy_survey_dates(survey_data)`. It should give you
   the same output as before.
8. Once it works fine, cut and paste the function into `R/functions.R`.
9. Add and commit the changes to the Git history with
   {{< var keybind.commit >}}. Then push to GitHub.

Now do the same thing for the pivoting code:

1. Convert the code into a function called `survey_to_long()`, with one
   argument `data` inside `function()`.
2. Copy and paste the code we just wrote into the body of the function.
   Replace the dataset name `survey_tidy` with the `data` argument and
   assign the output to `longer`.
3. Link the two packages used to the functions with `::`.
4. Add Roxygen documentation with {{< var keybind.roxygen >}}.
5. Run `{styler}` in your `docs/learning.qmd` with
   {{< var keybind.style >}}.
6. Run the function with {{< var keybind.run-code >}} and test it out in
   the Console with `survey_to_long(survey_tidy)`. It should give you
7. Cut and paste the function into `R/functions.R`.
8. Add and commit the changes to the Git history with
   {{< var keybind.commit >}}. Then push to GitHub.

In the next exercise, you'll use these new functions in the
`docs/cleaning.qmd` file to process the survey data.

```{r}
#| label: solution-add-tidyr
#| filename: "Console"
#| eval: false

usethis::use_package("tidyr")
```

```{r}
#| label: solution-tidy-survey-dates
#| output: false

#' Tidy up the dates in the survey results
#'
#' @param data A data frame of the survey results data.
#'
#' @return A data frame.
tidy_survey_dates <- function(data) {
  tidied <- data |>
  dplyr::mutate(
    date = lubridate::mdy(date),
    start_datetime = lubridate::as_datetime(paste(date, start_time)),
    end_datetime = lubridate::as_datetime(paste(date, end_time)),
    datetime_id = start_datetime,
    .before = start_time
  ) |>
  dplyr::select(-c(date, start_time, end_time, duration))
  return(tidied)
}

survey_data |>
  tidy_survey_dates()
```

```{r}
#| label: solution-survey-to-long
#| output: false

#' Pivot survey data to longer format, with only IDs and datetimes
#'
#' @param data A data frame of the data from `tidy_survey_dates()`.
#'
#' @returns A data frame.
survey_to_long <- function(data) {
  longer <- data |>
    dplyr::select(id, datetime_id, start_datetime, end_datetime) |>
    tidyr::pivot_longer(
      c(start_datetime, end_datetime),
      names_to = NULL,
      values_to = "collection_datetime"
    ) |>
    dplyr::group_by(dplyr::pick(-collection_datetime)) |>
    tidyr::complete(
      collection_datetime = seq(
        min(collection_datetime),
        max(collection_datetime),
        by = 60
      )
    ) |>
    dplyr::ungroup()
  return(longer)
}

survey_data |>
  tidy_survey_dates() |>
  survey_to_long()
```

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Add survey processing to `docs/cleaning.qmd`

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Once everyone has finished this exercise, walkthrough it to make sure
everyone is aligned and briefly explain again why we use the
`docs/cleaning.qmd` file (because it's the more "central" file to clean
and prepare the data, while the `docs/learning.qmd` file is more to use
as a sandbox).
:::

**Time: \~10 minutes.**

Now that you've tidying up the next piece, time to build up the
`docs/cleaning.qmd` more. So, open up the `docs/cleaning.qmd` file (you
can open files quickly with {{< var keybind.open-file >}}).

1. On the bottom of `docs/cleaning.qmd`, create a new header called
   `## Process survey data`. Create a new code chunk below that with
   {{< var keybind.chunk >}}.
2. In that code chunk, pipe `here::here()` to the
   `survey-results.csv.gz` file into `read()`, then pipe that into
   `tidy_survey_dates()`. Assign the output to `survey_tidy`.
3. Then below that code, pipe `survey_tidy` into `survey_to_long()`, and
   assign that to `survey_long`.
4. Style the code with {{< var keybind.style >}}.
5. Render the Quarto document with {{< var keybind.render >}} to confirm
   that everything works.
6. Once it works fine, add and commit the changes to the Git history
   with {{< var keybind.commit >}}. Then push to GitHub.

```{r}
#| label: solution-add-survey-processing
#| filename: "docs/cleaning.qmd"

survey_tidy <- here::here("data-raw/nurses-stress/survey-results.csv.gz") |>
  read() |>
  tidy_survey_dates()

survey_long <- survey_tidy |>
  survey_to_long()
```

{{< text_snippet sticky_up >}}

## :technologist: Extra exercise: Simplify code with `{tidyselect}` helpers

**Time: \~5 minutes.**

If you've finished the previous exercises early and have some time, try
simplify the code in the `survey_to_long()` function by using
`{tidyselect}` helpers rather than writing out the column names with
`c()`. To make it more challenging, we won't give anymore hints nor
solution code.

{{< text_snippet sticky_up >}}

## :book: Reading task: Pivoting data to a wider form

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Like with the pivoting to long section, let them read through this
section first and then go over it again to verbally explain it more,
making use of the graphs to help illustrate what is happening.
:::

**Time: \~10 minutes.**

While we won't use this function in this workshop, it's still useful to
read of `pivot_longer()`'s opposite function, `pivot_wider()`, which
converts to the wider format. Working in long form can be really useful
when doing certain types of wrangling, like you've read about above and
actually used. The majority of the time, you probably will only ever use
`pivot_longer()` and this is part of the reason why we don't cover it.
But, sometimes you need to pivot to a form that is a bit wider to do,
for example, when doing certain types of joins or processing.

The arguments for `pivot_wider()` are very similar to those in
`pivot_longer()`, except instead of `names_to` and `values_to`, they are
called `names_from` and `values_from`. Like with many `{tidyverse}`
functions, the first argument is the data and the other arguments are:

1. `id_cols`: This is optional as it will default to all column names.
   This argument tells `pivot_wider()` to use the given columns as the
   identifiers for when converting. Unlike `pivot_longer()` which
   doesn't require some type of "key" or "id" column to convert to long
   form, the conversion to wide form requires some type of "key" or "id"
   column because `pivot_wider()` needs to know which rows belong with
   each other.
2. `names_from`: Similar to the `pivot_longer()`, this is the name of
   the column that has the values that will make up the new columns.
   Unlike with the `names_to` argument in `pivot_longer()` which takes a
   character string as input, the column name for `names_from` must be
   *unquoted*, because you are selecting a column that already exists in
   the dataset (unlike `names_to` which is the name of a column that you
   *want to create* but that doesn't exist yet).
3. `values_from`: Same as `names_from`, this is the column name (that
   exists and must be given *unquoted*) for the values that will be in
   the new columns.

@fig-pivot-wider visually shows what's happening when using
`pivot_wider()`.

![Pivot wider in `{tidyr}`, where a set of stacked "groups" in the data
on the left are placed side-by-side as new columns in the output data on
the right. Notice how the values in the column `N`, which is used in the
`names_from` argument, are used as the names for the new columns `A` and
`B` in the new
data.](/images/pivot-wider.svg){#fig-pivot-wider
width="90%" fig-alt="A diagram showing the use of the pivot wider
function, where a table with three columns, id, name, and value, is
converted to a table with three columns, id, A and B. The values in the
column name are now the names of the columns in the new table: A and B.
The original id column still exists."}

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey. The previous exercise
will likely be challenging for them, so we will not do a discussion
activity this time.
:::

- Data is usually structured to varying degrees as wide or long format.
  Thinking of your data in terms of shifting it around in structure is a
  powerful and useful skill to have when doing data wrangling and
  analysis.
- When needing to convert to the longer form, use `pivot_longer()` from
  the `{tidyr}` package.
- If you need to convert to the wider form, use the `{tidyr}` function
  `pivot_wider()`.
- Seek out and test different functions within the `{tidyr}` package, it
  has many very helpful functions for tidying up data such as the
  `complete()` function.

```{r}
#| label: admin-save-session-code
#| include: false

save.image(here::here("_temp/pivots.RData"))
```

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
