# Pivoting your data from and to long or wide {#sec-pivots}

```{r setup}
#| include: false
# Used this only during writing to run code dependent on previous chapters.
library(here)
library(tidyverse)
load(here("_temp/split-apply-combine.RData"))
```

## Learning objectives

{{< include /includes/objectives/_pivots.qmd >}}

## :book: Reading task: Pivoting data for easier wrangling

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Briefly walk through it again after they've read it and explain it a bit
more, making use of the tables and graphs. Pivoting is often a difficult
concept to grasp, to it's good to go over it a few times. Emphasise the
difficulty of it, to remind them it is ok to not get it.
:::

**Time: \~6 minutes.**

Rarely is our data in the exact format we need or want it to be in to
effectively process, analyse, or visualize it. Often we'll need to do
some heavy reshaping. One very powerful way to reshape data is with
"pivoting". Pivoting is when you take a dataset's rows and convert them
to columns or convert columns to rows, and in many different
combinations of this.

The `{tidyr}` package within the `{tidyverse}` contains two wonderful
functions for pivoting: `pivot_longer()` and `pivot_wider()`. There is a
well written documentation on pivoting in the [tidyr
website](https://tidyr.tidyverse.org/articles/pivot.html) that can
explain more about it that you can read at another time.

Most use cases of pivoting are to convert to the longer form, which you
can do with `pivot_longer()` and is what we will cover first in this
session. There are many reasons pivoting longer is the most common pivot
direction.

An example is if you ever need to enter data. Entering data, especially
in a spreadsheet tool, in the wide form is much easier and more time
efficient than entering data in long form. Consider the scenario where
you are entering data from a participant for glucose values that were
measured every 30 minutes. So you might enter data like this:

| person_id | glucose_0 | glucose_30 | glucose_60 |
|:----------|:----------|:-----------|:-----------|
| 1         | 5.6       | 7.8        | 4.5        |
| 2         | 4.7       | 9.5        | 5.3        |
| 3         | 5.1       | 10.2       | 4.2        |

: This is a **wide** dataset that is useful for data entry.

If working with a spreadsheet, is this is to use tab to move to the next
cell and continue entering the data. However, when it comes time to
analyze the data, this wide form is very inefficient and difficult to
computationally and statistically work with. So, we do data entry in
wide and use functions like `pivot_longer()` to get the data ready for
analysis. See @fig-pivot-longer below that visually shows what happens
when you pivot from wide to long.

![Pivot longer in `{tidyr}`. New columns are called `name` and `value`.
Notice how the values in `A` and `B` columns are stacked on top of each
other in the newly created `V`
column.](/images/pivot-longer.png){#fig-pivot-longer width="90%"
fig-alt="A diagram showing the use of the pivot longer function, where a table with two columns, A and B, is converted to a table with two columns, name and value. The original columns values are stacked in the value column and the original columns names are stacked in the name column."}

If you had, for instance, an ID column for each participant, the
pivoting would look like what is shown in @fig-pivot-longer-id.

![Pivot longer in `{tidyr}`, excluding an `id` column. New columns are
called `name` and `value`, as well as the old `id` column. Notice how,
unlike the previous image, the `id` column is excluded when pivoting
into the data on the
right.](/images/pivot-longer-id.png){#fig-pivot-longer-id
width="90%"
fig-alt="A diagram showing the use of the pivot longer function, where a table with three columns, id, A and B, is converted to a table with three columns, id, name and value. The original columns values are stacked in the value column and the original columns names are stacked in the name column, but the original id column remains, but with the ids stacked on top of each other."}

Pivoting is a conceptually challenging thing to grasp, so don't be
discouraged if you can't understand how it works yet or of the ways you
could use. As you practice using it, you will slowly begin to understand
it. With `pivot_longer()`, the first argument is the data itself. The
other arguments are:

1.  `cols`: The columns to use to convert to long form. The input is a
    vector made using `c()` that contains the column names, like you
    would use in `select()` (e.g. you can use the `select_helpers` like
    `starts_with()`, or `-` minus to exclude).
2.  `names_to`: Optional, the default is `name`. If provided, it will be
    the name of the newly created column (as a quoted character) that
    contains the original column names.
3.  `values_to`: Optional, the default is `value`. Like `names_to`, sets
    the name of the new columns.

Both `pivot_longer()` and its opposite `pivot_wider()`, which we will
cover later in the session, are incredibly powerful functions. We can't
show close to everything it can do in this workshop, but if you want to
learn more, read up on the
[documentation](https://tidyr.tidyverse.org/articles/pivot.html) for it.

{{< text_snippet sticky_up >}}

## Pivot to long form

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Emphasise that there are a lot of really powerful ways you can use
`pivot_longer()` and to point out looking to the documentation to see
some examples of that.
:::

Our CGM data is already in a long format, but we eventually want to join
it with the participant details data. The participant details data is in
a wide format. Let's start with importing it in the `setup` code chunk
of our `docs/learning.qmd` file:

```{r import-participant-details}
#| filename: "docs/learning.qmd"
participant_details <- here("data-raw/dime/participant_details.csv") |>
  import_dime()
```

Run the code in the `setup` code chunk by running each line with
{{< var keybind.run-code >}}. Then, let's move to the end of our
`docs/learning.qmd` file and create a new header called
`## Pivot longer`. Below that, create a new code chunk with
{{< var keybind.chunk >}}. We'll start with seeing what the data looks
like:

```{r show-participant-details}
#| filename: "docs/learning.qmd"
participant_details
```

You can see that the data, while the data has the `intervention` column
in the long format, the dates are in the wide format, with `start_date`
and `end_date`. Our CGM and sleep only has a single `date` column, so we
can't yet join these datasets with `participant_details`. We first need
to pivot the data to the long format. Let's add a pipe to
`participant_details` to output into `pivot_longer()`:

```{r try-pivot-longer}
#| filename: "docs/learning.qmd"
#| error: true
participant_details |>
  pivot_longer()
```

This give an error because we haven't told it which columns to use yet.
We want to pivot only the date columns. We can do this using
`{tidyselect}` helpers, for instance the helper function `ends_with()`
is perfect for this case, since both date columns end with `date`.

```{r pivot-only-date-columns}
#| filename: "docs/learning.qmd"
participant_details |>
  pivot_longer(ends_with("date"))
```

Awesome! We now have a long format with dates in a single column. But we
also have new column names that are a bit ambiguous, like `name` and
`value`. We can tell `pivot_longer()` to give them more descriptive
names with `names_to` and `values_to`. But, we also don't need the
`name` column, since implicitly we know that the last date is the end
date and the first date is the start date. So, we can set
`names_to = NULL` to drop the new `name` column from being created.

```{r pivot-only-date-columns-renamed}
#| filename: "docs/learning.qmd"
participant_details |>
  pivot_longer(ends_with("date"), names_to = NULL, values_to = "date")
```

Great! But, maybe you may have noticed something else here. If we want
to join this data with the CGM data, we need to have a single row for
each day of the study that the participant has data for. Right now, the
`participant_details` data has only the date for the first day and for
the last day. So if we join the CGM or sleep data, we would only get the
data for those two dates. What we want is to have all the days included
between the start and end date for each participant. The `{tidyr}`
package has many wonderful functions that help tidy up data, including a
function for this particular situation! The function is called
`complete()` because it will fill in data where you want a sequence with
the missing dates in that sequence. For our situation, that is filling
in all the missing dates between the start and end date for each
participant.

Since we want to complete the data for each participant for each
intervention, we will need to group the data. We could select each
column to group by, but we could also use the `pick()` function we used
before to select all the columns except for the `date` column.

```{r group-pivoted-data}
#| filename: "docs/learning.qmd"
participant_details |>
  pivot_longer(ends_with("date"), names_to = NULL, values_to = "date") |>
  group_by(pick(-date))
```

We can now use `complete()` to fill in the dates. The `complete()`
function is used like `mutate()` or `summarise()`. Unlike these other
functions, you need to provide it with a column that exists, and you
can't create new columns. And you also need to provide it with the
values that you want completed. In our case, we want each date from the
start date to the end date, one for each day. We can use the `seq()`
function to create a sequence of dates, from the `min()` to the `max()`.
So we'd write it like:

```{r complete-dates}
#| filename: "docs/learning.qmd"
participant_details |>
  pivot_longer(ends_with("date"), names_to = NULL, values_to = "date") |>
  group_by(pick(-date)) |>
  complete(
    date = seq(min(date), max(date), by = "1 day")
  )
```

We now have a dataset that we can join with the other datasets! Time to
make it into a function :grin:

## :technologist: Exercise: Convert the date completion code into a function

**Time: \~10 minutes.**

{{< include /includes/solutions/_pivots-date-completion.qmd >}}

Time to convert it into a function :grin: You've practiced this several
times before, and the more you practice, the easier it will be to do! At
the end of the exercise, you should create the function so that it can
be used like this in the `setup` code chunk:

```{r}
#| filename: "docs/learning.qmd"
participant_details <- here("data-raw/dime/participant_details.csv") |>
  import_dime() |>
  clean_participant_details()
```

And if you were to run this code in the **Console**, it would look like
this:

```{r}
#| filename: "Console"
participant_details
```

1.  Since we are using `{tidyr}`, use `usethis::use_package()` to add it
    as a dependency to the `DESCRIPTION` file.
2.  Create a new function named `clean_participant_details` by assigning
    `function()` to it.
3.  Include only `data` as an argument in `function()`.
4.  Move the code into the function body and replace the
    `participant_details` variable with the `data` argument.
5.  Assign it to a new variable called `cleaned` and `return()` it at
    the end of the function.
6.  Find out which packages all the functions come from by using `?` and
    explicitly assign the package to the function with `::`. There
    should be four packages used in this code.
7.  Create Roxygen documentation for the function with
    {{< var keybind.roxygen >}}. Remember, the documentation doesn't
    have to be perfect or complete. Just to practice using it.
8.  Test that the code works by running the function and using it like
    shown above.
9.  Once it works as intended, cut and paste it over into
    `R/functions.R`.
10. Pipe the `import_dime()` output into this new function in the
    `setup` code chunk of `docs/learning.qmd`.
11. Style both `R/functions.R` and `docs/learning.qmd` with
    {{< var keybind.style >}}.
12. Render the document with {{< var keybind.render >}} to make sure it
    all works.
13. Once everything works fine, commit your changes to the Git history
    with {{< var keybind.commit >}} and push it to your GitHub.

{{< text_snippet sticky_up >}}

## :book: Reading task: Pivoting data to a wider form

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Like with the pivoting to long section, let them read through this
section first and then go over it again to verbally explain it more,
making use of the graphs to help illustrate what is happening.
:::

**Time: \~10 minutes.**

Working in long form can be really useful when doing certain types of
wrangling, like you've seen and read about above. The majority of the
time, you probably will only ever use `pivot_longer()`. But, sometimes
you need to pivot to a form that is a bit wider to do, for example, when
doing certain types of joins or processing.

For us, the sleep data is in a very long form, where each hour of sleep
has several rows for the different types of sleep. Joining the data in
this form to the other datasets, while possible, won't produce usable
output. We want to convert it so there is only one row per hour of a
participant's day, rather than multiple rows per hour. For that, we need
to convert to the longer form.

The arguments for `pivot_wider()` are very similar to those in
`pivot_longer()`, except instead of `names_to` and `values_to`, they are
called `names_from` and `values_from`. Like with many `{tidyverse}`
functions, the first argument is the data and the other arguments are:

1.  `id_cols`: This is optional as it will default to all column names.
    This argument tells `pivot_wider()` to use the given columns as the
    identifiers for when converting. Unlike `pivot_longer()` which
    doesn't require some type of "key" or "id" column to convert to long
    form, the conversion to wide form requires some type of "key" or
    "id" column because `pivot_wider()` needs to know which rows belong
    with each other.
2.  `names_from`: Similar to the `pivot_longer()`, this is the name of
    the column that has the values that will make up the new columns.
    Unlike with the `names_to` argument in `pivot_longer()` which takes
    a character string as input, the column name for `names_from` must
    be *unquoted*, because you are selecting a column that already
    exists in the dataset (unlike `names_to` which is the name of a
    column that you *want to create* but that doesn't exist yet).
3.  `values_from`: Same as `names_from`, this is the column name (that
    exists and must be given *unquoted*) for the values that will be in
    the new columns.

@fig-pivot-wider visually shows what's happening when using
`pivot_wider()`.

![Pivot wider in `{tidyr}`, where a set of stacked "groups" in the data
on the left are placed side-by-side as new columns in the output data on
the right. Notice how the values in the column `N`, which is used in the
`names_from` argument, are used as the names for the new columns `A` and
`B` in the new data.](/images/pivot-wider.png){#fig-pivot-wider
width="90%"
fig-alt="A diagram showing the use of the pivot wider function, where a table with three columns, id, name, and value, is converted to a table with three columns, id, A and B. The values in the column name are now the names of the columns in the new table: A and B. The original id column still exists."}

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Create a new function to pivot sleep to wider

**Time: \~25 minutes.**

{{< include /includes/solutions/_pivots-sleep-types-to-wider.qmd >}}

This will a bit more of a challenging exercise since we haven't used the
`pivot_wider()` function before. But the way you use it is very similar
to `pivot_longer()`. We've had you practice the process many times and
this will be an exercise to really challenge what you've learned so far!
Which is part of the learning journey :grin:

For this exercise, you will create a new function that converts the
sleep data into a wider form. To test it out, it should look like this
and should output something similar:

```{r}
#| filename: "docs/learning.qmd"
sleep_data |>
  sleep_types_to_wider()
```

Once you've made this new function, move it into the `clean_sleep()` so
it looks like:

```{r admin-explanation-clean-sleep}
#| filename: "R/functions.R"
clean_sleep <- function(data) {
  cleaned <- data |>
    get_participant_id() |>
    dplyr::rename(datetime = date) |>
    prepare_dates(datetime) |>
    # The function used to sum summarise may be different for you.
    summarise_column(seconds, list(sum = sum)) |>
    # This is the new function you will add below.
    sleep_types_to_wider()
  return(cleaned)
}
```

Then, you go to your `setup` code chunk in `docs/learning.qmd` and run
all the code in it, especially:

```{r}
#| filename: "docs/learning.qmd"
sleep_data <- here("data-raw/dime/sleep/") |>
  import_csv_files() |>
  clean_sleep()
```

It should like this if you run it in the **Console**:

```{r}
#| filename: "Console"
sleep_data
```

1.  Look into the documentation for `pivot_wider()` to get a better idea
    for how it works.
2.  Create a new header called `## Pivot wider` in the
    `docs/learning.qmd` file, then create a new code chunk below it with
    {{< var keybind.chunk >}}.
3.  Before making the function, prototype and test out code by piping
    the `sleep_data` into `pivot_wider()`. Try using `names_from` and
    `values_from` to get the right output. You can use the argument
    `names_prefix` to add text to the start of the new `names` columns.

After you've written some code that works, use the same process you've
used before to make a function:

1.  Create a new function called `sleep_types_to_wider` and assign
    `function()` to it.
2.  Include `data` as an argument in `function()`.
3.  Replace the `sleep_data` variable with the `data` argument.
4.  Assign it to a new variable called `wider` and `return()` it at the
    end of the function.
5.  Explicitly indicate the package the function uses by using
    `tidyr::`.
6.  Create Roxygen documentation for the function with
    {{< var keybind.roxygen >}}.
7.  Test that the code works by running the function, using it like
    `sleep_types_to_wider(sleep_data)`.
8.  When it works, cut and paste it over into `R/functions.R`.
9.  Include the new function in the `clean_sleep()` function in
    `R/functions.R` so that it looks like the above.
10. Style both `R/functions.R` and `docs/learning.qmd` with
    {{< var keybind.style >}}.
11. Render the document with {{< var keybind.render >}} to make sure it
    all works.
12. And lastly, commit your changes to the Git history with
    {{< var keybind.commit >}} and push it to your GitHub.

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey. The previous exercise
will likely be challenging for them, so we will not do a discussion
activity this time.
:::

-   Data is usually structured to varying degrees as wide or long
    format. Thinking of your data in terms of shifting it around in
    structure is a powerful and useful skill to have when doing data
    wrangling and analysis.
-   When needing to convert to the longer form, use `pivot_longer()`
    from the `{tidyr}` package.
-   If you need to convert to the wider form, use the `{tidyr}` function
    `pivot_wider()`.
-   Seek out and test different functions within the `{tidyr}` package,
    it has many very helpful functions for tidying up data such as the
    `complete()` function.

```{r admin-save-session-code}
#| include: false
save.image(here::here("_temp/pivots.RData"))
```

{{< include /_extensions/rostools/r3-theme/includes/code-appendix-r.qmd >}}
