# Processing strings with regular expressions {#sec-regex}

```{r}
#| label: setup
#| include: false

library(here)
library(fs)
library(tidyverse)
load(here("_temp/functionals.RData"))
source(here("R/functions.R"))
hr_data <- read_all("HR.csv.gz")
survey_data <- here("data-raw/nurses-stress/survey-results.csv.gz") |>
  read()
```

We've read a lot of our data into R so far. But we now have several
individual data frames that we want to be able to analyse together. If
you recall from the discussion activity we had in the introduction
session, we want to have a single data frame that has variables for data
like HR and BVP as well as the nurses' stress levels and other details.
That way, we can analyse the relationships between these variables. For
example, what is the relationship between HR and stress? Which means we
want to join them together, but the question is, how do we do that?

## Learning objectives

{{< include /includes/objectives/_regex.qmd >}}

## :technologist: Exercise: Inspect and compare the data frames

**Time: \~5 minutes.**

You haven't yet read in the `survey-results.csv.gz` file, which is the
main file that connects everything else together. So, read it into R and
compare the data values and structure to the HR data. Try to identify
the issues that prevent you from joining the datasets together right
now. Use this code as a scaffold.

```r
hr_data <- read_all("HR.csv.gz")

survey_data <- ___("___") |>
  ___()

hr_data
survey_data
```

1. In your `docs/learning.qmd` file, go to the bottom of the file and
   create a new header called `## Inspecting the data`. Then create a
   new code chunk below that with {{< var keybind.chunk >}}.
2. Copy and paste the code scaffold above into the code chunk you
   created.
3. Fill in the blanks so that you can `read()` in the
   `data-raw/nurses-stress/survey-results.csv.gz` file and assign it to
   an object called `survey_data`. (Hint: you can use `here()` to pipe
   in the file path.)
4. Run the code to see the two data frames in the **Console** and
   compare them to identify the issues that prevent you from joining
   them together right now.

::: {.callout-note collapse="true" appearance="minimal" title="**Click for the solution**. Only click if you are struggling or are out of time."}
```{r}
#| label: solution-compare-data
#| filename: "docs/learning.qmd"
#| output: false

hr_data <- read_all("HR.csv.gz")

survey_data <- here("data-raw/nurses-stress/survey-results.csv.gz") |>
  read()

hr_data
survey_data

```
:::

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: Why can't we join datasets together yet?

::: {.callout-note collapse="true"}
## :teacher: Teacher note

The issues stopping us from joining the datasets together are:

- HR-type data doesn't have an `id`, but does have a `file_path_id` with
  the participant ID contained within it.
- HR-type data has `collection_datetime` in long format, but the survey
  data has `start_time`, `end_time`, and `date`, but all in one row in
  wide format.
- The survey data `date` is a character, not a date.
- The survey data time columns have the smallest unit of time in minutes
  but the HR-type data has the smallest unit of time in seconds.
- The HR-type data is in long format but the survey data is in wide
  format.

:::

**Time: \~6 minutes.**

You've inspected the data frames to try and identify why we can't join
them together right now. With a neighbour, discuss what you've found and
see if you can come to a consensus on what the issues are. Then we will
share together.

2. For about 3 minutes, discuss what you've found when comparing the
   datasets with your neighbour
3. Then for about 3 minutes we will all share and discuss together.

## :book: Reading task: Cleaning character data with regular expressions

**Time: \~5 minutes.**

Before we go into joining datasets together, we have to do a bit of
processing first. Specifically, we want to get the participant ID from
the `file_path_id` character data since all the datasets have an ID (in
some form). Whenever you are processing and cleaning data, you will very
likely encounter and deal with character data. A wonderful package to
use for working with character data is called `{stringr}`, which we'll
use to get the participant ID from the `file_path_id` column. The
package is called `{stringr}` because in programming, a "string" is
defined as any collection of character objects. So we will be using
"string" from now on.

The main "engine" behind the functions in `{stringr}` is called [regular
expressions](https://en.wikipedia.org/wiki/Regular_expression) (or regex
for short). You've already used regex when you used `dir_ls()` with the
`regexp` argument in the previous session. These expressions are
powerful, very concise ways of finding patterns in text. But they are
also *very very difficult* to learn, write, and read, even for
experienced users. That's because certain strings like `[` or `?` have
special meanings. For instance, `[aeiou]` is regex for "find one string
in the object that is either a, e, i, o, or u". The `[]` in this case
means "find the string in between the two brackets". We won't cover
regex too much in this workshop, but some great resources for learning
them are the [R for Data Science regex
section](https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions),
the [stringr regex
page](https://stringr.tidyverse.org/articles/regular-expressions.html),
as well as in the help doc `?regex`. There's also a nice [cheat
sheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf) on
it. But we will cover some of the more commonly used regex patterns,
like in @tab-regex:

  | regex       | Meaning                                                                                                         |
  | ----------- | --------------------------------------------------------------------------------------------------------------- | --------------------------- | --------------------------------------------------- | ----------------------------------------------- |
  | `\\d`       | find one digit (0-9)                                                                                            |
  | `\\D`       | find one non-digit                                                                                              |
  | `[0-9]`     | find one digit between 0 and 9 (the `-` is a range)                                                             |
  | `[a-z]`     | find one letter between a and z (lowercase)                                                                     |
  | `[A-Z]`     | find one letter between A and Z (uppercase)                                                                     |
  | `[a-zA-Z]`  | find one letter between a and z or A and Z                                                                      |
  | `\\.`       | find one dot string                                                                                             |
  | `$`         | find the end of the string, e.g. `\\.$` means find a dot that is at the end of the string                       |
  | `^`         | find the start of the string, e.g. `^\\d` means find a digit at the start of the string                         |
  | `?`         | maybe find the string before the `?`, e.g. `\\d?` means maybe find one digit                                    |
  | `*`         | find the string before it, but maybe many times, e.g. `\\d*` means find zero or many digits one after the other |
  | `+`         | find the string one or more times before `{}`, e.g. `\\d+` means find one or more digits one after the other    |
  | `{n}`       | find the same string `n` times before `{}`, e.g. `\\d{2}` means find two digits one after the other             |
  | `[:alpha:]` | find one letter (a-z or A-Z)                                                                                    |
  | `[:digit:]` | find one digit (0-9)                                                                                            |
  | `[:alnum:]` | find one alphanumeric (a-z, A-Z, 0-9)                                                                           |
  | `[:upper:]` | find one uppercase letter (A-Z)                                                                                 |
  | `[:lower:]` | find one lowercase letter (a-z)                                                                                 |
  | `(A         | B)`                                                                                                             | find *either* `A` or `B` (` | ` means "or" and the `()` groups them), e.g., `(\\D | \\d)` means find either a non-digit or a digit. |

  : A list of some common regex patterns and their meanings.
    {#tab-regex}

::: callout-note
The `\` string is a special string in regex, but in R it means "escape
or use this literally as the string". For instance, `\"` in R means "use
the `"` string", like if you wanted to write a message like:

```{r}
print("Analysis has \"nearly\" finished.")
```

Without the `\`, R would see this instead:

```r
print("Analysis has "nearly" finished.")
```

And think `"Analysis has "` is the first string, that `nearly` is an
object, and the `" finished"` is the second string object.

But in regex, the `\` string is a special command that can mean
different things. For instance, `\d` means "find a digit" (0-9) and `\D`
means "find a non-digit". So when using it in R, it would be `\\d` and
`\\D` so that regex "sees", e.g., `\d`
:::

{{< text_snippet sticky_up >}}

## :speech_balloon: Discussion activity: Brainstorm a regex for getting the participant ID

**Time: \~8 minutes.**

Using what you just read from above, try to think of a regex that would
get the participant ID from the file path. Try not to look ahead nor at
the potential solutions! :grin: :wink:

When the time is up, we'll share some ideas and go over what the regex
*could* be (there are a few different ways to do this). With your
neighbour, do the following tasks:

1. Looking at the `file_path_id` column, list what is similar in the
   user ID between rows and what is different.
2. Discuss and verbally describe (in English, not regex) what text
   pattern you might use to extract the user ID.
3. Using what you just read about regex and the tips below, think about
   how you might convert the English description of the text pattern to
   a regex. This will probably be very hard, but try anyway.
   - Use `[]` to find one possible string of the several between the
     brackets. E.g. `[12]` means 1 *or* 2 or `[ab]` means "a" or "b". To
     find a range of numbers or letters, use `-` in between the start
     and end ranges, e.g. `[1-3]` means 1 to 3 or `[a-c]` means "a" to
     "c".
   - Use `{n}` to find exactly `n` occurrences of the previous pattern.
     E.g. `a{2}` means "aa" or `\\d{3}` means three digits in a row.

Once you've done these tasks, we'll discuss all together and go over
what the regex could be to extract the user ID.

::: {.callout-note appearance="minimal" collapse="true"}
## Potential regex solutions

- `(\\d|\\D)+`
- `(\\d|\\D){2}`
- `[0-9A-Z]+`
- `[0-9A-Z]{2}`
- `[:alnum:]+`
- `[:alnum:]{2}`
- `([:upper:]|[:digit:]){2}`

To be very explicit and certain, we could add `/stress/` to the start of
the regex and `/` to the end to make certain we get the correct part of
the path that has the ID. The simplest one from above is `[:alnum:]{2}`
since it is easier to read compared to the others. With the `/stress/`
and `/` added, it would be `/stress/[:alnum:]{2}/`. Adding those things
at the start and end does mean we will need to have a couple other steps
to remove those after extracting the ID. But there are nice functions
from `{stringr}` to help out.

:::

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Make sure to reinforce that while regex is incredibly complicated, there
are some basic things you can do with it that are quite powerful.

More or less, this section and exercise are to introduce the idea and
concept of regex, but not to really teach it since that is well beyond
the scope of this workshop and this time frame.

Go over the solution. The explanation is that the pattern will find
anything that has two of a letter and/or a number that is preceded by
`/stress/` and followed by `/`.
:::

## Using regular expressions to extract text

Now that we've identified a possible regex to use to extract the
participant ID, let's try it out on the `hr_data`.

We want to create a new column called `id` for the participant ID rather
than modify the current `file_path_id` column, so we will use the
`mutate()` function from the `{dplyr}` package to create this new
column. We'll use the `str_extract()` function from the `{stringr}`
package to "extract a string" by using the regex `/stress/[:alnum:]{2}/`
that we discussed from the activity above. We'll also use a few
`str_remove()` functions to remove the `/stress/` and `/` from the
extracted string so that we just have the ID.

We're also using an argument to `mutate()` you might not have seen
previously, called `.before`. This will insert the new `id` column
before the column we use and we do this entirely for visual reasons,
since it is easier to see the newly created column when we run the code.
In your `docs/learning.qmd` file, create a new header called
`## Using regex for ID` at the bottom of the document, and create a new
code chunk below that with {{< var keybind.chunk >}}.

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Walk through writing this code, briefly explain/remind how to use
mutate, and about the `{stringr}` function.
:::

First we'll pipe `hr_data` into `mutate()` to create a new column called
`id` that will contain the participant ID. We'll start with using the
`str_extract()` function to extract the participant ID from the
`file_path_id` column using the regex `/stress/[:alnum:]{2}/`.

```{r}
#| label: get-id
#| filename: "docs/learning.qmd"
#| eval: false

# Note: your file paths and data may look slightly different.
hr_data |>
  mutate(
    id = str_extract(
      file_path_id,
      # "/stress/[:alnum:]{2}/"
      "(?<=/stress/)[:alnum:]{2}(?=/)"
    ),
    .before = file_path_id
  )
```

```{r}
#| label: admin-get-id
#| echo: false

hr_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "/stress/[:alnum:]{2}/"
    ),
    .before = file_path_id
  )
```

Cool! But we have the `/stress/` and `/` that we don't want. We can use
the `str_remove()` function to remove both from the string. So we can
pipe our output into that function (twice):

```{r}
#| label: get-id-remove-csv
#| filename: "docs/learning.qmd"
#| eval: false

hr_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "/stress/[:alnum:]{2}/"
    ) |>
      str_remove("/stress/") |>
      str_remove("/"),
    .before = file_path_id
  )
```

```{r}
#| label: admin-get-id-remove-csv
#| echo: false

hr_data |>
  trim_filepath_for_book() |>
  mutate(
    id = str_extract(
      file_path_id,
      "stress/[:alnum:]{2}/"
    ) |>
      str_remove("stress/") |>
      str_remove("/"),
    .before = file_path_id
  )
```

Nice! Almost there. We don't actually want to keep the `file_path_id`
column, so we can use the `select()` function to drop it. We can do this
by using the `-` operator to drop the column. So we can pipe our output
into that function:

```{r}
#| label: drop-file-path-id
#| filename: "docs/learning.qmd"

hr_data |>
  mutate(
    id = str_extract(
      file_path_id,
      "/stress/[:alnum:]{2}/"
    ) |>
      str_remove("/stress/") |>
      str_remove("/"),
    .before = file_path_id
  ) |>
  select(-file_path_id)
```

We've done it! We now have a new column called `id` that contains the
participant ID from the `file_path_id` column. Now, time to make it into
a function!

## :technologist: Exercise: Convert 'get ID' code into a function

**Time: \~10 minutes.**

You now have code that takes the data that has the `file_path_id` column
and gets the participant ID from it. You want to be able to easily use
that on any of the HR-type data frames. So, as you've done, convert this
code in the `docs/learning.qmd` file into a function.

By the end of this exercise, the code to read in any of the HR-type data
should look like this:

```r
hr_data |>
  get_participant_id()
# Or with one of the other datasets:
read_all("BVP.csv.gz") |>
  get_participant_id()
```

So, convert the code we just wrote into a function by following these
steps:

1. Assign `function()` to a new named function called
   `get_participant_id` and include one argument in `function()` with
   the name `data`.
2. Put the code we just wrote into the body of the function. Make sure
   to `return()` the output at the end of the function and to create
   Roxygen documentation. Replace all the relevant variables with the
   `data` argument.
3. Add `stringr` and `dplyr` as a package dependency by using
   `usethis::use_package()` in the **Console**.
4. Explicitly link the functions you are using in this new function to
   their package by using `::` (e.g. `dplyr::` and `stringr::`).
   Remember that you can find which package a function belongs by using
   the `?` help documentation and looking at the very top to see the
   package name.
5. Test that the function works by using `get_participant_id(hr_data)`.
6. After creating the function and testing that it works, move (**cut**
   and paste) the function into `R/functions.R`.
7. Run `{styler}` in both the `R/functions.R` file and
   `docs/learning.qmd` with {{< var keybind.style >}}.
8. Render the `docs/learning.qmd` file to make sure things remain
   reproducible with {{< var keybind.render >}}.
9. Restart your R session with {{< var keybind.restart-r >}} to make
   sure later work is in a clean state.
10. Add and commit the changes to the Git history with
    {{< var keybind.commit >}}. Then push to GitHub.

::: {.callout-note collapse="true" appearance="minimal" title="**Click for the solution**. Only click if you are struggling or are out of time."}
The function should look like this in the `R/functions.R` file:

```{r}
#| label: solution-participant-id-function
#| filename: "R/functions.R"
#| output: false

#' Get the participant ID from the file path column.
#'
#' @param data Data with `file_path_id` column.
#'
#' @returns A data frame/tibble.
#'
get_participant_id <- function(data) {
  data_with_id <- data |>
    dplyr::mutate(
      id = stringr::str_extract(
        file_path_id,
        "/stress/[:alnum:]{2}/"
      ) |>
        stringr::str_remove("/stress/") |>
        stringr::str_remove("/"),
      .before = file_path_id
    ) |>
    dplyr::select(-file_path_id)
  return(data_with_id)
}
```

And you should be able to use it like this in either the Console or in
the `docs/learning.qmd` file:

```{r}
#| label: solution-use-get-id-function
#| filename: "Console"

# How it will be used.
get_participant_id(hr_data)
```
:::

{{< text_snippet sticky_up >}}

## :technologist: Extra exercise: Simplify the regex with look arounds

**Time: \~8 minutes.**

If you finished the previous exercise early, you can try this more
challenging exercise. Try to simplify the regex so that it only extracts
the two alphanumeric characters without also getting the `/stress/` and
`/`. That way, you won't need to use the `str_remove()` functions to
remove those parts afterwards. This requires looking in the regex
documentation about something called [look
arounds](https://stringr.tidyverse.org/articles/regular-expressions.html#look-arounds).
Modify the internals of `get_participant_id()` to use these look arounds
so that you only use one `str_extract()` function and no `str_remove()`
functions. The output should be the same as before, but the code should
be simpler.

1. Look up the documentation for regex look arounds and try to
   understand how they work.
2. Modify the regex in the `str_extract()` function to use look arounds
   so that it finds the two alphanumeric characters that are preceded by
   `/stress/` (look behind) and followed by `/` (look ahead), without
   including those in the output.
3. Remove the two `str_remove()` functions since they should no longer
   be needed.
4. Test out the updates to the function by `source()`'ing the
   `R/functions.R` file with {{< var keybind.source >}} and running
   `get_participant_id(hr_data)` in the **Console**.
5. Once it works, add and commit the changes to the Git history with
   {{< var keybind.commit >}}. Then push to GitHub.

::: {.callout-note collapse="true" appearance="minimal" title="**Click for the solution**. Only click if you are struggling or are out of time."}
The function should look like this in the `R/functions.R` file:

```{r}
#| label: solution-look-arounds
#| filename: "R/functions.R"
#| output: false

get_participant_id <- function(data) {
  data_with_id <- data |>
    dplyr::mutate(
      id = stringr::str_extract(
        file_path_id,
        "(?<=/stress/)[:alnum:]{2}(?=/)"
      ),
      .before = file_path_id
    ) |>
    dplyr::select(-file_path_id)
  return(data_with_id)
}
```

It should still be usable in the same way as before:

```{r}
#| label: solution-use-look-arounds
#| filename: "Console"
#| output: false

# How it will be used.
get_participant_id(hr_data)
```
:::

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- While very difficult to learn and use, regular expressions (regex or
  regexp) are incredibly powerful at processing character (also called
  strings) data. Some strings have special meanings like `[]`, `?`, `*`,
  and `+` that can be used to find patterns in string data.
- Use the `{stringr}` package to work with string data as it has many
  functions to help use regex to process string data, like
  `str_extract()` or `str_remove()`.

```{r}
#| label: admin-store-session-code
#| include: false

save.image(here::here("_temp/regex.RData"))
```

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
