# Using split-apply-combine to help in processing {#sec-split-apply-combine}

```{r}
#| label: setup
#| include: false

library(here)
library(tidyverse)
load(here("_temp/regex.RData"))
```

In the last session, we used regular expressions to add a `id` column to
the HR-type data so that we have an identifier between the HR-type
datasets and the survey data. By having this identifier, we're one step
closer to joining them together. But we still have several other things
to process first. The next one we will do it process the HR-type data so
that the date times (in seconds) match the time units in the survey data
(in minutes). For that, we need to be able to summarise the HR-type data
into a value per minute.

## Learning objectives

{{< include /includes/objectives/_split-apply-combine.qmd >}}

## :book: Reading task: Split-apply-combine technique and functionals

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Verbally cover this section before moving on to using `across()` while
summarizing. Specifically, review the split-apply-combine technique via
the image below as well as in the context of using `summarise()` on
multiple columns and multiple functions.
:::

**Time: \~10 minutes.**

We're taking a quick detour to briefly talk about a concept that
perfectly illustrates how vectorization and functionals fit into doing
data analysis. The concept is called the
[split-apply-combine](https://r-cubed-intro.rostools.org/sessions/split-apply-combine.html)
technique, which we covered in the beginner R workshop. The method is:

1. Split the data into groups (e.g. diabetes status).
2. Apply some analysis or statistics to each group (e.g. finding the
   mean of age).
3. Combine the results to present them together (e.g. into a data frame
   that you can use to make a plot or table).

So when you split data into multiple groups, you create a list (or a
*vector*) that you can then apply (e.g. with the *map* functional) a
statistical technique to each group through *vectorization*, and where
you finally combine (e.g. with join that we will cover later or with
`list_rbind()`). This technique works really well for a range of tasks,
including for our task of summarizing some of the DIME data so we can
merge it all into one dataset.

![A diagram showing how a data frame is split up, an action is applied to the splits that outputs a new result, and the results are combined back together. Taken from [Software Carpentries R for Reproducible Scientific Analysis](https://unlhcc.github.io/r-novice-gapminder/16-plyr/).](/images/split-apply-combine.png)

Functionals and vectorization are integral components of how R works and
they appear throughout many of R's functions and packages. They are
particularly used throughout the `{tidyverse}` packages like `{dplyr}`.
Let's get into some more advanced features of `{dplyr}` functions that
work as functionals.

There are many "verbs" in `{dplyr}`, like `select()`, `rename()`,
`mutate()`, `summarise()`, and `group_by()` (covered in more detail in
the
[introductory](https://r-cubed-intro.rostools.org/sessions/wrangling)
workshop). The common usage of these verbs is through acting on and
directly using the column names (e.g. without `"` quotes around the
column name). Like most `{tidyverse}` functions, `{dplyr}` verbs are
designed with a strong functional programming approach. But many
`{dplyr}` verbs can also be used as functionals like we covered in
previous sessions with `map()`, where they take functions as input. For
instance, `summarise()` uses several functional programming concepts:
Create a new column using an action that may or may not be based on
other columns and output a single value from that action. Using an
example with our DIME data, to calculate the mean of a column and create
a new column from that mean for the glucose values, you would do:

```{r}
hr_data |>
  summarise(mean_hr = mean(hr))
```

This is *declarative* because we ask R to give us the mean of `hr` using
the `mean()` function and to output a data frame with only one row using
`summarise()`. If we wanted to calculate another summary statistic, like
the standard deviation, we would just add another line to the
`summarise()` function:

```{r}
hr_data |>
  summarise(
    mean_hr = mean(hr),
    sd_hr = sd(hr)
  )
```

Repeating this pattern works fine with a few columns and maybe one or
two summary statistics. But what if you wanted to calculate the mean,
standard deviation, the median, and maybe also the maximum and minimum
values? And what if you wanted to this for several different columns? If
you repeated the above code, that will be a lot of (annoying)
repetition. That's where the power of functional programming and
applying functionals (also called higher-order functions) shows its
strength.

In the [functionals session](/sessions/functionals.qmd) in
@sec-functionals, you used `map()` to apply a function to a list of
paths to import the data. In `{dplyr}`, there is a function that works
almost exactly like `map()`, but is used specifically for applying
functions to columns in a data frame. This function is called
`across()`.

Unlike in `map()`, where you have to give it a list that contains
different objects, in `across()` you give it a vector of columns
instead. Just like `map()`, you also give `across()` a function (as an
object without `()`) to apply to the columns.

While `map()` is a more general-purpose function, `across()` is
specifically designed to *only* work in `{dplyr}` verbs like `mutate()`
or `summarise()` and within the context of a data frame.

{{< text_snippet sticky_up >}}

## Summarising with `across()`

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Open up `?across` in the Console and go over the arguments with the
learners. Go over the first two arguments again, reinforcing what they
read.

Also, before coding again, remind everyone that we still only import the
first 100 rows of each data file. So if some of the data itself seems
weird, that is the reason why. Remind them that we do this to more
quickly prototype and test code out.
:::

Before we start using `across()`, let's look at the help page for it. In
the **Console**, type `?across` and hit enter. This will open the help
page for `across()`. For this workshop, we will only go over the first
two arguments. The first argument is the columns you want to work on.
You can use `c()` to combine multiple columns together. The second
argument is the function you want to apply to those columns. You can use
`list()` to combine multiple functions together.

Let's try out using `across()` on the `hr_data`. Just to confirm that
we're using the same data, let's read in the HR data and run
`get_participant_id()` on it again. So, go to the bottom of your
`docs/learning.qmd` file and create a new header called
`## Summarising with across` and create a code chunk below that with
{{< var keybind.chunk >}}.

```{r}
#| label: read-hr-with-id
#| filename: "docs/learning.qmd"

hr_data <- read_all("HR.csv.gz") |>
  get_participant_id()
```

Then, below that, we will do a very simple example of using `across()`
to calculate the mean of the `hr` values.

```{r}
#| label: simple-summarise-mean
#| filename: "docs/learning.qmd"

hr_data |>
  summarise(across(hr, mean))
```

This is nice, but let's try also calculating the median. The help
documentation of `across()` says that you have to wrap multiple
functions into a `list()`. Let's try that:

```{r}
#| label: summarise-mean-median
#| filename: "docs/learning.qmd"

hr_data |>
  summarise(across(hr, list(mean, median)))
```

It works, but, the column names are not helpful. The column names are
`hr_1` and `hr_2`, which doesn't tell us which is which. We can add the
names of the functions by using a named list. In our case, a named list
would look like:

::: {.callout-note collapse="true"}
## :teacher: Teacher note

They don't need to write this code in the Console, just explain to them
what the code means and does, while showing the output. Same with the
code below with the anonymous function.
:::

```{r}
#| label: named-list-console
#| filename: "Console"

list(mean = mean)
# or
list(average = mean)
# or
list(ave = mean)
```

See how the left hand side of the `=` is the name and the right hand
side is the function? This named list is what `across()` can use to add
the name to the end of the column name. Just like with `map()` we can
also give it anonymous functions if our function is a bit more complex.
For instance, if we needed to remove `NA` values from the calculations,
we would do:

```{r}
#| label: named-list-console-anonymous
#| filename: "Console"

list(mean = \(x) mean(x, na.rm = TRUE))
```

For now, we don't need to use anonymous functions. Let's all try using a
named list with only one function, like `list(mean = mean)`:

```{r}
#| label: example-summarise-mean-named-list
#| filename: "docs/learning.qmd"

hr_data |>
  summarise(across(hr, list(mean = mean)))
```

Then, we can add more functions to the named list. Let's add the median
and standard deviation:

```{r}
#| label: example-summarise-multiple-functions
#| filename: "docs/learning.qmd"

hr_data |>
  summarise(
    across(
      hr,
      list(mean = mean, sd = sd, median = median)
    )
  )
```

The great thing about using `across()` is that you can also use it with
all the `{tidyselect}` functions in the first argument. In our case, we
have datasets with columns that have different names. So if we made this
code into a function, it would only work for datasets that have a `hr`
column. That's not very useful. Instead, we can use `{tidyselect}`
functions to tell `across()` to apply the functions to columns based on
some condition. Since all of our HR-type data have columns that are
numeric, we can use the `where()` function with `is.numeric` (notice the
lack of `()`). Let's try it out by replacing `hr` with
`where(is.numeric)`:

```{r}
#| label: use-where
#| filename: "docs/learning.qmd"

hr_data |>
  summarise(
    across(
      where(is.numeric),
      list(mean = mean, sd = sd, median = median)
    )
  )
```

Running that code gives us the same output as before, but now it's
generic enough that we can use it with any other the other datasets.
Neat :grin:

But, having only one row for the summary statistic for us isn't very
useful. Summarising is very powerful when combined with grouping. Which
is what we'd like to do in order to effectively join the data. But
before we can do that, we need to talk about dates and times.

## Finding the right date

Working with datetime data can be surprisingly difficult and tricky.
There's timezones, daylight savings time, leap years, and different date
formats that all can cause a dozen different problems. On top of that,
but datetimes can have nanosecond and second level precision. In R,
there is a wonderful package called `{lubridate}` that helps with
working with dates and times. It can't solve all date and time problems,
but it can help with a many of them. And, thankfully, `{lubridate}` is
part of the `{tidyverse}` ecosystem, so we don't need to explicitly load
it.

In the `docs/learning.qmd` file, at the bottom of the document create a
new code chunk called `## Working with dates` and below that create a
new code chunk with {{< var keybind.chunk >}}. We'll use `mutate()`
again to create a column called `datetime`, as well as `.before` to put
it before the other columns.

Our `collection_datetime` column is already in a datetime format, but it
is in seconds, while the survey data is in minutes. So, we need to
convert the seconds into minutes. There are many ways to do this, but
the easiest is to use the `round_date()` function from `{lubridate}`.
This function rounds a datetime to the nearest unit of time. So, we can
use it to round the `collection_datetime` to the nearest minute. The
`round_date()` function takes two arguments: the datetime column and the
unit of time to round to (e.g. "minute", "hour", "day", etc.). It can
even be used to round to the nearest 5 minutes (by using `"5 mins"`) or
30 seconds (with `"30 secs"`). Let's round to the nearest minute by
using `round_date()` with the `unit` argument set to `"minute"`:

```{r}
#| label: round-date-minute

hr_data |>
  mutate(
    datetime = lubridate::round_date(collection_datetime, unit = "minute"),
    .before = collection_datetime
  )
```

Our next step is to summarise by this new `datetime` column, so we have
the measurement values per minute rather than per second. First, let's
add `{lubridate}` as a package dependency:

```{r}
#| label: add-lubridate-dependency
#| filename: Console
#| eval: false

usethis::use_package("lubridate")
```

## Summarising by groups

We used `summarise()` with the functional `across()` to calculate
summary statistics for columns based on a condition for the
"apply-combine" part. But now, we can do the "split" part for the
datetimes by minute. In `{dplyr}`, one way to use the
split-apply-combine technique is by using the `summarise()` function
with its `.by` argument.

::: callout-tip
You can also pipe `group_by()` into `summarise()`, but you have to
remember to use `ungroup()` afterwards. Using `group_by()` can be useful
if you want to do multiple actions on the grouped dataset, since
`{dplyr}` functions `mutate()` and `summarise()` also work on grouped
data. At the end, you'd always have to `ungroup()` and if you forget, it
can give wrong output that you might not realise.

If you only want to do a single action on groups, you can instead use
the `.by` argument in `summarise()` or in `mutate()`. This is a more
concise way of doing the `group_by()` and `ungroup()` workflow, since
the grouping only happens in the `summarise()` or `mutate()` function.
You can learn more about `.by` in the help page by running `?dplyr_by`
in the Console.
:::

Let's combine what we did with `round_date()` and what we did using
`summarise()` and `across()`. At the bottom of the `docs/learning.qmd`
file, create a new header called `## Summarising by groups` and create a
code chunk below that with {{< var keybind.chunk >}}. Rather than create
a new column for a variable called `datetime`, let's replace the
existing `collection_datetime` column with the rounded version. And at
the end of the `summarise()` function, let's add the `.by` argument and
set it to both `id` and `collection_datetime` within `c()`. This will
split the data up by the `id` and `collection_datetime` columns, apply
the summary statistics to each group, and then combine the results back
together to then output that new data frame.

```{r}
#| label: remove-device-timestamp
#| filename: "docs/learning.qmd"

hr_data |>
  mutate(
    collection_datetime = lubridate::round_date(collection_datetime, unit = "minute")
  ) |>
  summarise(
    across(
      where(is.numeric),
      list(mean = mean, sd = sd, median = median)
    ),
    .by = c(id, collection_datetime)
  )
```

The `.by` argument can be used similar to how `select()` is used, so you
can use `{tidyselect}` functions to select the columns you want to group
by. If you wanted to group by more than one column, you'd wrap the
columns in `c()`. For our purposes, we don't need to do that.

Amazing! Before continuing to the exercise, let's run `{styler}` with
{{< var keybind.style >}} on both `docs/learning.qmd` and on
`R/functions.R`. Then we will render the Quarto document with
{{< var keybind.render >}} to confirm that everything runs as it should.
If the rendering works, switch to the Git interface and **add and
commit** the changes so far with {{< var keybind.commit >}} and push to
GitHub.

## :technologist: Exercise: Convert the summarising code into a function

**Time: \~10 minutes.**

It's your turn to convert the code into a function so that you can use
it on the other datasets. It should be able to work like this for the HR
data:

```r
hr_data <- read_all("HR.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()
```

You should be able to do the same thing for the other datasets (IBI,
TEMP, BVP, and EDA) using this function.

Here's a scaffold of the function to help you get started:

```r
summarise_by_datetime <- function(___) {
  summarised_data <- ___ |>
    # Fill in below with the code we just wrote.
    ___
  return(summarised_data)
}
```

1. Take the code we just wrote and convert it into a function. Use the
   code above as a scaffold.
2. Inside the `function()`, add one argument called `data`.
3. Copy and paste the code we just wrote into the body of the function.
   Replace the dataset name `hr_data` with the `data` argument.
4. Explicitly link each function you use inside the
   `summarise_by_datetime()` function to its package with `::`. There
   should be three packages used within the function.
5. There's one package you haven't added to the dependencies. Add it
   with `usethis::use_package()`.
6. Add Roxygen documentation with {{< var keybind.roxygen >}}.
7. Run `{styler}` in your `docs/learning.qmd` with
   {{< var keybind.style >}}.
8. Run the function with {{< var keybind.run-code >}} and test it out in
   the Console with `summarise_by_datetime(hr_data)`. It should give you
   the same output as before.
9. Once it works fine, cut and paste the function into `R/functions.R`.
10. Add and commit the changes to the Git history with
    {{< var keybind.commit >}}. Then push to GitHub.

```{r}
#| label: solution-summarise-by-datetime-function
#| echo: true

# Add the package dependency for tidyselect
# usethis::use_package("tidyselect")

#' Summarise values in a data frame by a rounded datetime
#'
#' @param data A data frame with at least a `collection_datetime` column and
#'  some numeric columns to summarise.
#'
#' @returns A summarised data frame.
#'
summarise_by_datetime <- function(data) {
  summarised_data <- data |>
    dplyr::mutate(
      collection_datetime = lubridate::round_date(collection_datetime, unit = "minute")
    ) |>
    dplyr::summarise(
      dplyr::across(
        tidyselect::where(is.numeric),
        list(mean = mean, sd = sd, median = median)
      ),
      .by = c(id, collection_datetime)
    )
  return(summarised_data)
}

hr_data <- read_all("HR.csv.gz") |>
  get_participant_id() |>
  summarise_by_datetime()

```

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Start filling in `docs/cleaning.qmd`

**Time: \~10 minutes.**

You're now at a point where you have functions to read in and process
all of the HR-type data. So let's start filling in the
`docs/cleaning.qmd` file. This is the main document you'll use to keep
all the code that will clean and process the data into a single dataset.
So, open up the `docs/cleaning.qmd` file (you can open files quickly
with {{< var keybind.open-file >}}).

1. In the `setup` code chunk, add the
   `source(here::here("R/functions.R"))` line to load in the functions
   you have created so far.
2. On the bottom of `docs/cleaning.qmd`, create a new header called
   `## Prepare HR data`. Create a new code chunk below that with
   {{< var keybind.chunk >}}.
3. In that code chunk, use `read_all()`, `get_participant_id()`, and
   `summarise_by_datetime()` to read in and process the HR data by
   piping them together with `|>`. Assign the output to a new variable
   called `hr_data`.
4. Run the code chunk with {{< var keybind.run-code >}} to confirm it
   works.
5. For the other datasets (IBI, TEMP, BVP, and EDA), repeat steps 2-4 to
   read in and process those datasets as well. Their variable names
   should be `ibi_data`, `temp_data`, `bvp_data`, and `eda_data`
   respectively.
6. Render the Quarto document with {{< var keybind.render >}} to confirm
   that everything runs as it should.
7. Once it works fine, add and commit the changes to the Git history
   with {{< var keybind.commit >}}. Then push to GitHub.

In later exercises, you'll come back to this `docs/cleaning.qmd` file to
add more code to continue processing the data and eventually join it all
together.

<!-- TODO: Add solution chunk -->

{{< text_snippet sticky_up >}}

## :technologist: Exercise: Create a function to read in any HR-type dataset

**Time: \~10 minutes.**

If you've finished the previous exercises early and want to do something
a bit harder, try out this exercise. You might have noticed that the
code you wrote above to read in each of the HR-type datasets is very
similar. The only thing that changes is the filename. So, you can create
a function that takes in a filename as an argument and then reads in the
data, adds the participant id, and summarises it by datetime. This
function should be able to work with any of the HR-type datasets. Name
the new function `read_sensor_data()`.

::: callout-tip
If you did the extra exercise in one of the previous sessions and added
the `max_rows` argument to the `read_all()` function, add that argument
to this new function as well and set it to `100` by default. Make sure
to pass this argument to the `read_all()` function within the body of
the new function.
:::

Make sure to follow workflows we've been using throughout this workshop:
styling the R code, testing the code, committing, and pushing to GitHub.
Because this is an optional, extra exercise, we don't include a solution
for this exercise.

<!-- TODO: Add potential solution -->

{{< text_snippet sticky_up >}}

## :technologist: Extra exercise: Modify the function so you can set your own unit of time

**Time: \~8 minutes.**

If you've finished the previous exercises early and want to do something
a bit harder, try out this exercise. Modify the
`summarise_by_datetime()` function in the `R/functions.R` file so that
you can set your own unit of time to round to. For instance, you might
want to round to the nearest 5 minutes or 30 seconds. You can do this by
adding a new argument to the function called `unit` and then using that
argument in the `round_date()` function. This new argument should have a
default value of `"minute"` so that when we eventually join the datasets
together later on, than you don't have issues with the datasets not
matching up.

Make sure to add this new argument to the `read_sensor_data()` function
as well, so you can set the unit of time from there as you read in the
data.

Make sure to follow the workflows we've been using throughout this
workshop: styling the R code, testing the code, committing, and pushing
to GitHub. Because this is an optional, extra exercise, we don't include
a solution for this exercise.

<!-- TODO: Add potential solution -->

{{< text_snippet sticky_up >}}

## :technologist: Extra exercise: Modify the function so you can set your own summarising functions

**Time: \~5 minutes.**

If you've finished the previous exercises and the optional extra
exercise and there's some time left, try out this exercise if you want
to try something harder. Modify the `summarise_by_datetime()` in the
`R/functions.R` file so that you can set your own summarising functions.
To make it challenging, we won't give you more hints aside from showing
you how this function might look like when you use it:

```r
summarise_by_datetime(
  data = hr_data,
  fns = list(median = median)
)

summarise_by_datetime(
  data = hr_data,
  fns = mean
)

summarise_by_datetime(
  data = hr_data,
  fns = list(max = max, min = min)
)
```

Make sure to follow workflows we've been using throughout this workshop:
styling the R code, testing the code, committing, and pushing to GitHub.
Because this is an optional, extra exercise, we don't include a solution
for this exercise.

Good luck! :tada:

{{< text_snippet sticky_up >}}

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this and get them to do the survey before moving on to the
discussion activity.
:::

- The split-apply-combine technique is a powerful way to summarise and
  analyse data in R. It allows you to split your data into groups, apply
  a function to each group, and then combine the results back together.
  Re-framing how you think about your data using this technique can
  substantially help you in your data analysis.
- Use `summarise()`, its `.by` argument, and `across()` to use the
  split-apply-combine technique when needing to do an action on groups
  within the data.
- Build small functions that you pipe together to do more complex tasks.
- Use the `{lubridate}` package when working with dates and times, as
  there are many functions in it to help simplify date processing, like
  `round_date()`.

```{r}
#| label: admin-store-session-code
#| include: false

save.image(here::here("_temp/split-apply-combine.RData"))
```

{{< include /includes/_survey.qmd >}}

{{< include /includes/_code-appendix.qmd >}}
