# Create an R Project {#sec-create-project}

```{r}
#| label: setup
#| include: false

library(here)
# To trigger auto-linking
library(prodigenr)
library(fs)
temp_path <- fs::path_temp()
project_path <- fs::path(temp_path, "LearnR3")
```

One of the first, basic steps to ensuring a reproducible and modern data
analysis workflow is to keep everything self-contained in a single
location. In RStudio, this is done by using R Projects. Read all of this
[reading task](https://r-cubed-intro.rostools.org/sessions/setup-project#reading-task-what-is-a-project-and-why-use-it)
from the introduction workshop to learn about R Projects and how they
help keeping things self-contained.
*You don't need to do any of the exercises or activities*.

<!-- TODO: Convert this into a snippet? -->

::: callout-caution
Depending on your institution's policies or procedures, you may have
some type of file synching software like OneDrive or Dropbox installed
on your computer and your IT may have set it up so that (nearly) all
your files and folders are synched through this software. Sometimes IT
has set it up so it is fairly straightforward to not use it, but often,
it can be extremely difficult to find a folder that isn't being synched.
This is usually the case for Windows users and less for MacOS users, but
they can sometimes have these programs installed and setup for them too.

Unfortunately, these types of software can cause very subtle, small
issues when working with Git, R, and RStudio that are very difficult to
troubleshoot, for both you and for us as the teachers. They cause these
subtle issues because they are constantly working to synchronize files
and folders. This can sometimes cause conflicts with files when using
Git and especially when doing more data intensive work that creates and
deletes many files often. You may also notice that your computer is
slower when working with R and RStudio while using these synched
folders. That's because of the constant synching of files. It isn't the
fault of R and RStudio, but is instead because of the synching software.

A related issue is when you store your files on a shared drive, like
`H:` or `E:` or `U:` drives on Windows. This is again because your
institution may have a procedure for using these folders to "more easily
share with colleagues" or to keep everything on their servers. But these
shared drives have a very similar issue as using Dropbox or OneDrive,
because they also have to continually upload and download over the
internet from your computer to your work's remote server. Any time your
computer needs to regularly communicate over the internet, like when
using files stored on a server in the shared drives, things will be
slower. So, you may feel like RStudio is slow, but it is actually
because of the way your computer has been setup by your IT.

So in general, we tend to strongly recommend not creating your R
projects in these synched folders or in these shared drives. We've
learned that many, common problems we encounter during the workshops are
because of these synched folders or shared drives. That's why we will be
getting you to create your R Project on your Desktop *on your computer*
(not shared drive). Usually, but not always, this folder is not tracked
by these synching programs.

So find the path to your Desktop folder at
`C:\Users\yourusername\Desktop\` for Windows users and
`/Users/yourusername/Desktop/` for MacOS and use that as the location
for your R Project.
:::

There are several ways to organise a project folder. You'll be using the
structure from the package `{prodigenr}`. The project setup can be done
by either:

1. Using RStudio's New Project menu item: "File -\> New Project -\> New
   Directory", scroll down to "Scientific Analysis Project using
   prodigenr" and name the project "LearnR3" in the Directory Name,
   saving it to the "Desktop" as described above with Browse. When you
   use this method, the new RStudio Project will automatically open up
   for you.
2. Or, running the function
   `prodigenr::setup_project("~/Desktop/LearnR3")` in the R Console. The
   `~` is a shortcut for your home directory, so this function will
   create a new folder called `LearnR3` on your Desktop. When you use
   this method, you will need to use your file manager to navigate to
   the Desktop folder and open the `LearnR3.Rproj` file inside of the
   newly created `LearnR3/` folder.

```{r}
#| label: admin-create-project
#| filename: Console
#| include: false

prodigenr::setup_project(project_path)
```

After opening the new RStudio Project, run this next function in the R
Console to create a new R script file called `functions.R` in the `R/`
folder. This is where you will be moving all the functions that you will
create during the workshop.

```{r}
#| label: use-r-functions
#| eval: false
#| filename: "Console"

usethis::use_r("functions", open = FALSE)
```

```{r}
#| label: admin-add-setup-files-to-project
#| include: false

usethis::with_project(
  path = project_path,
  {
    usethis::use_r("functions", open = FALSE)
  }
)
```

Here you use the `{usethis}` package to help set things up. `{usethis}`
is an extremely useful package for managing R Projects and we highly
recommend checking it out more to see how you can use it more in your
own work.

## Quarto

We teach and use [Quarto](https://quarto.org) (which is a more powerful,
next generation version of R Markdown) because it is one of the very
first steps to being reproducible and because it is a very powerful tool
for doing data analysis. If you haven't used Quarto before or would like
a refresher, please read over the Quarto
[section](https://r-cubed-intro.rostools.org/sessions/reproducible-documents.html#what-is-quarto)
of the introduction workshop.

If you are not already in the `LearnR3` project, open it up by either
clicking the `LearnR3.Rproj` file in the `LearnR3/` folder or by using
the "File -\> Open Project" menu. Then, run the two functions below in
the Console when RStudio is in the `LearnR3` project, which will create
two new files called `learning.qmd` and `cleaning.qmd` in the `docs/`
folder. The `learning.qmd` file is where you will be building up your
code, prototyping functions, writing notes to yourself, and playing
around. The `cleaning.qmd` file is where you will be moving code that
works and using the functions you'll make during the workshop, to
ultimately make a final working dataset.

```{r}
#| label: create-qmd-doc
#| filename: "Console"
#| eval: false

r3::create_learning_qmd()
r3::create_cleaning_qmd()
```

```{r}
#| label: admin-create-qmd
#| include: false

usethis::with_project(
  path = project_path,
  {
    r3::create_learning_qmd()
    r3::create_cleaning_qmd()
  }
)
```

Throughout the workshop, you will use the `docs/learning.qmd` document
to build up and run your code, as well as to prototype the functions
you'll create before you move them into the `R/functions.R` file and
ultimately create the main file for processing and cleaning the dataset
in the `docs/cleaning.qmd` document.
