# Download the data {#sec-download-data}

```{r setup}
#| include: false
library(here)
library(fs)
temp_path <- fs::path_temp()
project_path <- fs::path(temp_path, "LearnR3")
```

In general, you want to store your original raw data that you'd use for
your project within a folder in your R Project and *never change it*.
This is because you want to keep the original data intact, so you can
always go back to it if you need to. Since the raw data is rarely
directly used for analyses, you'd use R to process the data and save the
processed data to a new file that you'd then use for your analysis. This
workflow is visually represented below in @fig-download-to-data-raw

In R Projects, the common convention is to keep the raw data in the
`data-raw/` folder, which is what you've already done during the
pre-workshop tasks. This is where you would download or copy the data
files you need for your project.

```{mermaid}
%%| echo: false
%%| label: "fig-download-to-data-raw"
%%| fig-cap: "Workflow for downloading or copying data to the `data-raw/` folder and processing it into the `data/` folder."
%%| fig-alt: "A diagram representing the workflow for downloading or copying data to the data-raw folder, then processed with R, and finally saved in the data folder."
flowchart TD
  data_file("Original<br>data file(s)") -- "Download<br>or copy" --> data_raw[/"data-raw/<br>folder"/]
  data_raw -- "Read only" --> process("Process data with R,<br>raw data not changed")
  process -- "Save"--> data[/"data/<br>folder"/]
```

To best demonstrate the concepts in the workshop, it's best to use a
real dataset to apply what you're going to learn. So for this workshop,
you're going to use an openly licensed dataset from a study looking at
how bioactive components in food impact our gut microbiota, which is
called the [DIME](https://zenodo.org/records/8268744) study
[@Traka2023].

Unfortunately, like most research data, it isn't the most tidy. So it's
been cleaned up a bit for you to use in this workshop.

::: callout-tip
Much of the concepts and skills we will cover in this workshop were used
to process and prepare this dataset for you to use in this workshop. If
you want to see the code "in action", check out the
[`data-raw/dime.R`](https://github.com/rostools/r-cubed-intermediate/blob/main/data-raw/dime.R)
script in this workshop's GitHub repository.
:::

Since this workshop is about being reproducible and applying modern
approaches to data analysis, you're going to start by writing and saving
R code to download the dataset your R project and save it to a folder
called `data-raw/`. During the workshop we'll be continuing to process
and prepare this dataset for analysis, though we won't do any
statistical analysis during this workshop.

The end goal for this workshop is that you will have a sequence of R
code written within the Quarto document that will take this raw data and
process it into a cleaned and prepared dataset. Processing data for
cleaning is something that you will do a lot of in your own work, since
most research data is not directly ready to be analysed. In the process
of processing the data, you'll also learn concepts and skills that can
be applied to many different tasks and situations.

For now, read the description of the DIME dataset that was taken from
the [DIME](https://zenodo.org/records/8268744) dataset page:

> "The DIME study consists of a randomised 2x2 cross-over human
> intervention where healthy participants (n = 20) are subjected to a
> diet high in bioactive-rich food for two weeks and a diet low in
> bioactive-rich food. There is a four-week washout between the two
> interventions.
>
> The continuous glucose monitoring was achieved using the Abbott
> \[freestyle\] libre flash glucose device. The baseline of the
> participants were determined 7 days before the start of the
> intervention, followed by the first arm and second arm. The period
> between the two arms (washout) was not recorded.
>
> We also included sleep data which consists of the amount of time spent
> in bed and during that time the amount of time spent in light, deep
> and rem in all 20 participants during the course of the dietary
> intervention, both the high and low bioactive diet. that was captured
> using \[FitBit\] wearables during both stages of the dietary
> intervention"

::: callout-note
You may be wondering, how can we be using this dataset that has human
health data? Isn't that against GDPR? The short answer is no, it isn't.

If you went to the dataset page, you may have noticed that the dataset
is openly licensed and public, which means you can use it for your own
purposes. Yes, it is possible to openly license and make health data
publicly available! And no it doesn't conflict with privacy laws like
GDPR. Unfortunately, the way that universities have communicated about
GDPR's impact on and role in research data hasn't been done very well.
They've communicated things in a very alarmist, risk-averse way and made
their recommendations or policies so generalised to the point of not
being useful to specific situations.

So, while GDPR does make it more strict on what can and cannot be shared
and requires more detailed information on how the participant's data
will be used when getting prior informed consent, it *does not prohibit
sharing it or making it public*! GDPR and open data are not strictly in
conflict with one another.
:::

The standard and conventional way of saving raw data in R is to save it
to the `data-raw/` folder. There is a helper function for that from
`{usethis}`. While in your `LearnR3` R Project, go to the R Console pane
in RStudio and type out:

```{r use-data-raw-dime}
#| filename: "Console"
#| eval: false
usethis::use_data_raw("dime")
```

```{r admin-create-data-raw}
#| include: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_data_raw("dime", open = FALSE)
  }
)
```

What this function does is create a new folder called `data-raw/` and
creates an R script called `dime.R` in that folder. This is where we
will store the mostly raw (though partly cleaned up) DIME data that
you'll get from the workshop's GitHub repository. After running this
function, the R script should open up for you, otherwise, go into the
`data-raw/` folder and open up the new `dime.R` script.

The first thing you want do is **delete all the code in the script**
that is added there by default. Then write this `library()` function at
the top of the file:

```{r data-raw-script-add-here}
#| filename: "data-raw/dime.R"
#| output: false
library(here)
```

::: callout-tip
The `{here}` package was described in the [Import data
session](https://r-cubed-intro.rostools.org/sessions/import-data) of the
introductory workshop and makes it easier to refer to other files in an
R project. Read through the [section about the here
package](https://r-cubed-intro.rostools.org/sessions/import-data#reading-task-relative-paths-in-the-r-project)
in the introductory R workshop to learn more.
:::

R runs code while working in the current *working directory*, which you
can see on the top of the RStudio Console pane and is shown in the red
box inside the image below.

![The folder location that R does it's "work", called the "working
directory", highlighted by the red
box.](/images/rstudio-console-working-directory.png){fig-alt="An image of the RStudio IDE with the Console pane open and a red outline highlighting the working directory location."}

When in an RStudio R Project, the working directory is the folder where
the `.Rproj` file is located. When you run scripts in R with `source()`
or when you render a Quarto document, sometimes the working directory
will be set to the folder that the R script or Quarto document is
located in. So you can sometimes encounter problems with finding files
if your code needs to load or use a file and you don't use a function
like `here()` to link to it. When you use `here()` it tells R to start
searching for files from the `.Rproj` location.

Let's use an example. Below is the list of the folders and files you
have so far in your R Project.

```{r admin-print-dir-tree}
#| echo: false
withr::with_dir(
  new = project_path,
  {
    fs::dir_tree(recurse = 1)
  }
)
```

::: callout-note
You don't need to run the below code. But if you want to see the
structure and content of a directory, you can use the `dir_tree()`
function from the `{fs}` package, which means "filesystem", by running
the following code in the R Console:

```{r print-dir-tree-project}
#| filename: "Console"
#| eval: false
# To print the file list.
fs::dir_tree("~/Desktop/LearnR3", recurse = 1)
```
:::

If we open up RStudio with the `LearnR3.Rproj` file and run code in the
`data-raw/dime.R`, R runs the commands assuming everything starts in the
`LearnR3/` folder. But! If we run the code that is in the `dime.R`
script by, e.g. not while in RStudio, not in the R Project, or using
`source()`, R runs the code in that file by assuming the working
directory is the folder that the file is in, which is in the `data-raw/`
folder. So if you try to load in data in another folder like `data/`, R
will *think* you mean the `data/` folder in the `data-raw/` folder,
which doesn't exist! So this can make things tricky. What `here()` does
is to tell R to first look for the `.Rproj` file and then start looking
for the file we actually want. This might not make sense yet, but as we
go through the workshop, you will see why this is important to consider.

Alright, the next step is to download the dataset. Paste this code into
the `data-raw/dime.R` script:

```{r dime-data-link}
#| filename: "data-raw/dime.R"
#| eval: false
dime_link <- "https://github.com/rostools/r-cubed-intermediate/raw/refs/heads/main/data/dime.zip"
```

Then we're going to write out the function `download.file()` to download
and save the DIME dataset zip file. We're going to save the zip file to
`data-raw/dime.zip` with the `destfile` argument. This code should be
written in the `data-raw/dime.R` file. Run all the lines of code in this
script to download the dataset.

```{r download-dime}
#| filename: "data-raw/dime.R"
#| eval: false
download.file(dime_link, destfile = here("data-raw/dime.zip"))
```

```{r admin-copy-zip-data}
#| include: false
# To the temporary project, to test how this works there.
fs::file_copy(
  here::here("data/dime.zip"),
  fs::path(project_path, "data-raw", "dime.zip"),
  overwrite = TRUE
)

# To make the next code works as we've kept it in `data`,
# but below we do it in `data-raw/`.
fs::file_copy(
  here::here("data/dime.zip"),
  here::here("data-raw", "dime.zip"),
  overwrite = TRUE
)
```

*After* the zip file is downloaded (it should be called `dime.zip` in
the `data-raw/` folder), comment out this line, so that inside the
`data-raw/dime.R` script it looks like:

```{r comment-out-download-dime}
#| filename: "data-raw/dime.R"
#| eval: false
# Comment out so it doesn't download again.
# download.file(dime_link, destfile = here("data-raw/dime.zip"))
```

We want you to do this because we don't want you to accidentally run
this code again, since you've already downloaded it.

Because the original dataset is stored elsewhere online, you don't need
to save it to your Git history. Add the zip file to the Git ignore list
by typing out and running this code in the Console. You only need to do
this once.

```{r use-git-ignore-dime-zip}
#| filename: "Console"
#| eval: false
usethis::use_git_ignore("data-raw/dime.zip")
```

```{r admin-ignore-dime-zip}
#| include: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_git_ignore("data-raw/dime.zip")
  }
)
```

Part of reproducibility is ensuring you have a written record of what
you did to your data to get to what you are claiming from your analysis.
And part of that principle is to *"keep your raw data raw"*, meaning not
to directly edit your raw data. Instead, let code do it for you so you
have a record of what happened. When you have raw data for a specific
project, store it in the `data-raw/` folder.

Eventually, you will use R code to unzip the file, but the way that R
works is if you unzip to an already existing folder, there can be some
issues or errors. And if you `source()` or re-run your code multiple
times, it will always unzip the file. So before we begin, you'll write
code to always first delete that unzipped output folder. In order to
delete the folder, you'll use the `{fs}` package to handle filesystem
actions. Load the `{fs}` package with `library()` right below the other
`library()` function at the top of the script:

```{r}
#| filename: "data-raw/dime.R"
library(here)
library(fs)
```

Now you can unzip the zip files by using the `unzip()` function by
writing it in `data-raw/dime.R`. The first argument of `unzip()` is the
zip file you want to unzip and the other important argument is called
`exdir` that tells `unzip()` which folder you want to extract the files
to. You'll want to save the output of the zip file to `data-raw/dime/`,
so write this code at the bottom of the `data-raw/dime.R` script:

```{r admin-dime-dir-exists}
#| include: false
skip <- !fs::dir_exists(here::here("data-raw/dime/"))
```

```{r unzip-dime-file}
#| filename: "data-raw/dime.R"
#| eval: !expr skip
unzip(
  here("data-raw/dime.zip"),
  exdir = here("data-raw/dime/")
)
```

```{r admin-unzip-dime-file}
#| include: false
usethis::with_project(
  path = project_path,
  {
    unzip(
      "data-raw/dime.zip",
      exdir = "data-raw/"
    )
  }
)
```

Notice the indentation and spacing of the code. Like writing any
language, code should follow a [style
guide](https://style.tidyverse.org). An easy way of following a specific
style in R is by using the `{styler}` package. We will be using this
package regularly during the workshop. You can run it by using
{{< var keybind.style >}} while in RStudio. Try it right now while you
are working in the `data-raw/dime.R` script.

There probably won't be anything that changes since you haven't written
much code yet and maybe have copy and pasted it from this website. But
since you'll use this package regularly throughout the workshop, this is
a good chance to become more familiar with it.

To make sure you're aligned with what is described in this pre-workshop
tasks, let's check to make sure your files and folders in `data-raw/`
align with the files and folders that should be there. You can do this
by running the `dir_tree()` function from the `{fs}` package, but you
can also manually check the files by using the RStudio file pane or by
using your file browser.

```{r print-tree-data-raw}
#| filename: "Console"
#| eval: false
# NOTE: You don't need to run this code,
# its here to show how we got the file list.
fs::dir_tree("data-raw/")
```

```{r admin-print-data-raw-after-unzip}
#| echo: false
#| message: false
#| warning: false
usethis::with_project(
  path = project_path,
  {
    fs::dir_tree("data-raw")
  }
)
```

If your files and folders in the `data-raw/` folder do not look like
this, start over by deleting all the files *except for* the `dime.R` and
`dime.zip` files. Then re-run the code from beginning to end.

Since you have an R script that downloads the data and processes it for
you, you don't need to have Git track it. So, in the Console, type out
and run this command:

```{r use-git-ignore-data-raw-dime}
#| filename: "Console"
#| eval: false
usethis::use_git_ignore("data-raw/dime/")
```

```{r admin-git-ignore-dime}
#| include: false
usethis::with_project(
  path = project_path,
  {
    usethis::use_git_ignore("data-raw/dime/")
  }
)
```

You now have the data ready for the workshop! At this point, please run
this function in the Console:

```{r check-project-setup}
#| filename: "Console"
#| eval: false
r3::check_project_setup_intermediate()
```

```
~/Desktop/LearnR3
├── DESCRIPTION
├── LearnR3.Rproj
├── R
│   ├── README.md
│   └── functions.R
├── README.md
├── TODO.md
├── data
│   └── README.md
├── data-raw
│   ├── README.md
│   ├── dime
│   │   ├── cgm
│   │   ├── sleep
│   │   └── participant_details.csv
│   ├── dime.zip
│   └── dime.R
└── docs
    ├── README.md
    └── learning.qmd
```

The output should look something a bit like the above text. If it
doesn't, start over by deleting all *but* the `dime.R` and `dime.zip`
files and running the code from the beginning again. If your output
looks a bit like this, then copy and paste the output from the check
into the survey question at the end.

```{r admin-save-project}
#| include: false
withr::with_dir(
  project_path,
  {
    tar(
      tarfile = here::here("_temp", "project.tar.gz"),
      files = fs::dir_ls(".", all = TRUE, recurse = TRUE)
    )
  }
)
```

{{< include /includes/_code-appendix.qmd >}}
