# Git basics and setup {#sec-git-basics}

Since Git has already been covered in the [introduction
workshop](https://r-cubed-intro.rostools.org), we won't cover *learning
it* during this workshop. However, since version control is a
fundamental component of any modern and reproducible data analysis
workflow and *should be used*, we **will be using it throughout the
workshop**.

::: callout-important
If you have used or currently use Git, you can skip this section and
move on to @sec-confirm-git-setup.
:::

Git itself isn't too difficult to use. But the *concepts* that Git uses
are very difficult to learn and use. That's because Git requires you to
view your files and folders and computer *very differently* to how you
may normally be used to working with them. And because we will be using
Git throughout the workshop, to ensure you have the smoothest
experience, we want you to at least be aware of and have read a bit
about Git. So, from the introduction workshop, please read these
sections:

1.  [What is version
    control?](https://r-cubed-intro.rostools.org/sessions/version-control#sec-what-is-version-control)
2.  [What is
    Git?](https://r-cubed-intro.rostools.org/sessions/version-control#sec-what-is-git)
3.  [Basics of
    Git](https://r-cubed-intro.rostools.org/sessions/version-control#sec-basics-of-git)

After reading them, in your RStudio in the R Console, run this function
that should open up a pop-up window to type in your name and email:

```{r setup-git-config}
#| eval: false
#| filename: Console
r3::setup_git_config()
```

For most of the workshop, we will be using Git as shown in the [Using
Git in
RStudio](https://r-cubed-intro.rostools.org/sessions/version-control#using-git-in-rstudio)
section. Later on during the workshop, we might connect our projects to
GitHub, which is described in the [Synchronizing with
GitHub](https://r-cubed-intro.rostools.org/sessions/github#synchronizing-with-github)
section.

## Confirming your Git setup {#sec-confirm-git-setup}

Since you will be using Git regularly throughout the workshop, please
run this check to ensure everything is setup correctly. Type and run
this function in the R Console of RStudio:

```{r check-git-setup}
#| filename: Console
#| eval: false
r3::check_setup()
```

The output you'll get for success should look something like this:

```
Checking R version:
✔ Your R is at the latest version of 4.4.0!
Checking RStudio version:
✔ Your RStudio is at the latest version of 2024.12.1.563!
Checking Git config settings:
✔ Your Git configuration is all setup!
  Git now knows that:
  - Your name is 'Luke W. Johnston'
  - Your email is 'lwjohnst@gmail.com'
```

**Eventually you will need to copy and paste the output into one of the
survey questions**.

::: callout-note
While GitHub is a natural extension to using Git, given the limited time
available in the workshop, we will not be going over how to use GitHub.
If you want to learn about using GitHub, check out this
[session](https://r-cubed-intro.rostools.org/sessions/github) about it
in the introduction workshop.
:::
