# Git basics and setup {#sec-git}

Since Git has already been covered in the [introduction
workshop](https://r-cubed-intro.rostools.org), we won't cover *learning
it* during this workshop. However, since version control is a
fundamental component of any modern and reproducible data analysis
workflow and *should be used*, **we will be using it throughout the
workshop**.

::: callout-important
If you have used or currently use Git, you can skip this section and
move on to @sec-confirm-git-setup.
:::

Git itself isn't too difficult to use. But the *concepts* that Git uses
are very difficult to learn and use. That's because Git requires you to
view your files and folders and computer *very differently* to how you
may normally be used to working with them. And because we will be using
Git throughout the workshop, to ensure you have the smoothest
experience, we want you to at least be aware of and have read a bit
about Git. So, from the introduction workshop, please read these
sections:

1. [What is version
   control?](https://r-cubed-intro.rostools.org/sessions/version-control#sec-what-is-version-control)
2. [What is
   Git?](https://r-cubed-intro.rostools.org/sessions/version-control#sec-what-is-git)
3. [Basics of
   Git](https://r-cubed-intro.rostools.org/sessions/version-control#sec-basics-of-git)

After reading them, open up your RStudio and type out (or paste) this
function in the R Console, then run it by pressing Enter. This function
should open up a pop-up window to type in your name and email:

```{r}
#| eval: false
#| filename: "Console"

r3::setup_git_config()
```

For most of the workshop, we will be using Git as shown in the [Using
Git in
RStudio](https://r-cubed-intro.rostools.org/sessions/version-control#using-git-in-rstudio)
section. Later in the [pre-workshop tasks](github.com) in @sec-github,
you'll read about GitHub and connect your project to it.

## Confirming your Git setup {#sec-confirm-git-setup}

Since you will be using Git regularly throughout the workshop, please
run this check to ensure everything is setup correctly. Type and run
this function in the R Console of RStudio:

```{r}
#| eval: false
#| filename: "Console"

r3::check_setup()
```

The output you'll get for success should look something like this:

```text
Checking R version:
✔ Your R is at the latest version of {{< meta versions.r >}}!
Checking RStudio version:
✔ Your RStudio is at the latest version of {{< meta versions.rstudio >}}!
Checking Git config settings:
✔ Your Git configuration is all setup!
  Git now knows that:
  - Your name is 'Luke W. Johnston'
  - Your email is 'lwjohnst@gmail.com'
```

**Eventually you will need to copy and paste the output into one of the
survey questions**.
