# GitHub basics and setup {#sec-github}

```{r}
#| include: false
# To trigger auto-linking
library(downlit)
```

## The basics of GitHub

Since we will be using Git to track changes to our files during the
workshop and using GitHub to store our projects as well as to
collaborate during the project work, you'll need to get set up with
GitHub.

While GitHub is a natural extension to using Git, given the limited time
available, we will only be going over aspects of GitHub that relate to
storing your project Git repository and working together. If you want to
learn more about using GitHub, check out the
[session](https://r-cubed-intro.rostools.org/sessions/version-control.html)
on it in the introduction workshop. However, you'll briefly read about
it in this section.

GitHub is a popular online service for hosting Git repositories. It also
makes collaborating on projects much easier. Keep in mind, that GitHub
is a company that provides a service for storing Git repositories, while
Git is software you install on your computer. They are two different
things that can sometimes be confused with each other.

A version control system like Git that didn't include a type of external
backup wouldn't be a very good system, because if something happened to
your computer, you'd lose your Git repository. In Git, this "external"
backup is called a "remote", meaning it is something that is separate
and in a different location, usually online, than the main repository.
The remote repository is essentially a duplicate copy of the *history*,
found in the `.git/` folder, of your *local* repository on your
computer. So when you synchronize with the remote, as illustrated in
@fig-git-remotes, it only copies over the changes made as commits in the
history.

One of the biggest reasons why we teach Git is because of the popularity
of several Git repository hosting sites. The most popular one is
[GitHub](https://github.com/), which is what we will use during this
workshop and is also where we host the Git repository of this website
for the workshop.

```{mermaid}
%%| label: fig-git-remotes
%%| fig-cap: The 'remote' vs 'local' repository, or online vs on your computer.
%%| echo: false
%%| eval: true
graph TB
    linkStyle default interpolate basis
    A('Remote':<br>GitHub) --- B('Local':<br>Your computer)

    style A fill:White,stroke:DarkBlue,stroke-width:1.5px;
    style B fill:White,stroke:DarkBlue,stroke-width:1.5px;
```

::: {.callout-warning appearance="default"}
When using GitHub, especially in relation to health research, you need
to be mindful of what you save into the Git history and what you put up
online. Some things to think about are:

- **Do not** save any personal or sensitive data or files in your Git
  repository.
- Generally don't save very large files, like big image files or large
  (non-personal) datasets.

In both cases, it's better to use another tool to store files like that,
rather than through Git and GitHub.
:::

::: callout-note
Some research projects require working on restricted server environments
(such as [Denmark Statistics](https://www.dst.dk/en) when doing research
on the Danish register data), where access to the internet is not
available. This means that you can't use GitHub or any other online Git
repository hosting service. However, you can still use Git on those
servers without using a remote.
:::

The way you interact with GitHub is through Git, which you can use
through RStudio. You do that by uploading ("pushing") a local Git
repository (on your computer) to GitHub (the remote) and then
downloading ("pulling") it back to your computer whenever you want to
synchronise the changes between GitHub and your local Git repository,
which is shown in @fig-git-remotes-synch. Unlike OneDrive or Dropbox
where synchronisation is done automatically, in Git you need to manually
synchronise with the remote because Git is designed to leave you in
control. "Pushing" is when changes to the history are uploaded to GitHub
while "pulling" is when the history is downloaded from GitHub.

```{mermaid}
%%| label: fig-git-remotes-synch
%%| fig-cap: "Synchronizing with GitHub: 'Pushing' and 'pulling'."
%%| echo: false
%%| eval: true
graph TB
    linkStyle default interpolate basis
    A('Remote':<br>GitHub) -- Pull --> B('Local':<br>Your computer)
    B -- Push --> A

    style A fill:White,stroke:DarkBlue,stroke-width:1.5px;
    style B fill:White,stroke:DarkBlue,stroke-width:1.5px;
```

## Setting up GitHub and connecting to RStudio

Interacting with GitHub through RStudio requires you to use something
called a "personal access token" (also called PAT or simply "token").
But first, you need to have a GitHub account in order to make a token.
If you don't have an account, go to
[github.com/join](https://github.com/join) to create an account. After
that, you can connect (or rather, authenticate) your computer with
GitHub.

Any time you do anything on the Internet, there is some risk to having
your information maliciously hacked. This is no different when using
GitHub. So if you can, you should try to be more secure with what you
send across the internet. In fact, most functions that relate to Git or
using GitHub *require* using more secure features in order to work.
`{usethis}` simplifies this with some helper functions. The `{usethis}`
website has a really
[well written guide](https://usethis.r-lib.org/articles/articles/git-credentials.html)
on being more secure when working with GitHub. Here is a very simplified
version of what they recommend that is relevant for what we are doing in
this course.

- Use tokens (PATs) when interacting with your GitHub remote
  repositories while outside of the GitHub website (e.g. when using R or
  `{usethis}`). Tokens are like temporary passwords that provide limited
  access to your GitHub account, like being able to read or write to
  your GitHub repositories, but not being able to delete them. They are
  useful because you can easily delete the token if you feel someone got
  access to it and prevent it from being used, unlike your own password
  which you would have to manually change if it was stolen.

- Use a password manager to save the token for later use. Using password
  managers is basically a requirement for having secure online accounts,
  because they can generate random and long passwords that you don't
  have to remember.

- Use packages like `{gitcreds}` to give `{usethis}` access to the token
  and to interact with your GitHub repositories. You normally would use
  `{gitcreds}` every time you restart your computer or after a certain
  amount of time passes since you last used it.

::: {.callout-note appearance="default"}
## What is a password manager?

A password manager is an app or web service that let's you save or
create passwords for all your accounts, like banking or social media.
Instead of having to remember multiple passwords used across multiple
accounts, or the very insecure approach of one or two passwords for all
your accounts, you instead need to remember only one very secure
password that contains all your other very secure passwords. You can
google for "password manager" and your operating system (Windows, MacOS)
to find possible ones to install or use.

Want a recommendation? [Bitwarden](https://bitwarden.com/) is a very
good password manager that is easy to use and the free version has
everything you need to manage, store, and create passwords.
:::

You might have created a token if you took the introductory workshop.
You can check if you have a token already set by using the `{usethis}`
package. Open up RStudio and in the Console type:

``` {.r filename="Console"}
usethis::gh_token_help()
```

You should see something like the below:

```text
• GitHub host: 'https://github.com'
• Personal access token for 'https://github.com': <unset>
• To create a personal access token, call `create_github_token()`
• To store a token for current and future use, call `gitcreds::gitcreds_set()`
ℹ Read more in the 'Managing Git(Hub) Credentials' article:
  https://usethis.r-lib.org/articles/articles/git-credentials.html
```

If the output says that the token is `<unset>` like the above text does,
that means you need to make Git and `{usethis}` aware of the token on
your computer. You do that by typing the next function in the Console to
create the token on GitHub (if you haven't created one already).

``` {.r filename="Console"}
usethis::create_github_token()
```

This function sends you to the GitHub "Generate new token" webpage with
all the necessary settings checked. Set the "Expiry date" to 90 days
(this is a good security feature). Then, click the green button at the
bottom called "Generate token" and you'll have a very long string
generated for you that starts with `ghp_`.
**Save this token in your password manager** (see note above).

This is the token you will use every time you open up RStudio and
interact with GitHub through R. You do **not** need to create a new
token for each R project or package you make, you only need to create
one after your current token expires (typically every couple of months),
if you've forgotten the token or lost it, or if you've changed to a new
computer.

In the Console, run:

``` {.r filename="Console"}
gitcreds::gitcreds_set()
```

And then copy and paste your token into the prompt in the Console. This
token usually gets saved for the day (it gets cached), but after
restarting your computer, you will need to run the action again. If it
asks to replace an existing one, select the "yes" option. Doing this is
a bit like using the two-factor authentication (2FA) you often have to
do when, for instance, accessing your online bank account or other
government website. In this case, you are telling GitHub (when
interacting to it through RStudio, like uploading and downloading your
changes) that you are who you claim to digitally be.

::: {.callout-tip appearance="default"}
There is another great helper function that runs a lot of checks and
gives some advice when it finds potential problems.

``` {.r filename="Console"}
usethis::git_sitrep()
```

Just to be aware, using this function outputs a *lot* of stuff, most of
which you probably don't even need to know or don't even know what it
means. That's ok, since it is meant as a diagnostic tool.
:::

## Linking your project to GitHub

You've authenticated your computer with GitHub, now is the time to
upload your project to GitHub. Before we do that though, you need to do
a few things. First, let's tell Git to ignore some other files, like the
`.html` files as well as the `_files/` folder that Quarto automatically
creates when making an HTML document. And then, you'll need to commit
your changes to Git, so that you can eventually push them to GitHub.

While in the `LearnR3` repository in RStudio, go to the Console and run
this function to ignore the auto-generated Quarto files.

```{r ignore-html-files}
#| filename: "Console"
#| eval: false
usethis::use_git_ignore(c("*.html", "*_files"))
```

As mentioned in the [Git](git.qmd) (@sec-git) section, we covered Git in
the introductory workshop and won't go into much detail here. During the
workshop, we'll take our time with using the Git interface, as we will
use it regularly. For this pre-workshop tasks though, you can either use
the Git interface or use R functions in the Console.

**If you are familiar enough with using the Git interface**, then open
the Git interface with either the Git icon at the top near the menu bar
or with {{< var keybind.commit >}}. While in the interface, select all
the current changes by clicking the checkbox beside the files and stage
them. Then write a commit message in the text box on the right,
something like `Setting up the project`. Click the "Commit" button, and
finally close the Git interface.

**If you have never used the Git interface before**, you can instead use
R functions to do the same thing. In the Console, run:

```{r add-to-git-history}
#| filename: "Console"
#| eval: false
gert::git_add(".")
gert::git_commit("Setting up the project")
```

This will do the same thing as the Git interface, which is to save all
the changes to the Git history.

After adding the changes to the Git history, this should be the first
commit for everyone. You're now ready to connect your project to GitHub.
In the Console, run:

```{r use-github}
#| filename: "Console"
#| eval: false
usethis::use_github()
```

::: callout-note
You may have to manually enter your username and password, even though
you used `gitcreds::gitcreds_set()`.

If you have troubles logging in, you may need to update Git.
:::

::: callout-tip
You might notice the word `origin` when referring to remotes. The word
*origin* is the default short name used to refer to the location of the
remote (the GitHub URL). You will probably see this word in many other
places to refer to a remote.
:::

The `use_github()` function will take your project and upload it to
GitHub. A bunch of text should pop up and it should open a browser with
your project on GitHub. As you work on this project during the workshop,
you'll be synchronising changes regularly.

As you use Git for your project and save your changes to the Git
history, you can use the "Push" command to send your history of changes
to GitHub. The diagram in @fig-local-vs-remote shows how it conceptually
looks like.

```{mermaid}
%%| label: fig-local-vs-remote
%%| fig-cap: Schematic showing a local repository connected to GitHub's remote repository.
%%| fig-alt: Flow diagram with two boxes, one for the local repository and one for the remote repository on GitHub. An arrow labeled "Push" goes from the local to GitHub, and an arrow labeled "Pull" goes from GitHub to the local.
graph TB;
    yours(Your local<br>repository) -->|Push| github(GitHub remote<br>repository)
    github -->|Pull| yours
```

The "Your local" is your own computer. Whenever you "push" to GitHub, it
means it will upload your file changes (like synchronizing in Dropbox).
Whenever you "pull" from GitHub, it takes any changes made on GitHub and
downloads them to your "Local" computer.

Using GitHub (because of Git) is one of the most effective ways to
collaborate on a project. Hundreds of companies and hundreds of
thousands of workers use Git and services like GitHub to work together
on massive projects. The way collaboration works would conceptually look
like:

```{mermaid}
%%| label: fig-local-collab-and-remote
%%| fig-cap: Schematic showing a local repository, GitHub's remote repository, and a collaborator's repository.
graph TB;
    yours(Your local<br>repository) -->|Push| github(GitHub remote<br>repository)
    github -->|Pull| yours
    others(Collaborator's<br>local repository) -->|Push| github
    github -->|Pull| others
```

This approach to collaborating makes it much easier to contribute
*directly* (not through emails) to projects and to more easily help
others out with issues.

You'll get more practice using GitHub during the workshop, and you'll
have lots of chances to ask questions about it.
