<!-- NOTE: This comes from `template-workshop` template. -->

# Installing R Packages {#sec-installing-r-packages}

```{r}
#| include: false

# To trigger auto-linking
library(pak)
library(remotes)
```

We will be using specific R packages for the workshop, so you will need
to install them. To make it easier, we've created a custom R package
called `r3` that will install all the necessary packages for you for
this workshop. Run these two functions in the Console by copying and
pasting them to install our custom `r3` package as well as the
`{tidyverse}` package.

```{r}
#| eval: false
#| filename: "Console"

install.packages("r3", repos = c("https://rostools.r-universe.dev", "https://cloud.r-project.org"))
install.packages("tidyverse")
```

::: {.callout-tip appearance="default"}
Installing packages is tricky, especially on some work laptops, and
doesn't always work as expected. Here are some other methods to try if
you have trouble installing the packages with the code above.

For Windows users, you may have to install
[RTools](https://cran.r-project.org/bin/windows/Rtools/) to use either
the `{pak}` or `{remotes}` method. The version of RTools you install
must match the version of R that you have installed. For example, if you
have R version 4.4, you need to install RTools 4.4.

First, try installing the `r3` package directly from GitHub with
`{pak}`:

```{r}
#| eval: false
#| filename: "Console"

install.packages("pak")
pak::pak("rostools/r3")
```

If that doesn't work, try with `{remotes}`:

```{r}
#| eval: false
#| filename: "Console"

install.packages("remotes")
remotes::install_github("rostools/r3")
```

If you encounter any errors with either of these functions, try to
restart R by going to "Sessions -> Restart R" and re-run them again. If
that also doesn't work, try to complete the other tasks as best you can,
complete the survey, and let us know you have a problem in the survey.
We'll email you at least a week before the workshop starts to try to
help you get it working.
:::

::: {.callout-tip appearance="default"}
When you see a command like `something::something()`, for example with
`pak::pak()`, you would "read" this as:

> R, can you please use the `pak()` function from the `{pak}` package.

The normal way of doing this would be to load the package with
`library(pak)` and then running the command (`pak()`). But by using the
`::`, we tell R to directly use a function from a package, without
needing to load the package and all of its other functions too. We use
this trick because we only want to use the `pak()` command from the
`{pak}` package and not have to load all the other functions as well. In
this workshop we will be using `::` often.
:::
