# Pre-course tasks {#sec-pre-course}

```{r setup}
#| include: false
temp_path <- fs::path_temp()
project_path <- fs::path(temp_path, "LearnR3")
```

In order to participate in this course, you must complete this section
for the pre-course tasks and finish with **completing the
[survey](https://docs.google.com/forms/d/e/1FAIpQLSe2gpbN9ExCXyeTfJLgu_poemF0iUtkcihnCunXl3YA4e1hHQ/viewform?usp=sf_link)**
at the end. These tasks are designed to make it easier for everyone to
start the course with everything set up. For some of the tasks, you
might not understand why you need to do them, but you will likely
understand why once the course begins.

Depending on your skills and knowledge, these tasks could take between
**3-7 hrs to finish**, so we suggest planning a full day to complete
them. Depending on your institution and how they handle installing
software on work computers, you also might have to contact IT very early
to make sure everything is properly installed and set up.

## List of tasks

Here's the list of tasks you need to do. Specific details about them are
found as you work through the tasks.

1.  Read the Learning objectives in @sec-pre-course-learning-objectives
    for the pre-course tasks (below).
2.  Read the ["How to read the
    website"](https://guides.rostools.org/pre-course.html#reading-website)
    (on a new website).
3.  Follow the [installation
    instructions](https://guides.rostools.org/pre-course.html#installation-instructions)
    (on a new website). Install a version of R, RStudio, and Git that is
    as updated as possible. For some people, depending on their
    institution, this task can take the **longest amount of time**
    because you have to contact your IT to install these packages.
4.  Install the necessary R packages in @sec-install-r-packages.
5.  Read about Git from the introduction course and configure Git on
    your computer in @sec-setup-git. If you haven't used Git before,
    this task could take a while because of the reading. Run the checks
    in this subsection to see if everything works. You'll later need to
    paste this output into the survey.
6.  Create an R Project in @sec-create-r-project, including the folder
    and files.
7.  Create an R Markdown / Quarto file in @sec-quarto.
8.  Write R code to download the data in @sec-download-data and save it
    to your computer. This task will probably take up the most time,
    maybe 30-90 minutes. Run a check in this subsection to see that
    everything is as expected. You'll later need to paste this output
    into the survey.
9.  Read about the basic course details in @sec-course-details.
10. Read the syllabus in @sec-syllabus.
11. Read the [Code of
    Conduct](https://guides.rostools.org/conduct.html).
12. **Complete the pre-course
    [survey](https://docs.google.com/forms/d/e/1FAIpQLSe2gpbN9ExCXyeTfJLgu_poemF0iUtkcihnCunXl3YA4e1hHQ/viewform?usp=sf_link)**.
    This survey is pretty quick, maybe \~10 minutes.

Check each section for exact details on completing these tasks.

## Learning objectives {#sec-pre-course-learning-objectives}

In general, these pre-course tasks are meant to help prepare you for the
course and make sure everything is setup properly so the first session
runs smoothly. However, some of these tasks are meant for learning as
well as for general setup, so we have defined the following learning
objectives for this page:

1.  Learn about making reproducible documents with R Markdown.
2.  Learn about filesystems, relative and absolute paths, and how to
    make use of the `{fs}` R package to navigate files in your project.
3.  Learn where to store your raw data so that you can use scripts as a
    record of what was done to process the data before analyzing it, and
    why that's important.

## Installing the R packages {#sec-install-r-packages}

We will be using specific R packages for the course, so you will need to
install them. A detailed walkthrough for installing the necessary
packages is available on the [pre-course tasks for installing
packages](https://r-cubed-intro.rostools.org/preamble/pre-course.html#installing-r-packages)
section of the introduction course, however, you only need to install
the r3 helper package in order to install all the necessary packages by
running these commands in the R Console:

1.  Install the `{pak}` package:

    ``` r
    install.packages("pak")
    ```

2.  Install the r3 helper package for this course:

    ``` r
    pak::pkg_install("rostools/r3", upgrade = TRUE)
    ```

3.  Install the necessary packages for the course:

    ``` r
    r3::install_packages_intermediate()
    ```

*Note*: When you see a command like `something::something()`, for
example with `pak::pkg_install()`, you would "read" this as:

> R, can you please use the `pkg_install` function from the pak package.

The normal way of doing this would be to load the package with
`library(pak)` and then running the command (`pkg_install()`). But by
using the `::`, we tell R to directly use a function from a package,
without needing to load the package and all of its other functions too.
We use this trick because we only want to use the `pkg_install()`
command from the `{pkg}` package and not have to load all the other
functions as well. In this course we will be using `::` often.

## Setting up Git {#sec-setup-git}

Since Git has already been covered in the Introduction course, we won't
cover *learning it* during this course. However, since version control
is a fundamental component of any modern data analysis workflow and
*should be used*, we **will be using it throughout the course**. If you
have used or currently use Git, you can skip this section. If you
haven't used it, please do these tasks:

1.  Follow the [pre-course tasks for
    Git](https://r-cubed-intro.rostools.org/preamble/pre-course.html#setting-up-git-and-github)
    (*not* the GitHub tasks) from the introduction course. Specifically,
    type in the RStudio Console:

    ``` r
    # A pop-up to type in your name (first and last), 
    # as well as your email
    r3::setup_git_config()
    ```

2.  **Please read** through the [Version Control
    lesson](https://r-cubed-intro.rostools.org/sessions/version-control.html)
    of the introduction course. *You don't need to do any of the
    exercises or activities*, but you are welcome to do them if it will
    help you learn or understand it better. For most of the course, we
    will be using Git as shown in the [Using Git in
    RStudio](https://r-cubed-intro.rostools.org/sessions/version-control.html#using-git-in-rstudio)
    section. Later on during the course, we might connect our projects
    to GitHub, which is described in the [Synchronizing with
    GitHub](https://r-cubed-intro.rostools.org/sessions/version-control.html#synchronizing-with-github)
    section.

Regardless of whether you've done the steps above or not, *everyone*
needs to run:

``` r
r3::check_setup()
```

The output you'll get for success will look something like this:

    Checking R version:
    ✔ Your R is at the latest version of 4.2.0!
    Checking RStudio version:
    ✔ Your RStudio is at the latest version of 2022.2.2.485!
    Checking Git config settings:
    ✔ Your Git configuration is all setup!
      Git now knows that:
      - Your name is 'Luke W. Johnston'
      - Your email is 'lwjohnst@gmail.com'

**Eventually you will need to copy and paste the output into one of the
survey questions**. *Note* that while GitHub is a natural connection to
using Git, given the limited time available, we will not be going over
how to use GitHub. If you want to learn about using GitHub, check out
the
[session](https://r-cubed-intro.rostools.org/sessions/version-control.html)
on it in the introduction course.

## Create an R Project

One of the basic steps to reproducibility and modern workflows in data
analysis is to keep everything contained in a single location. In
RStudio, this is done with R Projects. Please read all of [Section
7.1](https://r-cubed-intro.rostools.org/sessions/project-management.html#what-is-a-project-and-why-use-it)
from the introduction course to learn about R Projects and how they help
keeping things self-contained. *You don't need to do any of the
exercises or activities*.

There are several ways to organise a project folder. We'll be using the
structure from the package
[prodigenr](https://rostools.github.io/prodigenr/). The project setup
can be done by either:

1.  Using RStudio's New Project menu item: "File -\> New Project -\> New
    Directory", scroll down to "Scientific Analysis Project using
    prodigenr" and name the project "LearnR3" in the Directory Name,
    saving it to the "Desktop" with Browse.
2.  Or, running the command
    `prodigenr::setup_project("~/Desktop/LearnR3")` in the R Console.

```{r create-project}
#| include: false
prodigenr::setup_project(project_path)
```

When the RStudio Project opens up again, run these three commands in the
R Console:

``` r
prodigenr::setup_with_git()
usethis::use_blank_slate()
usethis::use_r("functions", open = FALSE)
```

```{r admin-add-setup-files-to-project}
#| echo: false
usethis::with_project(
  path = project_path,
  {
    prodigenr::setup_with_git()
    usethis::use_blank_slate()
    usethis::use_r("functions", open = FALSE)
  }
)
```

Here we use the [usethis](https://usethis.r-lib.org/) package to help
set things up. usethis is an extremely useful package for managing R
Projects and we highly recommend checking it out more to see how you can
use it more in your own work.

## R Markdown / Quarto {#sec-quarto}

<!-- TODO: switch to Quarto -->

We teach and use R Markdown / [Quarto](https://quarto.org) because it is
one of the very first steps to being reproducible and because it is a
very powerful tool to doing data analysis. Please do these two tasks:

1.  Please read over the [R Markdown
    section](https://r-cubed-intro.rostools.org/sessions/reproducible-documents.html#what-is-r-markdown-and-quarto)
    of the introduction course. If you use R Markdown / Quarto already,
    you can skip this step.

2.  Open up the `LearnR3` project, either by clicking the
    `LearnR3.Rproj` file or by using the "File -\> Open Project" menu.
    Run the function below in the Console when RStudio is in the
    `LearnR3` project, which will create a new file called
    `learning.qmd` in the `doc/` folder.

    ``` r
    r3::create_qmd_doc()
    ```

```{r admin-create-qmd}
#| include: false
usethis::with_project(
  path = project_path,
  {r3::create_qmd_doc()}
)
```

## Download the course data {#sec-download-data}

To best demonstrate the concepts in the course, we ideally should work
on a real dataset to apply what we're going to learn. So for this
course, we're going to use an openly licensed dataset on monitoring
sleep and activity (MMASH) [@Rossi2020; @Rossi2020a]. To begin learning
about being reproducible and applying modern approaches to data
analysis, we're going to write and save R code to download a dataset,
prepare it a bit so it's at least a little usable, and than save it to
your computer. The goal at the end of the course is to create a pipeline
to download the data, process and clean it, and save it in a form that
makes it easier to analyze. Why don't we get you to download an already
cleaned and prepared dataset? Because in the real world, the data you
get is rarely all cleaned up and ready for you, and this course is about
learning more advanced tools to do the data wrangling. Look over these
tasks and than switch over to the [MMASH
website](https://physionet.org/content/mmash/1.0.0/):

1.  Look through the Data Description to get familiar with the dataset
    and see what is contained inside of it. We'll refer back to the Data
    Description throughout the course as well as in the exercises.
2.  Look over the open license that allows you to re-use it, even for
    research purposes. Note: GDPR makes it stricter on how to share and
    use personal data, but it *does not prohibit sharing it or making it
    public*! GDPR and Open Data are not in conflict.

> *Note*: Sometimes the PhysioNet website is slow. If that's the case,
> use [**this alternative link**](resources/mmash-page.html) instead.

After looking over the MMASH website, you need to setup where to store
the dataset to prepare it for later processing. While in your `LearnR3`
R Project, go to the Console pane in RStudio and type out:

``` r
usethis::use_data_raw("mmash")
```

```{r admin-create-data-raw}
#| echo: false
usethis::with_project(
  path = project_path,
  {usethis::use_data_raw("mmash")}
)
```

What this function does is create a new folder called `data-raw/` and
creates an R script called `mmash.R` in that folder. This is where we
will store the raw, original MMASH data that we'll get from the website.
The R script should have opened up for you, otherwise, go into the
`data-raw/` folder and open up the new `mmash.R` script.

The first thing we want do is delete all the code in the script that is
added there by default. Then we'll create a new line at the top and type
out:

```{r data-raw-script-add-here}
library(here)
```

The [here](https://here.r-lib.org/) package was described in the
[Management of R
Projects](https://r-cubed-intro.rostools.org/sessions/project-management.html)
of the introductory course and makes it easier to refer to other files
in an R project. Read through the [section about the here
package](https://r-cubed-intro.rostools.org/sessions/project-management.html#packages-data-and-file-paths)
in the introductory R course.

R works based on the current *working directory*, which you can see on
the top of the RStudio Console pane, which you can see in the red box
inside the image below. When in an RStudio R Project, the working
directory is the folder where the `.Rproj` file is located. When you run
scripts in R with `source()`, sometimes the working directory will be
set to where the R script is located. So you can sometimes encounter
problems with finding files. Instead, when you use `here()` R knows to
start searching for files from the `.Rproj` location.

![The folder location that R does it's "work", called the "working
directory", highlighted by the red
box.](images/rstudio-console-working-directory.png)

Let's use an example. Below is the tree of the folders and files you
have so far. If we open up RStudio with the `LearnR3.Rproj` file and run
code in the `data-raw/mmash.R`, R runs the commands assuming everything
starts in the `LearnR3/` folder. But! If we run the code in the
`mmash.R` script by other ways (e.g. not with RStudio, not in an R
Project, or with `source()`), R runs everything assuming it starts in
the `data-raw/` folder. This can make things tricky. What `here()` does
is tell R to first look for the `.Rproj` file and then start looking for
the file we actually want. This might not make sense yet, but as we go
through the course, you will see why this is important to consider.

::: callout-note
You don't need to run the below code. But if you want to see the
structure and content of a directory, you can use the `dir_tree()`
function from the `{fs}` package, which means "filesystem", by running
the following code in the R Console:

``` r
# To print the file list.
fs::dir_tree("~/Desktop/LearnR3", recurse = 1)
```

```{r admin-print-dir-tree}
#| echo: false
withr::with_dir(
  new = project_path,
  {fs::dir_tree(recurse = 1)}
)
```
:::

Alright, the next step is to download the dataset. Paste this code into
the `data-raw/mmash.R` script:

``` r
mmash_link <- "https://physionet.org/static/published-projects/mmash/multilevel-monitoring-of-activity-and-sleep-in-healthy-people-1.0.0.zip"
```

> *Note*: Sometimes the PhysioNet website is slow. If that's the case,
> use the `r3::mmash_data_link` instead of the link used above. In this
> case, it will look like `mmash_link <- r3::mmash_data_link`.

Then we're going to write out the function `download.file()` to download
and save the zip dataset. We're going to save the zip file to
`data-raw/mmash-data.zip` with the `destfile` argument. This code should
be written in the `data-raw/mmash.R` file. Run these lines of code to
download the dataset.

```{r download-data, eval=FALSE}
download.file(mmash_link, destfile = here("data-raw/mmash-data.zip"))
```

*After* downloading the zip file (it should be called `mmash-data.zip`
in the `data-raw/` folder), comment out this line, so that inside the
`data-raw/mmash.R` script it looks like:

```{r download-data-comment-out, eval=FALSE}
# download.file(mmash_link, destfile = here("data-raw/mmash-data.zip"))
```

We do this because we don't want to accidentally run this code again,
since we already downloaded the file.

Because the original dataset is stored elsewhere on a website, we don't
need to save it to our Git history. Add the zip file to the Git ignore
list by typing out and run this code in the Console. You only need to do
this once.

``` r
usethis::use_git_ignore("data-raw/mmash-data.zip")
```

```{r admin-ignore-mmash-zip}
#| echo: false
usethis::with_project(
  path = project_path,
  {usethis::use_git_ignore("data-raw/mmash-data.zip")}
)
```

Next, open up the zip file with your File Manager and look at what is
inside. There should be the license file, another file to check if the
download worked correctly (the SHA file, which you don't need to worry
about), and another zip file of the dataset. Because we are starting
with the original raw `mmash-data.zip`, we should record exactly how we
process the data set for use. This relates to the key principal of
*"keep your raw data raw"*, as in don't edit or touch your raw data, let
R or another programming language process it. This lets you have a
history of what was done to the raw data. During data collection and
entry, programs like Excel or Google Sheets are incredibly powerful. But
after collection is done, don't make edits directly to the data unless
absolutely necessary.

A quick comment about whether you should save your *raw* data in
`data-raw/`. A general guideline is:

-   *Do* store it to `data-raw/` if the data will only be used for the
    one project. Use the `data-raw/` R script to be the record for how
    you processed your data for the final analysis work.
-   *Don't* save it to `data-raw/` if: 1) there is a central dataset
    that multiple people use for multiple projects; or 2) you have the
    data online. Instead, use the `data-raw/` R script to be the record
    for which website or central location you extracted it from and how
    you later processed it.
-   *Don't* save it to a project-specific `data-raw/` folder if you will
    use the raw data for multiple projects. Instead, create a central
    location for the data *for yourself* so that you can point all other
    projects to it and use their individual `data-raw/` R scripts as the
    record for how you processed the raw data.

Unzip the zip files by using the `unzip()` function and writing it in
`data-raw/mmash.R` below the `download.file()` function. The main
argument for `unzip()` is the zip file and the next important one is
called `exdir` that tells `unzip()` which folder you want to extract the
files to. The argument `junkpaths` is used here to tell `unzip()` to
extract everything to the `data-raw/` folder (no idea why it's called
"junkpaths"). This code should be written and executed in the
`data-raw/mmash.R` script.

```{r unzip-main-file}
unzip(here("data-raw/mmash-data.zip"), 
      exdir = here("data-raw"),
      junkpaths = TRUE)
```

Notice the indentations and spacings of the code. Like writing any
language, code should follow a [style
guide](https://style.tidyverse.org). An easy way of following a style is
by selecting your code and using RStudio's builtin style fixer with
either `Ctrl-Shift-A` or `Code -> Reformat Code` menu item. Next, we'll
extract the new `data-raw/MMASH.zip` file using `unzip()` again. Because
we want to keep the folder structure inside this zip file, we won't use
`junkpaths`. Write and execute this code in the `data-raw/mmash.R`
script.

```{r unzip-data-file}
unzip(here("data-raw/MMASH.zip"),
      exdir = here("data-raw"))
```

Almost done! There are several files left over that we don't need, so
you'll need to write code in the script code that removes them. We'll
use the [fs](https://fs.r-lib.org/) package to work with files. Before
you change any files, look into the `data-raw/` folder and confirm that
the below listed files and folders are there.

```{r, eval=FALSE}
# NOTE: You don't need to run this code, 
# its here to show how we got the file list.
fs::dir_tree("data-raw", recurse = 1)
```

    data-raw
    ├── README.md
    ├── LICENSE.txt
    ├── MMASH.zip
    ├── SHA256SUMS.txt
    ├── DataPaper
    │   ├── user_1
    │   ├── user_10
    │   ├── ...
    │   ├── user_8
    │   └── user_9
    ├── mmash-data.zip
    └── mmash.R

If your files and folders in the `data-raw/` folder do not look like
this, start over by deleting all the files *except for* the `mmash.R`
and `mmash-data.zip` files. Then re-run the code from beginning to end.

To tidy up the files, first, use the `file_delete()` function from fs to
delete all the files we originally extracted (`LICENSE.txt`,
`SHA256SUMS.txt`, and `MMASH.zip`). Then use `file_move()` to rename the
new folder `data-raw/DataPaper/` to something more explicit like
`data-raw/mmash/`. Add these lines of code to the `data-raw/mmash.R`
script and run them:

```{r delete-rename-files}
library(fs)
file_delete(here(c("data-raw/MMASH.zip", 
                   "data-raw/SHA256SUMS.txt",
                   "data-raw/LICENSE.txt")))
file_move(here("data-raw/DataPaper"), here("data-raw/mmash"))
```

Afterwards, the files and folders in `data-raw/` will look like:

```         
data-raw
├── README.md
├── mmash
│   ├── user_1
│   ├── user_10
│   ├── ...
│   ├── user_8
│   └── user_9
├── mmash-data.zip
└── mmash.R
```

Like before, if your files and folders inside `data-raw/` don't look
like those listed above, start over again (making sure to delete all
*but* the `mmash.R` and `mmash-data.zip` files).

Since we have an R script that downloads the data and processes it for
us, we don't need to have Git track it. So, in the Console, type out and
run this command:

``` r
usethis::use_git_ignore("data-raw/mmash/")
```

```{r admin-git-ignore-mmash}
#| echo: false
usethis::with_project(
  path = project_path,
  {usethis::use_git_ignore("data-raw/mmash/")}
)
```

The `data-raw/mmash.R` script should look like this at this point:

```{r pre-course-state-of-script}
#| file: ../R/pre-course-script.R
#| eval: false
```

```{r admin-pre-course-state-of-script}
#| echo: false
fs::file_copy(
  here::here("R/pre-course-script.R"),
  fs::path(project_path, "data-raw", "mmash.R")
)
```

You now have the data ready for the course! At this point, please run
this function in the Console:

``` r
r3::check_project_setup()
```

```         
~/Desktop/LearnR3
├── DESCRIPTION
├── LearnR3.Rproj
├── R
│   ├── README.md
│   └── functions.R
├── README.md
├── TODO.md
├── data
│   └── README.md
├── data-raw
│   ├── README.md
│   ├── mmash
│   │   ├── user_1
│   │   ├── user_10
│   │   ├── user_11
│   │   ├── user_12
│   │   ├── user_13
│   │   ├── user_14
│   │   ├── user_15
│   │   ├── user_16
│   │   ├── user_17
│   │   ├── user_18
│   │   ├── user_19
│   │   ├── user_2
│   │   ├── user_20
│   │   ├── user_21
│   │   ├── user_22
│   │   ├── user_3
│   │   ├── user_4
│   │   ├── user_5
│   │   ├── user_6
│   │   ├── user_7
│   │   ├── user_8
│   │   └── user_9
│   ├── mmash-data.zip
│   └── mmash.R
└── doc
    ├── README.md
    └── learning.qmd
```

The output should look something a bit like the above text. If it
doesn't, start over by deleting all *but* the `mmash.R` and
`mmash-data.zip` files and running the code from the beginning again. If
your output looks a bit like this, than copy and paste the output into
the survey question at the end.

## Basic course details {#sec-course-details}

Most of the course description is found in the syllabus in
@sec-syllabus. If you haven't read it, **please read it now**. Read over
what the course will cover, what we expect you to learn at the end of
it, and what our basic assumptions are about who you are and what you
know. The final pre-course task is a survey that asks some questions on
if you've read and understood it.

One goal of the course is to teach about open science, and true to our
mission, we practice what we preach. The course material is publicly
accessible (all on this website) and [openly
licensed](index.html#licensing) so you can use and re-use it for free!
The material and table of contents on the side is listed in the order
that we will cover in the course.

We have a [Code of Conduct](https://guides.rostools.org/conduct.html).
If you haven't read it, **read it now**. The survey at the end will ask
about Conduct. We want to make sure this course is a supportive and safe
environment for learning, so this Code of Conduct is important.

You're almost done. Please fill out the [**pre-course
survey**](https://docs.google.com/forms/d/e/1FAIpQLSe2gpbN9ExCXyeTfJLgu_poemF0iUtkcihnCunXl3YA4e1hHQ/viewform?usp=sf_link)
to finish this assignment.

See you at the course!
