# Extra material {#sec-extra-material}

## Debugging functions

Debugging is one of activities that seems really scary and difficult,
but once you try it and use it, is not nearly as intimidating as it
seemed. To debug, which means to find and fix problems in your code,
there are several ways, the simplest of which is by inserting the
`browser()` function into the the start of your function, re-running the
function by either manually running it or using source with
{{< var keybind.source >}}, and using it again.

For instance, we have a function like this:

```{r}
test_debugging <- function(number) {
  new_number <- number + number
  return(new_number)
}
test_debugging(5)
```

To start the debugger, insert the `browser()` function into your
function:

```{r}
#| eval: false

test_debugging <- function(number) {
  browser()
  new_number <- number + number
  return(new_number)
}
test_debugging(5)
```

And re-run and use the function again, which will pop up a new debugging
panel in RStudio. Sadly, we can't show this on the website since it only
works in RStudio. When you are in RStudio, the debugger will open up and
it will show a few things:

- A yellow line will highlight the code in the function, along with a
  green arrow on the left of the line number.
- The Console will now start with `Browse[1]>` and will have text like
  `debug at ...`.
- There will be new buttons on the top of the Console like "Next",
  "Continue", and "Stop".
- The Environment pane will be empty and will say "Traceback".

In this mode you can really investigate what is happening with your code
and how to fix it. The way to figure out what's wrong is by running the
code bit by bit. This debug environment is empty except for the actions
that occur within it, so it really can help figure things out.

## Functionals and for loops {#sec-alternative-loop-explanation}

The concept of functional programming can be difficult to grasp. As an
alternative to the above explanation, one could also explain functional
programming in relation to what it enables. For example, imagine that
you have a vector and would like to investigate whether any given number
inside this vector is above 5. This can be accomplished using a variety
of programming styles. First, we could create a function to check
whether a given number is above 5, and manually check each element of
the vector:

```{r}
# Create vector of 10 numbers
numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

# Function to check if a number is above 5
over_five <- function(number) {
  if (number > 5) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

# Check each element
over_five(numbers[1])
over_five(numbers[2])
over_five(numbers[3])
over_five(numbers[4])
over_five(numbers[5])
over_five(numbers[6])
over_five(numbers[7])
over_five(numbers[8])
over_five(numbers[9])
over_five(numbers[10])
```

Such an approach is verbose, not "functional" in style, and very
inflexible. Imagine if the vector had contained 100 numbers instead of
10. In such a case, it would have been unreasonable to manually check
each element. As an alternative we could accomplish the same by using a
for loop:

```{r}
# Initialize a vector to capture the output in the for loop
output <- vector("logical", length = length(numbers))

# Use seq_along to get for loop to start from 1 and end at the end of numbers
for (item in seq_along(numbers)) {
  output[item] <- over_five(numbers[item])
}
output
```

This for loop is taking each item from numbers, using `over_five()` on
it, and then saving the output into the object `output`. So this loop
allows us to do the same as we did manually, but more precisely and with
less code. If the vector has more items or less, it doesn't matter to
the for loop since it will run no matter the length. You might notice
already that there are several technical things going on, like the use
of `vector()`, `seq_along()`, and `output[item]`. We won't explain them
here, but this is meant to highlight that loops aren't easy to actually
use properly.

For some situations, for loops are the perfect solution. However, they
are not R's strength, but rather functional programming is. In this
case, we replace the for loop with the functional `map()`, which would
make the code shorter and more robust.

```{r}
library(tidyverse)
# Functional
map(numbers, over_five)
```

This works great as is, but sometimes you might not need to create a new
function to use inside of `map()`. Instead we could use an anonymous
function with `\()` (or either `function()` or `~`).

```{r}
# Anonymous function
map(numbers, \(x) if (x > 5) TRUE else FALSE)
```

## Using `case_when()` for nested conditionals

For simple fixes of categorical variables or for creating simple
categorical variables from existing ones, you would usually use
`if_else()`. But when it becomes more complex or when you think you need
to use nested `if_else()` statements, you should probably switch to
using `case_when()`.

This function works by giving it a series of logical conditions and an
associated output if the condition is true. Each condition is processed
*sequentially*, meaning that if a condition is `TRUE`, the output won't
be overridden by later conditions. The general form of `case_when()`
looks like:

```{r}
#| eval: false

case_when(
  variable1 == condition1 ~ output,
  variable2 == condition2 ~ output,
  # (Optional) Otherwise
  TRUE ~ final_output
)
```

The optional ending is only necessary if you want a certain output if
none of your conditions are met. Because conditions are processed
*sequentially* and because it is the *last* condition, by setting the
final condition to be `TRUE`, the `final_output` will used (in this
case). If this last `TRUE` condition is not used, then by default the
final output would be a missing value. A (silly) example using age might
be:

```{r}
#| eval: false

case_when(
  age > 20 ~ "old",
  age <= 20 ~ "young",
  # For final condition
  TRUE ~ "unborn!"
)
```

If instead you want one of the conditions to be `NA`, you need to set
the appropriate `NA` value (in this example, `NA_character_` is for
character data):

```{r}
#| eval: false

case_when(
  age > 20 ~ "old",
  age <= 20 ~ NA_character_,
  # For final condition
  TRUE ~ "unborn!"
)
```

Alternatively, if we want missing age values to output `NA` values at
the end (instead of `"unborn!"`), we would exclude the final condition:

```{r}
#| eval: false

case_when(
  age > 20 ~ "old",
  age <= 20 ~ "young"
)
```

With `{dplyr}` functions like `case_when()`, it requires you be explicit
about the type of output each condition has since all the outputs must
match (e.g. all character or all numeric). This prevents you from
accidentally mixing e.g. numeric output with character output. Missing
values also have data types:

- `NA_character_` (character)
- `NA_real_` (numeric)
- `NA_integer_` (integer)

## Key takeaways

::: {.callout-note appearance="minimal" collapse="true"}
## Teacher note

Quickly cover this before finishing the session and getting them to do
the survey.
:::

- Instead of using multiple `if_else()` statements, you can use
  `case_when()` to make your code cleaner and easier to read.
- Save final datasets into the `data/` to communicate to yourself and
  others that this is data to work with, not the data in `data-raw/`.

{{< include /includes/_code-appendix.qmd >}}
