# Extra material {#sec-extra-material}

## Functionals and for loops {#sec-alternative-loop-explanation}

The concept of functional programming can be difficult to grasp. As an
alternative to the above explanation, one could also explain functional
programming in relation to what it enables. For example, imagine that
you have a vector and would like to investigate whether any given number
inside this vector is above 5. This can be accomplished using a variety
of programming styles. First, we could create a function to check
whether a given number is above 5, and manually check each element of
the vector:

```{r}
# Create vector of 10 numbers
numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

# Function to check if a number is above 5
over_five <- function(number) {
  if (number > 5) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

# Check each element
over_five(numbers[1])
over_five(numbers[2])
over_five(numbers[3])
over_five(numbers[4])
over_five(numbers[5])
over_five(numbers[6])
over_five(numbers[7])
over_five(numbers[8])
over_five(numbers[9])
over_five(numbers[10])
```

Such an approach is verbose, not "functional" in style, and very
inflexible. Imagine if the vector had contained 100 numbers instead of
10. In such a case, it would have been unreasonable to manually check
each element. As an alternative we could accomplish the same by using a
for loop:

```{r}
# Initialize a vector to capture the output in the for loop
output <- vector("logical", length = length(numbers))

# Use seq_along to get for loop to start from 1 and end at the end of numbers
for (item in seq_along(numbers)) {
  output[item] <- over_five(numbers[item])
}
output
```

This for loop is taking each item from numbers, using `over_five()` on
it, and then saving the output into the object `output`. So this loop
allows us to do the same as we did manually, but more precisely and with
less code. If the vector has more items or less, it doesn't matter to
the for loop since it will run no matter the length. You might notice
already that there are several technical things going on, like the use
of `vector()`, `seq_along()`, and `output[item]`. We won't explain them
here, but this is meant to highlight that loops aren't easy to actually
use properly.

For some situations, for loops are the perfect solution. However, they
are not R's strength, but rather functional programming is. In this
case, we replace the for loop with the functional `map()`, which would
make the code shorter and more robust.

```{r}
library(tidyverse)
# Functional
map(numbers, over_five)
```

This works great as is, but sometimes you might not need to create a new
function to use inside of `map()`. Instead we could use an anonymous
function with either `function()` or `~`.

```{r}
# Anonymous function
map(numbers, function(x) if (x > 5) TRUE else FALSE)

# Anonymous function (with shortcut)
map(numbers, ~ if (.x > 5) TRUE else FALSE)
```

## Debugging functions

Debugging is one of activities that seems really scary and difficult,
but once you try it and use it, is not nearly as intimidating as it
seemed. To debug, which means to find and fix problems in your code,
there are several ways, the simplest of which is by inserting the
`browser()` function into the the start of your function, re-running the
function by either manually running it or using source with
{{< var keybind.source >}}, and using it again.

For instance, we have a function like this:

```{r}
test_debugging <- function(number) {
  number + number
}
```

To start the debugger, insert the `browser()` function into your
function:

```{r, eval=FALSE}
test_debugging <- function(number) {
  browser()
  number + number
}
```

And re-run and use the function again, which will pop up a new debugging
panel in RStudio. Sadly, we can't show this on the website since it only
works in RStudio (we may add a video as some point). When you are in the
debugger will open up and it will show a few things:

<!-- TODO: eventually add an image or video.. but for next year. -->

-   A yellow line will highlight the code in the function, along with a
    green arrow on the left of the line number.
-   The Console will now start with `Browse[1]>` and will have text like
    `debug at ...`.
-   There will be new buttons on the top of the Console like "Next",
    "Continue", and "Stop".
-   The Environment pane will be empty and will say "Traceback".

<!-- TODO: Add image or video of this or extend it from previous video? -->

<!-- TODO: Image of debugging changes, with descriptions -->

In this mode you can really investigate what is happening with your code
and how to fix it. The way to figure out what's wrong is by running the
code bit by bit. This debug environment is empty except for the actions
that occur within it, so it really can help figure things out.

## Non-standard evaluation (NSE)

Writing your own functions that use `{tidyverse}` functions, you may
eventually encounter an error that might not be very easy to figure out.
Here's an example where you want to use your own arguments in
`filter()`.

```{r test-nse, error=TRUE}
test_nse <- function(data, filter_condition) {
  data %>%
    dplyr::filter(filter_condition)
}

CO2 %>%
  tibble() %>%
  test_nse(conc > 100)
```

The error occurs of something called "[non-standard
evaluation](https://adv-r.had.co.nz/Computing-on-the-language.html)" (or
NSE). NSE is a feature of R and is used quite a lot throughout R (e.g.
`library()`), but is especially used in the `{tidyverse}` packages. It's
one of the first things computer scientists complain about when they use
R, because it is such a foreign thing in other programming languages.
But NSE is what allows you to use formulas (e.g. `y ~ x + x2` in
modeling) or allows you to type out `select(Gender, BMI)` or
`library(purrr)`. In "standard evaluation", these would instead be
`select("Gender", "BMI")` or `library("purrr")`. So NSE gives
flexibility and ease of use for the user (we don't have to type quotes
every time) when doing data analysis, but can give some headaches when
programming in R, like when making functions. There's more detail about
this on the [dplyr
website](https://dplyr.tidyverse.org/articles/programming.html#warm-up),
which will give a few options to deal with NSE while programming with
`{tidyverse}` packages, the simplest of which is to wrap the argument
with `{{}}`.

```{r test-nse-fixed}
#| error: true
test_nse <- function(data, filter_condition) {
  data %>%
    dplyr::filter({{ filter_condition }})
}

CO2 %>%
  tibble() %>%
  test_nse(conc > 100)

CO2 %>%
  tibble() %>%
  test_nse(uptake < 30)
```
